### 2019-11-07 ManuscriptTe.Rmd
Add example

---
title: "Manuscript Te example"
author: "Lauren Buckley"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Function testing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
##Load TrenchR package
library(TrenchR)
```

```{r}
#Grasshopper biophys
library(TrenchR)
library(maptools)
library(ggplot2)

count=function(x){length(na.omit(x))}

error.bars= function(bloc, thing, se, cols="black", ltys="solid"){
  arrows(bloc, thing-se, bloc, thing+se, code=3, angle=90,length=0, col=cols, lty=ltys)
}
#--------------------------
#Grasshopper data
setwd("/Volumes/GoogleDrive/Shared drives/TrEnCh/Projects/TrenchR/data/")
spec.dat=read.csv("SpecData.csv")

#read weather data
dat=read.csv("DatAll_Pace2011.csv", sep=",", na.strings = c("-49.99","NA"))
#match site data
sites=read.csv("SiteElevations.csv", sep=",")

#combine soil temperatures
dat$SoilTemp= rowMeans(dat[,c("SoilTemp","SoilTemperature")], na.rm = TRUE)

#ditch problematic temperatures
dat$Temp[dat$Temp< -10]=NA

#add elevation
match1= match(dat$site, sites$Site)
dat$Elev=sites[match1,"Elevation"]

#----------------

z<-0.001 #specify distance from ground

#calculate Julian day
date.ct <- strptime(dat$datetime, format="%m/%d/%Y %H:%M", tz="MST")
dat$J= round(as.numeric(julian(date.ct, origin = as.Date("2011-01-01") )))
dat$hour= date.ct$hour

#match site data
match1= match(dat$site, sites$Site) 
dat$lon= sites$Lon[match1]
dat$lat= sites$Lat[match1]

#Calculate zenith
dat$psi= zenith_angle(dat$J,lat=dat$lat, lon=dat$lon, hour=dat$hour)

#------------------------
##Calculate Te for each species
dat$Te<-Tb_grasshopper(T_a=dat$Temp, T_g=dat$SoilTemp, u=dat$Wind, H=dat$Rad, K_t=0.7, psi=dat$psi, L=21.1, Acondfact=0.3)
#fix infinite
dat$Te[is.infinite(dat$Te)]<-NA
#ditch big numbers for now
dat$Te[dat$Te>100]<-NA

#----------------------------------------
#PLOTS

#Subset to August
dat.dt<- strptime(dat$datetime, format="%m/%d/%Y %H:%M", tz="MST") #as.Date
dataug= subset(dat, dat.dt$mon==7)
#subset to day
dataug= subset(dataug, dataug$Rad>5)
#aggregate by day and hour
dat.dh= aggregate(dataug, by=list(dataug$site, dataug$hour), FUN=mean, na.rm=TRUE)

```

```{r}
library(reshape2)

#convert to long format
dat.s= dat.dh[,c("Rad","Temp","Wind","DodgeiTemp","SoilTemp","SanguinipesTemp","BivatatusTemp","Elev","J", "hour","psi","Te")]

dat.s= melt(dat.s, id = c("Elev", "J","hour","psi"))

#plot
dat.s= subset(dat.s, dat.s %in% c("Temp","SoilTemp","SanguinipesTemp","Te") )

ggplot(data=dat.s, aes(x=hour, y = value, color=variable))+geom_line( ) #+
#  facet_wrap(~Elev)

```

### 2019-11-08 ManuscriptTe.Rmd
Update Te example

@@
 library(TrenchR)
+library(maptools)
+library(ggplot2)
 ```
 
+#Microclimate calculations
+u_z=function(z, d=0.229, z_0=0.0208, u_star= 0.114, k=0.41) u_star/k*log((z-d)/z_0)
+
 ```{r}
-#Grasshopper biophys
-library(TrenchR)
-library(maptools)
-library(ggplot2)
+setwd("/Volumes/GoogleDrive/Shared drives/TrEnCh/Projects/TrenchR/data/")
+dat=read.csv("microclimate_C1.csv", sep=",", na.strings = c("-49.99","NA"))
+```
+```{r}
+#ANALYZE SURFACE ROUGHNESS AND PROFILES
+C1all$day=  as.POSIXlt(as.character(C1all$datetime),tz='MST', format="%m/%d/%Y %H:%M")$mday
+C1all$hour=as.POSIXlt(as.character(C1all$datetime),tz='MST', format="%m/%d/%Y %H:%M")$hour
 
-count=function(x){length(na.omit(x))}
+C1.dayhr<- aggregate(C1all[,c(2:8,13:22)], by=list(C1all$day, C1all$hour), FUN=mean, na.rm=TRUE)
+names(C1.dayhr)[1:2]=c("day","hour")
+C1.hr<- aggregate(C1all[,c(2:8,13:22)], by=list(C1all$hour), FUN=mean, na.rm=TRUE)
+names(C1.hr)[1]="hour"
 
-error.bars= function(bloc, thing, se, cols="black", ltys="solid"){
-  arrows(bloc, thing-se, bloc, thing+se, code=3, angle=90,length=0, col=cols, lty=ltys)
+C1.hr= as.matrix(C1.hr)
+#surface roughness
+z0= apply(C1.hr[, c("Wind0.25m","Wind0.5m","Wind1m","Wind1.5m")], FUN=surface_roughness, MARGIN=1, zr=c(0.25,0.5,1.5,1.5))
+
+z.seq <- seq(0, 2, by = 0.1)
+
+for(hr in 1:24){
+t.seq1 <- air_temp_profile_neutral(T_r=C1.hr[hr,"Air0.5m"], zr=0.5, z0=0.02, z=2, T_s=C1.hr[hr,"Soil"])
+t.seq2 <- air_temp_profile(T_r=C1.hr[hr,"Air0.5m"], u_r=C1.hr[hr,"Wind0.5m"], zr=0.5, z0=0.02, z=2, T_s=C1.hr[hr,"Soil"])
+t.seq3 <- air_temp_profile_segment(T_r=C1.hr[hr, c("Air0.25m","Air0.5m","Air1.0m","Air1.5m")],u_r=C1.hr[hr, c("Wind0.25m","Wind0.5m","Wind1m","Wind1.5m")], zr=c(0.25,0.5,1,1.5), z0=0.02, z=2, T_s=C1.hr[hr,"Soil"])
+
+u.seq1<- wind_speed_profile_neutral(u_r=C1.hr[hr,"Wind0.5m"], zr=0.5, z0=0.02, z=2)
+u.seq2<- wind_speed_profile(u_r=C1.hr[hr,"Wind0.5m"], zr=0.5, z0=0.02, z=2)
+u.seq3<- wind_speed_profile_segment(u_r=C1.hr[hr, c("Wind0.25m","Wind0.5m","Wind1m","Wind1.5m")], zr=c(0.25,0.5,1,1.5), z0=0.02, z=2)
 }
-#--------------------------
+
+#plot(t.seq, z.seq, type = "l", xlab = "Temperature (C)", ylab = "z (height above ground in m)")
+```
+
+#Te calculation
+```{r}
+
 #Grasshopper data
@@
 
-#------------------------
+#Subset to August
+dat.dt<- strptime(dat$datetime, format="%m/%d/%Y %H:%M", tz="MST") #as.Date
+dataug= subset(dat, dat.dt$mon==7)
+#subset to day
+dat= subset(dataug, dataug$Rad>5)
+
+#scale temperature to ground
+dat$Tgrass <- air_temp_profile(T_r=dat$Temp, u_r=dat$Wind, zr=0.5, z0=0.02, z=z, T_s=dat$SoilTemp)
+
+#cut nas
+dat= dat[,c("Rad","Temp","Wind","SoilTemp","SanguinipesTemp","Elev","J", "hour","psi","Tgrass")]
+dat= na.omit(dat)
+```
+
+```{r}
 ##Calculate Te for each species
-dat$Te<-Tb_grasshopper(T_a=dat$Temp, T_g=dat$SoilTemp, u=dat$Wind, H=dat$Rad, K_t=0.7, psi=dat$psi, L=21.1, Acondfact=0.3)
+dat$Te<-Tb_grasshopper(T_a=dat$Tgrass, T_g=dat$SoilTemp, u=dat$Wind, H=dat$Rad, K_t=0.7, psi=dat$psi, L=0.0211, Acondfact=0.01)  #T_a=dat$Tgrass
 #fix infinite
@@
 dat$Te[dat$Te>100]<-NA
+```
 
-#----------------------------------------
-#PLOTS
+```{r}
+#estimate convective heat transfer coefficient
 
-#Subset to August
-dat.dt<- strptime(dat$datetime, format="%m/%d/%Y %H:%M", tz="MST") #as.Date
-dataug= subset(dat, dat.dt$mon==7)
-#subset to day
-dataug= subset(dataug, dataug$Rad>5)
+#Estimate using empirical relationship
+dat$H_L=heat_transfer_coefficient(V=dat$Wind,D=0.0211,K=0.025, nu= 15.3 * 10^(-6) , taxa="cylinder")
+
+#Also illustrate estimations using a spherical approximation and a simplified version of the approximation.
+H_L1= heat_transfer_coefficient_approximation(V=dat$Wind,D=0.0211,K= 0.025, nu= 15.3 * 10^(-6), taxa="flyinginsect")
+H_L2= heat_transfer_coefficient_simple(V=dat$Wind,D=0.0211)
+
+#surface area
+A= sa_from_length(l=0.0211)
+```
+
+```{r}
+#Gates calculation
+TeGates= rep(NA, nrow(dat) )
+
+ for(ind in 1:nrow(dat) ){
+   out= try(Tb_Gates(A=A, D=0.0211/3, psa_dir=0.66, psa_ref=1-0.66, psa_air=1, psa_g=0.01, T_g=dat$SoilTemp[ind]+273, T_a=dat$Tgrass[ind]+273, Qabs=A*dat$Rad[ind], epsilon=0.95, H_L=dat$H_L[ind], ef=1.3, K=0.15), silent=TRUE)
+ if(is.numeric(out))TeGates[ind]=out
+ }
+  dat$TeGates=TeGates-273.15
+```
+
+```{r}
 #aggregate by day and hour
-dat.dh= aggregate(dataug, by=list(dataug$site, dataug$hour), FUN=mean, na.rm=TRUE)
+dat.dh= suppressWarnings(aggregate(dat, by=list(dat$Elev, dat$hour, dat$J), FUN=mean, na.rm=TRUE))
+```
 
+```{r}
+#aggregate by hour
+dat.h= suppressWarnings(aggregate(dat, by=list(dat$Elev, dat$hour), FUN=mean, na.rm=TRUE))
 ```
@@
 library(reshape2)
+#convert to long format
+dat.s= melt(dat.h, id = c("Elev", "J","hour","psi"))
+
+#plot
+dat.s= subset(dat.s, dat.s$variable %in% c("Temp","SoilTemp","SanguinipesTemp","Te","Tgrass","TeGates") )
+
+ggplot(data=dat.s, aes(x=hour, y = value, color=variable))+geom_line(aes(lty=variable) )+facet_wrap(~Elev)
+
+```
+```{r}
+#plot by date time
+dat.dh$Jhr= dat.dh$J +dat.dh$hour/24
 
 #convert to long format
-dat.s= dat.dh[,c("Rad","Temp","Wind","DodgeiTemp","SoilTemp","SanguinipesTemp","BivatatusTemp","Elev","J", "hour","psi","Te")]
-
-dat.s= melt(dat.s, id = c("Elev", "J","hour","psi"))
+dat.s= melt(dat.dh, id = c("Elev", "J","hour","psi","Jhr"))
 
 #plot
-dat.s= subset(dat.s, dat.s %in% c("Temp","SoilTemp","SanguinipesTemp","Te") )
+dat.s= subset(dat.s, dat.s$variable %in% c("Temp","SoilTemp","SanguinipesTemp","Te","Tgrass","TeGates") )
 
-ggplot(data=dat.s, aes(x=hour, y = value, color=variable))+geom_line( ) #+
-#  facet_wrap(~Elev)
+plot.t=ggplot(data=dat.s, aes(x=Jhr, y = value, color=variable))+geom_line(aes(lty=variable) )+facet_grid(Elev~1)
+```
+
+```{r}
+setwd("/Volumes/GoogleDrive/Shared drives/TrEnCh/Projects/TrenchR/figures/")
+
+pdf("TeTime.pdf", height = 10, width = 15)
+plot.t
+dev.off()
 

### 2019-11-12 ManuscriptTe.Rmd
Update examples

@@
 #Microclimate calculations
-u_z=function(z, d=0.229, z_0=0.0208, u_star= 0.114, k=0.41) u_star/k*log((z-d)/z_0)
-
 ```{r}
@@
 C1.hr= as.matrix(C1.hr)
+```
+
+We can estimate surface roughness across hours:
+```{r}
 #surface roughness
-z0= apply(C1.hr[, c("Wind0.25m","Wind0.5m","Wind1m","Wind1.5m")], FUN=surface_roughness, MARGIN=1, zr=c(0.25,0.5,1.5,1.5))
+z0= apply(C1.hr[, c("Wind0.25m","Wind0.5m","Wind1m","Wind1.5m")], FUN=surface_roughness, MARGIN=1, zr=c(0.25,0.5,1.0,1.5))
 
+```
+
+We calculate the temperature and windspeed profiles for a given hour using multiple functions in TrenchR.
+```{r}
 z.seq <- seq(0, 2, by = 0.1)
 
-for(hr in 1:24){
-t.seq1 <- air_temp_profile_neutral(T_r=C1.hr[hr,"Air0.5m"], zr=0.5, z0=0.02, z=2, T_s=C1.hr[hr,"Soil"])
-t.seq2 <- air_temp_profile(T_r=C1.hr[hr,"Air0.5m"], u_r=C1.hr[hr,"Wind0.5m"], zr=0.5, z0=0.02, z=2, T_s=C1.hr[hr,"Soil"])
-t.seq3 <- air_temp_profile_segment(T_r=C1.hr[hr, c("Air0.25m","Air0.5m","Air1.0m","Air1.5m")],u_r=C1.hr[hr, c("Wind0.25m","Wind0.5m","Wind1m","Wind1.5m")], zr=c(0.25,0.5,1,1.5), z0=0.02, z=2, T_s=C1.hr[hr,"Soil"])
+#subset to day and hour
+C1.sub= subset(C1.dayhr, C1.dayhr$day==10)
+hr=10
 
-u.seq1<- wind_speed_profile_neutral(u_r=C1.hr[hr,"Wind0.5m"], zr=0.5, z0=0.02, z=2)
-u.seq2<- wind_speed_profile(u_r=C1.hr[hr,"Wind0.5m"], zr=0.5, z0=0.02, z=2)
-u.seq3<- wind_speed_profile_segment(u_r=C1.hr[hr, c("Wind0.25m","Wind0.5m","Wind1m","Wind1.5m")], zr=c(0.25,0.5,1,1.5), z0=0.02, z=2)
-}
+#estimate profiles
+#temperature
+t.seq1= sapply(z.seq, FUN=air_temp_profile_neutral, T_r=C1.sub[hr,"Air0.5m"], z0=z0[hr], zr=0.5, T_s=C1.sub[hr,"Soil"])
 
-#plot(t.seq, z.seq, type = "l", xlab = "Temperature (C)", ylab = "z (height above ground in m)")
+t.seq2= sapply(z.seq, FUN=air_temp_profile, T_r=C1.sub[hr,"Air0.5m"], u_r=C1.hr[hr,"Wind0.5m"], zr=0.5, z0=z0[hr], T_s=C1.sub[hr,"Soil"])
+
+t.seq3= sapply(z.seq, FUN=air_temp_profile_segment, T_r=C1.sub[hr, c("Air0.25m","Air0.5m","Air1.0m","Air1.5m")],u_r=C1.sub[hr, c("Wind0.25m","Wind0.5m","Wind1m","Wind1.5m")], zr=c(0.25,0.5,1,1.5), z0=rep(z0[hr],4), T_s=C1.sub[hr,"Soil"])
+
+#windspeed
+u.seq1= sapply(z.seq, FUN=wind_speed_profile_neutral, u_r=C1.sub[hr,"Wind0.5m"], zr=0.5, z0=z0[hr])
+
+u.seq2= sapply(z.seq, FUN=wind_speed_profile, u_r=C1.sub[hr,"Wind0.5m"], zr=0.5, z0=z0[hr])
+
+u.seq3= sapply(z.seq, FUN=wind_speed_profile_segment, u_r=C1.sub[hr, c("Wind0.25m","Wind0.5m","Wind1m","Wind1.5m")], zr=c(0.25,0.5,1,1.5), z0=rep(z0[hr],4))
+
+#replace infinity with N
+t.seq1[which(!is.finite(t.seq1))]=NA
+```
+
+We then plot the temperature and windspeed profiles.
+```{r}
+plot(t.seq1, z.seq, type = "l", xlab = "Temperature (C)", ylab = "z (height above ground in m)")
+points(t.seq2, z.seq, type = "l", lty="dashed")
+points(t.seq3, z.seq, type = "l", lty="dotted")
+points(C1.sub[hr, c("Air0.25m","Air0.5m","Air1.0m","Air1.5m")], c(0.25,0.5,1,1.5), type = "l", col="red")
+
+plot(u.seq1, z.seq, type = "l", xlab = "windspeed (m/s)", ylab = "z (height above ground in m)")
+points(u.seq2, z.seq, type = "l", lty="dashed")
+points(u.seq3, z.seq, type = "l", lty="dotted")
+points(C1.sub[hr, c("Wind0.25m","Wind0.5m","Wind1m","Wind1.5m")], c(0.25,0.5,1,1.5), type = "l", col="red")
 ```

### 2020-01-30 ManuscriptTe.Rmd
bug fixes

@@
 setwd("/Volumes/GoogleDrive/Shared drives/TrEnCh/Projects/TrenchR/data/")
-dat=read.csv("microclimate_C1.csv", sep=",", na.strings = c("-49.99","NA"))
+C1all=read.csv("microclimate_C1.csv", sep=",", na.strings = c("-49.99","NA"))
 ```
+
 ```{r}
 #ANALYZE SURFACE ROUGHNESS AND PROFILES
-C1all$day=  as.POSIXlt(as.character(C1all$datetime),tz='MST', format="%m/%d/%Y %H:%M")$mday
-C1all$hour=as.POSIXlt(as.character(C1all$datetime),tz='MST', format="%m/%d/%Y %H:%M")$hour
+C1all$day=as.POSIXlt(as.character(C1all$datetime), tz='MST', format="%m/%d/%Y %H:%M")$mday
+C1all$hour=as.POSIXlt(as.character(C1all$datetime), tz='MST', format="%m/%d/%Y %H:%M")$hour
 
@@
 
-u.seq3= sapply(z.seq, FUN=wind_speed_profile_segment, u_r=C1.sub[hr, c("Wind0.25m","Wind0.5m","Wind1m","Wind1.5m")], zr=c(0.25,0.5,1,1.5), z0=rep(z0[hr],4))
+u.seq3= sapply(z.seq, FUN=wind_speed_profile_segment, u_r=C1.sub[hr, c("Wind0.5m","Wind1m","Wind1.5m")], zr=c(0.5,1,1.5), z0=rep(z0[hr],3))
 
@@
 ```
+
 ```{r}

### 2020-02-05 Manuscript_full.Rmd
added the manuscript and fixed some codes/sentences

---
title: 'TrenchR: an R package for transparent environmental and ecological biophysics'
author: "Lauren B. Buckley, Bryan Briones Ortiz, Aji John, Ofir Levy, Yutaro Sakairi"
date: "February 4, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Abstract

Much understanding of climate change responses relies on the assumption that organismal body temperatures are equal to temporally averaged air temperatures high above the ground. However, climate change responses often hinge on the occurrence of acute thermal stress events due to radiative heating of organisms near the ground. We introduce the TrenchR package, which aids in Translating Environmental Change into organismal responses (https://github.com/trenchproject/TrenchR). The package facilitates microclimate modelling to translate weather station data into the environmental conditions experienced by organisms and biophysical modelling to predict organismal body temperatures given the environmental conditions. The package aims to introduce and enable microclimate and biophysical modelling to improve ecological and evolutionary forecasting and includes tutorials and well as a series of educational modules introducing microclimate and biophysical modelling.

##Introduction
Many responses of species and ecosystems to climate change are inconsistent with predictive models (Maguire et al., 2015). Some of these predictive shortcomings stem from omitting spatial and temporal environmental variation and how it interacts with organismal phenotypes (e.g., size, coloration) (Nadeau et al., 2017; Potter et al., 2013). Many analyses assume that organismal body temperatures are equal to air temperate at weather station height (usually 2m). However, air temperatures near the ground where most organisms reside can be considerably warmer (XX-XXC) and absorption of solar radiation can raise body temperatures well above air temperatures [XXXX].  Acute thermal stress events are often more relevant to the physiology and energy balance of organisms than daily, monthly, or annually aggregated environmental conditions. Yet, daily and seasonal environmental variation is often neglected (Dillon and Woods, 2016; Dillon et al., 2016). We aim to help remedy these shortcoming by introducing the TrenchR package as a component of the TrEnCh project (https://trenchproject.github.io), which builds computational and visualization tools to Translate Environmental Change into organismal responses.

The mathematical tools for leveraging environmental data to predict organismal conditions have long lingered in books (Campbell and Norman, 2000; Gates, 1980) and articles (Porter and Gates, 1969; Porter and Tracy, 1983), but adoption of the tools has not kept pace with research on climate change responses. Adoptions of biophysical approaches are expanding with recognition of the importance of spatial and temporal environmental variation to determining climate change responses (Buckley et al., 2018; Kearney and Porter, 2017).  Increasing dissemination of computational tools is facilitating adoption. The TrenchR package is intended to complement the NicheMapR package (Kearney and Porter, 2017), which includes sophisticated microclimate and biophysical models developed by Warren Porter and colleagues over decades. Many of the approaches are similar between the packages, but TrenchR focuses on simple function to aid the understanding and transparency of biophysical approaches.  The TrenchR functions allow for modular use and easy adaptation to a variety of organismal and research questions.  Although many of the functions are general, some components of TrenchR are best suited for ectothermic animals. We have minimized inclusion of plant specific biophysical models since they are the focus of several R packages [tealeaves: https://www.biorxiv.org/content/biorxiv/early/2019/01/24/529487.full.pdf; plantecophys (Duursma, 2015)].

##Components
The TrenchR package aims to promote transparency and reproducibility and we welcome contributions and corrections from users. Our package is build using the devtools methodology (https://github.com/hadley/devtools) in Githib. Issues and feature requests can be contributed in Github [XXXX]. The package is maintained and checked using Travis CI as a tool for continuous integration. TrenchR functions are organized into the following categories:

**Utility functions** allow estimating environmental metrics that form the basis of microclimate and biophysical models (e.g., zenith and declination angles), facilitate modelling diurnal temperature variation, and estimate biologically relevant aggregate metrics such as degree days available for development.

**Microclimate functions** allow estimating the environmental conditions experienced by organisms. Temperature and wind profile functions scale environmental conditions from sensor to organismal height. Radiation functions allow estimating incoming solar radiation and partitioning the variation diurnally and across components (i.e., direct, diffuse, reflected).

**General biophysical functions** provide generalized models of heat exchanges between organisms and their environment so that users can build custom biophysical models. The functions allow implementing an energy balance of the following components of heat exchange between organisms and their environment: *radiative heat exchanges* of solar and thermal radiation; *convective heat exchanges* between organisms and their surrounding fluid (air or water) driven by fluid flow; *conductive heat exchanges* between organisms and the ground due to physical contact; and *evaporative heat exchanges* associated with organisms’ evaporative water loss. Functions serve to aggregate these forms of heat exchange into energy balances and to use the energy balances to predict operative environmental temperatures. Operative environmental temperatures are the predicted temperatures of organisms with specified physical properties based on the above components of heat exchange [XXXX]. [WATER BALANCE.]

**Specific biophysical functions** provide biophysical models that have been built for specific organisms based on their physical properties, behavior, environment, and life history. We currently provide published biophysical models for lizards, butterflies, and grasshoppers and plan to include models for marine taxa (mussels and limpets). 
APIs??

We do not present each function’s biophysical equations for tractability, but most functions include a reference deriving the equations. We have incorporated default parameters (e.g., organism emissivity or heat conduction rates) when general values are available. Many functions additionally include comments that describe potential parameterizations. Biophysical texts (Campbell and Norman, 2000; Gates, 1980) include many tables with parameter values and we provide several tables in TrenchR.
[Table 8.2]

##Tutorial and introductory modules
We introduce each function in categorically grouped tutorials.  A good place to start is the *Allometry and conversions* tutorial, which provides tools for preparing data such as estimating additional dimensions of organisms from measured dimensions. The *Estimating microclimates* tutorial provides resources for estimating the environmental conditions experienced by organisms.  This includes estimating solar radiation and its components, diurnal variation in temperature and radiation, temperature and windspeed profiles, and soil temperatures and profiles. Finally, the core biophysical modelling functions are described in a tutorial on *Using energy balances to estimate body temperatures*. Components of an energy budget can be estimated using individual functions and then operative environmental temperatures, $T_e$, can be solved for using either a generic energy balance or taxa specific biophysical models.

For broader background on microclimate and biophysical modelling, we have updated a series of modules focused on “Physical processes in ecosystems”.  We intend the modules to provide a more contained and accessible introduction to microclimate and biophysical modelling than that included in classic textbooks (e.g., Campbell and Norman, 2000; Gates, 1980). The modules provide less detailed coverage of many of the topics contained in the books. The series of 14 modules starts with an overview of the calculus and physics principles underlying the modelling. Thermodynamics and energy budget modelling are then introduced with detailed examples. We have revised the tutorials (which originated from an NSF training grant in 1979 and lacked broad distribution) to include R codes and utilization of TrenchR functions. We expanded the original series of tutorials to include introductions by Michael Kearney to the Microclim environmental data (Kearney et al., 2014) and integrating the Microclim data with biophysical modelling to examine broad scale climatic limits. The modules are provided in html form at a bookdown site and as a pdf. The R markdown files are available in Github at https://github.com/trenchproject/PhysicalProcessesEcosystems/.

##Implementation
We illustrate use of the TrenchR package by estimating an energy budget for a *Sceoporus* lizard on June 1, 2018 in Santa Fe, New Mexico, USA (35.69, -105.944, elevation: 2121m). The example is also incorporated in the *Using energy balances to estimate body temperatures* tutorial. We first estimate the energy budget with component functions and we then use an integrated biophysical model to estimate operative environmental temperatures. We start by generating environmental inputs.

We will use the energy budget to estimate body temperatures, $T_b$, of organisms can depart dramatically from the air temperatures due to energy exchange with the environment.  Heat is gained from absorbing solar and thermal radiation and from metabolic reactions. Heat is lost through the organism's emission of radiation and the evaporation of water. The organism exchanges heat with the surrounding air or water via convection and with the ground via conduction. The balance of these heat exchanges can be estimated and is referred to as operative environmental temperature, $T_e$. $T_e$ indicates the steady-state temperature of an organism with given physical properties in a particular environment (Bakken, 1992).

Let's assume the lizard is in an unshaded location where the daily air temperature varies from a minimum of 10°C to a maximum of 25°C, the soil surface temperature varies from a minimum of 15°C to a maximum of 30°C, and the windspeed is 0.5 m/s.

Assume that atmospheric transmissivity $\tau=0.7$, albedo $\rho=0.6$

WE ASSUME $T_b=T_a$ TO ILLUSTRATE CALCULATIONS

###Environmental data

Let us first prepare the environmental data for analysis. We will need to estimate hourly air and soil temperatures and radiation. We start by estimating day of year and the timing of sunrise and sunset:
```{r}
library(TrenchR)
#set up input data as variables
lat=35.69; lon= -105.944; elev=2121
Tmin=10; Tmax=25; Tmin_s=15; Tmax_s=30
V=0.5
#assumptions
tau=0.7; rho=0.6
Tb0=25

doy= day_of_year("2018-06-01", format= "%Y-%m-%d")
snoon= solar_noon(lon=lon, doy=doy)
dayl= daylength(lat=lat, doy=doy)
tr= snoon-dayl/2
ts= snoon+dayl/2
```

We can estimate hourly solar radiation as follows:
```{r}
psi_deg= sapply(1:24, FUN=zenith_angle, doy=doy, lat=lat, lon=lon)
psi_rad= degree_to_radian(psi_deg) #convert to radians
Srad= sapply(psi_rad, FUN=estimate_radiation, doy=doy, tau=tau, elev=elev, rho=rho)
#Separate into direct, diffuse, and reflected solar radiation
Sdir= Srad[1,]
Sdif= Srad[2,]
Sref= Srad[3,]
```

We can then calculate hourly air and soil temperatures. We use the sine-exponential model for air temperature and the sine model for surface temperature:
```{r}
Ta= sapply(1:24, diurnal_temp_variation_sineexp, T_max=Tmax, T_min=Tmin, t_r=tr, t_s=ts, alpha=2.59, beta= 1.55, gamma=2.2)
Ts= sapply(1:24, diurnal_temp_variation_sine, T_max=Tmax_s, T_min=Tmin_s)
```

###Energy balance
We will be solving the following energy balance to estimate $T_e$:
$$Q_{net} = Q_{abs} - Q_{emit} + Q_{conv} + Q_{cond} + Q_{met} - Q_{evap},$$
We will estimate each term on the right side of the equation in turn. Estimating $Q_{abs}$requires the surface area exposed to radation and the solar absorptivity of the animal surface ($a$ proportion). We model a 10g Sceloporus lizard with solar aborptivity $a=0.9$ [Gates 1980].

```{r}
mass=10
a=0.9 #solar absorptivity

#estimate surface area (m^2) from mass (g)
A= sa_from_mass(mass, "lizard")
#estimate projected (silhouette) area as a portion of surface area
psa= sapply(psi_deg, prop_silhouette_area, taxa= "lizard", posture= "prostrate")
#change negative values to zero
psa[psa<0]=0
```

We calculate the hourly radiation asborbed as follows:
```{r}
Qabs= rep(NA, 24)
for(hr in 1:24) Qabs[hr]= Qradiation_absorbed(a=a, A=A, psa_dir=psa[hr], psa_ref=1-psa[hr], S_dir=Sdir[hr], S_dif=Sdif[hr], a_s=rho)
```

We estimate thermal radiation $Q_{emit}$ for the lizard outdoors. We assume the surface emisivity of lizards, $epsilon_s=0.965$ [Bartlett & Gates 1967].

```{r}
epsilon_s=0.965

Qemit= rep(NA, 24)
for(hr in 1:24) Qemit[hr]=Qemitted_thermal_radiation(epsilon=epsilon_s, A=A, psa_dir=psa[hr], psa_ref=1-psa[hr], T_b=Ta[hr]+273.15, T_g=Ts[hr]+273.15, T_a=Ta[hr]+273.15, enclosed=FALSE)
```

We next estimate convection $Q_{conv}$ and conduction $Q_{cond}$. We will estimate the lizard's heat transfer coefficient, $H_L$ using an empirical relationship for lizards (`heat_transfer_coefficient()`), but also illustrate a function estimating $H_L$ using a spherical approximation (`heat_transfer_coefficient_approximation()`) and a simplified approximation (`heat_transfer_coefficient_simple()`).

```{r}
#Use DRYAIR in NicheMapR to estimate thermal conductivity of air and kinematic viscocity
#library(NicheMapR)

#estimate Barometric pressure (pascal)
ap= airpressure_from_elev(elev)*1000

DRYAIRout= DRYAIR(db=Ta, bp=ap, alt=elev)
K= DRYAIRout$thcond
nu= DRYAIRout$viskin #Kinematic viscosity (m2 s-1)

svl=0.006 #SVL for sceloporus

#We will use the average of K and nu across the day for simplicity and since there's not substantial variation
K= mean(K)
nu=mean(nu)

#Estimate using empirical relationship
H_L=heat_transfer_coefficient(V=V,D=svl,K= 25.7 * 10^(-3), nu= 15.3 * 10^(-6) , taxa="lizard_surface")

#Also illustrate estimations using a spherical approximation and a simplified version of the approximation.
heat_transfer_coefficient_approximation(V=V,D=svl,K= 25.7 * 10^(-3), nu= 15.3 * 10^(-6), taxa="lizard")
heat_transfer_coefficient_simple(V=0.5,D=0.05)

```

We estimate convective heat exchange between the animal and surrounding air using ithe following relationship:
$$Q_{conv}= ef*H_L*(A*proportion)*(T_a-T_b),$$
where an enhancement factor, $ef$, multiplier can be incorported to account for increases in heat exchange resulting from air turbulance in field conditions

The function is available in R assuming that 2/3 of the lizard's surface area is exchanging heat through convection. We note we currently estimate no convection since we are assuming $Tb=Ta$ for illustration purposes:
```{r}
Qconv= rep(NA, 24)
for(hr in 1:24) Qconv[hr]= Qconvection(T_a= Ta[hr]+273.15,T_b= Ta[hr]+273.15,H=H_L,A=A, proportion=0.67, ef=1.3)
```

We estimate conductance between the lizard and surface assuming conductance through the animal tissue is the rate limiting step as follows:
$$Qcond= A*proportion*K*(T_g-T_b)/d, $$
We conduct the estimate in R assuming that 1/3 of the lizard surface is in contact with the ground and a skin thinkness of $10^{-6}$:
```{r}
Qcond= rep(NA, 24)
for(hr in 1:24) Qcond[hr]=Qconduction_animal(T_g= Ts[hr]+273.15,T_b=Ta[hr]+273.15,d=10^-6,K=K,A=A, proportion=0.33)
```

We assume, as in generally done for lizards, that heat exchange associated with metabolism and evaporation is negliglible:
```{r}
Qmet= 0.1
Qevap= 0.1
```

The full heat budget can be calculated as follows: 
```{r}
Qnet_Gates(Qabs=Qabs, Qemit=Qemit, Qconv=Qconv, Qcond=Qcond, Qmet=Qmet, Qevap=Qevap)
```
We constructed the energy balance assuming $T_b=T_a$ for illustration. We now use a function based on the energy balance above to estimate body temperature given the environmental conditions:

```{r}
Te= rep(NA, 24)
for(hr in 1:24) Te[hr]= Tb_Gates(A=A, D=svl, psa_dir=psa[hr], psa_ref=1-psa[hr], psa_air=0.67, psa_g=0.33, T_g=Ts[hr]+273.15, T_a=Ta[hr]+273.15, Qabs=Qabs[hr], epsilon=epsilon_s, H_L=H_L, ef=1.3, K=K)
```

We also implement a similiar but simplied energy balance. The energy balance omits conduction with the ground and thus underestimates temperatures, but we implement it for comparison purposes:
```{r}
Te2= rep(NA, 24)
for(hr in 1:24) Te2[hr]= Tb_CampbellNorman(T_a=Ta[hr]+273.15, S=Qabs[hr]/svl, epsilon=epsilon_s, c_p=29.3, D=svl, V=V)
#CHECK
```

We additionally estimate $T_e$ using the specialized function for lizards:
```{r}
Te3= rep(NA, 24)
for(hr in 1:24) Te3[hr]= Tb_lizard(T_a=Ta[hr], T_g=Ts[hr], u=V, svl=svl*1000, m=mass, psi=psi_deg[hr], rho_S=rho, elev=elev, doy=doy, sun=TRUE, surface=TRUE, alpha_S=a, alpha_L=0.965, epsilon_s=epsilon_s, F_d=0.8, F_r=0.5, F_a=0.5, F_g=0.5)

Te4= rep(NA, 24)
for(hr in 1:24) Te4[hr]= Tb_Fei(TAH=Ta[hr]+273.15, T_s=Ts[hr]+273.15, T_a=Ts[hr]+273.15, UDT=Qabs[hr], GLW=Qabs[hr],  shade=0)
#FIX UDT?

```

###Plot
We next plot the diurnal trends in environmental data and the operative data.

```{r}
plot(1:24, (Te-273), type="l", xlab="Hour", ylab="Temperature (°C)", col="blue") #Gates
points(1:24, Te2-273, type="l", col="blue", lty="dotted") #Campbell
points(1:24, Te3, type="l", col="blue", lty="dashed") #Buckley 2008
#points(1:24, Te4, type="l", col="red", lty="solid") #Fei et al.
points(1:24, Ta, type="l", col="orange")
points(1:24, Ts, type="l", col="purple")

#add additional axis with radiation
par(new = T)
plot(1:24, Sdir, pch=16, axes=F, xlab=NA, ylab=NA, type="l", col="green")
axis(side = 4)
mtext(side = 4, line = 3, 'Radiation')
legend("topleft",
       legend=c("Ta", "Ts", "Te", "Radiation"),
       lty=1, pch=NA, col=c("orange", "purple", "blue","green"))

```

##Summary
New environmental data and tools make biophysical modelling more accessible. Expand use since we realize.

##Literature cited
Bakken, G. S. (1992). Measurement and application of operative and standard operative temperatures in ecology. Integr. Comp. Biol. 32, 194–216.
Buckley, L. B., Cannistra, A. F. and John, A. (2018). Leveraging organismal biology to forecast the effects of climate change. Integr. Comp. Biol. doi.org/10.1093/icb/icy018,.
Campbell, G. and Norman, J. (2000). An introduction to environmental biophysics. New York: Springer.
Dillon, M. E. and Woods, H. A. (2016). Introduction to the Symposium: Beyond the Mean: Biological Impacts of Changing Patterns of Temperature Variation. Integr. Comp. Biol. icw020.
Dillon, M. E., Woods, H. A., Wang, G., Fey, S. B., Vasseur, D. A., Telemeco, R. S., Marshall, K. and Pincebourde, S. (2016). Life in the frequency domain: the biological impacts of changes in climate variability at multiple time scales. Integr. Comp. Biol. 56, 14–30.
Duursma, R. A. (2015). Plantecophys - An R Package for Analysing and Modelling Leaf Gas Exchange Data. PLOS ONE 10, e0143346.
Gates, D. M. (1980). Biophysical ecology. New York: Springer-Verlag.
Kearney, M. R. and Porter, W. P. (2017). NicheMapR–an R package for biophysical modelling: the microclimate model. Ecography 40, 664–674.
Kearney, M. R., Isaac, A. P. and Porter, W. P. (2014). microclim: Global estimates of hourly microclimate based on long-term monthly climate averages. Sci. Data 1, 140006.
Maguire, K. C., Nieto-Lugilde, D., Fitzpatrick, M. C., Williams, J. W. and Blois, J. L. (2015). Modeling species and community responses to past, present, and future episodes of climatic and ecological change. Annu. Rev. Ecol. Evol. Syst. 46, 343–368.
Nadeau, C. P., Urban, M. C. and Bridle, J. R. (2017). Coarse climate change projections for species living in a fine-scaled world. Glob. Change Biol. 23, 12–24.
Porter, W. P. and Gates, D. M. (1969). Thermodynamic equilibria of animals with environment. Ecol. Monogr. 39, 227–244.
Porter, W. P. and Tracy, C. R. (1983). Biophysical analyses of energetics, time-space utilization, and distributional limits. Harvard University Press.
Potter, K. A., Woods, H. A. and Pincebourde, S. (2013). Microclimatic challenges in global change biology. Glob. Change Biol. 19, 2932–2939.

### 2020-02-05 Manuscript_full.Rmd
fixed plot axis label

@@
 library(TrenchR)
+
 #set up input data as variables
@@
 ```{r}
-Qmet= 0.1
-Qevap= 0.1
+Qmet= 0
+Qevap= 0
 ```
@@
 ```{r}
-plot(1:24, (Te-273), type="l", xlab="Hour", ylab="Temperature (°C)", col="blue") #Gates
-points(1:24, Te2-273, type="l", col="blue", lty="dotted") #Campbell
+par(mar=c(5, 4, 4, 8) + 0.1)
+plot(1:24, (Te-273.15), type="l", xlab="Hour", ylab="Temperature (°C)", col="blue") #Gates
+points(1:24, Te2-273.15, type="l", col="blue", lty="dotted") #Campbell
 points(1:24, Te3, type="l", col="blue", lty="dashed") #Buckley 2008

### 2020-10-06 Manuscript_full.Rmd
Minor fixes

@@
 date: "February 4, 2020"
-output: html_document
+output:
+  pdf_document: default
+  html_document: default
 ---
@@
 $$Q_{conv}= ef*H_L*(A*proportion)*(T_a-T_b),$$
-where an enhancement factor, $ef$, multiplier can be incorported to account for increases in heat exchange resulting from air turbulance in field conditions
+where an enhancement factor, $ef$, multiplier can be incorporated to account for increases in heat exchange resulting from air turbulance in field conditions
 
@@
 
-We assume, as in generally done for lizards, that heat exchange associated with metabolism and evaporation is negliglible:
+We assume, as in generally done for lizards, that heat exchange associated with metabolism and evaporation is negligible:
 ```{r}
@@
 
-We also implement a similiar but simplied energy balance. The energy balance omits conduction with the ground and thus underestimates temperatures, but we implement it for comparison purposes:
+We also implement a similar but simplified energy balance. The energy balance omits conduction with the ground and thus underestimates temperatures, but we implement it for comparison purposes:
 ```{r}
@@
 Te4= rep(NA, 24)
-for(hr in 1:24) Te4[hr]= Tb_Fei(TAH=Ta[hr]+273.15, T_s=Ts[hr]+273.15, T_a=Ts[hr]+273.15, UDT=Qabs[hr], GLW=Qabs[hr],  shade=0)
-#FIX UDT?
+for(hr in 1:24) Te4[hr]= Tb_Fei(T_a=Ta[hr]+273.15, T_g=Ts[hr]+273.15, H=Qabs[hr], lw=Qabs[hr], shade=0, m=mass, Acondfact =0.1, Agradfact =0.3)
+#CHECK
 

### 2020-10-09 Manuscript_full.Rmd
Added bibliography

@@
 output:
+  html_document: default
   pdf_document: default
-  html_document: default
+bibliography: bibliography.bib
 ---
@@
 
-##Abstract
+## Abstract
 
 Much understanding of climate change responses relies on the assumption that organismal body temperatures are equal to temporally averaged air temperatures high above the ground.
-However, climate change responses often hinge on the occurrence of acute thermal stress events due to radiative heating of organisms near the ground.
+However, climate change responses often hinge on the occurrence of acute thermal stress events due to the radiative heating of organisms near the ground.
 We introduce the TrenchR package, which aids in Translating Environmental Change into organismal responses (https://github.com/trenchproject/TrenchR).
-The package facilitates microclimate modelling to translate weather station data into the environmental conditions experienced by organisms and biophysical modelling to predict organismal body temperatures given the environmental conditions.
-The package aims to introduce and enable microclimate and biophysical modelling to improve ecological and evolutionary forecasting and includes tutorials and well as a series of educational modules introducing microclimate and biophysical modelling.
+The package facilitates microclimate modeling to translate weather station data into the environmental conditions experienced by organisms and biophysical modeling to predict organismal body temperatures given the environmental conditions.
+The package aims to introduce and enable microclimate and biophysical modeling to improve ecological and evolutionary forecasting and includes tutorials and well as a series of educational modules introducing microclimate and biophysical modeling.
 
-##Introduction
-Many responses of species and ecosystems to climate change are inconsistent with predictive models (Maguire et al., 2015).
-Some of these predictive shortcomings stem from omitting spatial and temporal environmental variation and how it interacts with organismal phenotypes (e.g., size, coloration) (Nadeau et al., 2017; Potter et al., 2013).
+## Introduction
+Many responses of species and ecosystems to climate change are inconsistent with predictive models [@maguire2015modeling].
+Some of these predictive shortcomings stem from omitting spatial and temporal environmental variation and how it interacts with organismal phenotypes (e.g., size, coloration) [@nadeau2017coarse; @potter2013microclimatic].
 Many analyses assume that organismal body temperatures are equal to air temperate at weather station height (usually 2m).
@@
 Acute thermal stress events are often more relevant to the physiology and energy balance of organisms than daily, monthly, or annually aggregated environmental conditions.
-Yet, daily and seasonal environmental variation is often neglected (Dillon and Woods, 2016; Dillon et al., 2016).
-We aim to help remedy these shortcoming by introducing the TrenchR package as a component of the TrEnCh project (https://trenchproject.github.io), which builds computational and visualization tools to Translate Environmental Change into organismal responses.
+Yet, daily and seasonal environmental variation is often neglected [@dillon2016introduction; @dillon2016life].
+We aim to help remedy these shortcomings by introducing the TrenchR package as a component of the TrEnCh project (https://trenchproject.github.io), which builds computational and visualization tools to Translate Environmental Change into organismal responses.
 
-The mathematical tools for leveraging environmental data to predict organismal conditions have long lingered in books (Campbell and Norman, 2000; Gates, 1980) and articles (Porter and Gates, 1969; Porter and Tracy, 1983), but adoption of the tools has not kept pace with research on climate change responses.
-Adoptions of biophysical approaches are expanding with recognition of the importance of spatial and temporal environmental variation to determining climate change responses (Buckley et al., 2018; Kearney and Porter, 2017).
-Increasing dissemination of computational tools is facilitating adoption.
-The TrenchR package is intended to complement the NicheMapR package (Kearney and Porter, 2017), which includes sophisticated microclimate and biophysical models developed by Warren Porter and colleagues over decades.
-Many of the approaches are similar between the packages, but TrenchR focuses on simple function to aid the understanding and transparency of biophysical approaches.
+The mathematical tools for leveraging environmental data to predict organismal conditions have long lingered in books [@campbell2000introduction; @gates1980biophysical] and articles [@porter1969thermodynamic; @porter1983biophysical], but adoption of the tools has not kept pace with research on climate change responses.
+Adoptions of biophysical approaches are expanding with recognition of the importance of spatial and temporal environmental variation to determining climate change responses [@buckley2018leveraging; @kearney2017nichemapr].
+Increasing the dissemination of computational tools is facilitating adoption.
+The TrenchR package is intended to complement the NicheMapR package [@kearney2017nichemapr], which includes sophisticated microclimate and biophysical models developed by Warren Porter and colleagues over decades.
+Many of the approaches are similar between the packages, but TrenchR focuses on simple functions to aid the understanding and transparency of biophysical approaches.
 The TrenchR functions allow for modular use and easy adaptation to a variety of organismal and research questions.
 Although many of the functions are general, some components of TrenchR are best suited for ectothermic animals.
-We have minimized inclusion of plant specific biophysical models since they are the focus of several R packages [tealeaves: https://www.biorxiv.org/content/biorxiv/early/2019/01/24/529487.full.pdf; plantecophys (Duursma, 2015)].
+We have minimized the inclusion of plant-specific biophysical models since they are the focus of several R packages [tealeaves: https://www.biorxiv.org/content/biorxiv/early/2019/01/24/529487.full.pdf; plantecophys [@duursma2015plantecophys]].
 
-##Components
+## Components
 The TrenchR package aims to promote transparency and reproducibility and we welcome contributions and corrections from users.
-Our package is build using the devtools methodology (https://github.com/hadley/devtools) in Githib.
+Our package is built using the devtools methodology (https://github.com/hadley/devtools) in Github.
 Issues and feature requests can be contributed in Github [XXXX].
@@
 
-**Utility functions** allow estimating environmental metrics that form the basis of microclimate and biophysical models (e.g., zenith and declination angles), facilitate modelling diurnal temperature variation, and estimate biologically relevant aggregate metrics such as degree days available for development.
+**Utility functions** allow estimating environmental metrics that form the basis of microclimate and biophysical models (e.g., zenith and declination angles), facilitate modeling diurnal temperature variation, and estimate biologically relevant aggregate metrics such as degree days available for development.
 
 **Microclimate functions** allow estimating the environmental conditions experienced by organisms.
-Temperature and wind profile functions scale environmental conditions from sensor to organismal height.
+Temperature and wind profile functions scale environmental conditions from sensors to organismal height.
 Radiation functions allow estimating incoming solar radiation and partitioning the variation diurnally and across components (i.e., direct, diffuse, reflected).
@@
 Many functions additionally include comments that describe potential parameterizations.
-Biophysical texts (Campbell and Norman, 2000; Gates, 1980) include many tables with parameter values and we provide several tables in TrenchR.
+Biophysical texts [@campbell2000introduction; @gates1980biophysical] include many tables with parameter values and we provide several tables in TrenchR.
 [Table 8.2]
 
-##Tutorial and introductory modules
+## Tutorial and introductory modules
 We introduce each function in categorically grouped tutorials.
@@
 This includes estimating solar radiation and its components, diurnal variation in temperature and radiation, temperature and windspeed profiles, and soil temperatures and profiles.
-Finally, the core biophysical modelling functions are described in a tutorial on *Using energy balances to estimate body temperatures*.
+Finally, the core biophysical modeling functions are described in a tutorial on *Using energy balances to estimate body temperatures*.
 Components of an energy budget can be estimated using individual functions and then operative environmental temperatures, $T_e$, can be solved for using either a generic energy balance or taxa specific biophysical models.
 
-For broader background on microclimate and biophysical modelling, we have updated a series of modules focused on “Physical processes in ecosystems”.
-We intend the modules to provide a more contained and accessible introduction to microclimate and biophysical modelling than that included in classic textbooks (e.g., Campbell and Norman, 2000; Gates, 1980).
+For a broader background on microclimate and biophysical modeling, we have updated a series of modules focused on “Physical processes in ecosystems”.
+We intend the modules to provide a more contained and accessible introduction to microclimate and biophysical modeling than that included in classic textbooks [@campbell2000introduction; @gates1980biophysical].
 The modules provide less detailed coverage of many of the topics contained in the books.
-The series of 14 modules starts with an overview of the calculus and physics principles underlying the modelling.
-Thermodynamics and energy budget modelling are then introduced with detailed examples.
+The series of 14 modules starts with an overview of the calculus and physics principles underlying the modeling.
+Thermodynamics and energy budget modeling are then introduced with detailed examples.
 We have revised the tutorials (which originated from an NSF training grant in 1979 and lacked broad distribution) to include R codes and utilization of TrenchR functions.
-We expanded the original series of tutorials to include introductions by Michael Kearney to the Microclim environmental data (Kearney et al., 2014) and integrating the Microclim data with biophysical modelling to examine broad scale climatic limits.
-The modules are provided in html form at a bookdown site and as a pdf.
+We expanded the original series of tutorials to include introductions by Michael Kearney to the Microclim environmental data [-@kearney2014microclim] and integrating the Microclim data with biophysical modeling to examine broad-scale climatic limits.
+The modules are provided in HTML form at a bookdown site and as a pdf.
 The R markdown files are available in Github at https://github.com/trenchproject/PhysicalProcessesEcosystems/.
 
-##Implementation
-We illustrate use of the TrenchR package by estimating an energy budget for a *Sceoporus* lizard on June 1, 2018 in Santa Fe, New Mexico, USA (35.69, -105.944, elevation: 2121m).
+## Implementation
+We illustrate the use of the TrenchR package by estimating an energy budget for a *Sceloporus* lizard on June 1, 2018, in Santa Fe, New Mexico, USA (35.69, -105.944, elevation: 2121m).
 The example is also incorporated in the *Using energy balances to estimate body temperatures* tutorial.
@@
 The balance of these heat exchanges can be estimated and is referred to as operative environmental temperature, $T_e$.
-$T_e$ indicates the steady-state temperature of an organism with given physical properties in a particular environment (Bakken, 1992).
+$T_e$ indicates the steady-state temperature of an organism with given physical properties in a particular environment [@bakken1992measurement].
 
@@
 
-WE ASSUME $T_b=T_a$ TO ILLUSTRATE CALCULATIONS
+We assume $T_b=T_a$ to illustrate calculations
 
-###Environmental data
+### Environmental data
 
@@
 We will need to estimate hourly air and soil temperatures and radiation.
-We start by estimating day of year and the timing of sunrise and sunset:
+We start by estimating the day of year and the timing of sunrise and sunset:
 ```{r}
@@
 
-###Energy balance
+### Energy balance
 We will be solving the following energy balance to estimate $T_e$:
@@
 We will estimate each term on the right side of the equation in turn.
-Estimating $Q_{abs}$requires the surface area exposed to radation and the solar absorptivity of the animal surface ($a$ proportion).
-We model a 10g Sceloporus lizard with solar aborptivity $a=0.9$ [Gates 1980].
+Estimating $Q_{abs}$requires the surface area exposed to radiation and the solar absorptivity of the animal surface ($a$ proportion).
+We model a 10g Sceloporus lizard with solar absorptivity $a=0.9$ [@gates1980biophysical].
 
@@
 
-We calculate the hourly radiation asborbed as follows:
+We calculate the hourly radiation absorbed as follows:
 ```{r}
@@
 
-We estimate thermal radiation $Q_{emit}$ for the lizard outdoors. We assume the surface emisivity of lizards, $epsilon_s=0.965$ [Bartlett & Gates 1967].
+We estimate thermal radiation $Q_{emit}$ for the lizard outdoors. We assume the surface emissivity of lizards, $epsilon_s=0.965$ [Bartlett & Gates 1967].
 
@@
 ```{r}
-#Use DRYAIR in NicheMapR to estimate thermal conductivity of air and kinematic viscocity
+#Use DRYAIR in NicheMapR to estimate the thermal conductivity of air and kinematic viscosity
 #library(NicheMapR)
@@
 
-#We will use the average of K and nu across the day for simplicity and since there's not substantial variation
+#We will use the average of K and nu across the day for simplicity and since there's not a substantial variation
 K= mean(K)
@@
 
-#Estimate using empirical relationship
+#Estimate using an empirical relationship
 H_L=heat_transfer_coefficient(V=V,D=svl,K= 25.7 * 10^(-3), nu= 15.3 * 10^(-6) , taxa="lizard_surface")
@@
 We estimate convective heat exchange between the animal and surrounding air using ithe following relationship:
-$$Q_{conv}= ef*H_L*(A*proportion)*(T_a-T_b),$$
+$$Q_{conv}= ef \cdot H_L(A\cdot \mbox{proportion})(T_a-T_b),$$
 where an enhancement factor, $ef$, multiplier can be incorporated to account for increases in heat exchange resulting from air turbulance in field conditions
@@
 We estimate conductance between the lizard and surface assuming conductance through the animal tissue is the rate limiting step as follows:
-$$Qcond= A*proportion*K*(T_g-T_b)/d, $$
+$$Q_{cond}= A \cdot \mbox{proportion} \cdot K(T_g-T_b)/d, $$
 We conduct the estimate in R assuming that 1/3 of the lizard surface is in contact with the ground and a skin thinkness of $10^{-6}$:
@@
 
-###Plot
+### Plot
 We next plot the diurnal trends in environmental data and the operative data.
@@
        legend=c("Ta", "Ts", "Te", "Radiation"),
-       lty=1, pch=NA, col=c("orange", "purple", "blue","green"))
+       lty=1, pch=NA, col=c("orange", "purple", "blue", "green"))
 
@@
 
-##Summary
-New environmental data and tools make biophysical modelling more accessible. Expand use since we realize.
+## Summary
+New environmental data and tools make biophysical modeling more accessible. Expand use since we realize.
 
-##Literature cited
-Bakken, G. S. (1992). Measurement and application of operative and standard operative temperatures in ecology. Integr. Comp. Biol. 32, 194–216.
-Buckley, L. B., Cannistra, A. F. and John, A. (2018). Leveraging organismal biology to forecast the effects of climate change. Integr. Comp. Biol. doi.org/10.1093/icb/icy018,.
-Campbell, G. and Norman, J. (2000). An introduction to environmental biophysics. New York: Springer.
-Dillon, M. E. and Woods, H. A. (2016). Introduction to the Symposium: Beyond the Mean: Biological Impacts of Changing Patterns of Temperature Variation. Integr. Comp. Biol. icw020.
-Dillon, M.
-E., Woods, H.
-A., Wang, G., Fey, S.
-B., Vasseur, D.
-A., Telemeco, R.
-S., Marshall, K.
-and Pincebourde, S.
-(2016).
-Life in the frequency domain: the biological impacts of changes in climate variability at multiple time scales.
-Integr.
-Comp.
-Biol.
-56, 14–30.
-Duursma, R. A. (2015). Plantecophys - An R Package for Analysing and Modelling Leaf Gas Exchange Data. PLOS ONE 10, e0143346.
-Gates, D. M. (1980). Biophysical ecology. New York: Springer-Verlag.
-Kearney, M. R. and Porter, W. P. (2017). NicheMapR–an R package for biophysical modelling: the microclimate model. Ecography 40, 664–674.
-Kearney, M. R., Isaac, A. P. and Porter, W. P. (2014). microclim: Global estimates of hourly microclimate based on long-term monthly climate averages. Sci. Data 1, 140006.
-Maguire, K.
-C., Nieto-Lugilde, D., Fitzpatrick, M.
-C., Williams, J.
-W.
-and Blois, J.
-L.
-(2015).
-Modeling species and community responses to past, present, and future episodes of climatic and ecological change.
-Annu.
-Rev.
-Ecol.
-Evol.
-Syst.
-46, 343–368.
-Nadeau, C. P., Urban, M. C. and Bridle, J. R. (2017). Coarse climate change projections for species living in a fine-scaled world. Glob. Change Biol. 23, 12–24.
-Porter, W. P. and Gates, D. M. (1969). Thermodynamic equilibria of animals with environment. Ecol. Monogr. 39, 227–244.
-Porter, W. P. and Tracy, C. R. (1983). Biophysical analyses of energetics, time-space utilization, and distributional limits. Harvard University Press.
-Potter, K. A., Woods, H. A. and Pincebourde, S. (2013). Microclimatic challenges in global change biology. Glob. Change Biol. 19, 2932–2939.
+## References

### 2020-11-05 Manuscript_full.Rmd
update text

@@
 title: 'TrenchR: an R package for transparent environmental and ecological biophysics'
-author: "Lauren B. Buckley, Bryan Briones Ortiz, Aji John, Ofir Levy, Yutaro Sakairi"
+author: "Lauren B. Buckley, Bryan Briones Ortiz, Aji John, Ofir Levy, Yutaro Sakairi,..."
 date: "February 4, 2020"
 output:
+  word_document: default
   html_document: default
@@
 Much understanding of climate change responses relies on the assumption that organismal body temperatures are equal to temporally averaged air temperatures high above the ground.
-However, climate change responses often hinge on the occurrence of acute thermal stress events due to the radiative heating of organisms near the ground.
+However, most organisms live near the ground and acute exposure to solar radiation can substantially elevate their body temperatures.
 We introduce the TrenchR package, which aids in Translating Environmental Change into organismal responses (https://github.com/trenchproject/TrenchR).
-The package facilitates microclimate modeling to translate weather station data into the environmental conditions experienced by organisms and biophysical modeling to predict organismal body temperatures given the environmental conditions.
-The package aims to introduce and enable microclimate and biophysical modeling to improve ecological and evolutionary forecasting and includes tutorials and well as a series of educational modules introducing microclimate and biophysical modeling.
+The package includes microclimate models to vertically scale weather station data to organismal heights.
+The biophysical modelling tools include both general models for heat flows and specific models to predict body temperatures for a variety of ectothermic taxa.
+Additional functions model and temporally partition air and soil temperatures and solar radiation.
+The package aims to introduce and enable microclimate and biophysical modelling to improve ecological and evolutionary forecasting and includes tutorials and well as a series of educational modules.
 
-## Introduction
-Many responses of species and ecosystems to climate change are inconsistent with predictive models [@maguire2015modeling].
+## Background
+Many responses of species and ecosystems to climate change are inconsistent with predictive most current models [@maguire2015modeling].
 Some of these predictive shortcomings stem from omitting spatial and temporal environmental variation and how it interacts with organismal phenotypes (e.g., size, coloration) [@nadeau2017coarse; @potter2013microclimatic].
 Many analyses assume that organismal body temperatures are equal to air temperate at weather station height (usually 2m).
-However, air temperatures near the ground where most organisms reside can be considerably warmer (XX-XXC) and absorption of solar radiation can raise body temperatures well above air temperatures [XXXX].
+However, air temperatures near the ground where most organisms reside can be considerably warmer and absorption of solar radiation can raise body temperatures well above air temperatures, resulting in potential discrepancies of tens of degrees C [Kearney et al., 2009; Sinclair et al., 2016; Sunday et al., 2014].
 Acute thermal stress events are often more relevant to the physiology and energy balance of organisms than daily, monthly, or annually aggregated environmental conditions.
@@
 Adoptions of biophysical approaches are expanding with recognition of the importance of spatial and temporal environmental variation to determining climate change responses [@buckley2018leveraging; @kearney2017nichemapr].
-Increasing the dissemination of computational tools is facilitating adoption.
-The TrenchR package is intended to complement the NicheMapR package [@kearney2017nichemapr], which includes sophisticated microclimate and biophysical models developed by Warren Porter and colleagues over decades.
+Increasing dissemination of computational tools is facilitating adoption.
+The TrenchR package is intended to complement the NicheMapR package [@kearney2017nichemapr], which includes sophisticated microclimate and biophysical models.
 Many of the approaches are similar between the packages, but TrenchR focuses on simple functions to aid the understanding and transparency of biophysical approaches.
 The TrenchR functions allow for modular use and easy adaptation to a variety of organismal and research questions.
-Although many of the functions are general, some components of TrenchR are best suited for ectothermic animals.
-We have minimized the inclusion of plant-specific biophysical models since they are the focus of several R packages [tealeaves: https://www.biorxiv.org/content/biorxiv/early/2019/01/24/529487.full.pdf; plantecophys [@duursma2015plantecophys]].
+Some energy budget calculation components of biophysical models are available in the ThermImage package (Tattersall, 2017).
+Although many of the TrenchR functions are general, some components of TrenchR are best suited for ectothermic animals.
+We have minimized inclusion of plant specific biophysical models since they are the focus of several R packages [tealeaves: (Muir, 2019); plantecophys [@duursma2015plantecophys]].
 
-## Components
+## Methods and features
 The TrenchR package aims to promote transparency and reproducibility and we welcome contributions and corrections from users.
 Our package is built using the devtools methodology (https://github.com/hadley/devtools) in Github.
-Issues and feature requests can be contributed in Github [XXXX].
+Issues and feature requests can be contributed in Github (https://github.com/trenchproject/TrenchR/).
 The package is maintained and checked using Travis CI as a tool for continuous integration.
+
+Functions were adapted from biophysical ecology textbooks [@campbell2000introduction; @gates1980biophysical] and research articles.
+The source reference is included in each function’s header.
+We have incorporated default parameters (e.g., organism emissivity or heat conduction rates) when general values are available.
+Many functions additionally include comments that describe potential parameterizations.
+The biophysical ecology texts [@campbell2000introduction; @gates1980biophysical] include many tables with parameter values and we provide several tables in TrenchR [Table 8.2XXXX].
+
+### Components
 TrenchR functions are organized into the following categories:
-
 **Utility functions** allow estimating environmental metrics that form the basis of microclimate and biophysical models (e.g., zenith and declination angles), facilitate modeling diurnal temperature variation, and estimate biologically relevant aggregate metrics such as degree days available for development.
@@
 The functions allow implementing an energy balance of the following components of heat exchange between organisms and their environment: *radiative heat exchanges* of solar and thermal radiation; *convective heat exchanges* between organisms and their surrounding fluid (air or water) driven by fluid flow; *conductive heat exchanges* between organisms and the ground due to physical contact; and *evaporative heat exchanges* associated with organisms’ evaporative water loss.
-Functions serve to aggregate these forms of heat exchange into energy balances and to use the energy balances to predict operative environmental temperatures.
-Operative environmental temperatures are the predicted temperatures of organisms with specified physical properties based on the above components of heat exchange [XXXX].
-[WATER BALANCE.]
+Additional functions aggregate these forms of heat exchange into energy balances and to use the energy balances to predict body temperatures.
 
 **Specific biophysical functions** provide biophysical models that have been built for specific organisms based on their physical properties, behavior, environment, and life history.
-We currently provide published biophysical models for lizards, butterflies, and grasshoppers and plan to include models for marine taxa (mussels and limpets).
-APIs??
+We currently provide published biophysical models for lizards, salamanders, butterflies, grasshoppers, limpets, mussels, and snails.
+Most models predict operative environmental temperatures, which are the equilibrium body temperatures of organisms with specified physical properties in a specific microenvironment and assume no heat exchange via metabolism or evaporation [@bakken1992measurement].
 
-We do not present each function’s biophysical equations for tractability, but most functions include a reference deriving the equations.
-We have incorporated default parameters (e.g., organism emissivity or heat conduction rates) when general values are available.
-Many functions additionally include comments that describe potential parameterizations.
-Biophysical texts [@campbell2000introduction; @gates1980biophysical] include many tables with parameter values and we provide several tables in TrenchR.
-[Table 8.2]
-
-## Tutorial and introductory modules
-We introduce each function in categorically grouped tutorials.
+### Vignettes and introductory modules
+We introduce the function in several vignettes.
 A good place to start is the *Allometry and conversions* tutorial, which provides tools for preparing data such as estimating additional dimensions of organisms from measured dimensions.
@@
 This includes estimating solar radiation and its components, diurnal variation in temperature and radiation, temperature and windspeed profiles, and soil temperatures and profiles.
-Finally, the core biophysical modeling functions are described in a tutorial on *Using energy balances to estimate body temperatures*.
-Components of an energy budget can be estimated using individual functions and then operative environmental temperatures, $T_e$, can be solved for using either a generic energy balance or taxa specific biophysical models.
+Finally, the core biophysical modelling functions are described in a tutorial on *Using energy balances to estimate body temperatures*.
+Components of an energy budget can be estimated using individual functions and then body temperatures can be solved for using either a generic energy balance or taxa specific biophysical models.
 
-For a broader background on microclimate and biophysical modeling, we have updated a series of modules focused on “Physical processes in ecosystems”.
-We intend the modules to provide a more contained and accessible introduction to microclimate and biophysical modeling than that included in classic textbooks [@campbell2000introduction; @gates1980biophysical].
+For additional background on microclimate and biophysical modelling, we have updated a series of educational modules entitled “Physical processes in ecosystems”.
+We intend the modules to provide a more contained and accessible introduction to microclimate and biophysical modelling than that included in classic textbooks [@campbell2000introduction; @gates1980biophysical].
 The modules provide less detailed coverage of many of the topics contained in the books.
-The series of 14 modules starts with an overview of the calculus and physics principles underlying the modeling.
-Thermodynamics and energy budget modeling are then introduced with detailed examples.
-We have revised the tutorials (which originated from an NSF training grant in 1979 and lacked broad distribution) to include R codes and utilization of TrenchR functions.
-We expanded the original series of tutorials to include introductions by Michael Kearney to the Microclim environmental data [-@kearney2014microclim] and integrating the Microclim data with biophysical modeling to examine broad-scale climatic limits.
-The modules are provided in HTML form at a bookdown site and as a pdf.
-The R markdown files are available in Github at https://github.com/trenchproject/PhysicalProcessesEcosystems/.
+The series of 14 modules starts with an overview of the calculus and physics principles underlying the modelling.
+Thermodynamics and energy budget modelling are then introduced with detailed examples.
+Modules address the climate space concept, operative temperatures, biophysical models for leaves and sheep, and heat flow in soils.
+We have revised the modules (which originated from an NSF training grant in 1979 and lacked broad distribution) to include R code and utilization of TrenchR functions.
+We expanded the original series of tutorials to include introductions by Michael Kearney to the Microclim environmental data [@kearney2014microclim] and integrating the Microclim data with biophysical modelling to examine broad scale climatic limits.
+The modules are provided in html form and available for download as a pdf at the bookdown server (https://bookdown.org/huckley/Physical_Processes_In_Ecosystems/).
+The R markdown files are available in Github at https://github.com/trenchproject/TrenchRmodules.
 
-## Implementation
+## Examples
 We illustrate the use of the TrenchR package by estimating an energy budget for a *Sceloporus* lizard on June 1, 2018, in Santa Fe, New Mexico, USA (35.69, -105.944, elevation: 2121m).
@@
 
-We will use the energy budget to estimate body temperatures, $T_b$, of organisms can depart dramatically from the air temperatures due to energy exchange with the environment.
+We will use the energy budget to estimate the lizard's body temperature, $T_b$, which can depart dramatically from the air temperatures due to energy exchange with the environment.
 Heat is gained from absorbing solar and thermal radiation and from metabolic reactions.
@@
 The organism exchanges heat with the surrounding air or water via convection and with the ground via conduction.
-The balance of these heat exchanges can be estimated and is referred to as operative environmental temperature, $T_e$.
-$T_e$ indicates the steady-state temperature of an organism with given physical properties in a particular environment [@bakken1992measurement].
+The balance of these heat exchanges can be estimated and often referred to as operative environmental temperature, $T_e$ [@bakken1992measurement].
 
 Let's assume the lizard is in an unshaded location where the daily air temperature varies from a minimum of 10°C to a maximum of 25°C, the soil surface temperature varies from a minimum of 15°C to a maximum of 30°C, and the windspeed is 0.5 m/s.
-
-Assume that atmospheric transmissivity $\tau=0.7$, albedo $\rho=0.6$
-
-We assume $T_b=T_a$ to illustrate calculations
+We assume that atmospheric transmissivity $\tau=0.7$ and albedo $\rho=0.6$
 
@@
 #set up input data as variables
-lat=35.69; lon= -105.944; elev=2121
-Tmin=10; Tmax=25; Tmin_s=15; Tmax_s=30
-V=0.5
+lat=35.69; lon= -105.944; elev=2121 #latitude and longitude (degrees) and elevation (m)
+Tmin=10; Tmax=25; Tmin_s=15; Tmax_s=30 #minimum and maximum of air and soil temperatures, respectively (C)
+V=0.5 #wind speed (m/s)
 #assumptions
-tau=0.7; rho=0.6
-Tb0=25
+tau=0.7; rho=0.6 #atmospheric transmissivity and albedo
+Tb0=25 #initial assumption of body temperature (C)
 
 doy= day_of_year("2018-06-01", format= "%Y-%m-%d") 
-snoon= solar_noon(lon=lon, doy=doy)
-dayl= daylength(lat=lat, doy=doy)
-tr= snoon-dayl/2
-ts= snoon+dayl/2
+snoon= solar_noon(lon=lon, doy=doy) #estimate solar noon
+dayl= daylength(lat=lat, doy=doy) #estimate daylength
+tr= snoon-dayl/2 #time of sunrise
+ts= snoon+dayl/2 #time of sunset
 ```
@@
 ```{r}
-psi_deg= sapply(1:24, FUN=zenith_angle, doy=doy, lat=lat, lon=lon)
+psi_deg= sapply(1:24, FUN=zenith_angle, doy=doy, lat=lat, lon=lon) #estimate zenith angle (degrees)
 psi_rad= degree_to_radian(psi_deg) #convert to radians
-Srad= sapply(psi_rad, FUN=estimate_radiation, doy=doy, tau=tau, elev=elev, rho=rho)
+Srad= sapply(psi_rad, FUN=estimate_radiation, doy=doy, tau=tau, elev=elev, rho=rho) #estimate radiation
 #Separate into direct, diffuse, and reflected solar radiation
@@
 We model a 10g Sceloporus lizard with solar absorptivity $a=0.9$ [@gates1980biophysical].
+We will initially assume $T_b=T_a$ to illustrate calculations.
 

### 2020-11-10 ManuscriptTe.Rmd
Update ms

@@
 date: "`r Sys.Date()`"
-output: rmarkdown::html_vignette
-vignette: >
-  %\VignetteIndexEntry{Function testing}
-  %\VignetteEngine{knitr::rmarkdown}
-  %\VignetteEncoding{UTF-8}
+output:
+  word_document: default
+  html_document: default
+  pdf_document: default
+  rmarkdown::html_vignette: default
 ---
@@
 library(ggplot2)
+library(reshape)
 ```
 
-#Microclimate calculations
+#APPENDIX S1
+We provide an additional example using microclimate data measured along an elevation gradient in CO, USA to illustrate use of TrenchR.
+We first use a times series of air temperatures and wind speeds collected at multiple heights to examine profiles and surface roughness.
+We then use a time series of air and surface temperature, windspeed, and solar radiation collected a a single height to implement and test a biophysical model for grasshoppers.
+
+#Microclimate models
 ```{r}
@@
 
+We first examine temperature and wind speed profiles and use them to estimate surface roughness.
 ```{r}
@@
 
-We can estimate surface roughness across hours:
+We can use wind speed variation across elevations to estimate surface roughness across hours.
 ```{r}
@@
 
-We calculate the temperature and windspeed profiles for a given hour using multiple functions in TrenchR.
+We calculate the temperature and wind speed profiles for a given hour using multiple functions in TrenchR.
 ```{r}
@@
 
-u.seq2= sapply(z.seq, FUN=wind_speed_profile, u_r=C1.sub[hr,"Wind0.5m"], zr=0.5, z0=z0[hr])
+#u.seq2= sapply(z.seq, FUN=wind_speed_profile, u_r=C1.sub[hr,"Wind0.5m"], zr=0.5, z0=z0[hr])
 
@@
 
-We then plot the temperature and windspeed profiles.
+We then plot the temperature and wind speed profiles.
 ```{r}
@@
 points(t.seq3, z.seq, type = "l", lty="dotted")
-points(C1.sub[hr, c("Air0.25m","Air0.5m","Air1.0m","Air1.5m")], c(0.25,0.5,1,1.5), type = "l", col="red")
+points(C1.sub[hr, c("Air0.25m","Air0.5m","Air1.0m","Air1.5m")], c(0.25,0.5,1,1.5), type = "p", col="red")
 
 plot(u.seq1, z.seq, type = "l", xlab = "windspeed (m/s)", ylab = "z (height above ground in m)")
-points(u.seq2, z.seq, type = "l", lty="dashed")
+#points(u.seq2, z.seq, type = "l", lty="dashed")
 points(u.seq3, z.seq, type = "l", lty="dotted")
-points(C1.sub[hr, c("Wind0.25m","Wind0.5m","Wind1m","Wind1.5m")], c(0.25,0.5,1,1.5), type = "l", col="red")
+points(C1.sub[hr, c("Wind0.25m","Wind0.5m","Wind1m","Wind1.5m")], c(0.25,0.5,1,1.5), type = "p", col="red")
 ```
 
-#Te calculation
+#Biophysical models
+We first read and process time series of environmental data.
 ```{r}
-
 #Grasshopper data
@@
 
-#ditch problematic temperatures
+#omit problematic temperatures
 dat$Temp[dat$Temp< -10]=NA
@@
 dat$Elev=sites[match1,"Elevation"]
-
-#----------------
-
-z<-0.001 #specify distance from ground
 
@@
 dat= subset(dataug, dataug$Rad>5)
+```
+
+We scale the air temperature data from 0.5m to grasshopper height (0.001m).
+```{r}
+z<-0.001 #specify distance from ground
 
@@
 
+We then use biophysical models to estimate body temperatures based on air and surface temperatures, wind speed, and solar radiation.
+We first use a biophysical model specifically configured for grasshoppers.
 ```{r}
 ##Calculate Te for each species
-dat$Te<-Tb_grasshopper(T_a=dat$Tgrass, T_g=dat$SoilTemp, u=dat$Wind, H=dat$Rad, K_t=0.7, psi=dat$psi, L=0.0211, Acondfact=0.01)  #T_a=dat$Tgrass
+dat$Te<-Tb_grasshopper(T_a=dat$Tgrass, T_g=dat$SoilTemp, u=dat$Wind, H=dat$Rad, K_t=0.7, psi=dat$psi, L=0.0211, Acondfact=0.01)
 #fix infinite
@@
 
+We then implement a generic energy budget. We first illustrate the estimation of convective heat transfer coefficients as a function of wind speed.
 ```{r}
-#estimate convective heat transfer coefficient
-
-#Estimate using empirical relationship
+#Estimate heat transfer coefficient using empirical relationship
 dat$H_L=heat_transfer_coefficient(V=dat$Wind,D=0.0211,K=0.025, nu= 15.3 * 10^(-6) , taxa="cylinder")
@@
 
+We then implement the energy budget as follows.
 ```{r}
@@
 
-```{r}
-#aggregate by day and hour
-dat.dh= suppressWarnings(aggregate(dat, by=list(dat$Elev, dat$hour, dat$J), FUN=mean, na.rm=TRUE))
-```
-
+We aggregate the data and plot average conditions across sites.
 ```{r}
@@
 dat.h= suppressWarnings(aggregate(dat, by=list(dat$Elev, dat$hour), FUN=mean, na.rm=TRUE))
-```
+#aggregate by day and hour
+dat.dh= suppressWarnings(aggregate(dat, by=list(dat$Elev, dat$hour, dat$J), FUN=mean, na.rm=TRUE))
 
-```{r}
-library(reshape2)
 #convert to long format
@@
 #plot
-dat.s= subset(dat.s, dat.s$variable %in% c("Temp","SoilTemp","SanguinipesTemp","Te","Tgrass","TeGates") )
+dat.s= subset(dat.s, dat.s$variable %in% c("Temp","SoilTemp","SanguinipesTemp","Te","Tgrass") )
+#,"TeGates"
 
-plot.t=ggplot(data=dat.s, aes(x=Jhr, y = value, color=variable))+geom_line(aes(lty=variable) )+facet_grid(Elev~1)
+plot.t= ggplot(data=dat.s, aes(x=Jhr, y = value, color=variable))+geom_line(aes(lty=variable))+facet_grid(Elev~1)
 ```
@@
 
-pdf("TeTime.pdf", height = 10, width = 15)
+pdf("TeTime2.pdf", height = 10, width = 15)
 plot.t

### 2020-11-10 Manuscript_full.Rmd
Update ms

@@
 ## Background
-Many responses of species and ecosystems to climate change are inconsistent with predictive most current models [@maguire2015modeling].
+Many responses of organisms and ecosystems to climate change are inconsistent with current predictive models [@maguire2015modeling].
 Some of these predictive shortcomings stem from omitting spatial and temporal environmental variation and how it interacts with organismal phenotypes (e.g., size, coloration) [@nadeau2017coarse; @potter2013microclimatic].
 Many analyses assume that organismal body temperatures are equal to air temperate at weather station height (usually 2m).
-However, air temperatures near the ground where most organisms reside can be considerably warmer and absorption of solar radiation can raise body temperatures well above air temperatures, resulting in potential discrepancies of tens of degrees C [Kearney et al., 2009; Sinclair et al., 2016; Sunday et al., 2014].
+However, air temperatures near the ground where most organisms reside can be considerably warmer and absorption of solar radiation can raise body temperatures well above air temperatures, resulting in potential discrepancies of tens of degrees C [@kearney2009potential; @sinclair2016can; @sunday2014thermal].
 Acute thermal stress events are often more relevant to the physiology and energy balance of organisms than daily, monthly, or annually aggregated environmental conditions.
@@
 Increasing dissemination of computational tools is facilitating adoption.
-The TrenchR package is intended to complement the NicheMapR package [@kearney2017nichemapr], which includes sophisticated microclimate and biophysical models.
+The TrenchR package is intended to complement the NicheMapR package, which includes sophisticated microclimate [@kearney2017nichemapr] and biophysical [@kearney2020nichemapr] models.
 Many of the approaches are similar between the packages, but TrenchR focuses on simple functions to aid the understanding and transparency of biophysical approaches.
 The TrenchR functions allow for modular use and easy adaptation to a variety of organismal and research questions.
-Some energy budget calculation components of biophysical models are available in the ThermImage package (Tattersall, 2017).
+Some energy budget calculation components of biophysical models are also available in the ThermImage package [@tattersall2019thermimage].
 Although many of the TrenchR functions are general, some components of TrenchR are best suited for ectothermic animals.
-We have minimized inclusion of plant specific biophysical models since they are the focus of several R packages [tealeaves: (Muir, 2019); plantecophys [@duursma2015plantecophys]].
+We have minimized inclusion of plant specific biophysical models since they are the focus of several R packages [tealeaves: [@muir2019tealeaves]; plantecophys [@duursma2015plantecophys]].
 
@@
 We start by generating environmental inputs.
+See Appendix S1 for another example examining a time series of microcliamte data and applying and testing the biophysical models.
 
@@
 
-Let's assume the lizard is in an unshaded location where the daily air temperature varies from a minimum of 10°C to a maximum of 25°C, the soil surface temperature varies from a minimum of 15°C to a maximum of 30°C, and the windspeed is 0.5 m/s.
+Let's assume the lizard is in an unshaded location where a weather station at standard height (2m) indicates that the daily air temperature varies from a minimum of 10°C to a maximum of 25°C and the wind speed averages 1 m/s.
+The soil surface temperature varies from a minimum of 15°C to a maximum of 30°C.
 We assume that atmospheric transmissivity $\tau=0.7$ and albedo $\rho=0.6$
@@
 
+Tb_CampbellNorman=function(T_a, T_g, S, alpha_L=0.96, epsilon=0.96, c_p=29.3, D, V){
+
+  stopifnot(T_a>200, T_a<400, epsilon>=0.5, epsilon<=1, c_p>=0, D>0, V>=0)
+
+  #Stefan-Boltzmann constant
+  sigma= 5.673*10^(-8) #W m^(-2) K^(-4)
+
+  #solar and thermal radiation absorbed
+  L_a=sigma*T_a^4  # (10.7) long wave flux densities from atmosphere 
+  L_g=sigma*T_g^4  # (10.7) long wave flux densities from ground
+  F_a=0.5; F_g=0.5 #proportion of organism exposure to air and ground, respectively
+  R_abs= S+alpha_L*(F_a*L_a+F_g*L_g) # (11.14) Absorbed radiation
+
+  #thermal radiation emitted
+  Qemit= epsilon*sigma*T_a^4
+
+  #conductance
+  g_Ha=1.4*0.135*sqrt(V/D) # boundary conductance, factor of 1.4 to account for increased convection (Mitchell 1976), assumes forced conduction
+  g_r= 4*epsilon*sigma*T_a^3/c_p # (12.7) radiative conductance
+
+  # operative environmental temperature
+  T_e=T_a+(R_abs-Qemit)/(c_p*(g_r+g_Ha))
+
+  return(T_e) 
+}
+```
+
+```{r}
 #set up input data as variables
@@
 Tmin=10; Tmax=25; Tmin_s=15; Tmax_s=30 #minimum and maximum of air and soil temperatures, respectively (C)
-V=0.5 #wind speed (m/s)
+V=1 #wind speed (m/s)
 #assumptions
@@
 
-We can estimate hourly solar radiation as follows:
+Although measured solar radiation is preferable if available, we can estimate hourly solar radiation by discounting incoming solar radiation as it moves through the atmosphere as follows:
 ```{r}
@@
 #Separate into direct, diffuse, and reflected solar radiation
-Sdir= Srad[1,]
-Sdif= Srad[2,]
-Sref= Srad[3,]
+Sdir= Srad[1,]; Sdif= Srad[2,]; Sref= Srad[3,]
 ```
 
-We can then calculate hourly air and soil temperatures. We use the sine-exponential model for air temperature and the sine model for surface temperature:
+We can then calculate hourly air and soil temperatures based on daily minimum and maximum temperatures.
+We select the sine-exponential model for air temperature and the sine model for surface temperature:
 ```{r}
-Ta= sapply(1:24, diurnal_temp_variation_sineexp, T_max=Tmax, T_min=Tmin, t_r=tr, t_s=ts, alpha=2.59, beta= 1.55, gamma=2.2)
-Ts= sapply(1:24, diurnal_temp_variation_sine, T_max=Tmax_s, T_min=Tmin_s)
+Ta= sapply(1:24, diurnal_temp_variation_sineexp, T_max=Tmax, T_min=Tmin, t_r=tr, t_s=ts, alpha=2.59, beta= 1.55, gamma=2.2) #air temperature (C)
+Ts= sapply(1:24, diurnal_temp_variation_sine, T_max=Tmax_s, T_min=Tmin_s) #soil temperature (C)
+```
+
+We use microclimate models to scale air temperature and wind speed from weather station height (2m) to lizard height (0.02m).
+We assume a surface roughness of $z_0=0.2$, which corresponds to bare sand and determines the turbulence of airflow.
+We implement neutral air temperature and wind speed profiles as well as an air temperature profile forced by wind speed.
+
+```{r}
+#scale air temperature to lizard height
+#z is reference height, z is organism height, z0 is surface roughness
+Ta_liz= air_temp_profile_neutral(T_r=Ta, zr=2, z0=0.2, z=0.02, T_s=Ts)
+Ta_liz2= air_temp_profile(T_r=Ta, u_r=V, zr=2, z0=0.2, z=0.02, T_s=Ts)
+
+#scale wind speed to lizard height
+V_liz= wind_speed_profile_neutral(u_r=V, zr=2, z0=0.2,z=0.02)
 ```
@@
 We will be solving the following energy balance to estimate $T_e$:
-$$Q_{net} = Q_{abs} - Q_{emit} + Q_{conv} + Q_{cond} + Q_{met} - Q_{evap},$$
+$$Q_{net} = Q_{abs} - Q_{emit} - Q_{conv} - Q_{cond} - Q_{met} - Q_{evap},$$
+where $Q_{net}$ is the net energy exchange with the environment (W), $Q_{abs}$ is the solar radiation absorbed (W), $Q_{emit}$ is the net thermal radiation emitted (W), $Q_{conv}$ is energy exchange due to convection (W), $Q_{cond}$ is energy exchange due to conduction (W), $Q_{met}$ is the energy emitted due to metabolism (W), and $Q_{evap}$ is the energy emitted due to evaporative water loss (W).
 We will estimate each term on the right side of the equation in turn.
-Estimating $Q_{abs}$requires the surface area exposed to radiation and the solar absorptivity of the animal surface ($a$ proportion).
+Estimating $Q_{abs}$ requires the surface area exposed to radiation and the solar absorptivity of the animal surface ($a$ proportion).
 We model a 10g Sceloporus lizard with solar absorptivity $a=0.9$ [@gates1980biophysical].
-We will initially assume $T_b=T_a$ to illustrate calculations.
+We will initially assume $T_b=T_a$ to illustrate the calculations before solving for $T_b$ givne the environmental conditions.
 
 ```{r}
-mass=10
-a=0.9 #solar absorptivity
+mass=10 #mass (g)
+a=0.9 #solar absorptivity (%)
 
@@
 
-We calculate the hourly radiation absorbed as follows:
+We calculate the hourly solar radiation absorbed (W) as follows:
 ```{r}
@@
 
-We estimate thermal radiation $Q_{emit}$ for the lizard outdoors. We assume the surface emissivity of lizards, $epsilon_s=0.965$ [Bartlett & Gates 1967].
+We estimate thermal radiation $Q_{emit}$ (W) for the lizard outdoors. We assume the surface emissivity of lizards, $epsilon_s=0.965$ [@barlett1967energy].
 
@@
 Qemit= rep(NA, 24)
-for(hr in 1:24) Qemit[hr]=Qemitted_thermal_radiation(epsilon=epsilon_s, A=A, psa_dir=psa[hr], psa_ref=1-psa[hr], T_b=Ta[hr]+273.15, T_g=Ts[hr]+273.15, T_a=Ta[hr]+273.15, enclosed=FALSE)
+for(hr in 1:24) Qemit[hr]=Qemitted_thermal_radiation(epsilon=epsilon_s, A=A, psa_dir=psa[hr], psa_ref=1-psa[hr], T_b=Ta_liz[hr]+273.15, T_g=Ts[hr]+273.15, T_a=Ta_liz[hr]+273.15, enclosed=FALSE)
 ```
 
-We next estimate convection $Q_{conv}$ and conduction $Q_{cond}$.
-We will estimate the lizard's heat transfer coefficient, $H_L$ using an empirical relationship for lizards (`heat_transfer_coefficient()`), but also illustrate a function estimating $H_L$ using a spherical approximation (`heat_transfer_coefficient_approximation()`) and a simplified approximation (`heat_transfer_coefficient_simple()`).
+We next estimate convection $Q_{conv}$ (W) and conduction $Q_{cond}$ (W).
+We will estimate the lizard's heat transfer coefficient, $H_L$ ($Wm^{-2}K^{-1}$) using an empirical relationship for lizards (`heat_transfer_coefficient()`), but also illustrate a function estimating $H_L$ using a spherical approximation (`heat_transfer_coefficient_approximation()`) and a simplified approximation (`heat_transfer_coefficient_simple()`).
 
 ```{r}
-#Use DRYAIR in NicheMapR to estimate the thermal conductivity of air and kinematic viscosity
-#library(NicheMapR)
-
-#estimate Barometric pressure (pascal)
-ap= airpressure_from_elev(elev)*1000
+#Use DRYAIR from NicheMapR to estimate the thermal conductivity of air and kinematic viscosity
+ap= airpressure_from_elev(elev)*1000 #estimate Barometric pressure (pascal)
 
 DRYAIRout= DRYAIR(db=Ta, bp=ap, alt=elev)
-K= DRYAIRout$thcond
-nu= DRYAIRout$viskin #Kinematic viscosity (m2 s-1)
+K= DRYAIRout$thcond #thermal conductivity (Wm^-2K^-1) 
+nu= DRYAIRout$viskin #kinematic viscosity (m2 s-1)
 
-svl=0.006 #SVL for sceloporus
+svl=0.006 #approximate snout vent length (m) for Sceloporus
 
-#We will use the average of K and nu across the day for simplicity and since there's not a substantial variation
+#We will use the average of K and nu across the day for simplicity and since there is not a substantial diurnal variation.
 K= mean(K)
@@
 
-#Estimate using an empirical relationship
-H_L=heat_transfer_coefficient(V=V,D=svl,K= 25.7 * 10^(-3), nu= 15.3 * 10^(-6) , taxa="lizard_surface")
+#Estimate the heat transfer coefficient using an empirical relationship for lizards
+H_L=heat_transfer_coefficient(V=V_liz,D=svl,K= 25.7 * 10^(-3), nu= 15.3 * 10^(-6) , taxa="lizard_surface")
 
 #Also illustrate estimations using a spherical approximation and a simplified version of the approximation.
-heat_transfer_coefficient_approximation(V=V,D=svl,K= 25.7 * 10^(-3), nu= 15.3 * 10^(-6), taxa="lizard")
-heat_transfer_coefficient_simple(V=0.5,D=0.05)
-
+H_L2= heat_transfer_coefficient_approximation(V=V_liz,D=svl,K= 25.7 * 10^(-3), nu= 15.3 * 10^(-6), taxa="lizard")
+H_L3= heat_transfer_coefficient_simple(V=V_liz,D=svl)
 ```
 
-We estimate convective heat exchange between the animal and surrounding air using ithe following relationship:
+We estimate convective heat exchange between the animal and surrounding air using the following relationship:
 $$Q_{conv}= ef \cdot H_L(A\cdot \mbox{proportion})(T_a-T_b),$$
-where an enhancement factor, $ef$, multiplier can be incorporated to account for increases in heat exchange resulting from air turbulance in field conditions
-
-The function is available in R assuming that 2/3 of the lizard's surface area is exchanging heat through convection.
+where an enhancement factor, $ef$, multiplier can be incorporated to account for increases in heat exchange resulting from air turbulence in field conditions.
+We implement the function in R assuming that 2/3 of the lizard's surface area is exchanging heat through convection.
 We note we currently estimate no convection since we are assuming $Tb=Ta$ for illustration purposes:
@@
 Qconv= rep(NA, 24)
-for(hr in 1:24) Qconv[hr]= Qconvection(T_a= Ta[hr]+273.15,T_b= Ta[hr]+273.15,H=H_L,A=A, proportion=0.67, ef=1.3)
+for(hr in 1:24) Qconv[hr]= Qconvection(T_a= Ta_liz[hr]+273.15, T_b= Ta_liz[hr]+273.15, H=H_L, A=A, proportion=0.67, ef=1.3)
 ```
 
-We estimate conductance between the lizard and surface assuming conductance through the animal tissue is the rate limiting step as follows:
-$$Q_{cond}= A \cdot \mbox{proportion} \cdot K(T_g-T_b)/d, $$
-We conduct the estimate in R assuming that 1/3 of the lizard surface is in contact with the ground and a skin thinkness of $10^{-6}$:
+We estimate conductive heat flow (W) from the lizard to the surface assuming conductance through the animal tissue is the rate limiting step as follows:
+$$Q_{cond}= A \cdot \mbox{proportion} \cdot K(T_g-T_b)/d. $$
+We implement the estimate in R assuming that 1/3 of the lizard surface is in contact with the ground and a skin thickness $d$ of $10^{-6}$:
 ```{r}
 Qcond= rep(NA, 24)
-for(hr in 1:24) Qcond[hr]=Qconduction_animal(T_g= Ts[hr]+273.15,T_b=Ta[hr]+273.15,d=10^-6,K=K,A=A, proportion=0.33)
+for(hr in 1:24) Qcond[hr]=Qconduction_animal(T_g= Ts[hr]+273.15, T_b=Ta_liz[hr]+273.15, d=10^-6, K=K, A=A, proportion=0.33)
 ```
@@
 ```{r}
-Qmet= 0
-Qevap= 0
+Qmet= 0; Qevap= 0
 ```
 
-The full heat budget can be calculated as follows: 
+The full heat budget can be calculated as follows [@gates1980biophysical]: 
 ```{r}
-Qnet_Gates(Qabs=Qabs, Qemit=Qemit, Qconv=Qconv, Qcond=Qcond, Qmet=Qmet, Qevap=Qevap)
+Qnet= Qnet_Gates(Qabs=Qabs, Qemit=Qemit, Qconv=Qconv, Qcond=Qcond, Qmet=Qmet, Qevap=Qevap)
 ```
-We constructed the energy balance assuming $T_b=T_a$ for illustration. We now use a function based on the energy balance above to estimate body temperature given the environmental conditions:
+We constructed the energy balance assuming $T_b=T_a$ for illustration.
+Thus, the energy balance is dominated by heat gain from contact with the warmer surface.
+We now use a function based on the energy balance above to estimate body temperature given the environmental conditions:
 
@@
 Te= rep(NA, 24)
-for(hr in 1:24) Te[hr]= Tb_Gates(A=A, D=svl, psa_dir=psa[hr], psa_ref=1-psa[hr], psa_air=0.67, psa_g=0.33, T_g=Ts[hr]+273.15, T_a=Ta[hr]+273.15, Qabs=Qabs[hr], epsilon=epsilon_s, H_L=H_L, ef=1.3, K=K)
+for(hr in 1:24) Te[hr]= Tb_Gates(A=A, D=svl, psa_dir=psa[hr], psa_ref=1-psa[hr], psa_air=0.67, psa_g=0.33, T_g=Ts[hr]+273.15, T_a=Ta_liz[hr]+273.15, Qabs=Qabs[hr], epsilon=epsilon_s, H_L=H_L, ef=1.3, K=K)
 ```
 
-We also implement a similar but simplified energy balance. The energy balance omits conduction with the ground and thus underestimates temperatures, but we implement it for comparison purposes:
+We also implement a similar but simplified energy balance [@campbell2000introduction]. The energy balance omits conduction with the ground:
 ```{r}
 Te2= rep(NA, 24)
-for(hr in 1:24) Te2[hr]= Tb_CampbellNorman(T_a=Ta[hr]+273.15, S=Qabs[hr]/svl, epsilon=epsilon_s, c_p=29.3, D=svl, V=V)
-#CHECK
+
+for(hr in 1:24) Te2[hr]= Tb_CampbellNorman(T_a=Ta_liz[hr]+273.15, T_g=Ts[hr]+273.15, S=Qabs[hr]/A, alpha_L=0.96, epsilon=epsilon_s, c_p=29.3, D=svl, V=V_liz)
+#S is solar radiation flux (W m^-2), so we divide by surface area, A
 ```
 
-We additionally estimate $T_e$ using the specialized function for lizards:
+We additionally estimate $T_e$ using a specialized function for lizards [@buckley2008linking]:
 ```{r}
 Te3= rep(NA, 24)
-for(hr in 1:24) Te3[hr]= Tb_lizard(T_a=Ta[hr], T_g=Ts[hr], u=V, svl=svl*1000, m=mass, psi=psi_deg[hr], rho_S=rho, elev=elev, doy=doy, sun=TRUE, surface=TRUE, alpha_S=a, alpha_L=0.965, epsilon_s=epsilon_s, F_d=0.8, F_r=0.5, F_a=0.5, F_g=0.5)
+for(hr in 1:24) Te3[hr]= Tb_lizard(T_a=Ta_liz[hr], T_g=Ts[hr], u=V_liz, svl=svl*1000, m=mass, psi=psi_deg[hr], rho_S=rho, elev=elev, doy=doy, sun=TRUE, surface=TRUE, alpha_S=a, alpha_L=0.965, epsilon_s=epsilon_s, F_d=0.8, F_r=0.5, F_a=0.5, F_g=0.5)
+```
 
-Te4= rep(NA, 24)
-for(hr in 1:24) Te4[hr]= Tb_Fei(T_a=Ta[hr]+273.15, T_g=Ts[hr]+273.15, H=Qabs[hr], lw=Qabs[hr], shade=0, m=mass, Acondfact =0.1, Agradfact =0.3)
-#CHECK
+### Plot
+We next plot the diurnal trends in environmental data and predicted body temperatures.
+
+```{r}
+par(mar=c(5, 4, 4, 8) + 0.1)
+plot(1:24, (Te-273.15), type="l", xlab="Hour", ylab="Temperature (°C)", col="#5DC863FF", ylim=c(10,60)) #Gates
+points(1:24, Te2-273.15, type="l", col="#5DC863FF", lty="dotted") #Campbell and Norman
+points(1:24, Te3, type="l", col="#5DC863FF", lty="dashed") #Lizard, Buckley 2008
+points(1:24, Ta, type="l", col="#440154FF")
+points(1:24, Ta_liz, type="l", col="#3B528BFF")
+points(1:24, Ts, type="l", col="#21908CFF")
+
+#add additional axis with radiation
+par(new = T)
+plot(1:24, Sdir, pch=16, axes=F, xlab=NA, ylab=NA, type="l", col="#FDE725FF")
+axis(side = 4)
+mtext(side = 4, line = 3, 'Radiation (W/m^2)')
+legend("topleft",
+       legend=c("Ta 2m", "Ta 0.02m", "Ts", "Te", "Radiation"),
+       lty=1, pch=NA, col=c("#440154FF", "#3B528BFF", "#21908CFF", "#5DC863FF", "#FDE725FF"))
 
@@
 
-### Plot
-We next plot the diurnal trends in environmental data and the operative data.
+*Figure 1*.
+Predicted lizard body temperatures (Te) as a function of air temperatures (Ta) at 2m and lizard height (0.02m), surface (Ts) temperatures (°C), and solar radiation ($W/m^2$).
+We depict body temperatures using two general energy budgets [solid: Tb_Gates(); dotted: Tb_CampbellNorman()] and a lizard specific biophysical model [dashed: Tb_lizard()].
 
-```{r}
-par(mar=c(5, 4, 4, 8) + 0.1)
-plot(1:24, (Te-273.15), type="l", xlab="Hour", ylab="Temperature (°C)", col="blue") #Gates
-points(1:24, Te2-273.15, type="l", col="blue", lty="dotted") #Campbell
-points(1:24, Te3, type="l", col="blue", lty="dashed") #Buckley 2008
-#points(1:24, Te4, type="l", col="red", lty="solid") #Fei et al.
-points(1:24, Ta, type="l", col="orange")
-points(1:24, Ts, type="l", col="purple")
-
-#add additional axis with radiation
-par(new = T)
-plot(1:24, Sdir, pch=16, axes=F, xlab=NA, ylab=NA, type="l", col="green")
-axis(side = 4)
-mtext(side = 4, line = 3, 'Radiation')
-legend("topleft",
-       legend=c("Ta", "Ts", "Te", "Radiation"),
-       lty=1, pch=NA, col=c("orange", "purple", "blue", "green"))
-
-```
+The microclimate models indicate that air temperatures at lizard height are similar to surface temperatures (Figure 1).
+The biophysical models indicate that solar radiation will elevate lizard body temperatures far above air temperatures and that the lizard will face thermal stress if it is unable to seek shade (Figure 1).
+The three biophysical models predict different body temperatures during peak period of solar radiation because they model interactions with the ground differently and users are encouraged to review the details of each biophysical model before selection.
 
 ## Summary
-New environmental data and tools make biophysical modeling more accessible. Expand use since we realize.
+TrenchR is intended to promote understanding of how organisms interact with their environment and the consequences for physiology, energetics, behavior, and demography.
+Our example implementation highlights the importance of considering organismal body temperatures, rather than air temperatures, when examining thermal stress and other responses to environmental variability and change.
+TrenchR currently focuses on heat balances but may be expanded to include water balances.
+Simple functions can be combined as needed to produce comprehensive and transparent models for biophysical ecology and evolution.
+The resultant models are likely to be sufficiently detailed for many applications, but users are referred to NicheMapR for more detailed biophysical models [@kearney2017nichemapr; @kearney2020nichemapr].
+Making classic biophysical ecology techniques more accessible will allow researchers to take advantage of rapidly accumulating data on environmental conditions and organismal traits to understand and predict ecological and evolutionary responses to variable and changing environments.
 

### 2021-04-27 ManuscriptTe.Rmd
Update format and references

@@
 output:
+  pdf_document: default
+  html_document: default
   word_document: default
-  html_document: default
-  pdf_document: default
   rmarkdown::html_vignette: default

### 2021-04-27 Manuscript_full.Rmd
Update format and references

@@
 title: 'TrenchR: an R package for transparent environmental and ecological biophysics'
-author: "Lauren B. Buckley, Bryan Briones Ortiz, Aji John, Ofir Levy, Yutaro Sakairi,..."
-date: "February 4, 2020"
+author: "Lauren B. Buckley, Bryan Briones Ortiz, Isaac Caruso, Aji John, Ofir Levy, Abby Meyer, Yutaro Sakairi,..."
 output:
+  html_document: default
   word_document: default
-  html_document: default
   pdf_document: default
 bibliography: bibliography.bib
+csl: methods-in-ecology-and-evolution.csl
+editor_options: 
+  markdown: 
+    wrap: 72
 ---
@@
 
+-   Within the word count (6000-7000 words for Standard Articles, 3000
+    words for Applications and Practical Tools)
+
+-   Continuous line and page numbering throughout
+
+Lauren B. Buckley^1^, Bryan Briones Ortiz^1^, Isaac Caruso^1^, Aji
+John^1^, Ofir Levy^2^, Abby Meyer^1^, Yutaro Sakairi^1^,...
+
+1.  Department of Biology, University of Washington, Seattle, WA
+    98195-1800, USA
+2.
+
+Corresponding Author: Lauren Buckley,
+[lbuckley\@uw.edu](mailto:lbuckley@uw.edu){.email}
+
+Running head: TrenchR microclimate and biophysical models
+
+Type: applications
+
 ## Abstract
 
-Much understanding of climate change responses relies on the assumption that organismal body temperatures are equal to temporally averaged air temperatures high above the ground.
-However, most organisms live near the ground and acute exposure to solar radiation can substantially elevate their body temperatures.
-We introduce the TrenchR package, which aids in Translating Environmental Change into organismal responses (https://github.com/trenchproject/TrenchR).
-The package includes microclimate models to vertically scale weather station data to organismal heights.
-The biophysical modelling tools include both general models for heat flows and specific models to predict body temperatures for a variety of ectothermic taxa.
-Additional functions model and temporally partition air and soil temperatures and solar radiation.
-The package aims to introduce and enable microclimate and biophysical modelling to improve ecological and evolutionary forecasting and includes tutorials and well as a series of educational modules.
+1.  Much understanding of climate change responses relies on the
+    assumption that organismal body temperatures are equal to temporally
+    averaged air temperatures high above the ground. However, most
+    organisms live near the ground and acute exposure to solar radiation
+    can substantially elevate their body temperatures.
+2.  We introduce the TrenchR package, which aids in Translating
+    Environmental Change into organismal responses
+    (<https://github.com/trenchproject/TrenchR>). The package includes
+    microclimate models to vertically scale weather station data to
+    organismal heights. The biophysical modelling tools include both
+    general models for heat flows and specific models to predict body
+    temperatures for a variety of ectothermic taxa. Additional functions
+    model and temporally partition air and soil temperatures and solar
+    radiation.
+3.  [Main results]
+4.  The package aims to introduce and enable microclimate and
+    biophysical modelling to improve ecological and evolutionary
+    forecasting and includes tutorials and well as a series of
+    educational modules.
 
-## Background
-Many responses of organisms and ecosystems to climate change are inconsistent with current predictive models [@maguire2015modeling].
-Some of these predictive shortcomings stem from omitting spatial and temporal environmental variation and how it interacts with organismal phenotypes (e.g., size, coloration) [@nadeau2017coarse; @potter2013microclimatic].
-Many analyses assume that organismal body temperatures are equal to air temperate at weather station height (usually 2m).
-However, air temperatures near the ground where most organisms reside can be considerably warmer and absorption of solar radiation can raise body temperatures well above air temperatures, resulting in potential discrepancies of tens of degrees C [@kearney2009potential; @sinclair2016can; @sunday2014thermal].
-Acute thermal stress events are often more relevant to the physiology and energy balance of organisms than daily, monthly, or annually aggregated environmental conditions.
-Yet, daily and seasonal environmental variation is often neglected [@dillon2016introduction; @dillon2016life].
-We aim to help remedy these shortcomings by introducing the TrenchR package as a component of the TrEnCh project (https://trenchproject.github.io), which builds computational and visualization tools to Translate Environmental Change into organismal responses.
+**Keywords**: biophysical, climate change, ecotherm, energy budget,
+mechanistic models, microclimate, scaling, temperature
 
-The mathematical tools for leveraging environmental data to predict organismal conditions have long lingered in books [@campbell2000introduction; @gates1980biophysical] and articles [@porter1969thermodynamic; @porter1983biophysical], but adoption of the tools has not kept pace with research on climate change responses.
-Adoptions of biophysical approaches are expanding with recognition of the importance of spatial and temporal environmental variation to determining climate change responses [@buckley2018leveraging; @kearney2017nichemapr].
-Increasing dissemination of computational tools is facilitating adoption.
-The TrenchR package is intended to complement the NicheMapR package, which includes sophisticated microclimate [@kearney2017nichemapr] and biophysical [@kearney2020nichemapr] models.
-Many of the approaches are similar between the packages, but TrenchR focuses on simple functions to aid the understanding and transparency of biophysical approaches.
-The TrenchR functions allow for modular use and easy adaptation to a variety of organismal and research questions.
-Some energy budget calculation components of biophysical models are also available in the ThermImage package [@tattersall2019thermimage].
-Although many of the TrenchR functions are general, some components of TrenchR are best suited for ectothermic animals.
-We have minimized inclusion of plant specific biophysical models since they are the focus of several R packages [tealeaves: [@muir2019tealeaves]; plantecophys [@duursma2015plantecophys]].
+## Introduction
+
+Many responses of organisms and ecosystems to climate change are
+inconsistent with current predictive models [@maguire2015modeling]. Some
+of these predictive shortcomings stem from omitting spatial and temporal
+environmental variation and how it interacts with organismal phenotypes
+(e.g., size, coloration) [@nadeau2017coarse; @potter2013microclimatic].
+Many analyses assume that organismal body temperatures are equal to air
+temperate at weather station height (usually 2m). However, air
+temperatures near the ground where most organisms reside can be
+considerably warmer and absorption of solar radiation can raise body
+temperatures well above air temperatures, resulting in potential
+discrepancies of tens of degrees C [@sinclair2016; @sunday2014;
+@kearney2009]. Acute thermal stress events are often more relevant to
+the physiology and energy balance of organisms than daily, monthly, or
+annually aggregated environmental conditions. Yet, daily and seasonal
+environmental variation is often neglected [@dillon2016introduction;
+@dillon2016life]. We aim to help remedy these shortcomings by
+introducing the TrenchR package as a component of the TrEnCh project
+(<https://trenchproject.github.io>), which builds computational and
+visualization tools to Translate Environmental Change into organismal
+responses.
+
+The mathematical tools for leveraging environmental data to predict
+organismal conditions have long lingered in books
+[@campbell2000introduction; @gates1980biophysical] and articles
+[@porter1969thermodynamic; @porter1983biophysical], but adoption of the
+tools has not kept pace with research on climate change responses.
+Adoptions of biophysical approaches are expanding with recognition of
+the importance of spatial and temporal environmental variation to
+determining climate change responses [@buckley2018leveraging;
+@kearney2017nichemapr]. Increasing dissemination of computational tools
+is facilitating adoption. The TrenchR package is intended to complement
+the NicheMapR package, which includes sophisticated microclimate
+[@kearney2017nichemapr] and biophysical [@kearney2020] models. Many of
+the approaches are similar between the packages, but TrenchR focuses on
+simple functions to aid the understanding and transparency of
+biophysical approaches. The TrenchR functions allow for modular use and
+easy adaptation to a variety of organismal and research questions. Some
+energy budget calculation components of biophysical models are also
+available in the ThermImage package [@tattersall2017]. Although many of
+the TrenchR functions are general, some components of TrenchR are best
+suited for ectothermic animals. We have minimized inclusion of plant
+specific biophysical models since they are the focus of several R
+packages [tealeaves: [@muir2019tea]; plantecophys:
+[@duursma2015plantecophys]].
 
 ## Methods and features
-The TrenchR package aims to promote transparency and reproducibility and we welcome contributions and corrections from users.
-Our package is built using the devtools methodology (https://github.com/hadley/devtools) in Github.
-Issues and feature requests can be contributed in Github (https://github.com/trenchproject/TrenchR/).
-The package is maintained and checked using Travis CI as a tool for continuous integration.
 
-Functions were adapted from biophysical ecology textbooks [@campbell2000introduction; @gates1980biophysical] and research articles.
-The source reference is included in each function’s header.
-We have incorporated default parameters (e.g., organism emissivity or heat conduction rates) when general values are available.
-Many functions additionally include comments that describe potential parameterizations.
-The biophysical ecology texts [@campbell2000introduction; @gates1980biophysical] include many tables with parameter values and we provide several tables in TrenchR [Table 8.2XXXX].
+The TrenchR package aims to promote transparency and reproducibility and
+we welcome contributions and corrections from users. Our package is
+built using the devtools methodology
+(<https://github.com/hadley/devtools>) in Github. Issues and feature
+requests can be contributed in Github
+(<https://github.com/trenchproject/TrenchR/>). The package is maintained
+and checked using Travis CI as a tool for continuous integration.
+
+Functions were adapted from biophysical ecology textbooks
+[@campbell2000introduction; @gates1980biophysical] and research
+articles. The source reference is included in each function's header. We
+have incorporated default parameters (e.g., organism emissivity or heat
+conduction rates) when general values are available. Many functions
+additionally include comments that describe potential parameterizations.
+The biophysical ecology texts [@campbell2000introduction;
+@gates1980biophysical] include many tables with parameter values and we
+provide several tables in TrenchR [Table 8.2XXXX].
 
 ### Components
+
 TrenchR functions are organized into the following categories:
-**Utility functions** allow estimating environmental metrics that form the basis of microclimate and biophysical models (e.g., zenith and declination angles), facilitate modeling diurnal temperature variation, and estimate biologically relevant aggregate metrics such as degree days available for development.
 
-**Microclimate functions** allow estimating the environmental conditions experienced by organisms.
-Temperature and wind profile functions scale environmental conditions from sensors to organismal height.
-Radiation functions allow estimating incoming solar radiation and partitioning the variation diurnally and across components (i.e., direct, diffuse, reflected).
+**Utility functions** allow estimating environmental metrics that form
+the basis of microclimate and biophysical models (e.g., zenith and
+declination angles), facilitate modeling diurnal temperature variation,
+and estimate biologically relevant aggregate metrics such as degree days
+available for development.
 
-**General biophysical functions** provide generalized models of heat exchanges between organisms and their environment so that users can build custom biophysical models.
-The functions allow implementing an energy balance of the following components of heat exchange between organisms and their environment: *radiative heat exchanges* of solar and thermal radiation; *convective heat exchanges* between organisms and their surrounding fluid (air or water) driven by fluid flow; *conductive heat exchanges* between organisms and the ground due to physical contact; and *evaporative heat exchanges* associated with organisms’ evaporative water loss.
-Additional functions aggregate these forms of heat exchange into energy balances and to use the energy balances to predict body temperatures.
+**Microclimate functions** allow estimating the environmental conditions
+experienced by organisms. Temperature and wind profile functions scale
+environmental conditions from sensors to organismal height. Radiation
+functions allow estimating incoming solar radiation and partitioning the
+variation diurnally and across components (i.e., direct, diffuse,
+reflected).
 
-**Specific biophysical functions** provide biophysical models that have been built for specific organisms based on their physical properties, behavior, environment, and life history.
-We currently provide published biophysical models for lizards, salamanders, butterflies, grasshoppers, limpets, mussels, and snails.
-Most models predict operative environmental temperatures, which are the equilibrium body temperatures of organisms with specified physical properties in a specific microenvironment and assume no heat exchange via metabolism or evaporation [@bakken1992measurement].
+**General biophysical functions** provide generalized models of heat
+exchanges between organisms and their environment so that users can
+build custom biophysical models. The functions allow implementing an
+energy balance of the following components of heat exchange between
+organisms and their environment: *radiative heat exchanges* of solar and
+thermal radiation; *convective heat exchanges* between organisms and
+their surrounding fluid (air or water) driven by fluid flow; *conductive
+heat exchanges* between organisms and the ground due to physical
+contact; and *evaporative heat exchanges* associated with organisms'
+evaporative water loss. Additional functions aggregate these forms of
+heat exchange into energy balances and to use the energy balances to
+predict body temperatures.
+
+**Specific biophysical functions** provide biophysical models that have
+been built for specific organisms based on their physical properties,
+behavior, environment, and life history. We currently provide published
+biophysical models for lizards, salamanders, butterflies, grasshoppers,
+limpets, mussels, and snails. Most models predict operative
+environmental temperatures, which are the equilibrium body temperatures
+of organisms with specified physical properties in a specific
+microenvironment and assume no heat exchange via metabolism or
+evaporation [@bakken1992measurement].
 
 ### Vignettes and introductory modules
-We introduce the function in several vignettes.
-A good place to start is the *Allometry and conversions* tutorial, which provides tools for preparing data such as estimating additional dimensions of organisms from measured dimensions.
-The *Estimating microclimates* tutorial provides resources for estimating the environmental conditions experienced by organisms.
-This includes estimating solar radiation and its components, diurnal variation in temperature and radiation, temperature and windspeed profiles, and soil temperatures and profiles.
-Finally, the core biophysical modelling functions are described in a tutorial on *Using energy balances to estimate body temperatures*.
-Components of an energy budget can be estimated using individual functions and then body temperatures can be solved for using either a generic energy balance or taxa specific biophysical models.
 
-For additional background on microclimate and biophysical modelling, we have updated a series of educational modules entitled “Physical processes in ecosystems”.
-We intend the modules to provide a more contained and accessible introduction to microclimate and biophysical modelling than that included in classic textbooks [@campbell2000introduction; @gates1980biophysical].
-The modules provide less detailed coverage of many of the topics contained in the books.
-The series of 14 modules starts with an overview of the calculus and physics principles underlying the modelling.
-Thermodynamics and energy budget modelling are then introduced with detailed examples.
-Modules address the climate space concept, operative temperatures, biophysical models for leaves and sheep, and heat flow in soils.
-We have revised the modules (which originated from an NSF training grant in 1979 and lacked broad distribution) to include R code and utilization of TrenchR functions.
-We expanded the original series of tutorials to include introductions by Michael Kearney to the Microclim environmental data [@kearney2014microclim] and integrating the Microclim data with biophysical modelling to examine broad scale climatic limits.
-The modules are provided in html form and available for download as a pdf at the bookdown server (https://bookdown.org/huckley/Physical_Processes_In_Ecosystems/).
-The R markdown files are available in Github at https://github.com/trenchproject/TrenchRmodules.
+We introduce the function in several vignettes. A good place to start is
+the *Allometry and conversions* tutorial, which provides tools for
+preparing data such as estimating additional dimensions of organisms
+from measured dimensions. The *Estimating microclimates* tutorial
+provides resources for estimating the environmental conditions
+experienced by organisms. This includes estimating solar radiation and
+its components, diurnal variation in temperature and radiation,
+temperature and windspeed profiles, and soil temperatures and profiles.
+Finally, the core biophysical modelling functions are described in a
+tutorial on *Using energy balances to estimate body temperatures*.
+Components of an energy budget can be estimated using individual
+functions and then body temperatures can be solved for using either a
+generic energy balance or taxa specific biophysical models.
+
+For additional background on microclimate and biophysical modelling, we
+have updated a series of educational modules entitled "Physical
+processes in ecosystems". We intend the modules to provide a more
+contained and accessible introduction to microclimate and biophysical
+modelling than that included in classic textbooks
+[@campbell2000introduction; @gates1980biophysical]. The modules provide
+less detailed coverage of many of the topics contained in the books. The
+series of 14 modules starts with an overview of the calculus and physics
+principles underlying the modelling. Thermodynamics and energy budget
+modelling are then introduced with detailed examples. Modules address
+the climate space concept, operative temperatures, biophysical models
+for leaves and sheep, and heat flow in soils. We have revised the
+modules (which originated from an NSF training grant in 1979 and lacked
+broad distribution) to include R code and utilization of TrenchR
+functions. We expanded the original series of tutorials to include
+introductions by Michael Kearney to the Microclim environmental data
+[@kearney2014microclim] and integrating the Microclim data with
+biophysical modelling to examine broad scale climatic limits. The
+modules are provided in html form and available for download as a pdf at
+the bookdown server
+(<https://bookdown.org/huckley/Physical_Processes_In_Ecosystems/>). The
+R markdown files are available in Github at
+<https://github.com/trenchproject/TrenchRmodules>.
 
 ## Examples
-We illustrate the use of the TrenchR package by estimating an energy budget for a *Sceloporus* lizard on June 1, 2018, in Santa Fe, New Mexico, USA (35.69, -105.944, elevation: 2121m).
-The example is also incorporated in the *Using energy balances to estimate body temperatures* tutorial.
-We first estimate the energy budget with component functions and we then use an integrated biophysical model to estimate operative environmental temperatures.
-We start by generating environmental inputs.
-See Appendix S1 for another example examining a time series of microcliamte data and applying and testing the biophysical models.
 
-We will use the energy budget to estimate the lizard's body temperature, $T_b$, which can depart dramatically from the air temperatures due to energy exchange with the environment.
-Heat is gained from absorbing solar and thermal radiation and from metabolic reactions.
-Heat is lost through the organism's emission of radiation and the evaporation of water.
-The organism exchanges heat with the surrounding air or water via convection and with the ground via conduction.
-The balance of these heat exchanges can be estimated and often referred to as operative environmental temperature, $T_e$ [@bakken1992measurement].
+We illustrate the use of the TrenchR package by estimating an energy
+budget for a *Sceloporus* lizard on June 1, 2018, in Santa Fe, New
+Mexico, USA (35.69, -105.944, elevation: 2121m). The example is also
+incorporated in the *Using energy balances to estimate body
+temperatures* tutorial. We first estimate the energy budget with
+component functions and we then use an integrated biophysical model to
+estimate operative environmental temperatures. We start by generating
+environmental inputs. See Appendix S1 for another example examining a
+time series of microclimate data and applying and testing the
+biophysical models.
 
-Let's assume the lizard is in an unshaded location where a weather station at standard height (2m) indicates that the daily air temperature varies from a minimum of 10°C to a maximum of 25°C and the wind speed averages 1 m/s.
-The soil surface temperature varies from a minimum of 15°C to a maximum of 30°C.
-We assume that atmospheric transmissivity $\tau=0.7$ and albedo $\rho=0.6$
+We will use the energy budget to estimate the lizard's body temperature,
+$T_b$, which can depart dramatically from the air temperatures due to
+energy exchange with the environment. Heat is gained from absorbing
+solar and thermal radiation and from metabolic reactions. Heat is lost
+through the organism's emission of radiation and the evaporation of
+water. The organism exchanges heat with the surrounding air or water via
+convection and with the ground via conduction. The balance of these heat
+exchanges can be estimated and often referred to as operative
+environmental temperature, $T_e$ [@bakken1992measurement].
+
+Let's assume the lizard is in an unshaded location where a weather
+station at standard height (2m) indicates that the daily air temperature
+varies from a minimum of 10°C to a maximum of 25°C and the wind speed
+averages 1 m/s. The soil surface temperature varies from a minimum of
+15°C to a maximum of 30°C. We assume that atmospheric transmissivity
+$\tau=0.7$ and albedo $\rho=0.6$
 
@@
 
-Let us first prepare the environmental data for analysis.
-We will need to estimate hourly air and soil temperatures and radiation.
-We start by estimating the day of year and the timing of sunrise and sunset:
-```{r}
+Let us first prepare the environmental data for analysis. We will need
+to estimate hourly air and soil temperatures and radiation. We start by
+estimating the day of year and the timing of sunrise and sunset:
+
+```{r, echo=FALSE}
 library(TrenchR)
 
+#to be REMOVED
 Tb_CampbellNorman=function(T_a, T_g, S, alpha_L=0.96, epsilon=0.96, c_p=29.3, D, V){
@@
 #set up input data as variables
-lat=35.69; lon= -105.944; elev=2121 #latitude and longitude (degrees) and elevation (m)
-Tmin=10; Tmax=25; Tmin_s=15; Tmax_s=30 #minimum and maximum of air and soil temperatures, respectively (C)
-V=1 #wind speed (m/s)
+#latitude and longitude (degrees) and elevation (m)
+lat=35.69; lon= -105.944; elev=2121 
+#minimum and maximum of air and soil temperatures, respectively (C)
+Tmin=10; Tmax=25; Tmin_s=15; Tmax_s=30 
+#wind speed (m/s)
+V=1 
 #assumptions
@@
 
-Although measured solar radiation is preferable if available, we can estimate hourly solar radiation by discounting incoming solar radiation as it moves through the atmosphere as follows:
+Although measured solar radiation is preferable if available, we can
+estimate hourly solar radiation by discounting incoming solar radiation
+as it moves through the atmosphere as follows:
+
 ```{r}
-psi_deg= sapply(1:24, FUN=zenith_angle, doy=doy, lat=lat, lon=lon) #estimate zenith angle (degrees)
-psi_rad= degree_to_radian(psi_deg) #convert to radians
-Srad= sapply(psi_rad, FUN=estimate_radiation, doy=doy, tau=tau, elev=elev, rho=rho) #estimate radiation
+#estimate zenith angle (degrees)
+psi_deg= sapply(1:24, FUN=zenith_angle, doy=doy, lat=lat, lon=lon) 
+#convert to radians
+psi_rad= degree_to_radian(psi_deg) 
+#estimate radiation
+Srad= sapply(psi_rad, FUN=estimate_radiation, doy=doy, tau=tau, elev=elev, rho=rho)
 #Separate into direct, diffuse, and reflected solar radiation
@@
 
-We can then calculate hourly air and soil temperatures based on daily minimum and maximum temperatures.
-We select the sine-exponential model for air temperature and the sine model for surface temperature:
+We can then calculate hourly air and soil temperatures based on daily
+minimum and maximum temperatures. We select the sine-exponential model
+for air temperature and the sine model for surface temperature:
+
 ```{r}
-Ta= sapply(1:24, diurnal_temp_variation_sineexp, T_max=Tmax, T_min=Tmin, t_r=tr, t_s=ts, alpha=2.59, beta= 1.55, gamma=2.2) #air temperature (C)
-Ts= sapply(1:24, diurnal_temp_variation_sine, T_max=Tmax_s, T_min=Tmin_s) #soil temperature (C)
+#air temperature (C)
+Ta= sapply(1:24, diurnal_temp_variation_sineexp, T_max=Tmax, T_min=Tmin, t_r=tr, t_s=ts, alpha=2.59, beta= 1.55, gamma=2.2) 
+#soil temperature (C)
+Ts= sapply(1:24, diurnal_temp_variation_sine, T_max=Tmax_s, T_min=Tmin_s) 
 ```
 
-We use microclimate models to scale air temperature and wind speed from weather station height (2m) to lizard height (0.02m).
-We assume a surface roughness of $z_0=0.2$, which corresponds to bare sand and determines the turbulence of airflow.
-We implement neutral air temperature and wind speed profiles as well as an air temperature profile forced by wind speed.
+We use microclimate models to scale air temperature and wind speed from
+weather station height (2m) to lizard height (0.02m). We assume a
+surface roughness of $z_0=0.2$, which corresponds to bare sand and
+determines the turbulence of airflow. We implement neutral air
+temperature and wind speed profiles as well as an air temperature
+profile forced by wind speed.
 
@@
 ### Energy balance
+
 We will be solving the following energy balance to estimate $T_e$:
 $$Q_{net} = Q_{abs} - Q_{emit} - Q_{conv} - Q_{cond} - Q_{met} - Q_{evap},$$
-where $Q_{net}$ is the net energy exchange with the environment (W), $Q_{abs}$ is the solar radiation absorbed (W), $Q_{emit}$ is the net thermal radiation emitted (W), $Q_{conv}$ is energy exchange due to convection (W), $Q_{cond}$ is energy exchange due to conduction (W), $Q_{met}$ is the energy emitted due to metabolism (W), and $Q_{evap}$ is the energy emitted due to evaporative water loss (W).
-We will estimate each term on the right side of the equation in turn.
-Estimating $Q_{abs}$ requires the surface area exposed to radiation and the solar absorptivity of the animal surface ($a$ proportion).
-We model a 10g Sceloporus lizard with solar absorptivity $a=0.9$ [@gates1980biophysical].
-We will initially assume $T_b=T_a$ to illustrate the calculations before solving for $T_b$ givne the environmental conditions.
+where $Q_{net}$ is the net energy exchange with the environment (W),
+$Q_{abs}$ is the solar radiation absorbed (W), $Q_{emit}$ is the net
+thermal radiation emitted (W), $Q_{conv}$ is energy exchange due to
+convection (W), $Q_{cond}$ is energy exchange due to conduction (W),
+$Q_{met}$ is the energy emitted due to metabolism (W), and $Q_{evap}$ is
+the energy emitted due to evaporative water loss (W). We will estimate
+each term on the right side of the equation in turn. Estimating
+$Q_{abs}$ requires the surface area exposed to radiation and the solar
+absorptivity of the animal surface ($a$ proportion). We model a 10g
+*Sceloporus* lizard with solar absorptivity $a=0.9$
+[@gates1980biophysical]. We will initially assume $T_b=T_a$ to
+illustrate the calculations before solving for $T_b$ given the
+environmental conditions.
 
@@
 We calculate the hourly solar radiation absorbed (W) as follows:
+
 ```{r}
@@
 
-We estimate thermal radiation $Q_{emit}$ (W) for the lizard outdoors. We assume the surface emissivity of lizards, $epsilon_s=0.965$ [@barlett1967energy].
+We estimate thermal radiation $Q_{emit}$ (W) for the lizard outdoors. We
+assume the surface emissivity of lizards, $epsilon_s=0.965$
+[@barlett1967].
 
@@
 
-We next estimate convection $Q_{conv}$ (W) and conduction $Q_{cond}$ (W).
-We will estimate the lizard's heat transfer coefficient, $H_L$ ($Wm^{-2}K^{-1}$) using an empirical relationship for lizards (`heat_transfer_coefficient()`), but also illustrate a function estimating $H_L$ using a spherical approximation (`heat_transfer_coefficient_approximation()`) and a simplified approximation (`heat_transfer_coefficient_simple()`).
+We next estimate convection $Q_{conv}$ (W) and conduction $Q_{cond}$
+(W). We will estimate the lizard's heat transfer coefficient, $H_L$
+($Wm^{-2}K^{-1}$) using an empirical relationship for lizards
+(`heat_transfer_coefficient()`), but also illustrate a function
+estimating $H_L$ using a spherical approximation
+(`heat_transfer_coefficient_approximation()`) and a simplified
+approximation (`heat_transfer_coefficient_simple()`).
 
@@
 
-We estimate convective heat exchange between the animal and surrounding air using the following relationship:
-$$Q_{conv}= ef \cdot H_L(A\cdot \mbox{proportion})(T_a-T_b),$$
-where an enhancement factor, $ef$, multiplier can be incorporated to account for increases in heat exchange resulting from air turbulence in field conditions.
-We implement the function in R assuming that 2/3 of the lizard's surface area is exchanging heat through convection.
-We note we currently estimate no convection since we are assuming $Tb=Ta$ for illustration purposes:
+We estimate convective heat exchange between the animal and surrounding
+air using the following relationship:
+$$Q_{conv}= ef \cdot H_L(A\cdot \mbox{proportion})(T_a-T_b),$$ where an
+enhancement factor, $ef$, multiplier can be incorporated to account for
+increases in heat exchange resulting from air turbulence in field
+conditions. We implement the function in R assuming that 2/3 of the
+lizard's surface area is exchanging heat through convection. We note we
+currently estimate no convection since we are assuming $Tb=Ta$ for
+illustration purposes:
+
 ```{r}
@@
 
-We estimate conductive heat flow (W) from the lizard to the surface assuming conductance through the animal tissue is the rate limiting step as follows:
-$$Q_{cond}= A \cdot \mbox{proportion} \cdot K(T_g-T_b)/d. $$
-We implement the estimate in R assuming that 1/3 of the lizard surface is in contact with the ground and a skin thickness $d$ of $10^{-6}$:
+We estimate conductive heat flow (W) from the lizard to the surface
+assuming conductance through the animal tissue is the rate limiting step
+as follows: $$Q_{cond}= A \cdot \mbox{proportion} \cdot K(T_g-T_b)/d. $$
+We implement the estimate in R assuming that 1/3 of the lizard surface
+is in contact with the ground and a skin thickness $d$ of $10^{-6}$:
+
 ```{r}
@@
 
-We assume, as in generally done for lizards, that heat exchange associated with metabolism and evaporation is negligible:
+We assume, as in generally done for lizards, that heat exchange
+associated with metabolism and evaporation is negligible:
+
 ```{r}
@@
 
-The full heat budget can be calculated as follows [@gates1980biophysical]: 
+The full heat budget can be calculated as follows
+[@gates1980biophysical]:
+
 ```{r}
@@
 ```
+
 We constructed the energy balance assuming $T_b=T_a$ for illustration.
-Thus, the energy balance is dominated by heat gain from contact with the warmer surface.
-We now use a function based on the energy balance above to estimate body temperature given the environmental conditions:
+Thus, the energy balance is dominated by heat gain from contact with the
+warmer surface. We now use a function based on the energy balance above
+to estimate body temperature given the environmental conditions:
 
@@
 
-We also implement a similar but simplified energy balance [@campbell2000introduction]. The energy balance omits conduction with the ground:
+We also implement a similar but simplified energy balance
+[@campbell2000introduction]. The energy balance omits conduction with
+the ground:
+
 ```{r}
@@
 
-We additionally estimate $T_e$ using a specialized function for lizards [@buckley2008linking]:
+We additionally estimate $T_e$ using a specialized function for lizards
+[@buckley2008link]:
+
 ```{r}
@@
 ### Plot
-We next plot the diurnal trends in environmental data and predicted body temperatures.
 
-```{r}
+We next plot the diurnal trends in environmental data and predicted body
+temperatures.
+
+```{r, echo=FALSE}
 par(mar=c(5, 4, 4, 8) + 0.1)
@@
 
-*Figure 1*.
-Predicted lizard body temperatures (Te) as a function of air temperatures (Ta) at 2m and lizard height (0.02m), surface (Ts) temperatures (°C), and solar radiation ($W/m^2$).
-We depict body temperatures using two general energy budgets [solid: Tb_Gates(); dotted: Tb_CampbellNorman()] and a lizard specific biophysical model [dashed: Tb_lizard()].
+*Figure 1*. Predicted lizard body temperatures (Te) as a function of air
+temperatures (Ta) at 2m and lizard height (0.02m), surface (Ts)
+temperatures (°C), and solar radiation ($W/m^2$). We depict body
+temperatures using two general energy budgets [solid: Tb_Gates();
+dotted: Tb_CampbellNorman()] and a lizard specific biophysical model
+[dashed: Tb_lizard()].
 
-The microclimate models indicate that air temperatures at lizard height are similar to surface temperatures (Figure 1).
-The biophysical models indicate that solar radiation will elevate lizard body temperatures far above air temperatures and that the lizard will face thermal stress if it is unable to seek shade (Figure 1).
-The three biophysical models predict different body temperatures during peak period of solar radiation because they model interactions with the ground differently and users are encouraged to review the details of each biophysical model before selection.
+The microclimate models indicate that air temperatures at lizard height
+are similar to surface temperatures (Figure 1). The biophysical models
+indicate that solar radiation will elevate lizard body temperatures far
+above air temperatures and that the lizard will face thermal stress if
+it is unable to seek shade (Figure 1). The three biophysical models
+predict different body temperatures during peak period of solar
+radiation because they model interactions with the ground differently
+and users are encouraged to review the details of each biophysical model
+before selection. Differences in estimated body temperatures are
+accentuated by the high level of solar radiation.
 
-## Summary
-TrenchR is intended to promote understanding of how organisms interact with their environment and the consequences for physiology, energetics, behavior, and demography.
-Our example implementation highlights the importance of considering organismal body temperatures, rather than air temperatures, when examining thermal stress and other responses to environmental variability and change.
-TrenchR currently focuses on heat balances but may be expanded to include water balances.
-Simple functions can be combined as needed to produce comprehensive and transparent models for biophysical ecology and evolution.
-The resultant models are likely to be sufficiently detailed for many applications, but users are referred to NicheMapR for more detailed biophysical models [@kearney2017nichemapr; @kearney2020nichemapr].
-Making classic biophysical ecology techniques more accessible will allow researchers to take advantage of rapidly accumulating data on environmental conditions and organismal traits to understand and predict ecological and evolutionary responses to variable and changing environments.
+## Conclusions
+
+TrenchR is intended to promote understanding of how organisms interact
+with their environment and the consequences for physiology, energetics,
+behavior, and demography. Our example implementation highlights the
+importance of considering organismal body temperatures, rather than air
+temperatures, when examining thermal stress and other responses to
+environmental variability and change. TrenchR currently focuses on heat
+balances but may be expanded to include water balances. Simple functions
+can be combined as needed to produce comprehensive and transparent
+models for biophysical ecology and evolution. The resultant models are
+likely to be sufficiently detailed for many applications, but users are
+referred to NicheMapR for more detailed biophysical models
+[@kearney2017nichemapr; @kearney2020nichemapr]. Making classic
+biophysical ecology techniques more accessible will allow researchers to
+take advantage of rapidly accumulating data on environmental conditions
+and organismal traits to understand and predict ecological and
+evolutionary responses to variable and changing environments.
+
+## Acknowledgements
+
+We thank Brian Helmuth and Eric Riddell for providing and checking
+functions. We thank Ray Huey, Joel Kingsolver, Luke Miller, Warren
+Porter, and Art Woods for assistance and input. We particularly thank
+Michael Kearney for extensive input on TrenchR and its alignment with
+NicheMapR and for contributing two educational modules.
+
+[Check with Eric]
+
+## Authors' contributions
+
+LB conceived and designed the package and lead all aspect of the
+packages creation and dissemination; BBO, IC, AJ, OL, AM, YS wrote and
+edited functions and associated package resources; AJ configured the
+package; AM, BBO, YS, and LB converted and edited the educational
+modules. All authors contributed critically to the drafts and gave final
+approval for publication.
+
+## Data availability
+
+All code is available at <https://github.com/trenchproject/TrenchR>.
 

### 2021-04-30 ManuscriptTe.Rmd
Update ms

@@
 ---
-title: "Manuscript Te example"
-author: "Lauren Buckley"
-date: "`r Sys.Date()`"
+title: "Appendix S1"
+author: "Buckley et al."
 output:
+  html_document: default
   pdf_document: default
-  html_document: default
   word_document: default
@@
 
-```{r}
+```{r, include=FALSE}
 ##Load TrenchR package
@@
 
-#APPENDIX S1
+# APPENDIX S1
 We provide an additional example using microclimate data measured along an elevation gradient in CO, USA to illustrate use of TrenchR.
 We first use a times series of air temperatures and wind speeds collected at multiple heights to examine profiles and surface roughness.
-We then use a time series of air and surface temperature, windspeed, and solar radiation collected a a single height to implement and test a biophysical model for grasshoppers.
+We then use a time series of air and surface temperature, windspeed, and solar radiation collected at a single height to implement and test a biophysical model for grasshoppers.
 
-#Microclimate models
-```{r}
+# Microclimate models
+```{r, include=FALSE}
 setwd("/Volumes/GoogleDrive/Shared drives/TrEnCh/Projects/TrenchR/data/")
@@
 
-We first examine temperature and wind speed profiles and use them to estimate surface roughness.
-```{r}
+```{r, include=FALSE}
 #ANALYZE SURFACE ROUGHNESS AND PROFILES
@@
 
-We can use wind speed variation across elevations to estimate surface roughness across hours.
+We first examine how temperature and windspeed profiles can be used to scale conditions to the heights of organisms.
+We first use wind speed variation across hours and heights to estimate surface roughness.
 ```{r}
-#surface roughness
+#estimate surface roughness based on wind velocity (m/s) at a vector of reference heights, zr (m)
 z0= apply(C1.hr[, c("Wind0.25m","Wind0.5m","Wind1m","Wind1.5m")], FUN=surface_roughness, MARGIN=1, zr=c(0.25,0.5,1.0,1.5))
@@
 
-We calculate the temperature and wind speed profiles for a given hour using multiple functions in TrenchR.
+We then calculate the temperature and wind speed profiles for a given hour using multiple functions in TrenchR.
 ```{r}
-z.seq <- seq(0, 2, by = 0.1)
-
 #subset to day and hour
@@
 
-#estimate profiles
-#temperature
+#vector of heights
+z.seq <- seq(0, 2, by = 0.1)
+
+#estimate profiles using three different functions including the assumption of neutral and segmented profiles. 
+#T_r	and u_r are temperatures (C) and windspeeds (m/s), respectively, at reference height zr (m).
+#z0	is surface roughness (m); z is the height to scale to (m).
+#T_s	is surface temperature (C).
+
+#temperature profiles
 t.seq1= sapply(z.seq, FUN=air_temp_profile_neutral, T_r=C1.sub[hr,"Air0.5m"], z0=z0[hr], zr=0.5, T_s=C1.sub[hr,"Soil"])
@@
 
-#windspeed
+#windspeed profiles
 u.seq1= sapply(z.seq, FUN=wind_speed_profile_neutral, u_r=C1.sub[hr,"Wind0.5m"], zr=0.5, z0=z0[hr])
@@
 u.seq3= sapply(z.seq, FUN=wind_speed_profile_segment, u_r=C1.sub[hr, c("Wind0.5m","Wind1m","Wind1.5m")], zr=c(0.5,1,1.5), z0=rep(z0[hr],3))
+```
 
+The temperature and wind speed profiles illustrate how temperatures generally decrease and wind speeds generally increase with height above the ground.
+Profiles can be fit as a single smooth function or segments.
+```{r, echo=FALSE}
 #replace infinity with N
 t.seq1[which(!is.finite(t.seq1))]=NA
-```
 
-We then plot the temperature and wind speed profiles.
-```{r}
 plot(t.seq1, z.seq, type = "l", xlab = "Temperature (C)", ylab = "z (height above ground in m)")
@@
 
-#Biophysical models
-We first read and process time series of environmental data.
-```{r}
+# Biophysical models
+```{r, include=FALSE}
 #Grasshopper data
@@
 
-#Calculate zenith
-dat$psi= zenith_angle(dat$J,lat=dat$lat, lon=dat$lon, hour=dat$hour)
-
 #Subset to August
@@
 
+We next illustrate estimating operative environmental temperatures, the estimated body temperatures for grasshoppers in the specified environments.
+We first read and process time series of environmental data.
+We estimate the zenith angle, the angle between the sun's rays and the vertical, across Julian days and hours.
+```{r}
+#Calculate zenith
+dat$psi= zenith_angle(dat$J,lat=dat$lat, lon=dat$lon, hour=dat$hour)
+
+```
+
 We scale the air temperature data from 0.5m to grasshopper height (0.001m).
@@
 
-#scale temperature to ground
+#scale temperature T_r from reference height zr=0.5m to grasshopper height based on windspeed u_r, surface temperature T_s, and surface roughness z0.
 dat$Tgrass <- air_temp_profile(T_r=dat$Temp, u_r=dat$Wind, zr=0.5, z0=0.02, z=z, T_s=dat$SoilTemp)
+```
 
+```{r, include=FALSE}
 #cut nas
@@
 We first use a biophysical model specifically configured for grasshoppers.
-```{r}
-##Calculate Te for each species
+```{r, warning=FALSE}
+##Calculate operative temperature, Te
 dat$Te<-Tb_grasshopper(T_a=dat$Tgrass, T_g=dat$SoilTemp, u=dat$Wind, H=dat$Rad, K_t=0.7, psi=dat$psi, L=0.0211, Acondfact=0.01)
+
+#T_a	and T_g are air and surface temperatures (C), respectively; u	is wind speed (m/s); 
+#H is total (direct + diffuse) solar radiation flux (W/m^2); 
+#K_t	is the clearness index (dimensionless), which is the ratio of the global solar radiation measured at the surface to the total solar radiation at the top of the atmosphere;
+#psi is solar zenith angle (degrees); L	is grasshopper length (m);
+#Acondfact is the proportion of the grasshopper surface area in contact with the ground.
+```
+
+```{r, include=FALSE}
 #fix infinite
@@
 dat$H_L=heat_transfer_coefficient(V=dat$Wind,D=0.0211,K=0.025, nu= 15.3 * 10^(-6) , taxa="cylinder")
+#where V is	air velocity (m/s); D is the characteristic dimension (m);
+#K is the thermal conductivity of air (W m^-1 K^-1);
+#nu is the kinematic viscosity of air (m^2 s^-1)
 
-#Also illustrate estimations using a spherical approximation and a simplified version of the approximation.
+#Also estimate using a spherical approximation and a simplified version of the approximation.
 H_L1= heat_transfer_coefficient_approximation(V=dat$Wind,D=0.0211,K= 0.025, nu= 15.3 * 10^(-6), taxa="flyinginsect")
@@
 
-#surface area
+#surface area from length(m)
 A= sa_from_length(l=0.0211)
@@
 We then implement the energy budget as follows.
-```{r}
-#Gates calculation
+```{r, results = FALSE}
 TeGates= rep(NA, nrow(dat) )
@@
 for(ind in 1:nrow(dat) ){
-  out= try(Tb_Gates(A=A, D=0.0211/3, psa_dir=0.66, psa_ref=1-0.66, psa_air=1, psa_g=0.01, T_g=dat$SoilTemp[ind]+273, T_a=dat$Tgrass[ind]+273, Qabs=A*dat$Rad[ind], epsilon=0.95, H_L=dat$H_L[ind], ef=1.3, K=0.15), silent=TRUE)
+  out= try(Tb_Gates(A=A, D=0.0211/3, psa_dir=0.66, psa_ref=1-0.66, psa_air=0.99, psa_g=0.01, T_g=dat$SoilTemp[ind]+273, T_a=dat$Tgrass[ind]+273, Qabs=A*dat$Rad[ind], epsilon=0.95, H_L=dat$H_L[ind], ef=1.3, K=0.15), silent=TRUE)
   if(is.numeric(out))TeGates[ind]=out
 }
-dat$TeGates=TeGates-273.15
+#where psa_dir, psa_ref, psa_air, psa_g are the proportions surface area exposed to direct radiation from the sky (or enclosure), reflected radiation from the ground, and air, and in contact with the ground, respectively;
+#Qabs is the solar and thermal radiation absorbed (W);
+#epsilon is longwave infrared emissivity of skin (proportion);
+#H_L is the convective heat transfer coefficient (W m^-2 K^-1);
+#ef	is the enhancement factor, used to adjust H_L to field conditions;
+#K is thermal conductivity. 
 ```
@@
 We aggregate the data and plot average conditions across sites.
-```{r}
+```{r, echo=FALSE, warning=FALSE}
+dat$TeGates=TeGates-273.15
+
 #aggregate by hour
@@
 #plot
-dat.s= subset(dat.s, dat.s$variable %in% c("Temp","SoilTemp","SanguinipesTemp","Te","Tgrass","TeGates") )
+dat.pl= subset(dat.s, dat.s$variable %in% c("Temp","SoilTemp","SanguinipesTemp","Te","Tgrass","TeGates") )
+dat.pl$variable= as.character(dat.pl$variable)
 
-ggplot(data=dat.s, aes(x=hour, y = value, color=variable))+geom_line(aes(lty=variable) )+facet_wrap(~Elev)
+#change names
+dat.pl[which(dat.pl$variable=="Temp"),"variable"]<-"Ta"
+dat.pl[which(dat.pl$variable=="SoilTemp"),"variable"]<-"Ts"
+dat.pl[which(dat.pl$variable=="SanguinipesTemp"),"variable"]<-"Te observed"
+dat.pl[which(dat.pl$variable=="Te"),"variable"]<-"Te"
+dat.pl[which(dat.pl$variable=="Tgrass"),"variable"]<-"Ta grasshopper height"
+dat.pl[which(dat.pl$variable=="TeGates"),"variable"]<-"Te Gates"
+
+#make factor
+dat.pl$variable= factor(dat.pl$variable, levels=c("Ta","Ts","Ta grasshopper height","Te","Te Gates","Te observed"))
+#make label
+dat.pl$elev.lab= paste(dat.pl$Elev,"m", sep=" ")
+#make Ta, Te classification
+dat.pl$Ttype="input"
+dat.pl$Ttype[which(dat.pl$variable %in% c("Te","Te Gates","Te observed") )]="estimates"
+
+ggplot(data=dat.pl, aes(x=hour, y = value, color=variable))+geom_line(aes(lty=variable) )+facet_wrap(~elev.lab)+theme_classic()+ theme(legend.position="bottom")
 
@@
 
-```{r}
+```{r, include=FALSE}
 #plot by date time
@@
 
-```{r}
+```{r, include=FALSE}
 setwd("/Volumes/GoogleDrive/Shared drives/TrEnCh/Projects/TrenchR/figures/")

### 2021-04-30 Manuscript_full.Rmd
Update ms

@@
 
-1.  Much understanding of climate change responses relies on the
+1.  Much understanding of responses to climate change and variability relies on the
     assumption that organismal body temperatures are equal to temporally
     averaged air temperatures high above the ground. However, most
-    organisms live near the ground and acute exposure to solar radiation
+    organisms live near the ground and acute exposure to solar radiation and thermal extremes
     can substantially elevate their body temperatures.
 2.  We introduce the TrenchR package, which aids in Translating
-    Environmental Change into organismal responses
-    (<https://github.com/trenchproject/TrenchR>). The package includes
+    Environmental Change into organismal responses. The package includes
     microclimate models to vertically scale weather station data to
@@
     model and temporally partition air and soil temperatures and solar
-    radiation.
-3.  [Main results]
+    radiation. We offer utility functions to aid in estimating the organismal and environonmental parameters needed for biophysical ecology.
+3.  TrenchR focuses on relatively simple and modular functions so users can create transparent and flexible biophysical models.
 4.  The package aims to introduce and enable microclimate and
     biophysical modelling to improve ecological and evolutionary
-    forecasting and includes tutorials and well as a series of
-    educational modules.
+    forecasting. To further this aim, we offer a series of educational modules to introduce the field of biophysical ecology.
 
@@
 discrepancies of tens of degrees C [@sinclair2016; @sunday2014;
-@kearney2009]. Acute thermal stress events are often more relevant to
+@kearney2009].
+Applications are increasingly demonstrating the importance of accounting for microclimate variation [@bramer2018advances; @butikofer2020problem; @lembrechts2019comparing].
+Acute thermal stress events are often more relevant to
 the physiology and energy balance of organisms than daily, monthly, or
@@
 @dillon2016life]. We aim to help remedy these shortcomings by
-introducing the TrenchR package as a component of the TrEnCh project
-(<https://trenchproject.github.io>), which builds computational and
+introducing the TrenchR package (<https://github.com/trenchproject/TrenchR>) as a component of the TrEnCh project
+(<https:www.trenchproject.com>), which builds computational and
 visualization tools to Translate Environmental Change into organismal
@@
 the NicheMapR package, which includes sophisticated microclimate
-[@kearney2017nichemapr] and biophysical [@kearney2020] models. Many of
+[@kearney2017nichemapr] and biophysical [@kearney2020nichemapr] models. Many of
 the approaches are similar between the packages, but TrenchR focuses on
@@
 biophysical approaches. The TrenchR functions allow for modular use and
-easy adaptation to a variety of organismal and research questions. Some
+easy adaptation to a variety of organisms and research questions. Some
 energy budget calculation components of biophysical models are also
@@
 the TrenchR functions are general, some components of TrenchR are best
-suited for ectothermic animals. We have minimized inclusion of plant
+suited for ectothermic animals. We have limited inclusion of plant
 specific biophysical models since they are the focus of several R
@@
 @gates1980biophysical] include many tables with parameter values and we
-provide several tables in TrenchR [Table 8.2XXXX].
+provide several data tables in TrenchR.
 
@@
 
-We introduce the function in several vignettes. A good place to start is
-the *Allometry and conversions* tutorial, which provides tools for
+We introduce the functions in several vignettes. A good place to start is
+the *Allometry and conversions* vignette, which provides tools for
 preparing data such as estimating additional dimensions of organisms
-from measured dimensions. The *Estimating microclimates* tutorial
+from measured dimensions. The *Estimating microclimates* vignette
 provides resources for estimating the environmental conditions
@@
 For additional background on microclimate and biophysical modelling, we
-have updated a series of educational modules entitled "Physical
-processes in ecosystems". We intend the modules to provide a more
+have updated a series of educational modules entitled *Physical
+Processes in Ecosystems*. We intend the modules to provide a more
 contained and accessible introduction to microclimate and biophysical
@@
 
-## Examples
+## Implementation
 
 We illustrate the use of the TrenchR package by estimating an energy
-budget for a *Sceloporus* lizard on June 1, 2018, in Santa Fe, New
+budget for a *Sceloporus* lizard on June 1, 2020, in Santa Fe, New
 Mexico, USA (35.69, -105.944, elevation: 2121m). The example is also
@@
 
-We will use the energy budget to estimate the lizard's body temperature,
+We will use the energy budget to estimate body temperature,
 $T_b$, which can depart dramatically from the air temperatures due to
@@
 
-Let's assume the lizard is in an unshaded location where a weather
+Let us assume the lizard is in an unshaded location where a weather
 station at standard height (2m) indicates that the daily air temperature
@@
 
-Let us first prepare the environmental data for analysis. We will need
+We first prepare the environmental data for analysis. We will need
 to estimate hourly air and soil temperatures and radiation. We start by
@@
 
-doy= day_of_year("2018-06-01", format= "%Y-%m-%d") 
+doy= day_of_year("2020-06-01", format= "%Y-%m-%d") 
 snoon= solar_noon(lon=lon, doy=doy) #estimate solar noon
@@
 
-### Plot
-
-We next plot the diurnal trends in environmental data and predicted body
-temperatures.
+The microclimate models indicate that air temperatures at lizard height
+are similar to surface temperatures (Figure 1). The biophysical models
+indicate that solar radiation will elevate lizard body temperatures far
+above air temperatures and that the lizard will face thermal stress if
+it is unable to seek shade (Figure 1). The three biophysical models
+predict different body temperatures during peak period of solar
+radiation because they model interactions with the ground differently
+and users are encouraged to review the details of each biophysical model
+before selection. Differences in estimated body temperatures are
+accentuated by the high level of solar radiation.
 
@@
 
-*Figure 1*. Predicted lizard body temperatures (Te) as a function of air
-temperatures (Ta) at 2m and lizard height (0.02m), surface (Ts)
-temperatures (°C), and solar radiation ($W/m^2$). We depict body
-temperatures using two general energy budgets [solid: Tb_Gates();
-dotted: Tb_CampbellNorman()] and a lizard specific biophysical model
-[dashed: Tb_lizard()].
-
-The microclimate models indicate that air temperatures at lizard height
-are similar to surface temperatures (Figure 1). The biophysical models
-indicate that solar radiation will elevate lizard body temperatures far
-above air temperatures and that the lizard will face thermal stress if
-it is unable to seek shade (Figure 1). The three biophysical models
-predict different body temperatures during peak period of solar
-radiation because they model interactions with the ground differently
-and users are encouraged to review the details of each biophysical model
-before selection. Differences in estimated body temperatures are
-accentuated by the high level of solar radiation.
+* **Figure 1**.
+Body temperatures (Te) are predicted to drastically exceed air temperature when lizards are exposed to high levels of solar radiation.
+Air temperatures (Ta, °C) at lizard height (0.02m) are predicted to exceed air temperatures at 2m and to be similar to surface temperatures (Ts).
+We estimate body temperatures using two general energy budgets [solid: Tb_Gates();
+dotted: Tb_CampbellNorman()] and a lizard specific biophysical model [dashed: Tb_lizard()] that differ in how they model heat exchanges.*
 
@@
 TrenchR is intended to promote understanding of how organisms interact
-with their environment and the consequences for physiology, energetics,
+with their environment and consequences for physiology, energetics,
 behavior, and demography. Our example implementation highlights the
@@
 referred to NicheMapR for more detailed biophysical models
-[@kearney2017nichemapr; @kearney2020nichemapr]. Making classic
+[@kearney2017nichemapr; @kearney2020nichemapr].
+We focus on models that predict equilibrium conditions for simplicity.
+Such models generally do not present computational challenges so our models are not optimized for computational efficiency.
+Making classic
 biophysical ecology techniques more accessible will allow researchers to
@@
 and organismal traits to understand and predict ecological and
-evolutionary responses to variable and changing environments.
+evolutionary responses. Considering how organisms experience their environment is central to understanding responses to variable and changing environments [@suggitt2018extinction].
 

### 2021-05-03 ManuscriptTe.Rmd
Text updates.

@@
 We provide an additional example using microclimate data measured along an elevation gradient in CO, USA to illustrate use of TrenchR.
-We first use a times series of air temperatures and wind speeds collected at multiple heights to examine profiles and surface roughness.
-We then use a time series of air and surface temperature, windspeed, and solar radiation collected at a single height to implement and test a biophysical model for grasshoppers.
+We first use a time series of air temperatures and wind speeds collected at multiple heights to examine profiles and surface roughness.
+We then use a time series of air and surface temperatures, wind speeds, and solar radiation collected at a single height to implement a biophysical model for grasshoppers and compare estimates to observations.
 
@@
 
-We first examine how temperature and windspeed profiles can be used to scale conditions to the heights of organisms.
+We first examine how temperature and wind speed profiles can be used to scale conditions to the heights of organisms.
 We first use wind speed variation across hours and heights to estimate surface roughness.
@@
 
-We then calculate the temperature and wind speed profiles for a given hour using multiple functions in TrenchR.
+We then calculate the temperature and wind speed profiles for a given hour using multiple, alternative functions in TrenchR.
 ```{r}
@@
 
-The temperature and wind speed profiles illustrate how temperatures generally decrease and wind speeds generally increase with height above the ground.
+The temperature and wind speed profiles illustrate how temperatures generally decrease and wind speeds generally increase with height above the ground (Figure 1).
 Profiles can be fit as a single smooth function or segments.
@@
 ```
+
+*Figure 1.
+Temperature (top) and wind speed (bottom) profiles using similar continuous [solid: air_temp_profile_neutral(); dashed: air_temp_profile()] and segmented estimation [dotted: air_temp_profile_segment()].*
 
@@
 
-We next illustrate estimating operative environmental temperatures, the estimated body temperatures for grasshoppers in the specified environments.
+We next illustrate quantifying operative environmental temperatures, which are estimated body temperatures for grasshoppers in the specified environments.
 We first read and process time series of environmental data.
@@
 
-We scale the air temperature data from 0.5m to grasshopper height (0.001m).
+We scale the air temperature and wind speed from a reference height of 0.5m to grasshopper height (0.001m).
 ```{r}
@@
 dat$Tgrass <- air_temp_profile(T_r=dat$Temp, u_r=dat$Wind, zr=0.5, z0=0.02, z=z, T_s=dat$SoilTemp)
+
+#scale wind to grasshopper height
+dat$Ugrass= wind_speed_profile_neutral(u_r=dat$Wind, zr=0.5, z0=0.02,z=z)
 ```
@@
 #cut nas
-dat= dat[,c("Rad","Temp","Wind","SoilTemp","SanguinipesTemp","Elev","J", "hour","psi","Tgrass")]
+dat= dat[,c("Rad","Temp","Wind","SoilTemp","SanguinipesTemp","Elev","J", "hour","psi","Tgrass","Ugrass")]
 dat= na.omit(dat)
@@
 ```{r, warning=FALSE}
+
 ##Calculate operative temperature, Te
-dat$Te<-Tb_grasshopper(T_a=dat$Tgrass, T_g=dat$SoilTemp, u=dat$Wind, H=dat$Rad, K_t=0.7, psi=dat$psi, L=0.0211, Acondfact=0.01)
+dat$Te<-Tb_grasshopper(T_a=dat$Tgrass, T_g=dat$SoilTemp, u=dat$Ugrass, H=dat$Rad, K_t=1, psi=dat$psi, L=0.0211, Acondfact=0.0, abs=0.9, r_g=0.5)
 
@@
 #psi is solar zenith angle (degrees); L	is grasshopper length (m);
-#Acondfact is the proportion of the grasshopper surface area in contact with the ground.
+#Acondfact is the proportion of the grasshopper surface area in contact with the ground;
+#abs is absorptivity of grasshopper to solar radiation (proportion);
+#r_g is substrate solar reflectivity (proportion).
 ```
@@
 #Estimate heat transfer coefficient using empirical relationship
-dat$H_L=heat_transfer_coefficient(V=dat$Wind,D=0.0211,K=0.025, nu= 15.3 * 10^(-6) , taxa="cylinder")
+dat$H_L=heat_transfer_coefficient(V=dat$Ugrass,D=0.0211,K=0.025, nu= 15.3 * 10^(-6) , taxa="cylinder")
 #where V is	air velocity (m/s); D is the characteristic dimension (m);
@@
 #Also estimate using a spherical approximation and a simplified version of the approximation.
-H_L1= heat_transfer_coefficient_approximation(V=dat$Wind,D=0.0211,K= 0.025, nu= 15.3 * 10^(-6), taxa="flyinginsect")
-H_L2= heat_transfer_coefficient_simple(V=dat$Wind,D=0.0211)
+H_L1= heat_transfer_coefficient_approximation(V=dat$Ugrass,D=0.0211,K= 0.025, nu= 15.3 * 10^(-6), taxa="flyinginsect")
+H_L2= heat_transfer_coefficient_simple(V=dat$Ugrass,D=0.0211)
 
@@
 We then implement the energy budget as follows.
-```{r, results = FALSE}
+```{r, results = FALSE, warning=FALSE}
 TeGates= rep(NA, nrow(dat) )
@@
 for(ind in 1:nrow(dat) ){
-  out= try(Tb_Gates(A=A, D=0.0211/3, psa_dir=0.66, psa_ref=1-0.66, psa_air=0.99, psa_g=0.01, T_g=dat$SoilTemp[ind]+273, T_a=dat$Tgrass[ind]+273, Qabs=A*dat$Rad[ind], epsilon=0.95, H_L=dat$H_L[ind], ef=1.3, K=0.15), silent=TRUE)
+
+  out= try(Tb_Gates(A=A, D=0.0211/3, psa_dir=0.5, psa_ref=1-0.5, psa_air=1, psa_g=0.0, T_g=dat$SoilTemp[ind]+273, T_a=dat$Tgrass[ind]+273, Qabs=A*0.66*dat$Rad[ind], epsilon=0.95, H_L=dat$H_L[ind], ef=1.3, K=0.15), silent=TRUE)
   if(is.numeric(out))TeGates[ind]=out
@@
 
-We aggregate the data and plot average conditions across sites.
+We aggregate the data and plot average conditions across sites (Figure 2).
+The simplifying assumptions in the biophysical model implementation reduce alignment with the observed temperatures of grasshoppers, but the divergence of grasshopper temperatures from air temperatures at 2m and, to a lesser extent, air temperatures at grasshopper height demonstrate the importance of implementing biophysical models.
 ```{r, echo=FALSE, warning=FALSE}
 dat$TeGates=TeGates-273.15
+
+#drop NAs
+dat= dat[which(!is.na(dat$TeGates)),]
 
@@
 
-ggplot(data=dat.pl, aes(x=hour, y = value, color=variable))+geom_line(aes(lty=variable) )+facet_wrap(~elev.lab)+theme_classic()+ theme(legend.position="bottom")
+ggplot(data=dat.pl, aes(x=hour, y = value, color=variable))+geom_line(aes(lty=variable) )+facet_wrap(~elev.lab)+theme_classic()+ theme(legend.position="bottom")+ylim(15,65)
 
 ```
+
+*Figure 2.
+Diurnal temperature variation averaged across days in August varies across the elevations of CO sites (panels) and the type of temperature (C).
+Microclimate models indicate that the air temperatures experienced by grasshoppers near the ground (Ta grasshopper height) is generally above air temperatures (Ta) and close to surface temperatures (Ts).
+Body temperatures estimated by a grasshopper specific biophysical model (Te) are fairly similar to but sometimes less than the body temperatures of a physical grasshopper model implanted with a thermistor (Te observed).
+The general biophysical models that makes simplifying assumptions (Te Gates) tends to overestimate body temperatures.*
 

### 2021-05-03 Manuscript_full.Rmd
Text updates.

@@
 the NicheMapR package, which includes sophisticated microclimate
-[@kearney2017nichemapr] and biophysical [@kearney2020nichemapr] models. Many of
-the approaches are similar between the packages, but TrenchR focuses on
+[@kearney2017nichemapr] and biophysical [@kearney2020nichemapr] models.
+Although the Fortran source code was recently released, the complexity of the functions can make it difficult to understand and modify the source code.
+Many of the approaches are similar between the packages, but TrenchR focuses on
 simple functions to aid the understanding and transparency of
@@
 convection and with the ground via conduction. The balance of these heat
-exchanges can be estimated and often referred to as operative
+exchanges (omitting metabolism and evaporation, which are often negligible for ectotherms) can be estimated and often referred to as operative
 environmental temperature, $T_e$ [@bakken1992measurement].
@@
 for(hr in 1:24) Qabs[hr]= Qradiation_absorbed(a=a, A=A, psa_dir=psa[hr], psa_ref=1-psa[hr], S_dir=Sdir[hr], S_dif=Sdif[hr], a_s=rho)
+#where
+
 ```
@@
 for(hr in 1:24) Qemit[hr]=Qemitted_thermal_radiation(epsilon=epsilon_s, A=A, psa_dir=psa[hr], psa_ref=1-psa[hr], T_b=Ta_liz[hr]+273.15, T_g=Ts[hr]+273.15, T_a=Ta_liz[hr]+273.15, enclosed=FALSE)
+
+#where psa_dir and psi_ref are the proportions of surface area exposed to direct and reflected solar radiation, respectively.
 ```
@@
 
-svl=0.006 #approximate snout vent length (m) for Sceloporus
+svl=0.006 #approximate snout vent length (m) for Sceloporus, used as the characteristic dimension for heat transfer
 
@@
 conditions. We implement the function in R assuming that 2/3 of the
-lizard's surface area is exchanging heat through convection. We note we
+lizard's surface area is exchanging heat through convection. We
 currently estimate no convection since we are assuming $Tb=Ta$ for
@@
 
-We assume, as in generally done for lizards, that heat exchange
+We assume, as is generally done for lizards, that heat exchange
 associated with metabolism and evaporation is negligible:
@@
 for(hr in 1:24) Te3[hr]= Tb_lizard(T_a=Ta_liz[hr], T_g=Ts[hr], u=V_liz, svl=svl*1000, m=mass, psi=psi_deg[hr], rho_S=rho, elev=elev, doy=doy, sun=TRUE, surface=TRUE, alpha_S=a, alpha_L=0.965, epsilon_s=epsilon_s, F_d=0.8, F_r=0.5, F_a=0.5, F_g=0.5)
+#where F_d, F_r, F_a, and F_g are the view factors between the surface of the lizard and diffuse and reflected solar radiation and atmospheric and ground thermal radiation, respectively. 
 ```
@@
 
-* **Figure 1**.
+**Figure 1**.
 Body temperatures (Te) are predicted to drastically exceed air temperature when lizards are exposed to high levels of solar radiation.
@@
 We estimate body temperatures using two general energy budgets [solid: Tb_Gates();
-dotted: Tb_CampbellNorman()] and a lizard specific biophysical model [dashed: Tb_lizard()] that differ in how they model heat exchanges.*
+dotted: Tb_CampbellNorman()] and a lizard specific biophysical model [dashed: Tb_lizard()] that differ in how they model heat exchanges.
 

### 2022-02-10 Manuscript_full.Rmd
Updates
abby edits and comments

@@
 title: 'TrenchR: an R package for transparent environmental and ecological biophysics'
-author: "Lauren B. Buckley, Bryan Briones Ortiz, Isaac Caruso, Aji John, Ofir Levy, Abby Meyer, Yutaro Sakairi,..."
+author: "Lauren B. Buckley, Bryan Briones Ortiz, Isaac Caruso, Aji John, Ofir Levy, Abigail Meyer, Yutaro Sakairi,..."
 output:
@@
 Lauren B. Buckley^1^, Bryan Briones Ortiz^1^, Isaac Caruso^1^, Aji
-John^1^, Ofir Levy^2^, Abby Meyer^1^, Yutaro Sakairi^1^,...
+John^1^, Ofir Levy^2^, Abby Meyer^1^, Yutaro Sakairi^1^
 
@@
 
-1.  Much understanding of responses to climate change and variability relies on the
-    assumption that organismal body temperatures are equal to temporally
-    averaged air temperatures high above the ground. However, most
-    organisms live near the ground and acute exposure to solar radiation and thermal extremes
-    can substantially elevate their body temperatures.
+1.  Much understanding of responses to climate change and variability
+    relies on the assumption that organismal body temperatures are equal
+    to temporally averaged air temperatures high above the ground.
+    However, most organisms live near the ground and acute exposure to
+    solar radiation and thermal extremes can substantially elevate their
+    body temperatures.
 2.  We introduce the TrenchR package, which aids in Translating
@@
     model and temporally partition air and soil temperatures and solar
-    radiation. We offer utility functions to aid in estimating the organismal and environonmental parameters needed for biophysical ecology.
-3.  TrenchR focuses on relatively simple and modular functions so users can create transparent and flexible biophysical models.
+    radiation. We offer utility functions to aid in estimating the
+    organismal and environonmental parameters needed for biophysical
+    ecology.
+3.  TrenchR focuses on simple and modular functions so users
+    can create transparent and flexible biophysical models.
 4.  The package aims to introduce and enable microclimate and
     biophysical modelling to improve ecological and evolutionary
-    forecasting. To further this aim, we offer a series of educational modules to introduce the field of biophysical ecology.
+    forecasting. To further this aim, we offer a series of educational
+    modules to introduce the field of biophysical ecology.
 
@@
 Many analyses assume that organismal body temperatures are equal to air
-temperate at weather station height (usually 2m). However, air
+temperatures at weather station height (usually 2 meters). However, air
 temperatures near the ground where most organisms reside can be
-considerably warmer and absorption of solar radiation can raise body
+considerably warmer, and absorption of solar radiation can raise body
 temperatures well above air temperatures, resulting in potential
-discrepancies of tens of degrees C [@sinclair2016; @sunday2014;
-@kearney2009].
-Applications are increasingly demonstrating the importance of accounting for microclimate variation [@bramer2018advances; @butikofer2020problem; @lembrechts2019comparing].
-Acute thermal stress events are often more relevant to
-the physiology and energy balance of organisms than daily, monthly, or
-annually aggregated environmental conditions. Yet, daily and seasonal
+discrepancies of tens of degrees Celsius [@sinclair2016; @sunday2014;
+@kearney2009]. Applications are increasingly demonstrating the
+importance of accounting for microclimate variation
+[@bramer2018advances; @butikofer2020problem; @lembrechts2019comparing].
+Acute thermal stress events are often more relevant to the physiology
+and energy balance of organisms than daily, monthly, or annually
+averaged (*Abby: changed from aggregated*) environmental conditions. Yet, daily and seasonal
 environmental variation is often neglected [@dillon2016introduction;
 @dillon2016life]. We aim to help remedy these shortcomings by
-introducing the TrenchR package (<https://github.com/trenchproject/TrenchR>) as a component of the TrEnCh project
-(<https:www.trenchproject.com>), which builds computational and
-visualization tools to Translate Environmental Change into organismal
-responses.
+introducing the TrenchR package
+(<https://github.com/trenchproject/TrenchR>) as a component of the
+TrEnCh project (<https:www.trenchproject.com>), which builds
+computational and visualization tools to Translate Environmental Change
+into organismal responses.
 
@@
 [@campbell2000introduction; @gates1980biophysical] and articles
-[@porter1969thermodynamic; @porter1983biophysical], but adoption of the
+[@porter1969thermodynamic; @porter1983biophysical], but adoption of these
 tools has not kept pace with research on climate change responses.
-Adoptions of biophysical approaches are expanding with recognition of
+Adoption of biophysical approaches is expanding with recognition of
 the importance of spatial and temporal environmental variation to
 determining climate change responses [@buckley2018leveraging;
-@kearney2017nichemapr]. Increasing dissemination of computational tools
-is facilitating adoption. The TrenchR package is intended to complement
+@kearney2017nichemapr] and the increased dissemination of computational tools. 
+The TrenchR package is intended to complement
 the NicheMapR package, which includes sophisticated microclimate
 [@kearney2017nichemapr] and biophysical [@kearney2020nichemapr] models.
-Although the Fortran source code was recently released, the complexity of the functions can make it difficult to understand and modify the source code.
-Many of the approaches are similar between the packages, but TrenchR focuses on
-simple functions to aid the understanding and transparency of
-biophysical approaches. The TrenchR functions allow for modular use and
-easy adaptation to a variety of organisms and research questions. Some
-energy budget calculation components of biophysical models are also
-available in the ThermImage package [@tattersall2017]. Although many of
-the TrenchR functions are general, some components of TrenchR are best
-suited for ectothermic animals. We have limited inclusion of plant
-specific biophysical models since they are the focus of several R
-packages [tealeaves: [@muir2019tea]; plantecophys:
-[@duursma2015plantecophys]].
+Although the Fortran source code was recently released, the complexity
+of NicheMapR functions can make it difficult to understand and modify the
+source code. Many of the approaches are similar between the packages,
+but TrenchR provides simple functions to aid the understanding and
+transparency of biophysical approaches. The TrenchR functions are
+modular and easy adapt to a variety of organisms and research
+questions. Although many of the TrenchR functions are general, some
+components of TrenchR are best suited for ectothermic animals.
+
+We have limited inclusion of plant specific biophysical models since
+they are the focus of several R packages [tealeaves: [@muir2019tea];
+plantecophys: [@duursma2015plantecophys]]. Some energy budget
+calculation components of biophysical models are also available in the
+ThermImage package [@tattersall2017]. Microclimate models, with an
+emphasis on describing spatial variation, are also available in the
+Microclima package [@maclean2019microclima].
 
@@
 
-The TrenchR package aims to promote transparency and reproducibility and
-we welcome contributions and corrections from users. Our package is
+The TrenchR package aims to promote transparency and reproducibility.
+We welcome contributions and corrections from users. Our package is
 built using the devtools methodology
@@
 and checked using Travis CI as a tool for continuous integration.
+
+(*Abby: maybe add the link to the reference website? trenchproject.github.io/TrenchR/*)
 
@@
 thermal radiation; *convective heat exchanges* between organisms and
-their surrounding fluid (air or water) driven by fluid flow; *conductive
-heat exchanges* between organisms and the ground due to physical
+their surrounding fluid (air or water) driven by fluid flow; 
+*conductive heat exchanges* between organisms and the ground due to physical
 contact; and *evaporative heat exchanges* associated with organisms'
@@
 of organisms with specified physical properties in a specific
-microenvironment and assume no heat exchange via metabolism or
+microclimate and assume no heat exchange via metabolism or
 evaporation [@bakken1992measurement].
@@
 
-We introduce the functions in several vignettes. A good place to start is
-the *Allometry and conversions* vignette, which provides tools for
-preparing data such as estimating additional dimensions of organisms
+We introduce the functions in several vignettes. A good place to start
+is the *Allometry and conversions* vignette, which provides tools for
+preparing data, such as estimating additional dimensions of organisms
 from measured dimensions. The *Estimating microclimates* vignette
@@
 its components, diurnal variation in temperature and radiation,
-temperature and windspeed profiles, and soil temperatures and profiles.
+temperature and wind speed profiles, and soil temperatures and profiles.
 Finally, the core biophysical modelling functions are described in a
@@
 For additional background on microclimate and biophysical modelling, we
-have updated a series of educational modules entitled *Physical
-Processes in Ecosystems*. We intend the modules to provide a more
+have updated a series of educational modules entitled 
+*Physical Processes in Ecosystems*. We intend the modules to provide a more
 contained and accessible introduction to microclimate and biophysical
@@
 (<https://bookdown.org/huckley/Physical_Processes_In_Ecosystems/>). The
-R markdown files are available in Github at
-<https://github.com/trenchproject/TrenchRmodules>.
+R markdown files are available in Github (<https://github.com/trenchproject/TrenchRmodules>).
 
 ## Implementation
+(*Abby: I might title this "Example Application" or "Demonstration of Use" or other. I would assume implementation to mean how you implemented the functions overall*)
 
 We illustrate the use of the TrenchR package by estimating an energy
-budget for a *Sceloporus* lizard on June 1, 2020, in Santa Fe, New
-Mexico, USA (35.69, -105.944, elevation: 2121m). The example is also
-incorporated in the *Using energy balances to estimate body
-temperatures* tutorial. We first estimate the energy budget with
-component functions and we then use an integrated biophysical model to
-estimate operative environmental temperatures. We start by generating
-environmental inputs. See Appendix S1 for another example examining a
+budget for a *Sceloporus* lizard on June 1, 2020 in Santa Fe, New
+Mexico, USA (35.69, -105.944, elevation: 2121 meters). The example is also
+incorporated in the *Using energy balances to estimate body temperatures* 
+vignette. We start by generating environmental inputs. Using these inputs,
+we estimate the energy budget with component functions. Finally, we use 
+an integrated biophysical model to estimate operative environmental temperatures.  
+See Appendix S1 for another example examining a
 time series of microclimate data and applying and testing the
@@
 
-We will use the energy budget to estimate body temperature,
-$T_b$, which can depart dramatically from the air temperatures due to
-energy exchange with the environment. Heat is gained from absorbing
-solar and thermal radiation and from metabolic reactions. Heat is lost
-through the organism's emission of radiation and the evaporation of
-water. The organism exchanges heat with the surrounding air or water via
-convection and with the ground via conduction. The balance of these heat
-exchanges (omitting metabolism and evaporation, which are often negligible for ectotherms) can be estimated and often referred to as operative
+We will use the energy budget to estimate body temperature, $T_b$, which
+can depart dramatically from the air temperatures due to energy exchange
+with the environment. Heat is gained from absorbed solar and thermal
+radiation and from metabolic reactions. Heat is lost through the
+organism's emission of radiation and the evaporation of water. The
+organism exchanges heat with the surrounding air or water via convection
+and with the ground via conduction. The balance of these heat exchanges
+(omitting metabolism and evaporation, which are often negligible for
+ectotherms) can be estimated and often referred to as operative
 environmental temperature, $T_e$ [@bakken1992measurement].
+(*Abby: I'm confused at where the distinction is made between Tb and Te, for example if the energy exchanges are part of Tb or Te or both.*)
 
 Let us assume the lizard is in an unshaded location where a weather
-station at standard height (2m) indicates that the daily air temperature
+station at standard height (2 meters) reports the daily air temperature
 varies from a minimum of 10°C to a maximum of 25°C and the wind speed
@@
 15°C to a maximum of 30°C. We assume that atmospheric transmissivity
-$\tau=0.7$ and albedo $\rho=0.6$
+$\tau = 0.7$ and albedo $\rho = 0.6$
 
@@
 
-We first prepare the environmental data for analysis. We will need
-to estimate hourly air and soil temperatures and radiation. We start by
+We first prepare the environmental data for analysis. We estimate 
+hourly air and soil temperatures and radiation. We start by
 estimating the day of year and the timing of sunrise and sunset:
@@
 ```{r}
-#set up input data as variables
-#latitude and longitude (degrees) and elevation (m)
-lat=35.69; lon= -105.944; elev=2121 
-#minimum and maximum of air and soil temperatures, respectively (C)
-Tmin=10; Tmax=25; Tmin_s=15; Tmax_s=30 
-#wind speed (m/s)
-V=1 
-#assumptions
-tau=0.7; rho=0.6 #atmospheric transmissivity and albedo
-Tb0=25 #initial assumption of body temperature (C)
+# Set up input data as variables
+lat = 35.69; # Latitude (degrees)
+lon = -105.944; # Longitude (degrees)
+elev = 2121 # Elevation (meters)
 
-doy= day_of_year("2020-06-01", format= "%Y-%m-%d") 
-snoon= solar_noon(lon=lon, doy=doy) #estimate solar noon
-dayl= daylength(lat=lat, doy=doy) #estimate daylength
-tr= snoon-dayl/2 #time of sunrise
-ts= snoon+dayl/2 #time of sunset
+Tmin = 10; # Minimum air temperature (C)
+Tmax = 25; # Maximum air temperature (C)
+Tmin_s = 15; # Minimum soil temperature (C)
+Tmax_s = 30; # Maximum soil temperature (C)
+V = 1 # Wind speed (m/s)
+
+# Assumptions
+tau = 0.7; # Atmospheric transmissivity
+rho = 0.6; # Albedo
+Tb0 = 25; # Initial assumption of body temperature (C)
+
+doy = day_of_year("2020-06-01", format= "%Y-%m-%d") # Find Julian Date (DOY)
+snoon = solar_noon(lon = lon, doy = doy) # Estimate solar noon
+dayl = daylength(lat = lat, doy = doy) # Estimate day length
+tr = snoon - dayl / 2 # Time of sunrise
+ts = snoon + dayl / 2 # Time of sunset
 ```
@@
 ```{r}
-#estimate zenith angle (degrees)
-psi_deg= sapply(1:24, FUN=zenith_angle, doy=doy, lat=lat, lon=lon) 
-#convert to radians
-psi_rad= degree_to_radian(psi_deg) 
-#estimate radiation
-Srad= sapply(psi_rad, FUN=estimate_radiation, doy=doy, tau=tau, elev=elev, rho=rho)
-#Separate into direct, diffuse, and reflected solar radiation
-Sdir= Srad[1,]; Sdif= Srad[2,]; Sref= Srad[3,]
+# Estimate zenith angle (degrees)
+psi_deg = sapply(1:24, FUN = zenith_angle, doy = doy, lat = lat, lon = lon) 
+# Convert to radians
+psi_rad = degree_to_radian(psi_deg) 
+# Estimate radiation
+Srad = sapply(psi_rad, FUN = estimate_radiation, doy = doy, tau = tau, elev = elev, rho = rho)
+
+# Separate solar radiation into direct, diffuse, and reflected
+Sdir = Srad[1,]; # Direct solar radiation
+Sdif = Srad[2,]; # Diffuse solar radiation
+Sref = Srad[3,]; # Reflected solar radiation
 ```
 
-We can then calculate hourly air and soil temperatures based on daily
+We then calculate hourly air and soil temperatures based on daily
 minimum and maximum temperatures. We select the sine-exponential model
@@
 ```{r}
-#air temperature (C)
-Ta= sapply(1:24, diurnal_temp_variation_sineexp, T_max=Tmax, T_min=Tmin, t_r=tr, t_s=ts, alpha=2.59, beta= 1.55, gamma=2.2) 
-#soil temperature (C)
-Ts= sapply(1:24, diurnal_temp_variation_sine, T_max=Tmax_s, T_min=Tmin_s) 
+# Air temperature (˚C)
+Ta = sapply(1:24, diurnal_temp_variation_sineexp, T_max = Tmax, T_min = Tmin, t_r = tr, t_s = ts, alpha = 2.59, beta = 1.55, gamma = 2.2) 
+# Soil temperature (˚C)
+Ts = sapply(1:24, diurnal_temp_variation_sine, T_max = Tmax_s, T_min = Tmin_s) 
 ```
@@
 We use microclimate models to scale air temperature and wind speed from
-weather station height (2m) to lizard height (0.02m). We assume a
-surface roughness of $z_0=0.2$, which corresponds to bare sand and
-determines the turbulence of airflow. We implement neutral air
-temperature and wind speed profiles as well as an air temperature
-profile forced by wind speed.
+weather station height (2 meters) to lizard height (0.02 meters). We assume a
+surface roughness of $z_0 = 0.2$ (*Abby: units?*), which corresponds to bare sand and
+determines the turbulence of airflow. We implement a neutral air
+temperature profile, an air temperature
+profile forced by wind speed, and a neutral wind speed profile.
 
 ```{r}
-#scale air temperature to lizard height
-#z is reference height, z is organism height, z0 is surface roughness
-Ta_liz= air_temp_profile_neutral(T_r=Ta, zr=2, z0=0.2, z=0.02, T_s=Ts)
-Ta_liz2= air_temp_profile(T_r=Ta, u_r=V, zr=2, z0=0.2, z=0.02, T_s=Ts)
+# Scale air temperature to lizard height
+# T_r = reference temperature; zr = reference height; z0 is surface roughness; z = organism height; T_s = surface temperature
+Ta_liz = air_temp_profile_neutral(T_r = Ta, zr = 2, z0 = 0.2, z = 0.02, T_s = Ts) # Neutral air temperature profile
+Ta_liz2 = air_temp_profile(T_r = Ta, u_r = V, zr = 2, z0 = 0.2, z = 0.02, T_s = Ts) # Air temperature profile forced by wind speed
 
-#scale wind speed to lizard height
-V_liz= wind_speed_profile_neutral(u_r=V, zr=2, z0=0.2,z=0.02)
+# Scale wind speed to lizard height
+# u_r = reference wind speed; zr = reference height; z0 is surface roughness; z = organism height
+V_liz= wind_speed_profile_neutral(u_r = V, zr = 2, z0 = 0.2 , z = 0.02) # Neutral wind speed profile
 ```
@@
 $Q_{abs}$ requires the surface area exposed to radiation and the solar
-absorptivity of the animal surface ($a$ proportion). We model a 10g
-*Sceloporus* lizard with solar absorptivity $a=0.9$
-[@gates1980biophysical]. We will initially assume $T_b=T_a$ to
+absorptivity of the animal surface ($a$ proportion). We model a 10 gram
+*Sceloporus* lizard with solar absorptivity $a = 0.9$
+[@gates1980biophysical]. We will initially assume $T_b = T_a$ to
 illustrate the calculations before solving for $T_b$ given the
@@
 ```{r}
-mass=10 #mass (g)
-a=0.9 #solar absorptivity (%)
+mass = 10 # Mass (g)
+a = 0.9 # Solar absorptivity (proportion)
 
-#estimate surface area (m^2) from mass (g)
-A= sa_from_mass(mass, "lizard")
-#estimate projected (silhouette) area as a portion of surface area
-psa= sapply(psi_deg, prop_silhouette_area, taxa= "lizard", posture= "prostrate")
-#change negative values to zero
-psa[psa<0]=0
+# Estimate surface area and sihouette area
+A = sa_from_mass(mass, "lizard") # surface area (m^2)
+psa = sapply(psi_deg, prop_silhouette_area, taxa = "lizard", posture = "prostrate") # Projected silhouette area (m^2)
+
+# Change negative values to zero
+psa[psa < 0] = 0
 ```
@@
 ```{r}
-Qabs= rep(NA, 24)
-for(hr in 1:24) Qabs[hr]= Qradiation_absorbed(a=a, A=A, psa_dir=psa[hr], psa_ref=1-psa[hr], S_dir=Sdir[hr], S_dif=Sdif[hr], a_s=rho)
-#where
+Qabs = rep(NA, 24)
+for(hour in 1:24) {
 
+  Qabs[hour] = Qradiation_absorbed(a = a, A = A, psa_dir = psa[hour], psa_ref = 1 - psa[hour], S_dir = Sdir[hour], S_dif = Sdif[hour], a_s = rho)
+
+}
 ```
@@
 We estimate thermal radiation $Q_{emit}$ (W) for the lizard outdoors. We
-assume the surface emissivity of lizards, $epsilon_s=0.965$
+assume the surface emissivity of lizards, $epsilon_s = 0.965$
 [@barlett1967].
@@
 ```{r}
-epsilon_s=0.965
+epsilon_s = 0.965 # Surface emissivity of lizards
 
-Qemit= rep(NA, 24)
-for(hr in 1:24) Qemit[hr]=Qemitted_thermal_radiation(epsilon=epsilon_s, A=A, psa_dir=psa[hr], psa_ref=1-psa[hr], T_b=Ta_liz[hr]+273.15, T_g=Ts[hr]+273.15, T_a=Ta_liz[hr]+273.15, enclosed=FALSE)
+Qemit = rep(NA, 24)
+for(hour in 1:24) {
 
-#where psa_dir and psi_ref are the proportions of surface area exposed to direct and reflected solar radiation, respectively.
+  # psa_dir = proportions of surface area exposed to direct solar radiation
+  # psi_ref = proportions of surface area exposed to reflected solar radiation
+  # A = surface area (m^2)
+  Qemit[hour] = Qemitted_thermal_radiation(epsilon = epsilon_s, A = A, psa_dir = psa[hour], psa_ref = 1-psa[hour], T_b = Ta_liz[hour] + 273.15, T_g = Ts[hour] + 273.15, T_a = Ta_liz[hour] + 273.15, enclosed = FALSE)
+
+}
 ```
@@
 ```{r}
-#Use DRYAIR from NicheMapR to estimate the thermal conductivity of air and kinematic viscosity
-ap= airpressure_from_elev(elev)*1000 #estimate Barometric pressure (pascal)
+ap = airpressure_from_elev(elev) * 1000 # Barometric pressure (pascal)
 
-DRYAIRout= DRYAIR(db=Ta, bp=ap, alt=elev)
-K= DRYAIRout$thcond #thermal conductivity (Wm^-2K^-1) 
-nu= DRYAIRout$viskin #kinematic viscosity (m2 s-1)
+# Use DRYAIR from NicheMapR to estimate the thermal conductivity of air and kinematic viscosity
+DRYAIRout = DRYAIR(db = Ta, bp = ap, alt = elev)
+K = DRYAIRout$thcond # Thermal conductivity (Wm^-2K^-1) 
+nu = DRYAIRout$viskin # Kinematic viscosity (m2 s-1)
 
-svl=0.006 #approximate snout vent length (m) for Sceloporus, used as the characteristic dimension for heat transfer
+svl = 0.006 # Approximate snout vent length (meters) for Sceloporus, used as the characteristic dimension for heat transfer
 
-#We will use the average of K and nu across the day for simplicity and since there is not a substantial diurnal variation.
-K= mean(K)
-nu=mean(nu)
+# We will use the average of K and nu across the day for simplicity and since there is not a substantial diurnal variation.
+K = mean(K) # Average thermal conductivity across the day
+nu = mean(nu) # Average kinematic viscosity
 
-#Estimate the heat transfer coefficient using an empirical relationship for lizards
-H_L=heat_transfer_coefficient(V=V_liz,D=svl,K= 25.7 * 10^(-3), nu= 15.3 * 10^(-6) , taxa="lizard_surface")
+# Estimate the heat transfer coefficient using an empirical relationship for lizards
+H_L = heat_transfer_coefficient(V = V_liz, D = svl, K = 25.7 * 10^(-3), nu = 15.3 * 10^(-6), taxa = "lizard_surface")
 
-#Also illustrate estimations using a spherical approximation and a simplified version of the approximation.
-H_L2= heat_transfer_coefficient_approximation(V=V_liz,D=svl,K= 25.7 * 10^(-3), nu= 15.3 * 10^(-6), taxa="lizard")
-H_L3= heat_transfer_coefficient_simple(V=V_liz,D=svl)
+# Estimate the heat transfer coefficient using a spherical approximation
+H_L2= heat_transfer_coefficient_approximation(V = V_liz, D = svl, K = 25.7 * 10^(-3), nu = 15.3 * 10^(-6), taxa = "lizard")
+
+# Estimate the heat transfer coefficient using a simplified version of the approximation
+H_L3= heat_transfer_coefficient_simple(V = V_liz, D = svl, type = "Gates")
 ```
+*Abby: Added type=Gates above to fix error in H_L3, change if Gates is not the equation wanted. Also, maybe give a reason why one might want to use a spherical or simplified version*
 
@@
 air using the following relationship:
-$$Q_{conv}= ef \cdot H_L(A\cdot \mbox{proportion})(T_a-T_b),$$ where an
+$$Q_{conv} = ef \cdot H_L(A\cdot \mbox{proportion})(T_a-T_b),$$ where an
 enhancement factor, $ef$, multiplier can be incorporated to account for
@@
 lizard's surface area is exchanging heat through convection. We
-currently estimate no convection since we are assuming $Tb=Ta$ for
+currently estimate no convection since we are assuming $Tb = Ta$ for
 illustration purposes:
@@
 ```{r}
-Qconv= rep(NA, 24)
-for(hr in 1:24) Qconv[hr]= Qconvection(T_a= Ta_liz[hr]+273.15, T_b= Ta_liz[hr]+273.15, H=H_L, A=A, proportion=0.67, ef=1.3)
+Qconv = rep(NA, 24)
+for(hour in 1:24) {
+
+  Qconv[hour] = Qconvection(T_a = Ta_liz[hour] + 273.15, T_b = Ta_liz[hour] + 273.15, H = H_L, A = A, proportion = 0.67, ef = 1.3)
+
+}
 ```
@@
 assuming conductance through the animal tissue is the rate limiting step
-as follows: $$Q_{cond}= A \cdot \mbox{proportion} \cdot K(T_g-T_b)/d. $$
+as follows: $$Q_{cond} = A \cdot \mbox{proportion} \cdot K(T_g-T_b)/d. $$
 We implement the estimate in R assuming that 1/3 of the lizard surface
-is in contact with the ground and a skin thickness $d$ of $10^{-6}$:
+is in contact with the ground and a skin thickness $d = 10^{-6}$:
 
 ```{r}
-Qcond= rep(NA, 24)
-for(hr in 1:24) Qcond[hr]=Qconduction_animal(T_g= Ts[hr]+273.15, T_b=Ta_liz[hr]+273.15, d=10^-6, K=K, A=A, proportion=0.33)
+Qcond = rep(NA, 24)
+for(hr in 1:24) {
+
+  Qcond[hr] = Qconduction_animal(T_g = Ts[hr] + 273.15, T_b = Ta_liz[hr] + 273.15, d = 10^-6, K = K, A = A, proportion = 0.33)
+
+}
 ```
@@
 ```{r}
-Qmet= 0; Qevap= 0
+Qmet = 0; 
+Qevap= 0;
 ```
@@
 ```{r}
-Qnet= Qnet_Gates(Qabs=Qabs, Qemit=Qemit, Qconv=Qconv, Qcond=Qcond, Qmet=Qmet, Qevap=Qevap)
+Qnet = Qnet_Gates(Qabs = Qabs, Qemit = Qemit, Qconv = Qconv, Qcond = Qcond, Qmet = Qmet, Qevap = Qevap)
 ```
 
-We constructed the energy balance assuming $T_b=T_a$ for illustration.
+We constructed the energy balance assuming $T_b = T_a$ for illustration.
 Thus, the energy balance is dominated by heat gain from contact with the
@@
 ```{r}
-Te= rep(NA, 24)
-for(hr in 1:24) Te[hr]= Tb_Gates(A=A, D=svl, psa_dir=psa[hr], psa_ref=1-psa[hr], psa_air=0.67, psa_g=0.33, T_g=Ts[hr]+273.15, T_a=Ta_liz[hr]+273.15, Qabs=Qabs[hr], epsilon=epsilon_s, H_L=H_L, ef=1.3, K=K)
+Te = rep(NA, 24)
+for(hour in 1:24) {
+
+  Te[hour]= Tb_Gates(A = A, D = svl, psa_dir = psa[hour], psa_ref = 1 - psa[hour], psa_air = 0.67, psa_g = 0.33, T_g = Ts[hour] + 273.15, T_a = Ta_liz[hour] + 273.15, Qabs = Qabs[hour], epsilon = epsilon_s, H_L = H_L, ef = 1.3, K = K)
+
+}
 ```
@@
 ```{r}
-Te2= rep(NA, 24)
+Te2 = rep(NA, 24)
+for(hr in 1:24) {
 
-for(hr in 1:24) Te2[hr]= Tb_CampbellNorman(T_a=Ta_liz[hr]+273.15, T_g=Ts[hr]+273.15, S=Qabs[hr]/A, alpha_L=0.96, epsilon=epsilon_s, c_p=29.3, D=svl, V=V_liz)
-#S is solar radiation flux (W m^-2), so we divide by surface area, A
+  # S is solar radiation flux (W m^-2), so we divide by surface area, A
+  Te2[hr] = Tb_CampbellNorman(T_a = Ta_liz[hr] + 273.15, T_g = Ts[hr] + 273.15, S = Qabs[hr] / A, alpha_L = 0.96, epsilon = epsilon_s, c_p = 29.3, D = svl, V = V_liz)
+
+}
+
 ```
@@
 ```{r}
-Te3= rep(NA, 24)
-for(hr in 1:24) Te3[hr]= Tb_lizard(T_a=Ta_liz[hr], T_g=Ts[hr], u=V_liz, svl=svl*1000, m=mass, psi=psi_deg[hr], rho_S=rho, elev=elev, doy=doy, sun=TRUE, surface=TRUE, alpha_S=a, alpha_L=0.965, epsilon_s=epsilon_s, F_d=0.8, F_r=0.5, F_a=0.5, F_g=0.5)
-#where F_d, F_r, F_a, and F_g are the view factors between the surface of the lizard and diffuse and reflected solar radiation and atmospheric and ground thermal radiation, respectively. 
+Te3 = rep(NA, 24)
+for(hour in 1:24) {
+
+  # F_d, F_r, F_a, and F_g are the view factors between the surface of the lizard and diffuse solar radiation, reflected solar radiation, and atmospheric thermal radiation, and ground thermal radiation, respectively.
+  Te3[hour]= Tb_lizard(T_a = Ta_liz[hour], T_g = Ts[hour], u = V_liz, svl = svl * 1000, m = mass, psi = psi_deg[hour], rho_S = rho, elev = elev, doy = doy, sun = TRUE, surface = TRUE, alpha_S = a, alpha_L = 0.965, epsilon_s = epsilon_s, F_d = 0.8, F_r = 0.5, F_a = 0.5, F_g = 0.5)
+
+}
 ```
@@
 par(mar=c(5, 4, 4, 8) + 0.1)
-plot(1:24, (Te-273.15), type="l", xlab="Hour", ylab="Temperature (°C)", col="#5DC863FF", ylim=c(10,60)) #Gates
-points(1:24, Te2-273.15, type="l", col="#5DC863FF", lty="dotted") #Campbell and Norman
-points(1:24, Te3, type="l", col="#5DC863FF", lty="dashed") #Lizard, Buckley 2008
+plot(1:24, (Te-273.15), type="l", xlab="Hour", ylab="Temperature (°C)", col="#5DC863FF", ylim=c(10,60)) # Gates
+points(1:24, Te2-273.15, type="l", col="#5DC863FF", lty="dotted") # Campbell and Norman
+points(1:24, Te3, type="l", col="#5DC863FF", lty="dashed") # Lizard, Buckley 2008
 points(1:24, Ta, type="l", col="#440154FF")
@@
 legend("topleft",
-       legend=c("Ta 2m", "Ta 0.02m", "Ts", "Te", "Radiation"),
+       legend=c("Ta (2m)", "Ta (0.02m)", "Ts", "Te", "Radiation"),
        lty=1, pch=NA, col=c("#440154FF", "#3B528BFF", "#21908CFF", "#5DC863FF", "#FDE725FF"))
@@
 
-**Figure 1**.
-Body temperatures (Te) are predicted to drastically exceed air temperature when lizards are exposed to high levels of solar radiation.
-Air temperatures (Ta, °C) at lizard height (0.02m) are predicted to exceed air temperatures at 2m and to be similar to surface temperatures (Ts).
-We estimate body temperatures using two general energy budgets [solid: Tb_Gates();
-dotted: Tb_CampbellNorman()] and a lizard specific biophysical model [dashed: Tb_lizard()] that differ in how they model heat exchanges.
+**Figure 1**. Body temperatures (Te) are predicted to drastically exceed
+air temperature when lizards are exposed to high levels of solar
+radiation. Air temperatures (Ta, °C) at lizard height (0.02m) are
+predicted to exceed air temperatures at 2m and to be similar to surface
+temperatures (Ts). We estimate body temperatures using two general
+energy budgets [solid: Tb_Gates(); dotted: Tb_CampbellNorman()] and a
+lizard specific biophysical model [dashed: Tb_lizard()] that differ in
+how they model heat exchanges.
 
@@
 referred to NicheMapR for more detailed biophysical models
-[@kearney2017nichemapr; @kearney2020nichemapr].
-We focus on models that predict equilibrium conditions for simplicity.
-Such models generally do not present computational challenges so our models are not optimized for computational efficiency.
-Making classic
-biophysical ecology techniques more accessible will allow researchers to
-take advantage of rapidly accumulating data on environmental conditions
-and organismal traits to understand and predict ecological and
-evolutionary responses. Considering how organisms experience their environment is central to understanding responses to variable and changing environments [@suggitt2018extinction].
+[@kearney2017nichemapr; @kearney2020nichemapr]. We focus on models that
+predict equilibrium conditions for simplicity. Such models generally do
+not present computational challenges so our models are not optimized for
+computational efficiency. Making classic biophysical ecology techniques
+more accessible will allow researchers to take advantage of rapidly
+accumulating data on environmental conditions and organismal traits to
+understand and predict ecological and evolutionary responses.
+Considering how organisms experience their environment is central to
+understanding responses to variable and changing environments
+[@suggitt2018extinction].
 

### 2022-03-28 ManuscriptTe.Rmd
MS fixes

@@
 H_L1= heat_transfer_coefficient_approximation(V=dat$Ugrass,D=0.0211,K= 0.025, nu= 15.3 * 10^(-6), taxa="flyinginsect")
-H_L2= heat_transfer_coefficient_simple(V=dat$Ugrass,D=0.0211)
+H_L2= heat_transfer_coefficient_simple(V=dat$Ugrass,D=0.0211, type="Gates")
 
 #surface area from length(m)
-A= sa_from_length(l=0.0211)
+A= surface_area_from_length(L=0.0211)
 ```

### 2022-03-28 Manuscript_full.Rmd
name edit
MS fixes

@@
 title: 'TrenchR: an R package for transparent environmental and ecological biophysics'
-author: "Lauren B. Buckley, Bryan Briones Ortiz, Isaac Caruso, Aji John, Ofir Levy, Abigail Meyer, Yutaro Sakairi,..."
+author: "Lauren B. Buckley, Bryan A. Briones Ortiz, Isaac Caruso, Aji John, Ofir Levy, Abigail V. Meyer, Eric A. Riddell, and Yutaro Sakairi"
 output:
@@
 
-Lauren B. Buckley^1^, Bryan Briones Ortiz^1^, Isaac Caruso^1^, Aji
-John^1^, Ofir Levy^2^, Abby Meyer^1^, Yutaro Sakairi^1^
+Lauren B. Buckley^1^, Bryan A. Briones Ortiz^1^, Isaac Caruso^1^, Aji
+John^1^, Ofir Levy^2^, Abigail V. Meyer^1^, Eric A. Riddell^3^, Yutaro Sakairi^1^
 
@@
     98195-1800, USA
-2.
-
+2.  School of Zoology, Tel Aviv University, Tel Aviv, Israel
+3.  Department of Ecology, Evolution, and Organismal Biology, Iowa State University, 
+Ames, IA 50011-4009, USA
 Corresponding Author: Lauren Buckley,
@@
     However, most organisms live near the ground and acute exposure to
-    solar radiation and thermal extremes can substantially elevate their
+    solar and thermal radiation and thermal extremes can substantially elevate their
     body temperatures.
@@
 3.  TrenchR focuses on simple and modular functions so users
-    can create transparent and flexible biophysical models.
+    can create transparent and flexible models for biophysical applications.
 4.  The package aims to introduce and enable microclimate and
     biophysical modelling to improve ecological and evolutionary
-    forecasting. To further this aim, we offer a series of educational
+    forecasting. We further this aim through a series of educational
     modules to introduce the field of biophysical ecology.
@@
 
-Many responses of organisms and ecosystems to climate change are
+Responses of organisms and ecosystems to climate change are heterogeneous and thus
 inconsistent with current predictive models [@maguire2015modeling]. Some
@@
 temperatures near the ground where most organisms reside can be
-considerably warmer, and absorption of solar radiation can raise body
+considerably warmer, and absorption of solar and thermal radiation can raise body
 temperatures well above air temperatures, resulting in potential
@@
 and energy balance of organisms than daily, monthly, or annually
-averaged (*Abby: changed from aggregated*) environmental conditions. Yet, daily and seasonal
+averaged environmental conditions. Yet, daily and seasonal
 environmental variation is often neglected [@dillon2016introduction;
@@
 @kearney2017nichemapr] and the increased dissemination of computational tools.
+
 The TrenchR package is intended to complement
@@
 questions. Although many of the TrenchR functions are general, some
-components of TrenchR are best suited for ectothermic animals.
-
-We have limited inclusion of plant specific biophysical models since
+components of TrenchR are best suited for ectothermic animals. We have limited inclusion of plant specific biophysical models since
 they are the focus of several R packages [tealeaves: [@muir2019tea];
@@
 
-The TrenchR package aims to promote transparency and reproducibility.
+The TrenchR package (<http://trenchproject.github.io/TrenchR/>) aims to promote transparency and reproducibility.
 We welcome contributions and corrections from users. Our package is
@@
 requests can be contributed in Github
-(<https://github.com/trenchproject/TrenchR/>). The package is maintained
-and checked using Travis CI as a tool for continuous integration.
-
-(*Abby: maybe add the link to the reference website? trenchproject.github.io/TrenchR/*)
+(<https://github.com/trenchproject/TrenchR/>).
 
@@
 the basis of microclimate and biophysical models (e.g., zenith and
-declination angles), facilitate modeling diurnal temperature variation,
+declination angles, which describe the angles of incident sunlight), facilitate modeling diurnal temperature variation,
 and estimate biologically relevant aggregate metrics such as degree days
@@
 build custom biophysical models. The functions allow implementing an
-energy balance of the following components of heat exchange between
+energy balance including the following components of heat exchange between
 organisms and their environment: *radiative heat exchanges* of solar and
@@
 their surrounding fluid (air or water) driven by fluid flow; 
-*conductive heat exchanges* between organisms and the ground due to physical
+*conductive heat exchanges* between organisms and solid surfaces (generally the ground) due to physical
 contact; and *evaporative heat exchanges* associated with organisms'
@@
 functions and then body temperatures can be solved for using either a
-generic energy balance or taxa specific biophysical models.
+generic energy balance or taxa specific biophysical models. We additionally offer a *List of symbols* used in equations.
 
@@
 
-## Implementation
-(*Abby: I might title this "Example Application" or "Demonstration of Use" or other. I would assume implementation to mean how you implemented the functions overall*)
+## Example Application
 
@@
 ectotherms) can be estimated and often referred to as operative
-environmental temperature, $T_e$ [@bakken1992measurement].
-(*Abby: I'm confused at where the distinction is made between Tb and Te, for example if the energy exchanges are part of Tb or Te or both.*)
+environmental temperature, $T_e$ [@bakken1992measurement]. $T_e$ is an estimate of $T_b$ and the package functions refer to $T_b$ for simplicity.
 
 Let us assume the lizard is in an unshaded location where a weather
-station at standard height (2 meters) reports the daily air temperature
+station at standard height (2 meters) reports that the daily air temperature
 varies from a minimum of 10°C to a maximum of 25°C and the wind speed
@@
 
-We first prepare the environmental data for analysis. We estimate 
-hourly air and soil temperatures and radiation. We start by
+We first prepare the environmental data for analysis. We will estimate 
+hourly air and soil temperatures and radiation usinf a function describing diurnal temperature variation. We start by
 estimating the day of year and the timing of sunrise and sunset:
@@
 estimate hourly solar radiation by discounting incoming solar radiation
-as it moves through the atmosphere as follows:
+as it moves through the atmosphere as follows. The function partitions radiation into direct, diffuse, and reflected components:
 
@@
 We use microclimate models to scale air temperature and wind speed from
-weather station height (2 meters) to lizard height (0.02 meters). We assume a
-surface roughness of $z_0 = 0.2$ (*Abby: units?*), which corresponds to bare sand and
+weather station height (2m) to lizard height (0.02m). We assume a
+surface roughness of $z_0 = 0.2$ m, which corresponds to bare sand and
 determines the turbulence of airflow. We implement a neutral air
@@
 # Estimate surface area and sihouette area
-A = sa_from_mass(mass, "lizard") # surface area (m^2)
-psa = sapply(psi_deg, prop_silhouette_area, taxa = "lizard", posture = "prostrate") # Projected silhouette area (m^2)
+A = surface_area_from_mass(mass, "lizard") # surface area (m^2)
+psa = sapply(psi_deg, proportion_silhouette_area, taxon = "lizard", posture = "prostrate") # Projected silhouette area (m^2)
 
@@
 ($Wm^{-2}K^{-1}$) using an empirical relationship for lizards
-(`heat_transfer_coefficient()`), but also illustrate a function
+(`heat_transfer_coefficient()`). We also illustrate a function
 estimating $H_L$ using a spherical approximation
 (`heat_transfer_coefficient_approximation()`) and a simplified
-approximation (`heat_transfer_coefficient_simple()`).
+approximation (`heat_transfer_coefficient_simple()`) for cases when taxon specific relationships for estimating heat transfer coefficients are not available.
 
@@
 ```
-*Abby: Added type=Gates above to fix error in H_L3, change if Gates is not the equation wanted. Also, maybe give a reason why one might want to use a spherical or simplified version*
 
@@
 
-We additionally estimate $T_e$ using a specialized function for lizards
+We additionally estimate $T_b$ using a specialized function for lizards
 [@buckley2008link]:
@@
 ## Acknowledgements
-
-We thank Brian Helmuth and Eric Riddell for providing and checking
-functions. We thank Ray Huey, Joel Kingsolver, Luke Miller, Warren
-Porter, and Art Woods for assistance and input. We particularly thank
+We thank Ray Huey, Joel Kingsolver, Luke Miller, Warren
+Porter, and Art Woods for assistance and input and Brian Helmuth for providing functions. We particularly thank
 Michael Kearney for extensive input on TrenchR and its alignment with
 NicheMapR and for contributing two educational modules.
-
-[Check with Eric]
 
@@
 LB conceived and designed the package and lead all aspect of the
-packages creation and dissemination; BBO, IC, AJ, OL, AM, YS wrote and
+packages creation and dissemination; BBO, IC, AJ, OL, AM, ER, YS wrote and
 edited functions and associated package resources; AJ configured the

### 2022-05-06 ManuscriptTe.Rmd
Ms updates

--- 2021-05-03 ManuscriptTe.Rmd
@@
 ##Calculate operative temperature, Te
-dat$Te<-Tb_grasshopper(T_a=dat$Tgrass, T_g=dat$SoilTemp, u=dat$Ugrass, H=dat$Rad, K_t=1, psi=dat$psi, L=0.0211, Acondfact=0.0, abs=0.9, r_g=0.5)
+dat$Te<-Tb_grasshopper(T_a=dat$Tgrass, T_g=dat$SoilTemp, u=dat$Ugrass, H=dat$Rad, K_t=1, psi=dat$psi, l=0.0211, Acondfact=0.0, abs=0.9, r_g=0.5)
 
@@
 #K_t	is the clearness index (dimensionless), which is the ratio of the global solar radiation measured at the surface to the total solar radiation at the top of the atmosphere;
-#psi is solar zenith angle (degrees); L	is grasshopper length (m);
+#psi is solar zenith angle (degrees); l	is grasshopper length (m);
 #Acondfact is the proportion of the grasshopper surface area in contact with the ground;
@@
 #Estimate heat transfer coefficient using empirical relationship
-dat$H_L=heat_transfer_coefficient(V=dat$Ugrass,D=0.0211,K=0.025, nu= 15.3 * 10^(-6) , taxa="cylinder")
+dat$H_L=heat_transfer_coefficient(V=dat$Ugrass,D=0.0211,K=0.025, nu= 15.3 * 10^(-6) , taxon="cylinder")
 #where V is	air velocity (m/s); D is the characteristic dimension (m);
@@
 #Also estimate using a spherical approximation and a simplified version of the approximation.
-H_L1= heat_transfer_coefficient_approximation(V=dat$Ugrass,D=0.0211,K= 0.025, nu= 15.3 * 10^(-6), taxa="flyinginsect")
-H_L2= heat_transfer_coefficient_simple(V=dat$Ugrass,D=0.0211)
+H_L1= heat_transfer_coefficient_approximation(V=dat$Ugrass,D=0.0211,K= 0.025, nu= 15.3 * 10^(-6), taxon="flyinginsect")
+H_L2= heat_transfer_coefficient_simple(V=dat$Ugrass,D=0.0211, type="Gates")
 
 #surface area from length(m)
-A= sa_from_length(l=0.0211)
+A= surface_area_from_length(l=0.0211)
 ```

### 2022-05-06 Manuscript_full.Rmd
Ms updates

@@
 title: 'TrenchR: an R package for transparent environmental and ecological biophysics'
-author: "Lauren B. Buckley, Bryan A. Briones Ortiz, Isaac Caruso, Aji John, Ofir Levy, Abigail V. Meyer, Eric A. Riddell, and Yutaro Sakairi"
+author: "Lauren B. Buckley, Bryan A. Briones Ortiz, Isaac Caruso, Aji John, Ofir Levy, Abigail V. Meyer, Eric A. Riddell, Yutaro Sakairi, and Juniper Simonis"
 output:
@@
 Lauren B. Buckley^1^, Bryan A. Briones Ortiz^1^, Isaac Caruso^1^, Aji
-John^1^, Ofir Levy^2^, Abigail V. Meyer^1^, Eric A. Riddell^3^, Yutaro Sakairi^1^
+John^1^, Ofir Levy^2^, Abigail V. Meyer^1^, Eric A. Riddell^3^, Yutaro
+Sakairi^1^, and Juniper L. Simonis^4^
 
@@
 2.  School of Zoology, Tel Aviv University, Tel Aviv, Israel
-3.  Department of Ecology, Evolution, and Organismal Biology, Iowa State University, 
-Ames, IA 50011-4009, USA
+3.  Department of Ecology, Evolution, and Organismal Biology, Iowa State
+    University, Ames, IA 50011-4009, USA
+4.  DAPPER Stats, 3519 NE 15th Avenue, Suite 467, Portland, Oregon
+    97212, USA
+
 Corresponding Author: Lauren Buckley,
@@
 
-1.  Much understanding of responses to climate change and variability
-    relies on the assumption that organismal body temperatures are equal
-    to temporally averaged air temperatures high above the ground.
+1.  Much understanding of organimsal responses to climate change and
+    variability relies on the assumption that body temperatures are
+    equal to temporally averaged air temperatures high above the ground.
     However, most organisms live near the ground and acute exposure to
-    solar and thermal radiation and thermal extremes can substantially elevate their
-    body temperatures.
+    solar and thermal radiation and thermal extremes can substantially
+    elevate their body temperatures.
 2.  We introduce the TrenchR package, which aids in Translating
@@
     ecology.
-3.  TrenchR focuses on simple and modular functions so users
-    can create transparent and flexible models for biophysical applications.
+3.  TrenchR focuses on simple and modular functions so users can create
+    transparent and flexible models for biophysical applications.
 4.  The package aims to introduce and enable microclimate and
@@
 
-Responses of organisms and ecosystems to climate change are heterogeneous and thus
-inconsistent with current predictive models [@maguire2015modeling]. Some
-of these predictive shortcomings stem from omitting spatial and temporal
-environmental variation and how it interacts with organismal phenotypes
-(e.g., size, coloration) [@nadeau2017coarse; @potter2013microclimatic].
-Many analyses assume that organismal body temperatures are equal to air
-temperatures at weather station height (usually 2 meters). However, air
-temperatures near the ground where most organisms reside can be
-considerably warmer, and absorption of solar and thermal radiation can raise body
-temperatures well above air temperatures, resulting in potential
-discrepancies of tens of degrees Celsius [@sinclair2016; @sunday2014;
-@kearney2009]. Applications are increasingly demonstrating the
-importance of accounting for microclimate variation
-[@bramer2018advances; @butikofer2020problem; @lembrechts2019comparing].
-Acute thermal stress events are often more relevant to the physiology
-and energy balance of organisms than daily, monthly, or annually
-averaged environmental conditions. Yet, daily and seasonal
-environmental variation is often neglected [@dillon2016introduction;
-@dillon2016life]. We aim to help remedy these shortcomings by
-introducing the TrenchR package
+Responses of organisms and ecosystems to climate change are
+heterogeneous and thus inconsistent with current predictive models
+[@maguire2015modeling]. Some of these predictive shortcomings stem from
+omitting spatial and temporal environmental variation and how it
+interacts with organismal phenotypes (e.g., size, coloration)
+[@nadeau2017coarse; @potter2013microclimatic]. Many analyses assume that
+organismal body temperatures are equal to air temperatures at weather
+station height (usually 2 meters). However, air temperatures near the
+ground where most organisms reside are often considerably warmer, and
+absorption of solar and thermal radiation can raise body temperatures
+well above air temperatures, resulting in potential discrepancies of
+tens of degrees Celsius [@sinclair2016; @sunday2014; @kearney2009].
+Applications are increasingly demonstrating the importance of accounting
+for microclimate variation [@bramer2018advances; @butikofer2020problem;
+@lembrechts2019comparing]. Acute thermal stress events are often more
+relevant to the physiology and energy balance of organisms than daily,
+monthly, or annually averaged environmental conditions. Yet, daily and
+seasonal environmental variation is often neglected
+[@dillon2016introduction; @dillon2016life]. We aim to help remedy these
+shortcomings by introducing the TrenchR package
 (<https://github.com/trenchproject/TrenchR>) as a component of the
@@
 [@campbell2000introduction; @gates1980biophysical] and articles
-[@porter1969thermodynamic; @porter1983biophysical], but adoption of these
-tools has not kept pace with research on climate change responses.
-Adoption of biophysical approaches is expanding with recognition of
-the importance of spatial and temporal environmental variation to
+[@porter1969thermodynamic; @porter1983biophysical], but adoption of
+these tools has not kept pace with research on climate change responses.
+Adoption of biophysical approaches is expanding with recognition of the
+importance of spatial and temporal environmental variation to
 determining climate change responses [@buckley2018leveraging;
-@kearney2017nichemapr] and the increased dissemination of computational tools.
+@kearney2017nichemapr] and the increased dissemination of computational
+tools.
 
-The TrenchR package is intended to complement
-the NicheMapR package, which includes sophisticated microclimate
-[@kearney2017nichemapr] and biophysical [@kearney2020nichemapr] models.
-Although the Fortran source code was recently released, the complexity
-of NicheMapR functions can make it difficult to understand and modify the
-source code. Many of the approaches are similar between the packages,
-but TrenchR provides simple functions to aid the understanding and
-transparency of biophysical approaches. The TrenchR functions are
-modular and easy adapt to a variety of organisms and research
-questions. Although many of the TrenchR functions are general, some
-components of TrenchR are best suited for ectothermic animals. We have limited inclusion of plant specific biophysical models since
-they are the focus of several R packages [tealeaves: [@muir2019tea];
-plantecophys: [@duursma2015plantecophys]]. Some energy budget
-calculation components of biophysical models are also available in the
-ThermImage package [@tattersall2017]. Microclimate models, with an
-emphasis on describing spatial variation, are also available in the
+Estimating how organisms experience their thermal environment generally
+entails two classes of models [@campbell2000introduction;
+@gates1980biophysical] (Figure 1). Microclimate models allow scaling conditions
+from sensors to organism height. They can characterize heat and air
+transport to estimate vertical air and soil temperature and wind
+profiles. They can also characterize direct and diffuse solar radiation
+and longwave radiation emanating from the sky, ground, and surrounding
+objects like vegetation. Biophysical models estimate body temperatures
+using an energy budget to balance heat exchanges between organisms and
+their environment.
+
+```{r, echo=FALSE, fig.height=4, out.width = "75%"}
+knitr::include_graphics('figures/TrenchRBiophys.png')
+```
+
+**Figure 1** A) Microclimate models scale air and soil temperatures and windspeeds from sensor to organismal heights.
+B) Biophysical models balance heat exchanges between organisms and their environment to estimate body temperatures [@porter1973behavioral].
+Grasshopper body temperatures vary substantially from vegetation, air, and ground temperatures due to heat exchange.
+
+The TrenchR package is intended to complement the NicheMapR package,
+which includes sophisticated microclimate [@kearney2017nichemapr] and
+biophysical [@kearney2020nichemapr] models. Although the Fortran source
+code was recently released, the complexity of NicheMapR functions can
+make it difficult to understand and modify the source code. Many of the
+approaches are similar between the packages, but TrenchR provides simple
+functions to aid the understanding and transparency of biophysical
+approaches. The TrenchR functions are modular and easily adapt to a
+variety of organisms and research questions. Although many of the
+TrenchR functions are general, some components of TrenchR are best
+suited for ectothermic animals. We omit plant specific biophysical
+models since they are the focus of several R packages [tealeaves:
+[@muir2019tea]; plantecophys: [@duursma2015plantecophys]]. Some energy
+budget calculation components of biophysical models are also available
+in the ThermImage package [@tattersall2017]. Microclimate models, with
+an emphasis on describing spatial variation, are also available in the
 Microclima package [@maclean2019microclima].
@@
 
-The TrenchR package (<http://trenchproject.github.io/TrenchR/>) aims to promote transparency and reproducibility.
-We welcome contributions and corrections from users. Our package is
-built using the devtools methodology
-(<https://github.com/hadley/devtools>) in Github. Issues and feature
-requests can be contributed in Github
-(<https://github.com/trenchproject/TrenchR/>).
+The TrenchR package (<http://trenchproject.github.io/TrenchR/>) aims to
+promote transparency and reproducibility. We welcome contributions and
+corrections from users. Our package is built using the devtools
+methodology (<https://github.com/r-lib/devtools>) in Github. Issues and
+feature requests can be contributed in Github
+(<https://github.com/trenchproject/TrenchR>). The packageis available
+via Github and has been submitted to CRAN.
 
-Functions were adapted from biophysical ecology textbooks
+We adapted functions from biophysical ecology textbooks
 [@campbell2000introduction; @gates1980biophysical] and research
-articles. The source reference is included in each function's header. We
-have incorporated default parameters (e.g., organism emissivity or heat
+articles. Source references are included in function headers. We have
+incorporated default parameters (e.g., organism emissivity or heat
 conduction rates) when general values are available. Many functions
@@
 the basis of microclimate and biophysical models (e.g., zenith and
-declination angles, which describe the angles of incident sunlight), facilitate modeling diurnal temperature variation,
-and estimate biologically relevant aggregate metrics such as degree days
-available for development.
+declination angles, which describe the angles of incident sunlight),
+facilitate modeling diurnal temperature variation, and estimate
+biologically relevant aggregate metrics such as degree days available
+for development.
 
@@
 build custom biophysical models. The functions allow implementing an
-energy balance including the following components of heat exchange between
-organisms and their environment: *radiative heat exchanges* of solar and
-thermal radiation; *convective heat exchanges* between organisms and
-their surrounding fluid (air or water) driven by fluid flow; 
-*conductive heat exchanges* between organisms and solid surfaces (generally the ground) due to physical
-contact; and *evaporative heat exchanges* associated with organisms'
-evaporative water loss. Additional functions aggregate these forms of
-heat exchange into energy balances and to use the energy balances to
-predict body temperatures.
+energy balance including the following components of heat exchange
+between organisms and their environment: *radiative heat exchanges* of
+solar and thermal radiation; *convective heat exchanges* between
+organisms and their surrounding fluid (air or water) driven by fluid
+flow; *conductive heat exchanges* between organisms and solid surfaces
+(generally the ground) due to physical contact; and *evaporative heat
+exchanges* associated with organisms' evaporative water loss. Additional
+functions aggregate these forms of heat exchange into energy balances
+and to use the energy balances to predict body temperatures.
 
@@
 of organisms with specified physical properties in a specific
-microclimate and assume no heat exchange via metabolism or
-evaporation [@bakken1992measurement].
+microclimate and assume no heat exchange via metabolism or evaporation
+[@bakken1992measurement].
 
-### Vignettes and introductory modules
+### Vignettes and introductory tutorials
 
@@
 functions and then body temperatures can be solved for using either a
-generic energy balance or taxa specific biophysical models. We additionally offer a *List of symbols* used in equations.
+generic energy balance or taxa specific biophysical models. We
+additionally offer a *List of symbols* used in equations.
 
 For additional background on microclimate and biophysical modelling, we
-have updated a series of educational modules entitled 
-*Physical Processes in Ecosystems*. We intend the modules to provide a more
-contained and accessible introduction to microclimate and biophysical
-modelling than that included in classic textbooks
-[@campbell2000introduction; @gates1980biophysical]. The modules provide
-less detailed coverage of many of the topics contained in the books. The
-series of 14 modules starts with an overview of the calculus and physics
-principles underlying the modelling. Thermodynamics and energy budget
-modelling are then introduced with detailed examples. Modules address
-the climate space concept, operative temperatures, biophysical models
-for leaves and sheep, and heat flow in soils. We have revised the
-modules (which originated from an NSF training grant in 1979 and lacked
-broad distribution) to include R code and utilization of TrenchR
-functions. We expanded the original series of tutorials to include
-introductions by Michael Kearney to the Microclim environmental data
+have updated a series of tutorials entitled *Physical Processes in
+Ecosystems*. We intend the tutorials to provide a more contained and
+accessible introduction to microclimate and biophysical modelling than
+that included in classic textbooks [@campbell2000introduction;
+@gates1980biophysical]. The tutorials provide less detailed coverage of
+many of the topics contained in the books. The series of 14 tutorials
+starts with an overview of the calculus and physics principles
+underlying the modelling. Thermodynamics and energy budget modelling are
+then introduced with detailed examples. Tutorials address the climate
+space concept, operative temperatures, biophysical models for leaves and
+sheep, and heat flow in soils. We have revised the tutorials (which
+originated from an NSF training grant in 1979 and lacked broad
+distribution) to include R code and utilization of TrenchR functions. We
+expanded the original series to include tutorials contributed by M.
+Kearney introducing the Microclim environmental data
 [@kearney2014microclim] and integrating the Microclim data with
 biophysical modelling to examine broad scale climatic limits. The
-modules are provided in html form and available for download as a pdf at
-the bookdown server
+tutorials are provided in html form and available for download as a pdf
+at the bookdown server
 (<https://bookdown.org/huckley/Physical_Processes_In_Ecosystems/>). The
-R markdown files are available in Github (<https://github.com/trenchproject/TrenchRmodules>).
+R markdown files are available in Github
+(<https://github.com/trenchproject/TrenchRmodules>).
 
@@
 We illustrate the use of the TrenchR package by estimating an energy
-budget for a *Sceloporus* lizard on June 1, 2020 in Santa Fe, New
-Mexico, USA (35.69, -105.944, elevation: 2121 meters). The example is also
-incorporated in the *Using energy balances to estimate body temperatures* 
-vignette. We start by generating environmental inputs. Using these inputs,
-we estimate the energy budget with component functions. Finally, we use 
-an integrated biophysical model to estimate operative environmental temperatures.  
-See Appendix S1 for another example examining a
-time series of microclimate data and applying and testing the
-biophysical models.
+budget for a *Sceloporus* lizard on June 1, 2021 in Santa Fe, New
+Mexico, USA (35.69, -105.944, elevation: 2121 meters). The example is
+also incorporated in the *Using energy balances to estimate body
+temperatures* vignette. We start by generating environmental inputs.
+Using these inputs, we estimate the energy budget with component
+functions. Finally, we use an integrated biophysical model to estimate
+operative environmental temperatures. See Appendix S1 for another
+example examining a time series of microclimate data and applying and
+testing the biophysical models.
 
@@
 ectotherms) can be estimated and often referred to as operative
-environmental temperature, $T_e$ [@bakken1992measurement]. $T_e$ is an estimate of $T_b$ and the package functions refer to $T_b$ for simplicity.
+environmental temperature, $T_e$ [@bakken1992measurement]. $T_e$ is an
+estimate of $T_b$ and the package functions refer to $T_b$ for
+simplicity.
 
 Let us assume the lizard is in an unshaded location where a weather
-station at standard height (2 meters) reports that the daily air temperature
-varies from a minimum of 10°C to a maximum of 25°C and the wind speed
-averages 1 m/s. The soil surface temperature varies from a minimum of
-15°C to a maximum of 30°C. We assume that atmospheric transmissivity
-$\tau = 0.7$ and albedo $\rho = 0.6$
+station at standard height (2 meters) reports that the daily air
+temperature varies from a minimum of 10°C to a maximum of 25°C and the
+wind speed averages 1 m/s. The soil surface temperature varies from a
+minimum of 15°C to a maximum of 30°C. We assume that atmospheric
+transmissivity $\tau = 0.7$ and albedo $\rho = 0.6$
 
@@
 We first prepare the environmental data for analysis. We will estimate
-hourly air and soil temperatures and radiation usinf a function describing diurnal temperature variation. We start by
-estimating the day of year and the timing of sunrise and sunset:
+hourly air and soil temperatures and radiation using a function
+describing diurnal temperature variation. We start by estimating the day
+of year and the timing of sunrise and sunset:
 
@@
 library(TrenchR)
-
-#to be REMOVED
-Tb_CampbellNorman=function(T_a, T_g, S, alpha_L=0.96, epsilon=0.96, c_p=29.3, D, V){
-
-  stopifnot(T_a>200, T_a<400, epsilon>=0.5, epsilon<=1, c_p>=0, D>0, V>=0)
-
-  #Stefan-Boltzmann constant
-  sigma= 5.673*10^(-8) #W m^(-2) K^(-4)
-
-  #solar and thermal radiation absorbed
-  L_a=sigma*T_a^4  # (10.7) long wave flux densities from atmosphere 
-  L_g=sigma*T_g^4  # (10.7) long wave flux densities from ground
-  F_a=0.5; F_g=0.5 #proportion of organism exposure to air and ground, respectively
-  R_abs= S+alpha_L*(F_a*L_a+F_g*L_g) # (11.14) Absorbed radiation
-
-  #thermal radiation emitted
-  Qemit= epsilon*sigma*T_a^4
-
-  #conductance
-  g_Ha=1.4*0.135*sqrt(V/D) # boundary conductance, factor of 1.4 to account for increased convection (Mitchell 1976), assumes forced conduction
-  g_r= 4*epsilon*sigma*T_a^3/c_p # (12.7) radiative conductance
-
-  # operative environmental temperature
-  T_e=T_a+(R_abs-Qemit)/(c_p*(g_r+g_Ha))
-
-  return(T_e) 
-}
 ```
@@
 
-doy = day_of_year("2020-06-01", format= "%Y-%m-%d") # Find Julian Date (DOY)
+doy = day_of_year("2021-06-01", format= "%Y-%m-%d") # Find Julian Date (DOY)
 snoon = solar_noon(lon = lon, doy = doy) # Estimate solar noon
@@
 estimate hourly solar radiation by discounting incoming solar radiation
-as it moves through the atmosphere as follows. The function partitions radiation into direct, diffuse, and reflected components:
+as it moves through the atmosphere as follows. The function partitions
+radiation into direct, diffuse, and reflected components:
 
@@
 # Convert to radians
-psi_rad = degree_to_radian(psi_deg) 
+psi_rad = degrees_to_radians(psi_deg) 
 # Estimate radiation
-Srad = sapply(psi_rad, FUN = estimate_radiation, doy = doy, tau = tau, elev = elev, rho = rho)
+Srad = sapply(psi_rad, FUN = solar_radiation, doy = doy, tau = tau, elev = elev, rho = rho)
 
@@
 determines the turbulence of airflow. We implement a neutral air
-temperature profile, an air temperature
-profile forced by wind speed, and a neutral wind speed profile.
+temperature profile, an air temperature profile forced by wind speed,
+and a neutral wind speed profile.
 
@@
 
-# Estimate surface area and sihouette area
+# Estimate surface area and the proportion sihouette area
 A = surface_area_from_mass(mass, "lizard") # surface area (m^2)
-psa = sapply(psi_deg, proportion_silhouette_area, taxon = "lizard", posture = "prostrate") # Projected silhouette area (m^2)
-
+psa = sapply(psi_deg, proportion_silhouette_area, taxon = "lizard", posture = "prostrate") 
 # Change negative values to zero
@@
 (`heat_transfer_coefficient_approximation()`) and a simplified
-approximation (`heat_transfer_coefficient_simple()`) for cases when taxon specific relationships for estimating heat transfer coefficients are not available.
+approximation (`heat_transfer_coefficient_simple()`) for cases when
+taxon specific relationships for estimating heat transfer coefficients
+are not available.
 
@@
 # Estimate the heat transfer coefficient using an empirical relationship for lizards
-H_L = heat_transfer_coefficient(V = V_liz, D = svl, K = 25.7 * 10^(-3), nu = 15.3 * 10^(-6), taxa = "lizard_surface")
+H_L = heat_transfer_coefficient(V = V_liz, D = svl, K = 25.7 * 10^(-3), nu = 15.3 * 10^(-6), taxon = "lizard_surface")
 
 # Estimate the heat transfer coefficient using a spherical approximation
-H_L2= heat_transfer_coefficient_approximation(V = V_liz, D = svl, K = 25.7 * 10^(-3), nu = 15.3 * 10^(-6), taxa = "lizard")
+H_L2= heat_transfer_coefficient_approximation(V = V_liz, D = svl, K = 25.7 * 10^(-3), nu = 15.3 * 10^(-6), taxon = "lizard")
 
@@
 assuming conductance through the animal tissue is the rate limiting step
-as follows: $$Q_{cond} = A \cdot \mbox{proportion} \cdot K(T_g-T_b)/d. $$
-We implement the estimate in R assuming that 1/3 of the lizard surface
-is in contact with the ground and a skin thickness $d = 10^{-6}$:
+as follows:
+$$Q_{cond} = A \cdot \mbox{proportion} \cdot K(T_g-T_b)/d. $$ We
+implement the estimate in R assuming that 1/3 of the lizard surface is
+in contact with the ground and a skin thickness $d = 10^{-6}$:
 
@@
 The microclimate models indicate that air temperatures at lizard height
-are similar to surface temperatures (Figure 1). The biophysical models
+are similar to surface temperatures (Figure 2). The biophysical models
 indicate that solar radiation will elevate lizard body temperatures far
 above air temperatures and that the lizard will face thermal stress if
-it is unable to seek shade (Figure 1). The three biophysical models
+it is unable to seek shade (Figure 2). The three biophysical models
 predict different body temperatures during peak period of solar
@@
 
-**Figure 1**. Body temperatures (Te) are predicted to drastically exceed
+**Figure 2**. Body temperatures (Te) are predicted to drastically exceed
 air temperature when lizards are exposed to high levels of solar
@@
 ## Acknowledgements
-We thank Ray Huey, Joel Kingsolver, Luke Miller, Warren
-Porter, and Art Woods for assistance and input and Brian Helmuth for providing functions. We particularly thank
-Michael Kearney for extensive input on TrenchR and its alignment with
-NicheMapR and for contributing two educational modules.
+
+We thank A. Cannistra, R. Huey, J. Kingsolver, L. Miller, W. Porter, and
+H. A. Woods for assistance and input and B. Helmuth for providing
+functions. We particularly thank M. Kearney for extensive input on
+TrenchR and its alignment with NicheMapR and for contributing two
+tutorials.
 
@@
 LB conceived and designed the package and lead all aspect of the
-packages creation and dissemination; BBO, IC, AJ, OL, AM, ER, YS wrote and
-edited functions and associated package resources; AJ configured the
-package; AM, BBO, YS, and LB converted and edited the educational
-modules. All authors contributed critically to the drafts and gave final
-approval for publication.
+packages creation and dissemination; BBO, IC, AJ, OL, AM, ER, YS, and JS
+wrote and edited functions and associated package resources; AJ and JS
+configured the package; AM, BBO, YS, and LB converted and edited the
+educational modules. All authors contributed critically to the drafts
+and gave final approval for publication.
 

### 2022-05-06 ManuscriptTe.Rmd
Minor fixes

@@
 
-#calculate Julian day
+#calculate day of year
 date.ct <- strptime(dat$datetime, format="%m/%d/%Y %H:%M", tz="MST")
@@
 We first read and process time series of environmental data.
-We estimate the zenith angle, the angle between the sun's rays and the vertical, across Julian days and hours.
+We estimate the zenith angle, the angle between the sun's rays and the vertical, across days of the year and hours.
 ```{r}
@@
 We then implement the energy budget as follows.
-```{r, results = FALSE, warning=FALSE}
+```{r echo=T, results = 'hide', warning=FALSE}
 TeGates= rep(NA, nrow(dat) )

### 2022-05-06 Manuscript_full.Rmd
Minor fixes

@@
 
--   Within the word count (6000-7000 words for Standard Articles, 3000
+-   Within the word count (6000-7000 words for Standard Articles, 3000-4000
     words for Applications and Practical Tools)
@@
     forecasting. We further this aim through a series of educational
-    modules to introduce the field of biophysical ecology.
+    modules that introduce the field of biophysical ecology.
 
@@
 promote transparency and reproducibility. We welcome contributions and
-corrections from users. Our package is built using the devtools
+corrections from users. Our package was built using the devtools
 methodology (<https://github.com/r-lib/devtools>) in Github. Issues and
 feature requests can be contributed in Github
-(<https://github.com/trenchproject/TrenchR>). The packageis available
+(<https://github.com/trenchproject/TrenchR>). The package is available
 via Github and has been submitted to CRAN.
@@
 **Specific biophysical functions** provide biophysical models that have
-been built for specific organisms based on their physical properties,
+been built for particular organisms based on their physical properties,
 behavior, environment, and life history. We currently provide published
@@
 
-## Example Application
+## Examples
 

### 2022-06-10 ManuscriptTe.Rmd
MS updates

@@
 
-The temperature and wind speed profiles illustrate how temperatures generally decrease and wind speeds generally increase with height above the ground (Figure 1).
+The temperature and wind speed profiles illustrate how temperatures generally decrease and wind speeds generally increase with height above the ground (Figure S1).
 Profiles can be fit as a single smooth function or segments.
@@
 
-*Figure 1.
-Temperature (top) and wind speed (bottom) profiles using similar continuous [solid: air_temp_profile_neutral(); dashed: air_temp_profile()] and segmented estimation [dotted: air_temp_profile_segment()].*
+*Figure S1.
+Temperature (top) and wind speed (bottom) profiles using similar continuous [solid: air_temp_profile_neutral(); dashed: air_temp_profile()] and segmented estimation [dotted: air_temp_profile_segment()].
+The points indicate empirical points used for parameterization.*
 
@@
 ```{r, include=FALSE}
-#cut nas
+#cut NAs
 dat= dat[,c("Rad","Temp","Wind","SoilTemp","SanguinipesTemp","Elev","J", "hour","psi","Tgrass","Ugrass")]
@@
 #Acondfact is the proportion of the grasshopper surface area in contact with the ground;
-#abs is absorptivity of grasshopper to solar radiation (proportion);
+#abs is absorptivity of grasshopper to solar radiation (proportion); abs=0.9 corresponds to dark colors.
 #r_g is substrate solar reflectivity (proportion).
@@
 
-We aggregate the data and plot average conditions across sites (Figure 2).
+We aggregate the data and plot average conditions across sites (Figure S2).
 The simplifying assumptions in the biophysical model implementation reduce alignment with the observed temperatures of grasshoppers, but the divergence of grasshopper temperatures from air temperatures at 2m and, to a lesser extent, air temperatures at grasshopper height demonstrate the importance of implementing biophysical models.
@@
 
-*Figure 2.
+*Figure S2.
 Diurnal temperature variation averaged across days in August varies across the elevations of CO sites (panels) and the type of temperature (C).

### 2022-06-10 Manuscript_full.Rmd
MS updates

@@
 output:
+  word_document: default
   html_document: default
-  word_document: default
   pdf_document: default
@@
     radiation. We offer utility functions to aid in estimating the
-    organismal and environonmental parameters needed for biophysical
+    organismal and environmental parameters needed for biophysical
     ecology.
@@
 these tools has not kept pace with research on climate change responses.
-Adoption of biophysical approaches is expanding with recognition of the
+In recent decades, adoption of biophysical approaches is expanding with recognition of the
 importance of spatial and temporal environmental variation to
@@
 B) Biophysical models balance heat exchanges between organisms and their environment to estimate body temperatures [@porter1973behavioral].
-Grasshopper body temperatures vary substantially from vegetation, air, and ground temperatures due to heat exchange.
+A thermal image depicts how grasshopper body temperatures vary substantially from vegetation, air, and ground temperatures due to heat exchange.
 
@@
 functions to aid the understanding and transparency of biophysical
-approaches. The TrenchR functions are modular and easily adapt to a
+approaches. Some variable definitions and units are lacking in NicheMapR. The TrenchR functions are modular and easily adapt to a
 variety of organisms and research questions. Although many of the
@@
 biologically relevant aggregate metrics such as degree days available
-for development.
+for oranismal development.
 
@@
 [@bakken1992measurement].
+However, we also present an analytical function for humid operative temperature that incorporates the effects of evaporative cooling on operative temperatures for wet-skinned ectotherms, such as salamanders.
 
@@
 testing the biophysical models.
+**Flow cart of functions?
 
@@
 estimate of $T_b$ and the package functions refer to $T_b$ for
-simplicity.
+simplicity. Our examplar exstimation of $T_b$ assumes thermal equilibrium.
 
@@
 minimum of 15°C to a maximum of 30°C. We assume that atmospheric
-transmissivity $\tau = 0.7$ and albedo $\rho = 0.6$
+transmissivity $\tau = 0.7$ and albedo $\rho = 0.6$.
 
@@
 
-We first prepare the environmental data for analysis. We will estimate
+At the first stage, we prepare the environmental data for analysis. We will estimate
 hourly air and soil temperatures and radiation using a function
@@
 
-We then calculate hourly air and soil temperatures based on daily
+We then calculate hourly air and soil surface temperatures based on daily
 minimum and maximum temperatures. We select the sine-exponential model
@@
 
-We use microclimate models to scale air temperature and wind speed from
+At the second stage, we use microclimate models to scale air temperature and wind speed from
 weather station height (2m) to lizard height (0.02m). We assume a
@@
 determines the turbulence of airflow. We implement a neutral air
-temperature profile, an air temperature profile forced by wind speed,
+temperature profile driven by density differences, an air temperature profile forced by wind speed,
 and a neutral wind speed profile.
@@
 
-We will be solving the following energy balance to estimate $T_e$:
+Finally, we will use our microclimates estimates to solve the following energy balance to estimate $T_e$:
 $$Q_{net} = Q_{abs} - Q_{emit} - Q_{conv} - Q_{cond} - Q_{met} - Q_{evap},$$
@@
 # Estimate the heat transfer coefficient using an empirical relationship for lizards
-H_L = heat_transfer_coefficient(V = V_liz, D = svl, K = 25.7 * 10^(-3), nu = 15.3 * 10^(-6), taxon = "lizard_surface")
+H_L = heat_transfer_coefficient(V = V_liz, D = svl, K = K, nu = nu, taxon = "lizard_surface")
 
 # Estimate the heat transfer coefficient using a spherical approximation
-H_L2= heat_transfer_coefficient_approximation(V = V_liz, D = svl, K = 25.7 * 10^(-3), nu = 15.3 * 10^(-6), taxon = "lizard")
+H_L2= heat_transfer_coefficient_approximation(V = V_liz, D = svl, K = K, nu = nu, taxon = "lizard")
 
@@
 [@kearney2017nichemapr; @kearney2020nichemapr]. We focus on models that
-predict equilibrium conditions for simplicity. Such models generally do
-not present computational challenges so our models are not optimized for
-computational efficiency. Making classic biophysical ecology techniques
+predict equilibrium conditions for simplicity (that is, steady-state conditions).
+Such models generally do not present computational challenges so our models are not optimized for computational efficiency.
+Making classic biophysical ecology techniques
 more accessible will allow researchers to take advantage of rapidly

### 2022-06-14 ManuscriptTe.Rmd
Minor Te updates

@@
 We then use a time series of air and surface temperatures, wind speeds, and solar radiation collected at a single height to implement a biophysical model for grasshoppers and compare estimates to observations.
+R markdown files with the full code and the associated data are available at https://github.com/trenchproject/TrenchRmanuscript/.
 
@@
 ```{r, include=FALSE}
-setwd("/Volumes/GoogleDrive/Shared drives/TrEnCh/Projects/TrenchR/data/")
-C1all=read.csv("microclimate_C1.csv", sep=",", na.strings = c("-49.99","NA"))
+C1all=read.csv("./data/microclimate_C1.csv", sep=",", na.strings = c("-49.99","NA"))
 ```

### 2022-06-14 Manuscript_full.Rmd
minor edits

@@
 ```{r, echo=FALSE, fig.height=4, out.width = "75%"}
-knitr::include_graphics('figures/TrenchRBiophys.png')
+knitr::include_graphics('./figures/TrenchRBiophys.png')
 ```
@@
 code was recently released, the complexity of NicheMapR functions can
-make it difficult to understand and modify the source code. Many of the
+make it difficult to understand and modify the source code. The function documentation including variable definitions and units can also be challenging to follow. Many of the
 approaches are similar between the packages, but TrenchR provides simple
 functions to aid the understanding and transparency of biophysical
-approaches. Some variable definitions and units are lacking in NicheMapR. The TrenchR functions are modular and easily adapt to a
+approaches. The TrenchR functions are modular and easily adapt to a
 variety of organisms and research questions. Although many of the
@@
 [@campbell2000introduction; @gates1980biophysical] and research
-articles. Source references are included in function headers. We have
+articles. Source references are included in function headers and reference sections. We have
 incorporated default parameters (e.g., organism emissivity or heat
@@
 
-TrenchR functions are organized into the following categories:
+TrenchR functions are organized into the following categories (Figure 2):
 
@@
 for oranismal development.
+
+**Allometric functions** allow estimating the dimensions of organisms needed for energy balances and other analyses.
+The functions can convert between organismal length, mass, surface area, and volume.
+The functions can also estimate the silhoutte area, which describes the organismal area exposed to solar radiation.
 
@@
 However, we also present an analytical function for humid operative temperature that incorporates the effects of evaporative cooling on operative temperatures for wet-skinned ectotherms, such as salamanders.
+
+```{r, echo=FALSE, fig.height=4, out.width = "75%"}
+knitr::include_graphics('./figures/TrenchRFlow.png')
+```
+
+**Figure 2** An examplar aim of translating from environmental and phenotypic data to organismal conditions in a given environment (grey boxes) can be achieved in TrenchR via several categories of functions (white boxes).
+We list example functions for each category used in the example application below.
 
@@
 at the bookdown server
-(<https://bookdown.org/huckley/Physical_Processes_In_Ecosystems/>). The
-R markdown files are available in Github
+(<https://bookdown.org/huckley/Physical_Processes_In_Ecosystems/>).
+R markdown files for the tutorials are available in Github
 (<https://github.com/trenchproject/TrenchRmodules>).
 
-## Examples
+## Example application
 
@@
 budget for a *Sceloporus* lizard on June 1, 2021 in Santa Fe, New
-Mexico, USA (35.69, -105.944, elevation: 2121 meters). The example is
+Mexico, USA (35.69°, -105.944°, elevation: 2121 m). The example is
 also incorporated in the *Using energy balances to estimate body
-temperatures* vignette. We start by generating environmental inputs.
+temperatures* vignette. We start by generating environmental inputs (Figure 2).
 Using these inputs, we estimate the energy budget with component
@@
 testing the biophysical models.
-**Flow cart of functions?
 

### 2022-06-14 ManuscriptTe.Rmd
minor fixes

@@
 output:
+  word_document: default
+  pdf_document: default
   html_document: default
-  pdf_document: default
-  word_document: default
   rmarkdown::html_vignette: default
@@
 #Grasshopper data
-setwd("/Volumes/GoogleDrive/Shared drives/TrEnCh/Projects/TrenchR/data/")
-spec.dat=read.csv("SpecData.csv")
+spec.dat=read.csv("./data/SpecData.csv")
 
 #read weather data
-dat=read.csv("DatAll_Pace2011.csv", sep=",", na.strings = c("-49.99","NA"))
+dat=read.csv("./data/DatAll_Pace2011.csv", sep=",", na.strings = c("-49.99","NA"))
 #match site data
-sites=read.csv("SiteElevations.csv", sep=",")
+sites=read.csv("./data/SiteElevations.csv", sep=",")
 
@@
 ```{r, include=FALSE}
-setwd("/Volumes/GoogleDrive/Shared drives/TrEnCh/Projects/TrenchR/figures/")
-
-pdf("TeTime2.pdf", height = 10, width = 15)
+pdf("./figures/TeTime2.pdf", height = 10, width = 15)
 plot.t

### 2022-06-14 Manuscript_full.Rmd
minor fixes

@@
 The microclimate models indicate that air temperatures at lizard height
-are similar to surface temperatures (Figure 2). The biophysical models
+are similar to surface temperatures (Figure 3). The biophysical models
 indicate that solar radiation will elevate lizard body temperatures far
 above air temperatures and that the lizard will face thermal stress if
-it is unable to seek shade (Figure 2). The three biophysical models
+it is unable to seek shade (Figure 3). The three biophysical models
 predict different body temperatures during peak period of solar
@@
 
-**Figure 2**. Body temperatures (Te) are predicted to drastically exceed
+**Figure 3**. Body temperatures (Te) are predicted to drastically exceed
 air temperature when lizards are exposed to high levels of solar

### 2022-06-14 ManuscriptTe.Rmd
very minor fix

@@
 output:
+  html_document: default
+  pdf_document: default
   word_document: default
-  pdf_document: default
-  html_document: default
   rmarkdown::html_vignette: default
@@
 
-We then implement the energy budget as follows.
+We then implement the energy budget as follows. We note that the energy budget can not be solved for some time periods and errors are generated.
 ```{r echo=T, results = 'hide', warning=FALSE}

### 2022-06-14 Manuscript_full.Rmd
very minor fix

@@
 output:
+  html_document: default
   word_document: default
-  html_document: default
   pdf_document: default
@@
        lty=1, pch=NA, col=c("#440154FF", "#3B528BFF", "#21908CFF", "#5DC863FF", "#FDE725FF"))
-
 ```

### 2022-06-28 Manuscript_full.Rmd
render
first set of edits
minor typos, bringing lines of the markdown doc together, rendering all three doc types

@@
 
-Lauren B. Buckley^1^, Bryan A. Briones Ortiz^1^, Isaac Caruso^1^, Aji
-John^1^, Ofir Levy^2^, Abigail V. Meyer^1^, Eric A. Riddell^3^, Yutaro
-Sakairi^1^, and Juniper L. Simonis^4^
+Lauren B. Buckley^1^, Bryan A. Briones Ortiz^1^, Isaac Caruso^1^, Aji John^1^, Ofir Levy^2^, Abigail V. Meyer^1^, Eric A. Riddell^3^, Yutaro Sakairi^1^, and Juniper L. Simonis^4^
 
-1.  Department of Biology, University of Washington, Seattle, WA
-    98195-1800, USA
+1.  Department of Biology, University of Washington, Seattle, WA 98195-1800, USA
 2.  School of Zoology, Tel Aviv University, Tel Aviv, Israel
-3.  Department of Ecology, Evolution, and Organismal Biology, Iowa State
-    University, Ames, IA 50011-4009, USA
-4.  DAPPER Stats, 3519 NE 15th Avenue, Suite 467, Portland, Oregon
-    97212, USA
+3.  Department of Ecology, Evolution, and Organismal Biology, Iowa State University, Ames, IA 50011-4009, USA
+4.  DAPPER Stats, 3519 NE 15th Avenue, Suite 467, Portland, Oregon 97212, USA
 
-Corresponding Author: Lauren Buckley,
-[lbuckley\@uw.edu](mailto:lbuckley@uw.edu){.email}
+Corresponding Author: Lauren Buckley, [lbuckley\@uw.edu](mailto:lbuckley@uw.edu){.email}
 
@@
 
-1.  Much understanding of organimsal responses to climate change and
-    variability relies on the assumption that body temperatures are
-    equal to temporally averaged air temperatures high above the ground.
-    However, most organisms live near the ground and acute exposure to
-    solar and thermal radiation and thermal extremes can substantially
-    elevate their body temperatures.
-2.  We introduce the TrenchR package, which aids in Translating
-    Environmental Change into organismal responses. The package includes
-    microclimate models to vertically scale weather station data to
-    organismal heights. The biophysical modelling tools include both
-    general models for heat flows and specific models to predict body
-    temperatures for a variety of ectothermic taxa. Additional functions
-    model and temporally partition air and soil temperatures and solar
-    radiation. We offer utility functions to aid in estimating the
-    organismal and environmental parameters needed for biophysical
-    ecology.
-3.  TrenchR focuses on simple and modular functions so users can create
-    transparent and flexible models for biophysical applications.
-4.  The package aims to introduce and enable microclimate and
-    biophysical modelling to improve ecological and evolutionary
-    forecasting. We further this aim through a series of educational
-    modules that introduce the field of biophysical ecology.
+1.
+Much understanding of organismal responses to climate change and variability relies on the assumption that body temperatures are equal to temporally averaged air temperatures high above the ground.
+    However, most organisms live near the ground and acute exposure to solar and thermal radiation and thermal extremes can substantially elevate their body temperatures.
+2.  We introduce the TrenchR package, which aids in Translating Environmental Change into organismal responses. 
+    The package includes microclimate models to vertically scale weather station data to organismal heights. 
+    The biophysical modeling tools include both general models for heat flows and specific models to predict body temperatures for a variety of ectothermic taxa. 
+    Additional functions model and temporally partition air and soil temperatures and solar radiation. 
+    We offer utility functions to aid in estimating the organismal and environmental parameters needed for biophysical ecology.
+3.  TrenchR focuses on simple and modular functions so users can create transparent and flexible models for biophysical applications.
+4.  The package aims to introduce and enable microclimate and biophysical modeling to improve ecological and evolutionary forecasting. 
+    We further this aim through a series of educational modules that introduce the field of biophysical ecology.
 
@@
 
-Responses of organisms and ecosystems to climate change are
-heterogeneous and thus inconsistent with current predictive models
-[@maguire2015modeling]. Some of these predictive shortcomings stem from
-omitting spatial and temporal environmental variation and how it
-interacts with organismal phenotypes (e.g., size, coloration)
-[@nadeau2017coarse; @potter2013microclimatic]. Many analyses assume that
-organismal body temperatures are equal to air temperatures at weather
-station height (usually 2 meters). However, air temperatures near the
-ground where most organisms reside are often considerably warmer, and
-absorption of solar and thermal radiation can raise body temperatures
-well above air temperatures, resulting in potential discrepancies of
-tens of degrees Celsius [@sinclair2016; @sunday2014; @kearney2009].
-Applications are increasingly demonstrating the importance of accounting
-for microclimate variation [@bramer2018advances; @butikofer2020problem;
-@lembrechts2019comparing]. Acute thermal stress events are often more
-relevant to the physiology and energy balance of organisms than daily,
-monthly, or annually averaged environmental conditions. Yet, daily and
-seasonal environmental variation is often neglected
-[@dillon2016introduction; @dillon2016life]. We aim to help remedy these
-shortcomings by introducing the TrenchR package
-(<https://github.com/trenchproject/TrenchR>) as a component of the
-TrEnCh project (<https:www.trenchproject.com>), which builds
-computational and visualization tools to Translate Environmental Change
-into organismal responses.
+Responses of organisms and ecosystems to climate change are heterogeneous and thus inconsistent with current predictive models [@maguire2015modeling]. 
+Some of these predictive shortcomings stem from omitting spatial and temporal environmental variation and how it interacts with organismal phenotypes (e.g., size, coloration) [@nadeau2017coarse; @potter2013microclimatic].
+Many analyses assume that organismal body temperatures are equal to air temperatures at weather station height (usually 2 meters). 
+However, air temperatures near the ground where most organisms reside are often considerably warmer, and absorption of solar and thermal radiation can raise body temperatures well above air temperatures, resulting in potential discrepancies of tens of degrees Celsius [@sinclair2016; @sunday2014; @kearney2009].
+Applications are increasingly demonstrating the importance of accounting for microclimate variation [@bramer2018advances; @butikofer2020problem; @lembrechts2019comparing]. 
+Acute thermal stress events are often more relevant to the physiology and energy balance of organisms than daily, monthly, or annually averaged environmental conditions. 
+Yet, daily and seasonal environmental variation is often neglected [@dillon2016introduction; @dillon2016life]. 
+We aim to help remedy these shortcomings by introducing the TrenchR package (<https://github.com/trenchproject/TrenchR>) as a component of the TrEnCh project (<https:www.trenchproject.com>), which builds computational and visualization tools to Translate Environmental Change into organismal responses.
 
-The mathematical tools for leveraging environmental data to predict
-organismal conditions have long lingered in books
-[@campbell2000introduction; @gates1980biophysical] and articles
-[@porter1969thermodynamic; @porter1983biophysical], but adoption of
-these tools has not kept pace with research on climate change responses.
-In recent decades, adoption of biophysical approaches is expanding with recognition of the
-importance of spatial and temporal environmental variation to
-determining climate change responses [@buckley2018leveraging;
-@kearney2017nichemapr] and the increased dissemination of computational
-tools.
+The mathematical tools for leveraging environmental data to predict organismal conditions have long lingered in books [@campbell2000introduction; @gates1980biophysical] and articles [@porter1969thermodynamic; @porter1983biophysical], but adoption of these tools has not kept pace with research on climate change responses.
+In recent decades, adoption of biophysical approaches is expanding with recognition of the importance of spatial and temporal environmental variation to determining climate change responses [@buckley2018leveraging; @kearney2017nichemapr] and the increased dissemination of computational tools.
 
-Estimating how organisms experience their thermal environment generally
-entails two classes of models [@campbell2000introduction;
-@gates1980biophysical] (Figure 1). Microclimate models allow scaling conditions
-from sensors to organism height. They can characterize heat and air
-transport to estimate vertical air and soil temperature and wind
-profiles. They can also characterize direct and diffuse solar radiation
-and longwave radiation emanating from the sky, ground, and surrounding
-objects like vegetation. Biophysical models estimate body temperatures
-using an energy budget to balance heat exchanges between organisms and
-their environment.
+Estimating how organisms experience their thermal environment generally entails two classes of models [@campbell2000introduction; @gates1980biophysical] (Figure 1).
+Microclimate models allow scaling conditions from sensors to organism height. They can characterize heat and air transport to estimate vertical air and soil temperature and wind profiles. 
+They can also characterize direct and diffuse solar radiation and longwave radiation emanating from the sky, ground, and surrounding objects like vegetation. 
+Biophysical models estimate body temperatures using an energy budget to balance heat exchanges between organisms and their environment.
 
@@
 
-**Figure 1** A) Microclimate models scale air and soil temperatures and windspeeds from sensor to organismal heights.
+**Figure 1** A) Microclimate models scale air and soil temperatures and wind speeds from sensor to organismal heights.
 B) Biophysical models balance heat exchanges between organisms and their environment to estimate body temperatures [@porter1973behavioral].
@@
 
-The TrenchR package is intended to complement the NicheMapR package,
-which includes sophisticated microclimate [@kearney2017nichemapr] and
-biophysical [@kearney2020nichemapr] models. Although the Fortran source
-code was recently released, the complexity of NicheMapR functions can
-make it difficult to understand and modify the source code. The function documentation including variable definitions and units can also be challenging to follow. Many of the
-approaches are similar between the packages, but TrenchR provides simple
-functions to aid the understanding and transparency of biophysical
-approaches. The TrenchR functions are modular and easily adapt to a
-variety of organisms and research questions. Although many of the
-TrenchR functions are general, some components of TrenchR are best
-suited for ectothermic animals. We omit plant specific biophysical
-models since they are the focus of several R packages [tealeaves:
-[@muir2019tea]; plantecophys: [@duursma2015plantecophys]]. Some energy
-budget calculation components of biophysical models are also available
-in the ThermImage package [@tattersall2017]. Microclimate models, with
-an emphasis on describing spatial variation, are also available in the
-Microclima package [@maclean2019microclima].
+The TrenchR package is intended to complement the NicheMapR package, which includes sophisticated microclimate [@kearney2017nichemapr] and biophysical [@kearney2020nichemapr] models.
+Although the Fortran source code was recently released, the complexity of NicheMapR functions can make it difficult to understand and modify the source code.
+The function documentation including variable definitions and units can also be challenging to follow. 
+Many of the approaches are similar between the packages, but TrenchR provides simple functions to aid the understanding and transparency of biophysical approaches. 
+The TrenchR functions are modular and easily adapt to a variety of organisms and research questions. 
+Although many of the TrenchR functions are general, some components of TrenchR are best suited for ectothermic animals. 
+We omit plant specific biophysical models since they are the focus of several R packages [tealeaves: [@muir2019tea]; plantecophys: [@duursma2015plantecophys]]. 
+Some energy budget calculation components of biophysical models are also available in the ThermImage package [@tattersall2017]. 
+Microclimate models, with an emphasis on describing spatial variation, are also available in the Microclima package [@maclean2019microclima].
 
-## Methods and features
+## Methods and Features
 
-The TrenchR package (<http://trenchproject.github.io/TrenchR/>) aims to
-promote transparency and reproducibility. We welcome contributions and
-corrections from users. Our package was built using the devtools
-methodology (<https://github.com/r-lib/devtools>) in Github. Issues and
-feature requests can be contributed in Github
-(<https://github.com/trenchproject/TrenchR>). The package is available
-via Github and has been submitted to CRAN.
+The TrenchR package (<http://trenchproject.github.io/TrenchR/>) aims to promote transparency and reproducibility. 
+We welcome contributions and corrections from users. 
+Our package was built using the devtools methodology (<https://github.com/r-lib/devtools>) with version control managed in Github. 
+Issues and feature requests can be contributed in Github (<https://github.com/trenchproject/TrenchR>). 
+The package is available via Github and has been submitted to CRAN.
 
-We adapted functions from biophysical ecology textbooks
-[@campbell2000introduction; @gates1980biophysical] and research
-articles. Source references are included in function headers and reference sections. We have
-incorporated default parameters (e.g., organism emissivity or heat
-conduction rates) when general values are available. Many functions
-additionally include comments that describe potential parameterizations.
-The biophysical ecology texts [@campbell2000introduction;
-@gates1980biophysical] include many tables with parameter values and we
-provide several data tables in TrenchR.
+We adapted functions from biophysical ecology textbooks [@campbell2000introduction; @gates1980biophysical] and research articles. 
+Source references are included in function headers and reference sections. 
+We have incorporated default parameters (e.g., organism emissivity or heat conduction rates) when general values are available. 
+Many functions additionally include comments that describe potential parameterizations.
+The biophysical ecology texts [@campbell2000introduction; @gates1980biophysical] include many tables with parameter values and we provide several data tables in TrenchR.
 
@@
 
-**Utility functions** allow estimating environmental metrics that form
-the basis of microclimate and biophysical models (e.g., zenith and
-declination angles, which describe the angles of incident sunlight),
-facilitate modeling diurnal temperature variation, and estimate
-biologically relevant aggregate metrics such as degree days available
-for oranismal development.
+**Utility functions** aid calculation of environmental metrics that form the basis of microclimate and biophysical models (e.g., zenith and declination angles, which describe the angles of incident sunlight), facilitate modeling diurnal temperature variation, and estimate biologically relevant aggregate metrics such as degree days available for oranismal development.
 
@@
 
-**Microclimate functions** allow estimating the environmental conditions
-experienced by organisms. Temperature and wind profile functions scale
-environmental conditions from sensors to organismal height. Radiation
-functions allow estimating incoming solar radiation and partitioning the
-variation diurnally and across components (i.e., direct, diffuse,
-reflected).
+**Microclimate functions** facilitate calculating the environmental conditions experienced by organisms. 
+Temperature and wind profile functions scale environmental conditions from sensors to organismal height. 
+Radiation functions allow estimating incoming solar radiation and partitioning the variation diurnally and across components (i.e., direct, diffuse, reflected).
 
-**General biophysical functions** provide generalized models of heat
-exchanges between organisms and their environment so that users can
-build custom biophysical models. The functions allow implementing an
-energy balance including the following components of heat exchange
-between organisms and their environment: *radiative heat exchanges* of
-solar and thermal radiation; *convective heat exchanges* between
-organisms and their surrounding fluid (air or water) driven by fluid
-flow; *conductive heat exchanges* between organisms and solid surfaces
-(generally the ground) due to physical contact; and *evaporative heat
-exchanges* associated with organisms' evaporative water loss. Additional
-functions aggregate these forms of heat exchange into energy balances
-and to use the energy balances to predict body temperatures.
+**General biophysical functions** provide generalized models of heat exchanges between organisms and their environment so that users can build custom biophysical models. 
+The functions allow implementing an energy balance including the following components of heat exchange between organisms and their environment: 
+  *radiative heat exchanges* of solar and thermal radiation; 
+  *convective heat exchanges* between organisms and their surrounding fluid (air or water) driven by fluid flow; 
+  *conductive heat exchanges* between organisms and solid surfaces (generally the ground) due to physical contact; and 
+  *evaporative heat exchanges* associated with organisms' evaporative water loss. 
+Additional functions aggregate these forms of heat exchange into energy balances and to use the energy balances to predict body temperatures.
 
-**Specific biophysical functions** provide biophysical models that have
-been built for particular organisms based on their physical properties,
-behavior, environment, and life history. We currently provide published
-biophysical models for lizards, salamanders, butterflies, grasshoppers,
-limpets, mussels, and snails. Most models predict operative
-environmental temperatures, which are the equilibrium body temperatures
-of organisms with specified physical properties in a specific
-microclimate and assume no heat exchange via metabolism or evaporation
-[@bakken1992measurement].
+**Specific biophysical functions** provide biophysical models that have been built for particular organisms based on their physical properties, behavior, environment, and life history. 
+We currently provide published biophysical models for lizards, salamanders, butterflies, grasshoppers, limpets, mussels, and snails.
+Most models predict operative environmental temperatures, which are the equilibrium body temperatures of organisms with specified physical properties in a specific microclimate and assume no heat exchange via metabolism or evaporation [@bakken1992measurement].
 However, we also present an analytical function for humid operative temperature that incorporates the effects of evaporative cooling on operative temperatures for wet-skinned ectotherms, such as salamanders.
@@
 
-We introduce the functions in several vignettes. A good place to start
-is the *Allometry and conversions* vignette, which provides tools for
-preparing data, such as estimating additional dimensions of organisms
-from measured dimensions. The *Estimating microclimates* vignette
-provides resources for estimating the environmental conditions
-experienced by organisms. This includes estimating solar radiation and
-its components, diurnal variation in temperature and radiation,
-temperature and wind speed profiles, and soil temperatures and profiles.
-Finally, the core biophysical modelling functions are described in a
-tutorial on *Using energy balances to estimate body temperatures*.
-Components of an energy budget can be estimated using individual
-functions and then body temperatures can be solved for using either a
-generic energy balance or taxa specific biophysical models. We
-additionally offer a *List of symbols* used in equations.
+We introduce the functions in several vignettes. 
+A good place to start is the *Allometry and conversions* vignette, which provides tools for preparing data, such as estimating additional dimensions of organisms from measured dimensions.
+The *Estimating microclimates* vignette provides resources for estimating the environmental conditions experienced by organisms.
+This includes estimating solar radiation and its components, diurnal variation in temperature and radiation, temperature and wind speed profiles, and soil temperatures and profiles.
+Finally, the core biophysical modeling functions are described in a tutorial on *Using energy balances to estimate body temperatures*.
+Components of an energy budget can be estimated using individual functions and then body temperatures can be solved for using either a generic energy balance or taxa specific biophysical models. 
+We additionally offer a *List of symbols* used in equations.
 
-For additional background on microclimate and biophysical modelling, we
-have updated a series of tutorials entitled *Physical Processes in
-Ecosystems*. We intend the tutorials to provide a more contained and
-accessible introduction to microclimate and biophysical modelling than
-that included in classic textbooks [@campbell2000introduction;
-@gates1980biophysical]. The tutorials provide less detailed coverage of
-many of the topics contained in the books. The series of 14 tutorials
-starts with an overview of the calculus and physics principles
-underlying the modelling. Thermodynamics and energy budget modelling are
-then introduced with detailed examples. Tutorials address the climate
-space concept, operative temperatures, biophysical models for leaves and
-sheep, and heat flow in soils. We have revised the tutorials (which
-originated from an NSF training grant in 1979 and lacked broad
-distribution) to include R code and utilization of TrenchR functions. We
-expanded the original series to include tutorials contributed by M.
-Kearney introducing the Microclim environmental data
-[@kearney2014microclim] and integrating the Microclim data with
-biophysical modelling to examine broad scale climatic limits. The
-tutorials are provided in html form and available for download as a pdf
-at the bookdown server
-(<https://bookdown.org/huckley/Physical_Processes_In_Ecosystems/>).
-R markdown files for the tutorials are available in Github
-(<https://github.com/trenchproject/TrenchRmodules>).
+For additional background on microclimate and biophysical modeling, we have updated a series of tutorials entitled *Physical Processes in Ecosystems*. 
+We intend the tutorials to provide a more contained and accessible introduction to microclimate and biophysical modeling than that included in classic textbooks [@campbell2000introduction; @gates1980biophysical].
+The tutorials provide less detailed coverage of many of the topics contained in the books. 
+The series of 14 tutorials starts with an overview of the calculus and physics principles underlying the modeling. 
+Thermodynamics and energy budget modeling are then introduced with detailed examples. 
+Tutorials address the climate space concept, operative temperatures, biophysical models for leaves and sheep, and heat flow in soils. 
+We have revised the tutorials (which originated from an NSF training grant in 1979 and lacked broad distribution) to include R code and utilization of TrenchR functions. 
+We expanded the original series to include tutorials contributed by M.
+Kearney introducing the Microclim environmental data [@kearney2014microclim] and integrating the Microclim data with biophysical modeling to examine broad scale climatic limits.
+The tutorials are provided in html form and available for download as a pdf at the bookdown server (<https://bookdown.org/huckley/Physical_Processes_In_Ecosystems/>).
+R markdown files for the tutorials are available in Github (<https://github.com/trenchproject/TrenchRmodules>).
 
-## Example application
+## Example Application
 
-We illustrate the use of the TrenchR package by estimating an energy
-budget for a *Sceloporus* lizard on June 1, 2021 in Santa Fe, New
-Mexico, USA (35.69°, -105.944°, elevation: 2121 m). The example is
-also incorporated in the *Using energy balances to estimate body
-temperatures* vignette. We start by generating environmental inputs (Figure 2).
-Using these inputs, we estimate the energy budget with component
-functions. Finally, we use an integrated biophysical model to estimate
-operative environmental temperatures. See Appendix S1 for another
-example examining a time series of microclimate data and applying and
-testing the biophysical models.
+We illustrate the use of the TrenchR package by estimating an energy budget for a *Sceloporus* lizard on June 1, 2021 in Santa Fe, New Mexico, USA (35.69$^\circ$, -105.944$^\circ$, elevation: 2121 m).
+The example is also incorporated in the *Using energy balances to estimate body temperatures* vignette. 
+We start by generating environmental inputs (Figure 2).
+Using these inputs, we estimate the energy budget with component functions. 
+Finally, we use an integrated biophysical model to estimate operative environmental temperatures. 
+See Appendix S1 for another example examining a time series of microclimate data and applying and testing the biophysical models.
 
-We will use the energy budget to estimate body temperature, $T_b$, which
-can depart dramatically from the air temperatures due to energy exchange
-with the environment. Heat is gained from absorbed solar and thermal
-radiation and from metabolic reactions. Heat is lost through the
-organism's emission of radiation and the evaporation of water. The
-organism exchanges heat with the surrounding air or water via convection
-and with the ground via conduction. The balance of these heat exchanges
-(omitting metabolism and evaporation, which are often negligible for
-ectotherms) can be estimated and often referred to as operative
-environmental temperature, $T_e$ [@bakken1992measurement]. $T_e$ is an
-estimate of $T_b$ and the package functions refer to $T_b$ for
-simplicity. Our examplar exstimation of $T_b$ assumes thermal equilibrium.
+We will use the energy budget to estimate body temperature, $T_b$, which can depart dramatically from the air temperatures due to energy exchange with the environment. 
+Heat is gained from absorbed solar and thermal radiation and from metabolic reactions. 
+Heat is lost through the organism's emission of radiation and the evaporation of water. 
+The organism exchanges heat with the surrounding air or water via convection and with the ground via conduction. 
+The balance of these heat exchanges (omitting metabolism and evaporation, which are often negligible for ectotherms) can be estimated and often referred to as operative environmental temperature, $T_e$ [@bakken1992measurement].
+$T_e$ is an estimate of $T_b$ and the package functions refer to $T_b$ for simplicity. 
+Our exemplar estimation of $T_b$ assumes thermal equilibrium.
 
-Let us assume the lizard is in an unshaded location where a weather
-station at standard height (2 meters) reports that the daily air
-temperature varies from a minimum of 10°C to a maximum of 25°C and the
-wind speed averages 1 m/s. The soil surface temperature varies from a
-minimum of 15°C to a maximum of 30°C. We assume that atmospheric
-transmissivity $\tau = 0.7$ and albedo $\rho = 0.6$.
+Let us assume the lizard is in an unshaded location where a weather station at standard height (2 meters) reports that the daily air temperature varies from a minimum of 10$^\circ$C to a maximum of 25$^\circ$C and the wind speed averages 1 m/s.
+The soil surface temperature varies from a minimum of 15$^\circ$C to a maximum of 30$^\circ$C. 
+We assume that atmospheric transmissivity $\tau = 0.7$ and albedo $\rho = 0.6$.
 
@@
 
-At the first stage, we prepare the environmental data for analysis. We will estimate
-hourly air and soil temperatures and radiation using a function
-describing diurnal temperature variation. We start by estimating the day
-of year and the timing of sunrise and sunset:
+At the first stage, we prepare the environmental data for analysis. 
+We will estimate hourly air and soil temperatures and radiation using a function describing diurnal temperature variation. 
+We start by estimating the day of year and the timing of sunrise and sunset:
 
@@
 
-We then calculate hourly air and soil surface temperatures based on daily
-minimum and maximum temperatures. We select the sine-exponential model
-for air temperature and the sine model for surface temperature:
+We then calculate hourly air and soil surface temperatures based on daily minimum and maximum temperatures. 
+We select the sine-exponential model for air temperature and the sine model for surface temperature:
 
 ```{r}
-# Air temperature (˚C)
+# Air temperature ($^\circ$C)
 Ta = sapply(1:24, diurnal_temp_variation_sineexp, T_max = Tmax, T_min = Tmin, t_r = tr, t_s = ts, alpha = 2.59, beta = 1.55, gamma = 2.2) 
-# Soil temperature (˚C)
+# Soil temperature ($^\circ$C)
 Ts = sapply(1:24, diurnal_temp_variation_sine, T_max = Tmax_s, T_min = Tmin_s) 
@@
 
-At the second stage, we use microclimate models to scale air temperature and wind speed from
-weather station height (2m) to lizard height (0.02m). We assume a
-surface roughness of $z_0 = 0.2$ m, which corresponds to bare sand and
-determines the turbulence of airflow. We implement a neutral air
-temperature profile driven by density differences, an air temperature profile forced by wind speed,
-and a neutral wind speed profile.
+At the second stage, we use microclimate models to scale air temperature and wind speed from weather station height (2 m) to lizard height (0.02 m). 
+We assume a surface roughness of $z_0 = 0.2$ m, which corresponds to bare sand and determines the turbulence of airflow. 
+We implement a neutral air temperature profile driven by density differences, an air temperature profile forced by wind speed, and a neutral wind speed profile.
 
@@
 Finally, we will use our microclimates estimates to solve the following energy balance to estimate $T_e$:
+
 $$Q_{net} = Q_{abs} - Q_{emit} - Q_{conv} - Q_{cond} - Q_{met} - Q_{evap},$$
-where $Q_{net}$ is the net energy exchange with the environment (W),
-$Q_{abs}$ is the solar radiation absorbed (W), $Q_{emit}$ is the net
-thermal radiation emitted (W), $Q_{conv}$ is energy exchange due to
-convection (W), $Q_{cond}$ is energy exchange due to conduction (W),
-$Q_{met}$ is the energy emitted due to metabolism (W), and $Q_{evap}$ is
-the energy emitted due to evaporative water loss (W). We will estimate
-each term on the right side of the equation in turn. Estimating
-$Q_{abs}$ requires the surface area exposed to radiation and the solar
-absorptivity of the animal surface ($a$ proportion). We model a 10 gram
-*Sceloporus* lizard with solar absorptivity $a = 0.9$
-[@gates1980biophysical]. We will initially assume $T_b = T_a$ to
-illustrate the calculations before solving for $T_b$ given the
-environmental conditions.
+
+where $Q_{net}$ is the net energy exchange with the environment (W), $Q_{abs}$ is the solar radiation absorbed (W), $Q_{emit}$ is the net thermal radiation emitted (W), $Q_{conv}$ is energy exchange due to convection (W), $Q_{cond}$ is energy exchange due to conduction (W), $Q_{met}$ is the energy emitted due to metabolism (W), and $Q_{evap}$ is the energy emitted due to evaporative water loss (W).
+We will estimate each term on the right side of the equation in turn. 
+Estimating $Q_{abs}$ requires the surface area exposed to radiation and the solar absorptivity of the animal surface ($a$ proportion). 
+We model a 10 gram *Sceloporus* lizard with solar absorptivity $a = 0.9$ [@gates1980biophysical]. 
+We will initially assume $T_b = T_a$ to illustrate the calculations before solving for $T_b$ given the environmental conditions.
 
@@
 
-We estimate thermal radiation $Q_{emit}$ (W) for the lizard outdoors. We
-assume the surface emissivity of lizards, $epsilon_s = 0.965$
-[@barlett1967].
+We estimate thermal radiation $Q_{emit}$ (W) for the lizard outdoors. 
+We assume the surface emissivity of lizards, $epsilon_s = 0.965$ [@barlett1967].
 
@@
 
-We next estimate convection $Q_{conv}$ (W) and conduction $Q_{cond}$
-(W). We will estimate the lizard's heat transfer coefficient, $H_L$
-($Wm^{-2}K^{-1}$) using an empirical relationship for lizards
-(`heat_transfer_coefficient()`). We also illustrate a function
-estimating $H_L$ using a spherical approximation
-(`heat_transfer_coefficient_approximation()`) and a simplified
-approximation (`heat_transfer_coefficient_simple()`) for cases when
-taxon specific relationships for estimating heat transfer coefficients
-are not available.
+We next estimate convection $Q_{conv}$ (W) and conduction $Q_{cond}$ (W). 
+We will estimate the lizard's heat transfer coefficient, $H_L$ ($Wm^{-2}K^{-1}$) using an empirical relationship for lizards (`heat_transfer_coefficient()`). 
+We also illustrate a function estimating $H_L$ using a spherical approximation (`heat_transfer_coefficient_approximation()`) and a simplified approximation (`heat_transfer_coefficient_simple()`) for cases when taxon specific relationships for estimating heat transfer coefficients are not available.
 
@@
 
-We estimate convective heat exchange between the animal and surrounding
-air using the following relationship:
-$$Q_{conv} = ef \cdot H_L(A\cdot \mbox{proportion})(T_a-T_b),$$ where an
-enhancement factor, $ef$, multiplier can be incorporated to account for
-increases in heat exchange resulting from air turbulence in field
-conditions. We implement the function in R assuming that 2/3 of the
-lizard's surface area is exchanging heat through convection. We
-currently estimate no convection since we are assuming $Tb = Ta$ for
-illustration purposes:
+We estimate convective heat exchange between the animal and surrounding air using the following relationship:
+
+$$Q_{conv} = ef \cdot H_L(A\cdot \mbox{proportion})(T_a-T_b),$$
+
+where an enhancement factor, $ef$, multiplier can be incorporated to account for increases in heat exchange resulting from air turbulence in field conditions. 
+We implement the function in R assuming that 2/3 of the lizard's surface area is exchanging heat through convection. 
+We currently estimate no convection since we are assuming $Tb = Ta$ for illustration purposes:
 
@@
 
-We estimate conductive heat flow (W) from the lizard to the surface
-assuming conductance through the animal tissue is the rate limiting step
-as follows:
-$$Q_{cond} = A \cdot \mbox{proportion} \cdot K(T_g-T_b)/d. $$ We
-implement the estimate in R assuming that 1/3 of the lizard surface is
-in contact with the ground and a skin thickness $d = 10^{-6}$:
+We estimate conductive heat flow (W) from the lizard to the surface assuming conductance through the animal tissue is the rate limiting step as follows:
+
+$$Q_{cond} = A \cdot \mbox{proportion} \cdot K(T_g-T_b)/d. $$
+
+We implement the estimate in R assuming that 1/3 of the lizard surface is in contact with the ground and a skin thickness $d = 10^{-6}$:
 
@@
 
-We assume, as is generally done for lizards, that heat exchange
-associated with metabolism and evaporation is negligible:
+We assume, as is generally done for lizards, that heat exchange associated with metabolism and evaporation is negligible:
 
@@
 
-The full heat budget can be calculated as follows
-[@gates1980biophysical]:
+The full heat budget can be calculated as follows [@gates1980biophysical]:
 
@@
 We constructed the energy balance assuming $T_b = T_a$ for illustration. 
-Thus, the energy balance is dominated by heat gain from contact with the
-warmer surface. We now use a function based on the energy balance above
-to estimate body temperature given the environmental conditions:
+Thus, the energy balance is dominated by heat gain from contact with the warmer surface. 
+We now use a function based on the energy balance above to estimate body temperature given the environmental conditions:
 
@@
 
-We also implement a similar but simplified energy balance
-[@campbell2000introduction]. The energy balance omits conduction with
-the ground:
+We also implement a similar but simplified energy balance [@campbell2000introduction]. 
+The energy balance omits conduction with the ground:
 
@@
 
-We additionally estimate $T_b$ using a specialized function for lizards
-[@buckley2008link]:
+We additionally estimate $T_b$ using a specialized function for lizards [@buckley2008link]:
 
@@
 
-The microclimate models indicate that air temperatures at lizard height
-are similar to surface temperatures (Figure 3). The biophysical models
-indicate that solar radiation will elevate lizard body temperatures far
-above air temperatures and that the lizard will face thermal stress if
-it is unable to seek shade (Figure 3). The three biophysical models
-predict different body temperatures during peak period of solar
-radiation because they model interactions with the ground differently
-and users are encouraged to review the details of each biophysical model
-before selection. Differences in estimated body temperatures are
-accentuated by the high level of solar radiation.
+The microclimate models indicate that air temperatures at lizard height are similar to surface temperatures (Figure 3). 
+The biophysical models indicate that solar radiation will elevate lizard body temperatures far above air temperatures and that the lizard will face thermal stress if it is unable to seek shade (Figure 3).
+The three biophysical models predict different body temperatures during peak period of solar radiation because they model interactions with the ground differently and users are encouraged to review the details of each biophysical model before selection.
+Differences in estimated body temperatures are accentuated by the high level of solar radiation.
 
@@
 par(mar=c(5, 4, 4, 8) + 0.1)
-plot(1:24, (Te-273.15), type="l", xlab="Hour", ylab="Temperature (°C)", col="#5DC863FF", ylim=c(10,60)) # Gates
+plot(1:24, (Te-273.15), type="l", xlab="Hour", ylab= expression("Temperature ("*~degree*C*")"), col="#5DC863FF", ylim=c(10,60)) # Gates
 points(1:24, Te2-273.15, type="l", col="#5DC863FF", lty="dotted") # Campbell and Norman
@@
 
-**Figure 3**. Body temperatures (Te) are predicted to drastically exceed
-air temperature when lizards are exposed to high levels of solar
-radiation. Air temperatures (Ta, °C) at lizard height (0.02m) are
-predicted to exceed air temperatures at 2m and to be similar to surface
-temperatures (Ts). We estimate body temperatures using two general
-energy budgets [solid: Tb_Gates(); dotted: Tb_CampbellNorman()] and a
-lizard specific biophysical model [dashed: Tb_lizard()] that differ in
-how they model heat exchanges.
+**Figure 3**. Body temperatures (Te) are predicted to drastically exceed air temperature when lizards are exposed to high levels of solar radiation. 
+Air temperatures (Ta, $^\circ$C) at lizard height (0.02 m) are predicted to exceed air temperatures at 2 m and to be similar to surface temperatures (Ts). 
+We estimate body temperatures using two general energy budgets [solid: Tb_Gates(); dotted: Tb_CampbellNorman()] and a lizard specific biophysical model [dashed: Tb_lizard()] that differ in how they model heat exchanges.
 
@@
 
-TrenchR is intended to promote understanding of how organisms interact
-with their environment and consequences for physiology, energetics,
-behavior, and demography. Our example implementation highlights the
-importance of considering organismal body temperatures, rather than air
-temperatures, when examining thermal stress and other responses to
-environmental variability and change. TrenchR currently focuses on heat
-balances but may be expanded to include water balances. Simple functions
-can be combined as needed to produce comprehensive and transparent
-models for biophysical ecology and evolution. The resultant models are
-likely to be sufficiently detailed for many applications, but users are
-referred to NicheMapR for more detailed biophysical models
-[@kearney2017nichemapr; @kearney2020nichemapr]. We focus on models that
-predict equilibrium conditions for simplicity (that is, steady-state conditions).
+TrenchR is intended to promote understanding of how organisms interact with their environment and consequences for physiology, energetics, behavior, and demography. 
+Our example implementation highlights the importance of considering organismal body temperatures, rather than air temperatures, when examining thermal stress and other responses to environmental variability and change.
+TrenchR currently focuses on heat balances but may be expanded to include water balances. 
+Simple functions can be combined as needed to produce comprehensive and transparent models for biophysical ecology and evolution. 
+The resultant models are likely to be sufficiently detailed for many applications, but users are referred to NicheMapR for more detailed biophysical models [@kearney2017nichemapr; @kearney2020nichemapr].
+We focus on models that predict equilibrium conditions for simplicity (that is, steady-state conditions). 
 Such models generally do not present computational challenges so our models are not optimized for computational efficiency. 
-Making classic biophysical ecology techniques
-more accessible will allow researchers to take advantage of rapidly
-accumulating data on environmental conditions and organismal traits to
-understand and predict ecological and evolutionary responses.
-Considering how organisms experience their environment is central to
-understanding responses to variable and changing environments
-[@suggitt2018extinction].
+Making classic biophysical ecology techniques more accessible will allow researchers to take advantage of rapidly accumulating data on environmental conditions and organismal traits to understand and predict ecological and evolutionary responses.
+Considering how organisms experience their environment is central to understanding responses to variable and changing environments [@suggitt2018extinction].
 
@@
 
-We thank A. Cannistra, R. Huey, J. Kingsolver, L. Miller, W. Porter, and
-H. A. Woods for assistance and input and B. Helmuth for providing
-functions. We particularly thank M. Kearney for extensive input on
-TrenchR and its alignment with NicheMapR and for contributing two
-tutorials.
+We thank A. Cannistra, R. Huey, J. Kingsolver, L. Miller, W. Porter, and H. A. Woods for assistance and input and B. Helmuth for providing functions. 
+We particularly thank M. Kearney for extensive input on TrenchR and its alignment with NicheMapR and for contributing two tutorials.
 
-## Authors' contributions
+## Authors' Contributions
 
-LB conceived and designed the package and lead all aspect of the
-packages creation and dissemination; BBO, IC, AJ, OL, AM, ER, YS, and JS
-wrote and edited functions and associated package resources; AJ and JS
-configured the package; AM, BBO, YS, and LB converted and edited the
-educational modules. All authors contributed critically to the drafts
-and gave final approval for publication.
+LB conceived and designed the package and lead all aspect of the packages creation and dissemination; BBO, IC, AJ, OL, AM, ER, YS, and JS wrote and edited functions and associated package resources; AJ and JS configured the package; AM, BBO, YS, and LB converted and edited the educational modules.
+All authors contributed critically to the drafts and gave final approval for publication.
 
-## Data availability
+## Data Availability
 
-All code is available at <https://github.com/trenchproject/TrenchR>.
+All code and data are available at <https://github.com/trenchproject/TrenchR>.
 

### 2022-06-28 ManuscriptTe.Rmd
supp rendering and bib

@@
 # APPENDIX S1
+
 We provide an additional example using microclimate data measured along an elevation gradient in CO, USA to illustrate use of TrenchR. 
@@
 
-# Microclimate models
+# Microclimate Models
+
 ```{r, include=FALSE}
@@
 We first examine how temperature and wind speed profiles can be used to scale conditions to the heights of organisms. 
-We first use wind speed variation across hours and heights to estimate surface roughness.
+We start by using wind speed variation across hours and heights to estimate surface roughness.
+
 ```{r}
@@
 We then calculate the temperature and wind speed profiles for a given hour using multiple, alternative functions in TrenchR.
+
 ```{r}
@@
 Profiles can be fit as a single smooth function or segments.
+
 ```{r, echo=FALSE}
@@
 We next illustrate quantifying operative environmental temperatures, which are estimated body temperatures for grasshoppers in the specified environments. 
-We first read and process time series of environmental data.
-We estimate the zenith angle, the angle between the sun's rays and the vertical, across days of the year and hours.
+We read and process time series of environmental data to estimate the zenith angle, the angle between the sun's rays and the vertical, across days of the year and hours.
+
 ```{r}
@@
 We scale the air temperature and wind speed from a reference height of 0.5m to grasshopper height (0.001m).
+
 ```{r}
@@
 
-We then implement a generic energy budget. We first illustrate the estimation of convective heat transfer coefficients as a function of wind speed.
+We then implement a generic energy budget. 
+We first illustrate the estimation of convective heat transfer coefficients as a function of wind speed.
+
 ```{r}
@@
 
-We then implement the energy budget as follows. We note that the energy budget can not be solved for some time periods and errors are generated.
+We then implement the energy budget as follows. 
+We note that the energy budget can not be solved for some time periods and errors are generated.
+
 ```{r echo=T, results = 'hide', warning=FALSE}
@@
 The simplifying assumptions in the biophysical model implementation reduce alignment with the observed temperatures of grasshoppers, but the divergence of grasshopper temperatures from air temperatures at 2m and, to a lesser extent, air temperatures at grasshopper height demonstrate the importance of implementing biophysical models.
+
 ```{r, echo=FALSE, warning=FALSE}
@@
 
-*Figure S2.
-Diurnal temperature variation averaged across days in August varies across the elevations of CO sites (panels) and the type of temperature (C).
+*Figure S2. Diurnal temperature variation averaged across days in August varies across the elevations of CO sites (panels) and the type of temperature (C). 
 Microclimate models indicate that the air temperatures experienced by grasshoppers near the ground (Ta grasshopper height) is generally above air temperatures (Ta) and close to surface temperatures (Ts).

### 2022-06-28 Manuscript_full.Rmd
supp rendering and bib
more bib
adding the tex preamble for the listings package following the rmarkdown cookbook, which controls formatting of the code blocks

@@
 title: 'TrenchR: an R package for transparent environmental and ecological biophysics'
-author: "Lauren B. Buckley, Bryan A. Briones Ortiz, Isaac Caruso, Aji John, Ofir Levy, Abigail V. Meyer, Eric A. Riddell, Yutaro Sakairi, and Juniper Simonis"
+author: 
+  - "Lauren B. Buckley, Bryan A. Briones Ortiz, Isaac Caruso, Aji John, Ofir Levy,"
+  - "Abigail V. Meyer, Eric A. Riddell, Yutaro Sakairi, and Juniper Simonis"
 output:
@@
   word_document: default
-  pdf_document: default
+  pdf_document: 
+    pandoc_args: --listings
+    includes:
+      in_header: preamble.tex
 bibliography: bibliography.bib
@@
 knitr::opts_chunk$set(echo = TRUE)
+
+library(TrenchR)
+
 ```
@@
 
-The mathematical tools for leveraging environmental data to predict organismal conditions have long lingered in books [@campbell2000introduction; @gates1980biophysical] and articles [@porter1969thermodynamic; @porter1983biophysical], but adoption of these tools has not kept pace with research on climate change responses.
+The mathematical tools for leveraging environmental data to predict organismal conditions have long lingered in books [@gates1980biophysical; @campbell2000introduction] and articles [@porter1969thermodynamic; @porter1983biophysical], but adoption of these tools has not kept pace with research on climate change responses.
 In recent decades, adoption of biophysical approaches is expanding with recognition of the importance of spatial and temporal environmental variation to determining climate change responses [@buckley2018leveraging; @kearney2017nichemapr] and the increased dissemination of computational tools.
 
-Estimating how organisms experience their thermal environment generally entails two classes of models [@campbell2000introduction; @gates1980biophysical] (Figure 1).
+Estimating how organisms experience their thermal environment generally entails two classes of models [@gates1980biophysical; @campbell2000introduction] (Figure 1).
 Microclimate models allow scaling conditions from sensors to organism height. They can characterize heat and air transport to estimate vertical air and soil temperature and wind profiles. 
@@
 For additional background on microclimate and biophysical modeling, we have updated a series of tutorials entitled *Physical Processes in Ecosystems*. 
-We intend the tutorials to provide a more contained and accessible introduction to microclimate and biophysical modeling than that included in classic textbooks [@campbell2000introduction; @gates1980biophysical].
+We intend the tutorials to provide a more contained and accessible introduction to microclimate and biophysical modeling than that included in classic textbooks [@gates1980biophysical; @campbell2000introduction].
 The tutorials provide less detailed coverage of many of the topics contained in the books. 
@@
 
-We illustrate the use of the TrenchR package by estimating an energy budget for a *Sceloporus* lizard on June 1, 2021 in Santa Fe, New Mexico, USA (35.69$^\circ$, -105.944$^\circ$, elevation: 2121 m).
+We illustrate the use of the TrenchR package by estimating an energy budget for a *Sceloporus* lizard on June 1, 2021 in Santa Fe, New Mexico, USA (35.69$^\circ$N, -105.944$^\circ$W, elevation: 2121 m).
 The example is also incorporated in the *Using energy balances to estimate body temperatures* vignette. 
@@
 
-Let us assume the lizard is in an unshaded location where a weather station at standard height (2 meters) reports that the daily air temperature varies from a minimum of 10$^\circ$C to a maximum of 25$^\circ$C and the wind speed averages 1 m/s.
-The soil surface temperature varies from a minimum of 15$^\circ$C to a maximum of 30$^\circ$C. 
+Let us assume the lizard is in an unshaded location where a weather station at standard height (2 meters) reports that the daily air temperature varies from a minimum of 10 $^\circ$C to a maximum of 25 $^\circ$C and the wind speed averages 1 m/s.
+The soil surface temperature varies from a minimum of 15 $^\circ$C to a maximum of 30 $^\circ$C. 
 We assume that atmospheric transmissivity $\tau = 0.7$ and albedo $\rho = 0.6$.
@@
 We start by estimating the day of year and the timing of sunrise and sunset:
-
-```{r, echo=FALSE}
-library(TrenchR)
-```
 

### 2022-06-28 Manuscript_full.Rmd

@@
 
-```{r, echo=FALSE, fig.height=4, out.width = "75%"}
+```{r, echo=FALSE, fig.height=4, fig.align = 'center', out.width = "75%"}
 knitr::include_graphics('./figures/TrenchRBiophys.png')
@@
 
-**Utility functions** aid calculation of environmental metrics that form the basis of microclimate and biophysical models (e.g., zenith and declination angles, which describe the angles of incident sunlight), facilitate modeling diurnal temperature variation, and estimate biologically relevant aggregate metrics such as degree days available for oranismal development.
+**Utility functions** calculate environmental metrics that form the basis of microclimate and biophysical models (e.g., zenith and declination angles, which describe the angles of incident sunlight), facilitate modeling diurnal temperature variation, and estimate biologically relevant aggregate metrics such as degree days available for oranismal development.
 
 **Allometric functions** allow estimating the dimensions of organisms needed for energy balances and other analyses. 
-The functions can convert between organismal length, mass, surface area, and volume.
-The functions can also estimate the silhoutte area, which describes the organismal area exposed to solar radiation.
+Available functions can convert between organismal length, mass, surface area, and volume as well as estimate the silhoutte area, which describes the organismal area exposed to solar radiation.
 
@@
 The functions allow implementing an energy balance including the following components of heat exchange between organisms and their environment:
-  *radiative heat exchanges* of solar and thermal radiation; 
-  *convective heat exchanges* between organisms and their surrounding fluid (air or water) driven by fluid flow; 
-  *conductive heat exchanges* between organisms and solid surfaces (generally the ground) due to physical contact; and 
-  *evaporative heat exchanges* associated with organisms' evaporative water loss. 
+
+  - *radiative heat exchanges* of solar and thermal radiation; 
+  - *convective heat exchanges* between organisms and their surrounding fluid (air or water) driven by fluid flow; 
+  - *conductive heat exchanges* between organisms and solid surfaces (generally the ground) due to physical contact; and 
+  - *evaporative heat exchanges* associated with organisms' evaporative water loss.
+
 Additional functions aggregate these forms of heat exchange into energy balances and to use the energy balances to predict body temperatures.
@@
 
-```{r, echo=FALSE, fig.height=4, out.width = "75%"}
+```{r, echo=FALSE, fig.height=4, fig.align = 'center', out.width = "75%"}
 knitr::include_graphics('./figures/TrenchRFlow.png')
@@
 We introduce the functions in several vignettes. 
-A good place to start is the *Allometry and conversions* vignette, which provides tools for preparing data, such as estimating additional dimensions of organisms from measured dimensions.
-The *Estimating microclimates* vignette provides resources for estimating the environmental conditions experienced by organisms.
+A good place to start is the *Allometry and Conversions* vignette, which provides tools for preparing data, such as estimating additional dimensions of organisms from measured dimensions.
+The *Estimating Microclimates* vignette provides resources for estimating the environmental conditions experienced by organisms.
 This includes estimating solar radiation and its components, diurnal variation in temperature and radiation, temperature and wind speed profiles, and soil temperatures and profiles.
-Finally, the core biophysical modeling functions are described in a tutorial on *Using energy balances to estimate body temperatures*.
+Finally, the core biophysical modeling functions are described in a tutorial on *Using Energy Balances to Estimate Body Temperatures*.
 Components of an energy budget can be estimated using individual functions and then body temperatures can be solved for using either a generic energy balance or taxa specific biophysical models. 
-We additionally offer a *List of symbols* used in equations.
+We additionally offer a *List of Symbols* used in equations.
 
@@
 # Set up input data as variables
-lat = 35.69; # Latitude (degrees)
-lon = -105.944; # Longitude (degrees)
-elev = 2121 # Elevation (meters)
 
-Tmin = 10; # Minimum air temperature (C)
-Tmax = 25; # Maximum air temperature (C)
-Tmin_s = 15; # Minimum soil temperature (C)
-Tmax_s = 30; # Maximum soil temperature (C)
-V = 1 # Wind speed (m/s)
+lat  <- 35.69    # Latitude (degrees)
+lon  <- -105.944 # Longitude (degrees)
+elev <- 2121     # Elevation (meters)
+
+Tmin   <- 10 # Minimum air temperature (C)
+Tmax   <- 25 # Maximum air temperature (C)
+Tmin_s <- 15 # Minimum soil temperature (C)
+Tmax_s <- 30 # Maximum soil temperature (C)
+V      <- 1  # Wind speed (m/s)
 
 # Assumptions
-tau = 0.7; # Atmospheric transmissivity
-rho = 0.6; # Albedo
-Tb0 = 25; # Initial assumption of body temperature (C)
 
-doy = day_of_year("2021-06-01", format= "%Y-%m-%d") # Find Julian Date (DOY)
-snoon = solar_noon(lon = lon, doy = doy) # Estimate solar noon
-dayl = daylength(lat = lat, doy = doy) # Estimate day length
-tr = snoon - dayl / 2 # Time of sunrise
-ts = snoon + dayl / 2 # Time of sunset
+tau <- 0.7 # Atmospheric transmissivity
+rho <- 0.6 # Albedo
+Tb0 <- 25  # Initial assumption of body temperature (C)
+
+doy   <- day_of_year("2021-06-01", format = "%Y-%m-%d") # Julian Date (DOY)
+
+snoon <- solar_noon(lon = lon, doy = doy) # Estimate solar noon
+dayl  <- daylength(lat = lat, doy = doy)  # Estimate day length
+
+tr    <- snoon - dayl / 2 # Time of sunrise
+ts    <- snoon + dayl / 2 # Time of sunset
 ```
 
-Although measured solar radiation is preferable if available, we can
-estimate hourly solar radiation by discounting incoming solar radiation
-as it moves through the atmosphere as follows. The function partitions
-radiation into direct, diffuse, and reflected components:
+Although measured solar radiation is preferable if available, we can estimate hourly solar radiation by discounting incoming solar radiation as it moves through the atmosphere as follows. 
+The function partitions radiation into direct, diffuse, and reflected components:
 
@@
 # Estimate zenith angle (degrees)
-psi_deg = sapply(1:24, FUN = zenith_angle, doy = doy, lat = lat, lon = lon) 
+
+psi_deg <- sapply(1:24, FUN = zenith_angle, doy = doy, lat = lat, lon = lon)
+
 # Convert to radians
-psi_rad = degrees_to_radians(psi_deg) 
+
+psi_rad <- degrees_to_radians(psi_deg)
+
 # Estimate radiation
-Srad = sapply(psi_rad, FUN = solar_radiation, doy = doy, tau = tau, elev = elev, rho = rho)
+
+Srad <- sapply(psi_rad, FUN = solar_radiation, doy = doy, tau = tau, elev = elev, rho = rho)
 
 # Separate solar radiation into direct, diffuse, and reflected
-Sdir = Srad[1,]; # Direct solar radiation
-Sdif = Srad[2,]; # Diffuse solar radiation
-Sref = Srad[3,]; # Reflected solar radiation
+
+Sdir <- Srad[1,] # Direct solar radiation
+Sdif <- Srad[2,] # Diffuse solar radiation
+Sref <- Srad[3,] # Reflected solar radiation
 ```
@@
 ```{r}
-# Air temperature ($^\circ$C)
-Ta = sapply(1:24, diurnal_temp_variation_sineexp, T_max = Tmax, T_min = Tmin, t_r = tr, t_s = ts, alpha = 2.59, beta = 1.55, gamma = 2.2) 
-# Soil temperature ($^\circ$C)
-Ts = sapply(1:24, diurnal_temp_variation_sine, T_max = Tmax_s, T_min = Tmin_s) 
+# Air temperature (C)
+
+Ta <- sapply(1:24, diurnal_temp_variation_sineexp, T_max = Tmax, T_min = Tmin, t_r = tr, t_s = ts, alpha = 2.59, beta = 1.55, gamma = 2.2)
+
+# Soil temperature (C)
+
+Ts <- sapply(1:24, diurnal_temp_variation_sine, T_max = Tmax_s, T_min = Tmin_s) 
 ```
@@
 # Scale air temperature to lizard height
-# T_r = reference temperature; zr = reference height; z0 is surface roughness; z = organism height; T_s = surface temperature
-Ta_liz = air_temp_profile_neutral(T_r = Ta, zr = 2, z0 = 0.2, z = 0.02, T_s = Ts) # Neutral air temperature profile
-Ta_liz2 = air_temp_profile(T_r = Ta, u_r = V, zr = 2, z0 = 0.2, z = 0.02, T_s = Ts) # Air temperature profile forced by wind speed
+# T_r = reference temperature
+# zr  = reference height
+# z0  = surface roughness
+# z   = organism height
+# T_s = surface temperature
+
+Ta_liz  <- air_temp_profile_neutral(T_r = Ta, zr = 2, z0 = 0.2, z = 0.02, T_s = Ts)  # Neutral air temperature profile
+Ta_liz2 <- air_temp_profile(T_r = Ta, u_r = V, zr = 2, z0 = 0.2, z = 0.02, T_s = Ts) # Air temperature profile forced by wind speed
 
 # Scale wind speed to lizard height
-# u_r = reference wind speed; zr = reference height; z0 is surface roughness; z = organism height
-V_liz= wind_speed_profile_neutral(u_r = V, zr = 2, z0 = 0.2 , z = 0.02) # Neutral wind speed profile
+# u_r = reference wind speed
+# zr  = reference height
+# z0  = surface roughness
+# z   = organism height
+
+V_liz <- wind_speed_profile_neutral(u_r = V, zr = 2, z0 = 0.2 , z = 0.02) # Neutral wind speed profile
 ```
@@
 ```{r}
-mass = 10 # Mass (g)
-a = 0.9 # Solar absorptivity (proportion)
+mass <- 10  # Mass (g)
+a    <- 0.9 # Solar absorptivity (proportion)
 
-# Estimate surface area and the proportion sihouette area
-A = surface_area_from_mass(mass, "lizard") # surface area (m^2)
-psa = sapply(psi_deg, proportion_silhouette_area, taxon = "lizard", posture = "prostrate") 
+# Estimate surface area (m^2) and the proportion sihouette area
+
+A   <- surface_area_from_mass(mass, "lizard")  
+psa <- sapply(psi_deg, proportion_silhouette_area, taxon = "lizard", posture = "prostrate")
+
 # Change negative values to zero
@@
 ```{r}
-Qabs = rep(NA, 24)
-for(hour in 1:24) {
+Qabs <- rep(NA, 24)
+for (hour in 1:24) {
 
-  Qabs[hour] = Qradiation_absorbed(a = a, A = A, psa_dir = psa[hour], psa_ref = 1 - psa[hour], S_dir = Sdir[hour], S_dif = Sdif[hour], a_s = rho)
+  Qabs[hour] <- Qradiation_absorbed(a = a, A = A, psa_dir = psa[hour], psa_ref = 1 - psa[hour], S_dir = Sdir[hour], S_dif = Sdif[hour], a_s = rho)
 
@@
 ```{r}
-epsilon_s = 0.965 # Surface emissivity of lizards
+epsilon_s <- 0.965 # Surface emissivity of lizards
 
-Qemit = rep(NA, 24)
-for(hour in 1:24) {
+Qemit <- rep(NA, 24)
+for (hour in 1:24) {
 
@@
   # A       = surface area (m^2)
-  Qemit[hour] = Qemitted_thermal_radiation(epsilon = epsilon_s, A = A, psa_dir = psa[hour], psa_ref = 1-psa[hour], T_b = Ta_liz[hour] + 273.15, T_g = Ts[hour] + 273.15, T_a = Ta_liz[hour] + 273.15, enclosed = FALSE)
+
+  Qemit[hour] <- Qemitted_thermal_radiation(epsilon = epsilon_s, A = A, psa_dir = psa[hour], psa_ref = 1-psa[hour], T_b = Ta_liz[hour] + 273.15, T_g = Ts[hour] + 273.15, T_a = Ta_liz[hour] + 273.15, enclosed = FALSE)
 
@@
 ```{r}
-ap = airpressure_from_elev(elev) * 1000 # Barometric pressure (pascal)
+ap <- airpressure_from_elev(elev) * 1000 # Barometric pressure (pascal)
 
 # Use DRYAIR from NicheMapR to estimate the thermal conductivity of air and kinematic viscosity
-DRYAIRout = DRYAIR(db = Ta, bp = ap, alt = elev)
-K = DRYAIRout$thcond # Thermal conductivity (Wm^-2K^-1) 
-nu = DRYAIRout$viskin # Kinematic viscosity (m2 s-1)
 
-svl = 0.006 # Approximate snout vent length (meters) for Sceloporus, used as the characteristic dimension for heat transfer
+DRYAIRout <- DRYAIR(db = Ta, bp = ap, alt = elev)
+
+K  <- DRYAIRout$thcond # Thermal conductivity (Wm^-2K^-1) 
+nu <- DRYAIRout$viskin # Kinematic viscosity (m2 s-1)
+
+svl <- 0.006 # Approximate snout vent length (meters) for Sceloporus, used as the characteristic dimension for heat transfer
 
 # We will use the average of K and nu across the day for simplicity and since there is not a substantial diurnal variation.
-K = mean(K) # Average thermal conductivity across the day
-nu = mean(nu) # Average kinematic viscosity
+
+K  <- mean(K)  # Average thermal conductivity across the day
+nu <- mean(nu) # Average kinematic viscosity
 
 # Estimate the heat transfer coefficient using an empirical relationship for lizards
-H_L = heat_transfer_coefficient(V = V_liz, D = svl, K = K, nu = nu, taxon = "lizard_surface")
+
+H_L <- heat_transfer_coefficient(V = V_liz, D = svl, K = K, nu = nu, taxon = "lizard_surface")
 
 # Estimate the heat transfer coefficient using a spherical approximation
-H_L2= heat_transfer_coefficient_approximation(V = V_liz, D = svl, K = K, nu = nu, taxon = "lizard")
+
+H_L2 <- heat_transfer_coefficient_approximation(V = V_liz, D = svl, K = K, nu = nu, taxon = "lizard")
 
 # Estimate the heat transfer coefficient using a simplified version of the approximation
-H_L3= heat_transfer_coefficient_simple(V = V_liz, D = svl, type = "Gates")
+
+H_L3 <- heat_transfer_coefficient_simple(V = V_liz, D = svl, type = "Gates")
 ```
@@
 ```{r}
-Qconv = rep(NA, 24)
-for(hour in 1:24) {
+Qconv <- rep(NA, 24)
+for (hour in 1:24) {
 
-  Qconv[hour] = Qconvection(T_a = Ta_liz[hour] + 273.15, T_b = Ta_liz[hour] + 273.15, H = H_L, A = A, proportion = 0.67, ef = 1.3)
+  Qconv[hour] <- Qconvection(T_a = Ta_liz[hour] + 273.15, T_b = Ta_liz[hour] + 273.15, H = H_L, A = A, proportion = 0.67, ef = 1.3)
 
@@
 ```{r}
-Qcond = rep(NA, 24)
+Qcond <- rep(NA, 24)
 for(hr in 1:24) {
 
-  Qcond[hr] = Qconduction_animal(T_g = Ts[hr] + 273.15, T_b = Ta_liz[hr] + 273.15, d = 10^-6, K = K, A = A, proportion = 0.33)
+  Qcond[hr] <- Qconduction_animal(T_g = Ts[hr] + 273.15, T_b = Ta_liz[hr] + 273.15, d = 10^-6, K = K, A = A, proportion = 0.33)
 
@@
 ```{r}
-Qmet = 0; 
-Qevap= 0;
+Qmet  <- 0 
+Qevap <- 0
 ```
@@
 ```{r}
-Qnet = Qnet_Gates(Qabs = Qabs, Qemit = Qemit, Qconv = Qconv, Qcond = Qcond, Qmet = Qmet, Qevap = Qevap)
+Qnet <- Qnet_Gates(Qabs = Qabs, Qemit = Qemit, Qconv = Qconv, Qcond = Qcond, Qmet = Qmet, Qevap = Qevap)
 ```
@@
 ```{r}
-Te = rep(NA, 24)
-for(hour in 1:24) {
+Te <- rep(NA, 24)
+for (hour in 1:24) {
 
-  Te[hour]= Tb_Gates(A = A, D = svl, psa_dir = psa[hour], psa_ref = 1 - psa[hour], psa_air = 0.67, psa_g = 0.33, T_g = Ts[hour] + 273.15, T_a = Ta_liz[hour] + 273.15, Qabs = Qabs[hour], epsilon = epsilon_s, H_L = H_L, ef = 1.3, K = K)
+  Te[hour] <- Tb_Gates(A = A, D = svl, psa_dir = psa[hour], psa_ref = 1 - psa[hour], psa_air = 0.67, psa_g = 0.33, T_g = Ts[hour] + 273.15, T_a = Ta_liz[hour] + 273.15, Qabs = Qabs[hour], epsilon = epsilon_s, H_L = H_L, ef = 1.3, K = K)
 
@@
 ```{r}
-Te2 = rep(NA, 24)
-for(hr in 1:24) {
+Te2 <- rep(NA, 24)
+for (hr in 1:24) {
 
   # S is solar radiation flux (W m^-2), so we divide by surface area, A
-  Te2[hr] = Tb_CampbellNorman(T_a = Ta_liz[hr] + 273.15, T_g = Ts[hr] + 273.15, S = Qabs[hr] / A, alpha_L = 0.96, epsilon = epsilon_s, c_p = 29.3, D = svl, V = V_liz)
+
+  Te2[hr] <- Tb_CampbellNorman(T_a = Ta_liz[hr] + 273.15, T_g = Ts[hr] + 273.15, S = Qabs[hr] / A, alpha_L = 0.96, epsilon = epsilon_s, c_p = 29.3, D = svl, V = V_liz)
 
@@
 ```{r}
-Te3 = rep(NA, 24)
-for(hour in 1:24) {
+Te3 <- rep(NA, 24)
+for (hour in 1:24) {
 
   # F_d, F_r, F_a, and F_g are the view factors between the surface of the lizard and diffuse solar radiation, reflected solar radiation, and atmospheric thermal radiation, and ground thermal radiation, respectively.
-  Te3[hour]= Tb_lizard(T_a = Ta_liz[hour], T_g = Ts[hour], u = V_liz, svl = svl * 1000, m = mass, psi = psi_deg[hour], rho_S = rho, elev = elev, doy = doy, sun = TRUE, surface = TRUE, alpha_S = a, alpha_L = 0.965, epsilon_s = epsilon_s, F_d = 0.8, F_r = 0.5, F_a = 0.5, F_g = 0.5)
+
+  Te3[hour] <- Tb_lizard(T_a = Ta_liz[hour], T_g = Ts[hour], u = V_liz, svl = svl * 1000, m = mass, psi = psi_deg[hour], rho_S = rho, elev = elev, doy = doy, sun = TRUE, surface = TRUE, alpha_S = a, alpha_L = 0.965, epsilon_s = epsilon_s, F_d = 0.8, F_r = 0.5, F_a = 0.5, F_g = 0.5)
 
@@
 ```{r, echo=FALSE}
-par(mar=c(5, 4, 4, 8) + 0.1)
-plot(1:24, (Te-273.15), type="l", xlab="Hour", ylab= expression("Temperature ("*~degree*C*")"), col="#5DC863FF", ylim=c(10,60)) # Gates
-points(1:24, Te2-273.15, type="l", col="#5DC863FF", lty="dotted") # Campbell and Norman
-points(1:24, Te3, type="l", col="#5DC863FF", lty="dashed") # Lizard, Buckley 2008
-points(1:24, Ta, type="l", col="#440154FF")
-points(1:24, Ta_liz, type="l", col="#3B528BFF")
-points(1:24, Ts, type="l", col="#21908CFF")
+par(mar = c(5, 4, 4, 8) + 0.1)
 
-#add additional axis with radiation
-par(new = T)
-plot(1:24, Sdir, pch=16, axes=F, xlab=NA, ylab=NA, type="l", col="#FDE725FF")
-axis(side = 4)
+plot(1:24, Te - 273.15, type = "l", xlab = "Hour", ylab = expression("Temperature ("*~degree*C*")"), col = "#5DC863FF", ylim = c(10,60), las = 1) # Gates
+points(1:24, Te2 - 273.15, type ="l", col = "#5DC863FF", lty = "dotted") # Campbell and Norman
+points(1:24, Te3, type = "l", col = "#5DC863FF", lty = "dashed") # Lizard, Buckley 2008
+points(1:24, Ta, type = "l", col = "#440154FF")
+points(1:24, Ta_liz, type = "l", col = "#3B528BFF")
+points(1:24, Ts, type = "l", col = "#21908CFF")
+
+# add additional axis with radiation
+
+par(new = TRUE)
+plot(1:24, Sdir, pch = 16, axes = FALSE, xlab = NA, ylab = NA, type = "l", col = "#FDE725FF")
+axis(side = 4, las = 1)
 mtext(side = 4, line = 3, 'Radiation (W/m^2)')
 legend("topleft",
-       legend=c("Ta (2m)", "Ta (0.02m)", "Ts", "Te", "Radiation"),
-       lty=1, pch=NA, col=c("#440154FF", "#3B528BFF", "#21908CFF", "#5DC863FF", "#FDE725FF"))
+       legend = c("Ta (2m)", "Ta (0.02m)", "Ts", "Te", "Radiation"),
+       lty = 1, pch = NA, col = c("#440154FF", "#3B528BFF", "#21908CFF", "#5DC863FF", "#FDE725FF"))
 ```

### 2022-06-28 Manuscript_full.Rmd
Clean up code chunks

@@
 ---
-title: 'TrenchR: an R package for transparent environmental and ecological biophysics'
+title: 'TrenchR: an R package for transparent microclimate and biophysical ecology'
 author: 
@@
 
-doy   <- day_of_year("2021-06-01", format = "%Y-%m-%d") # Julian Date (DOY)
+doy   <- day_of_year("2021-06-01", format = "%Y-%m-%d") # Day of year
 
@@
 
-# Separate solar radiation into direct, diffuse, and reflected
+# Separate solar radiation into direct, diffuse, and reflected components
 
-Sdir <- Srad[1,] # Direct solar radiation
-Sdif <- Srad[2,] # Diffuse solar radiation
-Sref <- Srad[3,] # Reflected solar radiation
+Sdir <- Srad[1,] # Direct solar radiation (W/m2)
+Sdif <- Srad[2,] # Diffuse solar radiation  (W/m2)
+Sref <- Srad[3,] # Reflected solar radiation  (W/m2)
 ```
@@
 
-# Soil temperature (C)
+# Soil surface temperature (C)
 
@@
 
-At the second stage, we use microclimate models to scale air temperature and wind speed from weather station height (2 m) to lizard height (0.02 m). 
+At the second stage, we use microclimate models to scale air temperature ($T_r$) and wind speed ($u_r$) from weather station height (reference height $z_r$= 2 m) to lizard height (organism height $z$= 0.02 m).
 We assume a surface roughness of $z_0 = 0.2$ m, which corresponds to bare sand and determines the turbulence of airflow. 
-We implement a neutral air temperature profile driven by density differences, an air temperature profile forced by wind speed, and a neutral wind speed profile.
+We implement neutral air temperature and wind speed profiles driven by density differences but profiles forced by wind speed are also available.
 
 ```{r}
-# Scale air temperature to lizard height
-# T_r = reference temperature
-# zr  = reference height
-# z0  = surface roughness
-# z   = organism height
-# T_s = surface temperature
+# Neutral air temperature profile
 
-Ta_liz  <- air_temp_profile_neutral(T_r = Ta, zr = 2, z0 = 0.2, z = 0.02, T_s = Ts)  # Neutral air temperature profile
-Ta_liz2 <- air_temp_profile(T_r = Ta, u_r = V, zr = 2, z0 = 0.2, z = 0.02, T_s = Ts) # Air temperature profile forced by wind speed
+Ta_liz  <- air_temp_profile_neutral(T_r = Ta, zr = 2, z0 = 0.2, z = 0.02, T_s = Ts)
 
-# Scale wind speed to lizard height
-# u_r = reference wind speed
-# zr  = reference height
-# z0  = surface roughness
-# z   = organism height
+# Neutral wind speed profile
 
-V_liz <- wind_speed_profile_neutral(u_r = V, zr = 2, z0 = 0.2 , z = 0.02) # Neutral wind speed profile
+V_liz <- wind_speed_profile_neutral(u_r = V, zr = 2, z0 = 0.2 , z = 0.02) 
 ```
@@
 Estimating $Q_{abs}$ requires the surface area exposed to radiation and the solar absorptivity of the animal surface ($a$ proportion).
+We use zenith angle $psi$ to estimate the projected (silhouette) area as a portion of the surface area of the organism, which allows estimating absorbed solar radiation.
 We model a 10 gram *Sceloporus* lizard with solar absorptivity $a = 0.9$ [@gates1980biophysical]. 
@@
 
-We estimate thermal radiation $Q_{emit}$ (W) for the lizard outdoors. 
+We estimate thermal radiation $Q_{emit}$ (W) for the lizard outdoors, where $psa_{dir}$ and $psa_{ref}$ are the proportions of surface area $A$ ($m^2$) exposed to direct and reflected solar radiation, respectively.
 We assume the surface emissivity of lizards, $epsilon_s = 0.965$ [@barlett1967].
@@
 
-  # psa_dir = proportions of surface area exposed to direct solar radiation
-  # psi_ref = proportions of surface area exposed to reflected solar radiation
-  # A       = surface area (m^2)
-
   Qemit[hour] <- Qemitted_thermal_radiation(epsilon = epsilon_s, A = A, psa_dir = psa[hour], psa_ref = 1-psa[hour], T_b = Ta_liz[hour] + 273.15, T_g = Ts[hour] + 273.15, T_a = Ta_liz[hour] + 273.15, enclosed = FALSE)
@@
 We will estimate the lizard's heat transfer coefficient, $H_L$ ($Wm^{-2}K^{-1}$) using an empirical relationship for lizards (`heat_transfer_coefficient()`).
+We average thermal conductivity and kinematic viscosity across the day for simplicity and since there is not substantial diurnal variation.
 We also illustrate a function estimating $H_L$ using a spherical approximation (`heat_transfer_coefficient_approximation()`) and a simplified approximation (`heat_transfer_coefficient_simple()`) for cases when taxon specific relationships for estimating heat transfer coefficients are not available.
@@
 ```{r}
+# Use DRYAIR from NicheMapR to estimate the thermal conductivity of air and kinematic viscosity
+
 ap <- airpressure_from_elev(elev) * 1000 # Barometric pressure (pascal)
-
-# Use DRYAIR from NicheMapR to estimate the thermal conductivity of air and kinematic viscosity
 
@@
 
-K  <- DRYAIRout$thcond # Thermal conductivity (Wm^-2K^-1) 
-nu <- DRYAIRout$viskin # Kinematic viscosity (m2 s-1)
+K  <- mean(DRYAIRout$thcond) # Thermal conductivity (Wm^-2K^-1) 
+nu <- mean(DRYAIRout$viskin) # Kinematic viscosity (m2 s-1)
 
 svl <- 0.006 # Approximate snout vent length (meters) for Sceloporus, used as the characteristic dimension for heat transfer
-
-# We will use the average of K and nu across the day for simplicity and since there is not a substantial diurnal variation.
-
-K  <- mean(K)  # Average thermal conductivity across the day
-nu <- mean(nu) # Average kinematic viscosity
 
@@
 We implement the function in R assuming that 2/3 of the lizard's surface area is exchanging heat through convection. 
-We currently estimate no convection since we are assuming $Tb = Ta$ for illustration purposes:
+We currently estimate no convection since we are assuming $T_b = T_a$ for illustration purposes:
 
@@
 
-We additionally estimate $T_b$ using a specialized function for lizards [@buckley2008link]:
+We additionally estimate $T_b$ using a specialized function for lizards [@buckley2008link], where $F_d$, $F_r$, $F_a$, and $F_g$ are the view factors between the surface of the lizard and diffuse solar radiation, reflected solar radiation, atmospheric thermal radiation, and ground thermal radiation, respectively:
 
@@
 for (hour in 1:24) {
-
-  # F_d, F_r, F_a, and F_g are the view factors between the surface of the lizard and diffuse solar radiation, reflected solar radiation, and atmospheric thermal radiation, and ground thermal radiation, respectively.
 

### 2022-07-01 ManuscriptTe.Rmd
Update code chunks in appendix.

@@
 
-# APPENDIX S1
-
 We provide an additional example using microclimate data measured along an elevation gradient in CO, USA to illustrate use of TrenchR. 
@@
 We then use a time series of air and surface temperatures, wind speeds, and solar radiation collected at a single height to implement a biophysical model for grasshoppers and compare estimates to observations.
-R markdown files with the full code and the associated data are available at https://github.com/trenchproject/TrenchRmanuscript/.
+We omit the code to read and process the environmental data here for brevity.
+R markdown files with the full code and the associated environmental data are available at https://github.com/trenchproject/TrenchRmanuscript/.
 
@@
 We first examine how temperature and wind speed profiles can be used to scale conditions to the heights of organisms. 
-We start by using wind speed variation across hours and heights to estimate surface roughness.
+We start by using wind speeds ($u_r$, m/s) at a vector of reference heights ($z_r$, m) to estimate surface roughness ($z_0$).
 
 ```{r}
-#estimate surface roughness based on wind velocity (m/s) at a vector of reference heights, zr (m)
 z0= apply(C1.hr[, c("Wind0.25m","Wind0.5m","Wind1m","Wind1.5m")], FUN=surface_roughness, MARGIN=1, zr=c(0.25,0.5,1.0,1.5))
-
 ```
 
-We then calculate the temperature and wind speed profiles for a given hour using multiple, alternative functions in TrenchR.
+We then calculate the temperature and wind speed profiles for a given hour using multiple, alternative functions in TrenchR that assume both neutral and forced flows and treat the profile as a single or multiple segments.
+The functions require temperatures ($T_r$, $^\circ$C) and wind speeds ($U_r$, $m s^{-1}$) at reference heights ($z_r$, $m$), $z_0$, surface temperature ($T_s$, $^\circ$C), and the desired height for scaling ($z$, m).
 
@@
 
-#estimate profiles using three different functions including the assumption of neutral and segmented profiles. 
-#T_r	and u_r are temperatures (C) and windspeeds (m/s), respectively, at reference height zr (m).
-#z0	is surface roughness (m); z is the height to scale to (m).
-#T_s	is surface temperature (C).
-
 #temperature profiles
@@
 
-#windspeed profiles
+#wind speed profiles
 u.seq1= sapply(z.seq, FUN=wind_speed_profile_neutral, u_r=C1.sub[hr,"Wind0.5m"], zr=0.5, z0=z0[hr])
 
-#u.seq2= sapply(z.seq, FUN=wind_speed_profile, u_r=C1.sub[hr,"Wind0.5m"], zr=0.5, z0=z0[hr])
-
-u.seq3= sapply(z.seq, FUN=wind_speed_profile_segment, u_r=C1.sub[hr, c("Wind0.5m","Wind1m","Wind1.5m")], zr=c(0.5,1,1.5), z0=rep(z0[hr],3))
+u.seq2= sapply(z.seq, FUN=wind_speed_profile_segment, u_r=C1.sub[hr, c("Wind0.5m","Wind1m","Wind1.5m")], zr=c(0.5,1,1.5), z0=rep(z0[hr],3))
 ```
@@
 The temperature and wind speed profiles illustrate how temperatures generally decrease and wind speeds generally increase with height above the ground (Figure S1).
-Profiles can be fit as a single smooth function or segments.
 
@@
 plot(u.seq1, z.seq, type = "l", xlab = "windspeed (m/s)", ylab = "z (height above ground in m)")
-#points(u.seq2, z.seq, type = "l", lty="dashed")
-points(u.seq3, z.seq, type = "l", lty="dotted")
+points(u.seq2, z.seq, type = "l", lty="dotted")
 points(C1.sub[hr, c("Wind0.25m","Wind0.5m","Wind1m","Wind1.5m")], c(0.25,0.5,1,1.5), type = "p", col="red")
@@
 *Figure S1.
-Temperature (top) and wind speed (bottom) profiles using similar continuous [solid: air_temp_profile_neutral(); dashed: air_temp_profile()] and segmented estimation [dotted: air_temp_profile_segment()].
+Temperature (top) and wind speed (bottom) profiles using similar continuous [solid: `air_temp_profile_neutral()`; dashed: `air_temp_profile()`] and segmented estimation [dotted: `air_temp_profile_segment()`].
 The points indicate empirical points used for parameterization.*
@@
 We next illustrate quantifying operative environmental temperatures, which are estimated body temperatures for grasshoppers in the specified environments. 
-We read and process time series of environmental data to estimate the zenith angle, the angle between the sun's rays and the vertical, across days of the year and hours.
+Assuming we have read and processed the time series of environmental data, we first estimate the zenith angle ($psi$, $^\circ$), the angle between the sun's rays and the vertical, as a function of day of year ($doy$), latitude ($lat$, $^\circ$), longitude ($lon$, $^\circ$), and hour.
 
 ```{r}
-#Calculate zenith
-dat$psi= zenith_angle(dat$J,lat=dat$lat, lon=dat$lon, hour=dat$hour)
+
+dat$psi= zenith_angle(doy=dat$J,lat=dat$lat, lon=dat$lon, hour=dat$hour)
 
@@
 
-We scale the air temperature and wind speed from a reference height of 0.5m to grasshopper height (0.001m).
+We scale the air temperature ($T_r$, $^\circ$C) and wind speed ($u_r$, m/s) from a reference height of $z_r$=0.5m to grasshopper height ($z$=0.001m).
+The functions also require surface temperature ($T_s$, $^\circ$C), and surface roughness ($z0$, m).
 
@@
 
-#scale temperature T_r from reference height zr=0.5m to grasshopper height based on windspeed u_r, surface temperature T_s, and surface roughness z0.
+#scale temperature 
 dat$Tgrass <- air_temp_profile(T_r=dat$Temp, u_r=dat$Wind, zr=0.5, z0=0.02, z=z, T_s=dat$SoilTemp)
 
-#scale wind to grasshopper height
+#scale wind speed
 dat$Ugrass= wind_speed_profile_neutral(u_r=dat$Wind, zr=0.5, z0=0.02,z=z)
@@
 
-We then use biophysical models to estimate body temperatures based on air and surface temperatures, wind speed, and solar radiation.
-We first use a biophysical model specifically configured for grasshoppers.
+We then use biophysical models to estimate body temperatures ($T_b$, $^\circ$C) based on air ($T_a$, $^\circ$C) and surface temperatures ($T_g$, $^\circ$C), wind speed ($u$, m/s), and solar radiation ($H$, $W m^{-2}$).
+We first use a biophysical model specifically configured for grasshoppers where $K_t$	is the clearness index (dimensionless), which is the ratio of the global solar radiation measured at the surface to the total solar radiation at the top of the atmosphere; $l$	is grasshopper length (m); $A_{condfact}$ is the proportion of the grasshopper surface area in contact with the ground; $abs$ is absorptivity of grasshopper to solar radiation (proportion, $abs$=0.9 corresponds to dark colors); and $r_g$ is substrate solar reflectivity (proportion).
+
 ```{r, warning=FALSE}
 
-##Calculate operative temperature, Te
 dat$Te<-Tb_grasshopper(T_a=dat$Tgrass, T_g=dat$SoilTemp, u=dat$Ugrass, H=dat$Rad, K_t=1, psi=dat$psi, l=0.0211, Acondfact=0.0, abs=0.9, r_g=0.5)
 
-#T_a	and T_g are air and surface temperatures (C), respectively; u	is wind speed (m/s); 
-#H is total (direct + diffuse) solar radiation flux (W/m^2);
-#K_t	is the clearness index (dimensionless), which is the ratio of the global solar radiation measured at the surface to the total solar radiation at the top of the atmosphere;
-#psi is solar zenith angle (degrees); l	is grasshopper length (m);
-#Acondfact is the proportion of the grasshopper surface area in contact with the ground;
-#abs is absorptivity of grasshopper to solar radiation (proportion); abs=0.9 corresponds to dark colors.
-#r_g is substrate solar reflectivity (proportion).
 ```
@@
 We then implement a generic energy budget. 
-We first illustrate the estimation of convective heat transfer coefficients as a function of wind speed.
+We first illustrate the estimation of convective heat transfer coefficients ($H_L$, $W K^{-1} m^{-2}$) as a function of wind speed ($V$, $m s^{-1}$) based on empirical measurements, where $D$ is the characteristic dimension (m), $K$ is the thermal conductivity of air ($W m^{-1} K^{-1}$), and $nu$ is the kinematic viscosity of air ($m^2 s^{-1}$).
+We subsequently estimate $H_L$ using a spherical approximation and a simplified version of the approximation, respectively.
 
 ```{r}
-#Estimate heat transfer coefficient using empirical relationship
 dat$H_L=heat_transfer_coefficient(V=dat$Ugrass,D=0.0211,K=0.025, nu= 15.3 * 10^(-6) , taxon="cylinder")
-#where V is	air velocity (m/s); D is the characteristic dimension (m);
-#K is the thermal conductivity of air (W m^-1 K^-1);
-#nu is the kinematic viscosity of air (m^2 s^-1)
 
-#Also estimate using a spherical approximation and a simplified version of the approximation.
 H_L1= heat_transfer_coefficient_approximation(V=dat$Ugrass,D=0.0211,K= 0.025, nu= 15.3 * 10^(-6), taxon="flyinginsect")
+
 H_L2= heat_transfer_coefficient_simple(V=dat$Ugrass,D=0.0211, type="Gates")
 
-#surface area from length(m)
-A= surface_area_from_length(l=0.0211)
 ```
 
-We then implement the energy budget as follows. 
+We then implement the energy budget as follows, where $A$ is surface area ($m^2$), $Q_{abs}$ is the solar and thermal radiation absorbed (W), $epsilon$ is longwave infrared emissivity of skin (proportion), and $ef$ is the enhancement factor that is used to adjust $H_L$ to field conditions.
+The proportions $psa_{dir}$, $psa_{ref}$, $psa_{air}$, and $psa_{g}$ are the proportions surface area exposed to direct radiation from the sky (or enclosure), reflected radiation from the ground, and air, and in contact with the ground, respectively.
 We note that the energy budget can not be solved for some time periods and errors are generated.
@@
 TeGates= rep(NA, nrow(dat) )
+
+A= surface_area_from_length(l=0.0211) #surface area from length(m)
 
@@
 }
-#where psa_dir, psa_ref, psa_air, psa_g are the proportions surface area exposed to direct radiation from the sky (or enclosure), reflected radiation from the ground, and air, and in contact with the ground, respectively;
-#Qabs is the solar and thermal radiation absorbed (W);
-#epsilon is longwave infrared emissivity of skin (proportion);
-#H_L is the convective heat transfer coefficient (W m^-2 K^-1);
-#ef	is the enhancement factor, used to adjust H_L to field conditions;
-#K is thermal conductivity. 
+
 ```
@@
 
-*Figure S2. Diurnal temperature variation averaged across days in August varies across the elevations of CO sites (panels) and the type of temperature (C). 
+*Figure S2. Diurnal temperature variation averaged across days in August varies across the elevations of CO sites (panels) and the type of temperature ($^\circ$C). 
 Microclimate models indicate that the air temperatures experienced by grasshoppers near the ground (Ta grasshopper height) is generally above air temperatures (Ta) and close to surface temperatures (Ts).

### 2022-07-01 Manuscript_full.Rmd
Update code chunks in appendix.

@@
 
-LB conceived and designed the package and lead all aspect of the packages creation and dissemination; BBO, IC, AJ, OL, AM, ER, YS, and JS wrote and edited functions and associated package resources; AJ and JS configured the package; AM, BBO, YS, and LB converted and edited the educational modules.
-All authors contributed critically to the drafts and gave final approval for publication.
+LB conceived and designed the package and lead all aspect of the packages creation and dissemination.
+All authors wrote and edited functions and associated package resources, contributed critically to manuscript drafts, and gave final approval for publication.
+AJ and JS configured the package.
+AM, BBO, YS, and LB converted and edited the educational modules.
 

### 2022-07-05 ManuscriptTe.Rmd
ms edits

@@
   html_document: default
-  pdf_document: default
   word_document: default
-  rmarkdown::html_vignette: default
+  pdf_document: 
+    pandoc_args: --listings
+    includes:
+      in_header: preamble.tex
 ---
@@
 ```{r, include=FALSE}
-C1all=read.csv("./data/microclimate_C1.csv", sep=",", na.strings = c("-49.99","NA"))
+C1all <- read.csv("./data/microclimate_C1.csv", sep = ",", na.strings = c("-49.99","NA"))
 ```
@@
 ```{r, include=FALSE}
-#ANALYZE SURFACE ROUGHNESS AND PROFILES
-C1all$day=as.POSIXlt(as.character(C1all$datetime), tz='MST', format="%m/%d/%Y %H:%M")$mday
-C1all$hour=as.POSIXlt(as.character(C1all$datetime), tz='MST', format="%m/%d/%Y %H:%M")$hour
+# ANALYZE SURFACE ROUGHNESS AND PROFILES
+C1all$day <- as.POSIXlt(as.character(C1all$datetime), tz = 'MST', format = "%m/%d/%Y %H:%M")$mday
+C1all$hour <- as.POSIXlt(as.character(C1all$datetime), tz = 'MST', format = "%m/%d/%Y %H:%M")$hour
 
-C1.dayhr<- aggregate(C1all[,c(2:8,13:22)], by=list(C1all$day, C1all$hour), FUN=mean, na.rm=TRUE)
-names(C1.dayhr)[1:2]=c("day","hour")
-C1.hr<- aggregate(C1all[,c(2:8,13:22)], by=list(C1all$hour), FUN=mean, na.rm=TRUE)
-names(C1.hr)[1]="hour"
+C1.dayhr <- aggregate(C1all[,c(2:8,13:22)], by = list(C1all$day, C1all$hour), FUN = mean, na.rm = TRUE)
+names(C1.dayhr)[1:2] <- c("day","hour")
+C1.hr <- aggregate(C1all[,c(2:8,13:22)], by = list(C1all$hour), FUN = mean, na.rm = TRUE)
+names(C1.hr)[1] <- "hour"
 
-C1.hr= as.matrix(C1.hr)
+C1.hr <- as.matrix(C1.hr)
 ```
@@
 ```{r}
-z0= apply(C1.hr[, c("Wind0.25m","Wind0.5m","Wind1m","Wind1.5m")], FUN=surface_roughness, MARGIN=1, zr=c(0.25,0.5,1.0,1.5))
+z0 <- apply(C1.hr[, c("Wind0.25m", "Wind0.5m", "Wind1m", "Wind1.5m")], FUN = surface_roughness, MARGIN = 1, zr = c(0.25, 0.5, 1.0, 1.5))
 ```
@@
 ```{r}
-#subset to day and hour
-C1.sub= subset(C1.dayhr, C1.dayhr$day==10)
-hr=10
+# subset to day and hour
+C1.sub <- subset(C1.dayhr, C1.dayhr$day == 10)
+hr <- 10
 
-#vector of heights
+# vector of heights
 z.seq <- seq(0, 2, by = 0.1)
 
-#temperature profiles
-t.seq1= sapply(z.seq, FUN=air_temp_profile_neutral, T_r=C1.sub[hr,"Air0.5m"], z0=z0[hr], zr=0.5, T_s=C1.sub[hr,"Soil"])
+# temperature profiles
+t.seq1 <- sapply(z.seq, FUN = air_temp_profile_neutral, T_r = C1.sub[hr, "Air0.5m"], z0 = z0[hr], zr = 0.5, T_s = C1.sub[hr, "Soil"])
 
-t.seq2= sapply(z.seq, FUN=air_temp_profile, T_r=C1.sub[hr,"Air0.5m"], u_r=C1.hr[hr,"Wind0.5m"], zr=0.5, z0=z0[hr], T_s=C1.sub[hr,"Soil"])
+t.seq2 <- sapply(z.seq, FUN = air_temp_profile, T_r = C1.sub[hr, "Air0.5m"], u_r = C1.hr[hr, "Wind0.5m"], zr = 0.5, z0 = z0[hr], T_s = C1.sub[hr, "Soil"])
 
-t.seq3= sapply(z.seq, FUN=air_temp_profile_segment, T_r=C1.sub[hr, c("Air0.25m","Air0.5m","Air1.0m","Air1.5m")],u_r=C1.sub[hr, c("Wind0.25m","Wind0.5m","Wind1m","Wind1.5m")], zr=c(0.25,0.5,1,1.5), z0=rep(z0[hr],4), T_s=C1.sub[hr,"Soil"])
+t.seq3 <- sapply(z.seq, FUN = air_temp_profile_segment, T_r = C1.sub[hr, c("Air0.25m", "Air0.5m", "Air1.0m", "Air1.5m")], u_r = C1.sub[hr,  c("Wind0.25m", "Wind0.5m", "Wind1m", "Wind1.5m")], zr=c(0.25, 0.5, 1, 1.5), z0 = rep(z0[hr], 4), T_s = C1.sub[hr, "Soil"])
 
-#wind speed profiles
-u.seq1= sapply(z.seq, FUN=wind_speed_profile_neutral, u_r=C1.sub[hr,"Wind0.5m"], zr=0.5, z0=z0[hr])
+# wind speed profiles
+u.seq1 <- sapply(z.seq, FUN = wind_speed_profile_neutral, u_r = C1.sub[hr, "Wind0.5m"], zr = 0.5, z0 = z0[hr])
 
-u.seq2= sapply(z.seq, FUN=wind_speed_profile_segment, u_r=C1.sub[hr, c("Wind0.5m","Wind1m","Wind1.5m")], zr=c(0.5,1,1.5), z0=rep(z0[hr],3))
+u.seq2 <- sapply(z.seq, FUN = wind_speed_profile_segment, u_r = C1.sub[hr, c("Wind0.5m", "Wind1m", "Wind1.5m")], zr = c(0.5, 1, 1.5), z0 = rep(z0[hr], 3))
 ```
@@
 
-```{r, echo=FALSE}
-#replace infinity with N
-t.seq1[which(!is.finite(t.seq1))]=NA
+```{r, echo=FALSE, fig.height=4, fig.align = 'center', out.width = "75%"}
+# replace infinity with N
+t.seq1[which(!is.finite(t.seq1))] <- NA
 
 plot(t.seq1, z.seq, type = "l", xlab = "Temperature (C)", ylab = "z (height above ground in m)")
-points(t.seq2, z.seq, type = "l", lty="dashed")
-points(t.seq3, z.seq, type = "l", lty="dotted")
-points(C1.sub[hr, c("Air0.25m","Air0.5m","Air1.0m","Air1.5m")], c(0.25,0.5,1,1.5), type = "p", col="red")
+points(t.seq2, z.seq, type = "l", lty = "dashed")
+points(t.seq3, z.seq, type = "l", lty = "dotted")
+points(C1.sub[hr, c("Air0.25m", "Air0.5m", "Air1.0m", "Air1.5m")], c(0.25, 0.5, 1, 1.5), type = "p", col = "red")
 
 plot(u.seq1, z.seq, type = "l", xlab = "windspeed (m/s)", ylab = "z (height above ground in m)")
-points(u.seq2, z.seq, type = "l", lty="dotted")
-points(C1.sub[hr, c("Wind0.25m","Wind0.5m","Wind1m","Wind1.5m")], c(0.25,0.5,1,1.5), type = "p", col="red")
+points(u.seq2, z.seq, type = "l", lty = "dotted")
+points(C1.sub[hr, c("Wind0.25m", "Wind0.5m", "Wind1m", "Wind1.5m")], c(0.25, 0.5, 1, 1.5), type = "p", col = "red")
 ```
@@
 #Grasshopper data
-spec.dat=read.csv("./data/SpecData.csv")
+spec.dat <- read.csv("./data/SpecData.csv")
 
-#read weather data
-dat=read.csv("./data/DatAll_Pace2011.csv", sep=",", na.strings = c("-49.99","NA"))
+# read weather data
+dat <- read.csv("./data/DatAll_Pace2011.csv", sep = ",", na.strings = c("-49.99", "NA"))
 #match site data
-sites=read.csv("./data/SiteElevations.csv", sep=",")
+sites <- read.csv("./data/SiteElevations.csv", sep= ",")
 
-#combine soil temperatures
-dat$SoilTemp= rowMeans(dat[,c("SoilTemp","SoilTemperature")], na.rm = TRUE)
+# combine soil temperatures
+dat$SoilTemp <- rowMeans(dat[,c("SoilTemp", "SoilTemperature")], na.rm = TRUE)
 
-#omit problematic temperatures
-dat$Temp[dat$Temp< -10]=NA
+# omit problematic temperatures
+dat$Temp[dat$Temp< -10] <- NA
 
-#add elevation
-match1= match(dat$site, sites$Site)
-dat$Elev=sites[match1,"Elevation"]
+# add elevation
+match1 <- match(dat$site, sites$Site)
+dat$Elev <- sites[match1, "Elevation"]
 
-#calculate day of year
-date.ct <- strptime(dat$datetime, format="%m/%d/%Y %H:%M", tz="MST")
-dat$J= round(as.numeric(julian(date.ct, origin = as.Date("2011-01-01") )))
-dat$hour= date.ct$hour
+# calculate day of year
+date.ct <- strptime(dat$datetime, format = "%m/%d/%Y %H:%M", tz = "MST")
+dat$J <- round(as.numeric(julian(date.ct, origin = as.Date("2011-01-01"))))
+dat$hour <- date.ct$hour
 
-#match site data
-match1= match(dat$site, sites$Site) 
-dat$lon= sites$Lon[match1]
-dat$lat= sites$Lat[match1]
+# match site data
+match1 <- match(dat$site, sites$Site) 
+dat$lon <- sites$Lon[match1]
+dat$lat <- sites$Lat[match1]
 
-#Subset to August
-dat.dt<- strptime(dat$datetime, format="%m/%d/%Y %H:%M", tz="MST") #as.Date
-dataug= subset(dat, dat.dt$mon==7)
-#subset to day
-dat= subset(dataug, dataug$Rad>5)
+# Subset to August
+dat.dt<- strptime(dat$datetime, format = "%m/%d/%Y %H:%M", tz = "MST") # as.Date
+dataug <- subset(dat, dat.dt$mon == 7)
+
+# subset to day
+dat <- subset(dataug, dataug$Rad > 5)
 ```
@@
 
-dat$psi= zenith_angle(doy=dat$J,lat=dat$lat, lon=dat$lon, hour=dat$hour)
+dat$psi <- zenith_angle(doy = dat$J, lat = dat$lat, lon = dat$lon, hour = dat$hour)
 
@@
 
-We scale the air temperature ($T_r$, $^\circ$C) and wind speed ($u_r$, m/s) from a reference height of $z_r$=0.5m to grasshopper height ($z$=0.001m).
+We scale the air temperature ($T_r$, $^\circ$C) and wind speed ($u_r$, m/s) from a reference height of $z_r$ = 0.5m to grasshopper height ($z$ = 0.001m).
 The functions also require surface temperature ($T_s$, $^\circ$C), and surface roughness ($z0$, m).
@@
 ```{r}
-z<-0.001 #specify distance from ground
+z <- 0.001 #specify distance from ground
 
-#scale temperature 
-dat$Tgrass <- air_temp_profile(T_r=dat$Temp, u_r=dat$Wind, zr=0.5, z0=0.02, z=z, T_s=dat$SoilTemp)
+# scale temperature 
+dat$Tgrass <- air_temp_profile(T_r = dat$Temp, u_r = dat$Wind, zr = 0.5, z0 = 0.02, z = z, T_s = dat$SoilTemp)
 
-#scale wind speed
-dat$Ugrass= wind_speed_profile_neutral(u_r=dat$Wind, zr=0.5, z0=0.02,z=z)
+# scale wind speed
+dat$Ugrass <- wind_speed_profile_neutral(u_r = dat$Wind, zr = 0.5, z0 = 0.02, z = z)
 ```
@@
 ```{r, include=FALSE}
-#cut NAs
-dat= dat[,c("Rad","Temp","Wind","SoilTemp","SanguinipesTemp","Elev","J", "hour","psi","Tgrass","Ugrass")]
-dat= na.omit(dat)
+# cut NAs
+dat <- dat[,c("Rad", "Temp", "Wind", "SoilTemp", "SanguinipesTemp", "Elev", "J", "hour", "psi", "Tgrass", "Ugrass")]
+dat <- na.omit(dat)
 ```
@@
 
-dat$Te<-Tb_grasshopper(T_a=dat$Tgrass, T_g=dat$SoilTemp, u=dat$Ugrass, H=dat$Rad, K_t=1, psi=dat$psi, l=0.0211, Acondfact=0.0, abs=0.9, r_g=0.5)
+dat$Te <- Tb_grasshopper(T_a = dat$Tgrass, T_g = dat$SoilTemp, u = dat$Ugrass, H = dat$Rad, K_t = 1, psi = dat$psi, l = 0.0211, Acondfact = 0.0, abs = 0.9, r_g = 0.5)
 
@@
 ```{r, include=FALSE}
-#fix infinite
-dat$Te[is.infinite(dat$Te)]<-NA
-#ditch big numbers for now
-dat$Te[dat$Te>100]<-NA
+# fix infinite
+dat$Te[is.infinite(dat$Te)] <- NA
+
+# ditch big numbers for now
+dat$Te[dat$Te>100] <- NA
 ```
@@
 ```{r}
-dat$H_L=heat_transfer_coefficient(V=dat$Ugrass,D=0.0211,K=0.025, nu= 15.3 * 10^(-6) , taxon="cylinder")
+dat$H_L <- heat_transfer_coefficient(V = dat$Ugrass, D = 0.0211, K = 0.025, nu = 15.3 * 10^(-6) , taxon = "cylinder")
 
-H_L1= heat_transfer_coefficient_approximation(V=dat$Ugrass,D=0.0211,K= 0.025, nu= 15.3 * 10^(-6), taxon="flyinginsect")
+H_L1 <- heat_transfer_coefficient_approximation(V = dat$Ugrass, D = 0.0211, K = 0.025, nu = 15.3 * 10^(-6), taxon = "flyinginsect")
 
-H_L2= heat_transfer_coefficient_simple(V=dat$Ugrass,D=0.0211, type="Gates")
+H_L2 <- heat_transfer_coefficient_simple(V = dat$Ugrass, D = 0.0211, type = "Gates")
 
@@
 
-```{r echo=T, results = 'hide', warning=FALSE}
-TeGates= rep(NA, nrow(dat) )
+```{r, message=FALSE}
+TeGates <- rep(NA, nrow(dat))
 
-A= surface_area_from_length(l=0.0211) #surface area from length(m)
+# surface area from length(m)
+A <- surface_area_from_length(l = 0.0211)
 
-for(ind in 1:nrow(dat) ){
+for (ind in 1:nrow(dat)) {
 
-  out= try(Tb_Gates(A=A, D=0.0211/3, psa_dir=0.5, psa_ref=1-0.5, psa_air=1, psa_g=0.0, T_g=dat$SoilTemp[ind]+273, T_a=dat$Tgrass[ind]+273, Qabs=A*0.66*dat$Rad[ind], epsilon=0.95, H_L=dat$H_L[ind], ef=1.3, K=0.15), silent=TRUE)
-  if(is.numeric(out))TeGates[ind]=out
+  out <- Tb_Gates(A = A, D = 0.0211/3, psa_dir = 0.5, psa_ref = 1-0.5, psa_air = 1, psa_g = 0.0, T_g = dat$SoilTemp[ind] + 273, T_a = dat$Tgrass[ind] + 273, Qabs = A * 0.66 * dat$Rad[ind], epsilon = 0.95, H_L = dat$H_L[ind], ef = 1.3, K = 0.15)
+
+  if (is.numeric(out)) {
+    TeGates[ind] <- out
+  }
+
 }
@@
 ```{r, echo=FALSE, warning=FALSE}
-dat$TeGates=TeGates-273.15
+dat$TeGates <- TeGates - 273.15
 
-#drop NAs
-dat= dat[which(!is.na(dat$TeGates)),]
+# drop NAs
+dat <-  dat[which(!is.na(dat$TeGates)), ]
 
-#aggregate by hour
-dat.h= suppressWarnings(aggregate(dat, by=list(dat$Elev, dat$hour), FUN=mean, na.rm=TRUE))
-#aggregate by day and hour
-dat.dh= suppressWarnings(aggregate(dat, by=list(dat$Elev, dat$hour, dat$J), FUN=mean, na.rm=TRUE))
+# aggregate by hour
+dat.h <- suppressWarnings(aggregate(dat, by = list(dat$Elev, dat$hour), FUN = mean, na.rm = TRUE))
 
-#convert to long format
-dat.s= melt(dat.h, id = c("Elev", "J","hour","psi"))
+# aggregate by day and hour
+dat.dh <- suppressWarnings(aggregate(dat, by = list(dat$Elev, dat$hour, dat$J), FUN = mean, na.rm = TRUE))
 
-#plot
-dat.pl= subset(dat.s, dat.s$variable %in% c("Temp","SoilTemp","SanguinipesTemp","Te","Tgrass","TeGates") )
-dat.pl$variable= as.character(dat.pl$variable)
+# convert to long format
+dat.s <- melt(dat.h, id = c("Elev", "J", "hour", "psi"))
 
-#change names
-dat.pl[which(dat.pl$variable=="Temp"),"variable"]<-"Ta"
-dat.pl[which(dat.pl$variable=="SoilTemp"),"variable"]<-"Ts"
-dat.pl[which(dat.pl$variable=="SanguinipesTemp"),"variable"]<-"Te observed"
-dat.pl[which(dat.pl$variable=="Te"),"variable"]<-"Te"
-dat.pl[which(dat.pl$variable=="Tgrass"),"variable"]<-"Ta grasshopper height"
-dat.pl[which(dat.pl$variable=="TeGates"),"variable"]<-"Te Gates"
+# plot
+dat.pl <- subset(dat.s, dat.s$variable %in% c("Temp", "SoilTemp", "SanguinipesTemp", "Te", "Tgrass", "TeGates"))
+dat.pl$variable <- as.character(dat.pl$variable)
 
-#make factor
-dat.pl$variable= factor(dat.pl$variable, levels=c("Ta","Ts","Ta grasshopper height","Te","Te Gates","Te observed"))
-#make label
-dat.pl$elev.lab= paste(dat.pl$Elev,"m", sep=" ")
-#make Ta, Te classification
-dat.pl$Ttype="input"
-dat.pl$Ttype[which(dat.pl$variable %in% c("Te","Te Gates","Te observed") )]="estimates"
+# change names
+dat.pl[which(dat.pl$variable=="Temp"),"variable"] <- "Ta"
+dat.pl[which(dat.pl$variable=="SoilTemp"),"variable"] <- "Ts"
+dat.pl[which(dat.pl$variable=="SanguinipesTemp"),"variable"] <- "Te observed"
+dat.pl[which(dat.pl$variable=="Te"),"variable"] <- "Te"
+dat.pl[which(dat.pl$variable=="Tgrass"),"variable"] <- "Ta grasshopper height"
+dat.pl[which(dat.pl$variable=="TeGates"),"variable"] <- "Te Gates"
 
-ggplot(data=dat.pl, aes(x=hour, y = value, color=variable))+geom_line(aes(lty=variable) )+facet_wrap(~elev.lab)+theme_classic()+ theme(legend.position="bottom")+ylim(15,65)
+# make factor
+dat.pl$variable <- factor(dat.pl$variable, levels = c("Ta", "Ts", "Ta grasshopper height", "Te", "Te Gates", "Te observed"))
+
+# make label
+dat.pl$elev.lab <- paste(dat.pl$Elev, "m", sep = " ")
+
+# make Ta, Te classification
+dat.pl$Ttype <- "input"
+dat.pl$Ttype[which(dat.pl$variable %in% c("Te","Te Gates","Te observed") )] <- "estimates"
+
+ggplot(data = dat.pl, aes(x=hour, y = value, color = variable))+ geom_line(aes(lty = variable))+ facet_wrap( ~ elev.lab) + theme_classic()+ theme(legend.position = "bottom") + ylim(15, 65)
 
@@
 ```{r, include=FALSE}
-#plot by date time
-dat.dh$Jhr= dat.dh$J +dat.dh$hour/24
+# plot by date time
+dat.dh$Jhr <- dat.dh$J +dat.dh$hour/24
 
-#convert to long format
-dat.s= melt(dat.dh, id = c("Elev", "J","hour","psi","Jhr"))
+# convert to long format
+dat.s <- melt(dat.dh, id = c("Elev", "J", "hour", "psi", "Jhr"))
 
-#plot
-dat.s= subset(dat.s, dat.s$variable %in% c("Temp","SoilTemp","SanguinipesTemp","Te","Tgrass") )
-#,"TeGates"
+# plot
 
-plot.t= ggplot(data=dat.s, aes(x=Jhr, y = value, color=variable))+geom_line(aes(lty=variable))+facet_grid(Elev~1)
+dat.s <- subset(dat.s, dat.s$variable %in% c("Temp", "SoilTemp", "SanguinipesTemp", "Te", "Tgrass") )
+
+# "TeGates"
+
+plot.t <- ggplot(data = dat.s, aes(x = Jhr, y = value, color=variable)) + geom_line(aes(lty = variable)) + facet_grid(Elev ~ 1)
 ```

### 2022-07-28 ManuscriptTe.Rmd
minor updates

@@
 output:
+  word_document: default
   html_document: default
-  word_document: default
   pdf_document: 

### 2022-07-28 Manuscript_full.Rmd
Update link
minor updates

@@
 Issues and feature requests can be contributed in Github (<https://github.com/trenchproject/TrenchR>). 
-The package is available via Github and has been submitted to CRAN.
+The package is available via CRAN and Github.
 
@@
 
-**Utility functions** calculate environmental metrics that form the basis of microclimate and biophysical models (e.g., zenith and declination angles, which describe the angles of incident sunlight), facilitate modeling diurnal temperature variation, and estimate biologically relevant aggregate metrics such as degree days available for oranismal development.
+**Utility functions** calculate environmental metrics that form the basis of microclimate and biophysical models (e.g., zenith and declination angles, which describe the angles of incident sunlight), facilitate modeling diurnal temperature variation, and estimate biologically relevant aggregate metrics such as degree days available for organismal development.
 
@@
 ## Data Availability
-
-All code and data are available at <https://github.com/trenchproject/TrenchR>.
+TrenchR is available at CRAN and <https://github.com/trenchproject/TrenchR>. All text, code, and data for this manuscript are available at <https://github.com/trenchproject/TrenchRmanuscript/>.
 

### 2022-10-28 ManuscriptTe.Rmd
update Te and add appendix S1

@@
 ---
-title: "Appendix S1"
+title: "Appendix S2"
 author: "Buckley et al."
 output:
+  html_document: default
   word_document: default
-  html_document: default
   pdf_document: 
@@
 We then calculate the temperature and wind speed profiles for a given hour using multiple, alternative functions in TrenchR that assume both neutral and forced flows and treat the profile as a single or multiple segments.
-The functions require temperatures ($T_r$, $^\circ$C) and wind speeds ($U_r$, $m s^{-1}$) at reference heights ($z_r$, $m$), $z_0$, surface temperature ($T_s$, $^\circ$C), and the desired height for scaling ($z$, m).
+The functions require temperatures ($T_r$, $^\circ$C) and wind speeds ($u_r$, $m s^{-1}$) at reference heights ($z_r$, $m$), $z_0$, surface temperature ($T_s$, $^\circ$C), and the desired height for scaling ($z$, m).
 
@@
 # Subset to August
-dat.dt<- strptime(dat$datetime, format = "%m/%d/%Y %H:%M", tz = "MST") # as.Date
+dat.dt<- strptime(dat$datetime, format = "%m/%d/%Y %H:%M", tz = "MST") 
 dataug <- subset(dat, dat.dt$mon == 7)
@@
 
-dat$Te <- Tb_grasshopper(T_a = dat$Tgrass, T_g = dat$SoilTemp, u = dat$Ugrass, H = dat$Rad, K_t = 1, psi = dat$psi, l = 0.0211, Acondfact = 0.0, abs = 0.9, r_g = 0.5)
+dat$Te <- Tb_grasshopper(T_a = dat$Tgrass, T_g = dat$SoilTemp, u = dat$Ugrass, S = dat$Rad, K_t = 1, psi = dat$psi, l = 0.0211, Acondfact = 0.0, abs = 0.9, r_g = 0.5)
 
@@
 ```{r}
-dat$H_L <- heat_transfer_coefficient(V = dat$Ugrass, D = 0.0211, K = 0.025, nu = 15.3 * 10^(-6) , taxon = "cylinder")
+dat$H_L <- heat_transfer_coefficient(u = dat$Ugrass, D = 0.0211, K = 0.025, nu = 15.3 * 10^(-6) , taxon = "cylinder")
 
-H_L1 <- heat_transfer_coefficient_approximation(V = dat$Ugrass, D = 0.0211, K = 0.025, nu = 15.3 * 10^(-6), taxon = "flyinginsect")
+H_L1 <- heat_transfer_coefficient_approximation(u = dat$Ugrass, D = 0.0211, K = 0.025, nu = 15.3 * 10^(-6), taxon = "flyinginsect")
 
-H_L2 <- heat_transfer_coefficient_simple(V = dat$Ugrass, D = 0.0211, type = "Gates")
+H_L2 <- heat_transfer_coefficient_simple(u = dat$Ugrass, D = 0.0211, type = "Gates")
 
@@
 
-  out <- Tb_Gates(A = A, D = 0.0211/3, psa_dir = 0.5, psa_ref = 1-0.5, psa_air = 1, psa_g = 0.0, T_g = dat$SoilTemp[ind] + 273, T_a = dat$Tgrass[ind] + 273, Qabs = A * 0.66 * dat$Rad[ind], epsilon = 0.95, H_L = dat$H_L[ind], ef = 1.3, K = 0.15)
+  out <- Tb_Gates(A = A, D = 0.0211/3, psa_dir = 0.5, psa_ref = 1-0.5, psa_air = 1, psa_g = 0.0, T_g = dat$SoilTemp[ind], T_a = dat$Tgrass[ind], Qabs = A * 0.66 * dat$Rad[ind], epsilon = 0.95, H_L = dat$H_L[ind], ef = 1.3, K = 0.15)
 
@@
 ```{r, echo=FALSE, warning=FALSE}
-dat$TeGates <- TeGates - 273.15
-
 # drop NAs

### 2022-10-28 Manuscript_AppS1.Rmd
update Te and add appendix S1

---
title: "Appendix S1"
author: "Buckley et al."
output:
  html_document: default
  word_document: default
  pdf_document: 
    pandoc_args: --listings
    includes:
      in_header: preamble.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We thank Mike Kearney for making modular R functions corresponding to NicheMapR available and providing code comparing TrenchR and NicheMapR functions during the review process. Our adaptation of the provided code corresponds to the operative temperature estimation in the manuscript. The NicheMapR implementation relies on the micro_global environmental data available through NicheMapR. The modular NicheMapR functions are available in version 3.2.1 via GitHub (https://github.com/mrke/NicheMapR/releases).

```{r, include=FALSE}
##Load TrenchR and NicheMapR package
library(TrenchR)
library(NicheMapR)
library(devtools)

#source_url('https://github.com/mrke/NicheMapR/blob/master/R/GEOM_ecto.R')
#source_url('https://github.com/mrke/NicheMapR/blob/master/R/SOLAR_ecto.R')
```

We provide the input data corresponding to the manuscript example.

```{r}
# Set up input data
lat  <- 35.69    # Latitude (degrees)
lon  <- -105.944 # Longitude (degrees)
elev <- 2121     # Elevation (meters)

# Assumptions
tau <- 0.7 # Atmospheric transmissivity # MRK change from 0.7 to 0.75
rho <- 0.6 # Albedo
Tb0 <- 25  # Initial assumption of body temperature (C)
doy <- day_of_year("2021-06-15", format = "%Y-%m-%d") # Day of year, changed to mid June to correspond to micro_global data
z <- 0.02 # height of animal
psa_g <- 0.33 # fraction of animal surface contacting the ground
K_sub <- 0.5 # parameter for thermal conductivity of substrate, W/mC
```

We then run the NicheMapR microclimate model for clear sky conditions at this site, passing elevation (would otherwise have been 2257m), soil albedo and animal height used in the example. The microclimate model output is prepared as model input.

```{r}
#run microclimate model
micro <- micro_global(loc = c(lon, lat), elev = elev, clearsky = 1, REFL = rho, Usrhyt = z)
metout <- as.data.frame(micro$metout)
soil <- as.data.frame(micro$soil)

# choose mid-June
metout_sub <- subset(metout, DOY == 166)
soil_sub <- subset(soil, DOY == 166)

# use min and max values of air and soil temperature for this day
# and max wind speed
Tmin   <- min(metout_sub$TAREF) # Minimum air temperature (C)
Tmax   <- max(metout_sub$TAREF) # Maximum air temperature (C)
Tmin_s <- min(soil_sub$D0cm) # Minimum soil temperature (C)
Tmax_s <- max(soil_sub$D0cm) # Maximum soil temperature (C)
u      <- max(metout_sub$VREF) # Wind speed (m/s)

snoon <- solar_noon(lon = lon, doy = doy) # Estimate solar noon
dayl  <- daylength(lat = lat, doy = doy)  # Estimate day length

tr    <- snoon - dayl / 2 # Time of sunrise
ts    <- snoon + dayl / 2 # Time of sunset
```

We compare the estimation of zenith angles between TrenchR and NicheMapR.

```{r}
# Estimate zenith angle (degrees)
psi_deg <- sapply(0:23, FUN = zenith_angle, doy = doy, lat = lat, lon = lon)

# compare TrenchR and NMR
plot(seq(0, 23), psi_deg, xlab = 'hour of day')
points(seq(0, 23), metout_sub$ZEN, pch = 3, col = 2)
legend("bottomleft",
       legend = c("TrenchR", "NichemapR"),
       lty = NULL, pch = c(1,3), col = c("black","red"))

# Convert to radians
psi_rad <- degrees_to_radians(psi_deg)
```
We next compare the estimation of radiation, air temperatures, and radiation between the packages.

```{r}
# Estimate radiation
Srad <- sapply(psi_rad, FUN = solar_radiation, doy = doy, tau = tau, elev = elev, rho = rho)
# Separate solar radiation into direct, diffuse, and reflected components
Sdir <- Srad[1,] # Direct solar radiation (W/m2)
Sdif <- Srad[2,] # Diffuse solar radiation  (W/m2)
Sref <- Srad[3,] # Reflected solar radiation  (W/m2)

# compare TrenchR and NMR
plot(seq(0, 23), Sdir + Sdif, xlab = 'hour of day')
points(seq(0, 23), metout_sub$SOLR, pch = 3, col = 2)
legend("topleft",
       legend = c("TrenchR", "NichemapR"),
       lty = NULL, pch = c(1,3), col = c("black","red"))
```

We then compare air and soil temperatures. We note that TrenchR offers several different functions for estimating diurnal temperature variation and also offers more sophisticated functions for estimating soil temperatures than that used here.

```{r}
# Air temperature (C)
Ta <- sapply(1:24, diurnal_temp_variation_sineexp, T_max = Tmax, T_min = Tmin, t_r = tr, t_s = ts, alpha = 2.59, beta = 1.55, gamma = 2.2)
plot(seq(0, 23), Ta, ylim = c(Tmin, Tmax), xlab = 'hour of day')
points(seq(0, 23), metout_sub$TAREF, pch = 3, col = 2)
legend("topleft",
       legend = c("TrenchR", "NichemapR"),
       lty = NULL, pch = c(1,3), col = c("black","red"))

# Soil surface temperature (C)
Ts <- sapply(1:24, diurnal_temp_variation_sine, T_max = Tmax_s, T_min = Tmin_s)
plot(seq(0, 23), Ts, ylim = c(Tmin_s, Tmax_s), xlab = 'hour of day')
points(seq(0, 23), soil_sub$D0cm, pch = 3, col = 2)
legend("topleft",
       legend = c("TrenchR", "NichemapR"),
       lty = NULL, pch = c(1,3), col = c("black","red"))
```

We then compare air temperature and windspeed profiles associated with free convection. We note that TrenchR assumed fixed windspeed for simplicity in the manuscript.
```{r}
# Neutral air temperature profile
Ta_liz  <- air_temp_profile_neutral(T_r = Ta, zr = 2, z0 = 0.004, z = z, T_s = Ts) # MRK used z0=0.004 instead of 0.2
plot(seq(0, 23), Ta_liz, ylim = c(Tmin_s, Tmax_s), xlab = 'hour of day')
points(seq(0, 23), metout_sub$TALOC, pch = 3, col = 2)
legend("topleft",
       legend = c("TrenchR", "NichemapR"),
       lty = NULL, pch = c(1,3), col = c("black","red"))

# Neutral wind speed profile
V_liz <- wind_speed_profile_neutral(u_r = u, zr = 2, z0 = 0.004, z = z) # MRK used z0=0.004 instead of 0.2
plot(seq(0, 23), rep(V_liz, 24), xlab = 'hour of day', ylab = 'wind speed, m/s', ylim = c(0, 1.5))
points(seq(0, 23), metout_sub$VLOC, pch = 3, col = "red")
legend("topleft",
       legend = c("TrenchR", "NichemapR"),
       lty = NULL, pch = c(1,3), col = c("black","red"))
```
Next, set up input for estimating solar radiation and estimate silhouette area
```{r}
mass <- 10  # Mass (g)
a    <- 0.9 # Solar absorptivity (proportion)
# Estimate surface area (m^2) and the proportion sihouette area
A   <- surface_area_from_mass(mass, "lizard")

# use NicheMapR GEOM function
GEOM.out <- GEOM_ecto(AMASS = mass/1000, GEOMETRY = 3, PTCOND = psa_g, PMOUTH = 0)
A_NMR <- GEOM.out$AREA
ASILN <- GEOM.out$ASILN
ASILP <- GEOM.out$ASILP
psa_NMR <- 0.5 * (ASILN + ASILP) / A_NMR

# compare outputs
A
A_NMR

psa <- sapply(psi_deg, proportion_silhouette_area, taxon = "lizard", posture = "prostrate")
# Change negative values to zero
psa[psa < 0] = 0

```

```{r, echo=FALSE}
#plot
plot(seq(0, 23), psa, xlab = 'hour of day', ylab = 'silhouette area (fraction)', ylim = c(0, 0.4))
points(seq(0, 23), rep(psa_NMR, 24), pch = 3,col = 2)

legend("topleft",
       legend = c("TrenchR", "NichemapR"),
       lty = NULL, pch = c(1,3), col = c("black","red"))
```

We then estimate and compare absorption of solar radiation
```{r}
Qabs <- rep(NA, 24)
Qabs_NMR <- Qabs
p_dif <- Sdif / (Sdir + Sdif)
# MRK note that psa_dif has been added as a parameter, set to half, and psa_ref has been set to the bottom half minus fraction in contact with ground
Qradiation_absorbed2 <- function (a = 0.9, A, psa_dir, psa_dif, psa_ref, S_dir, S_dif, S_ref = NA, 
                                  a_s = NA) 
{
  stopifnot(a >= 0, a <= 1, A > 0, psa_dir >= 0, psa_dir <= 
              1, psa_ref >= 0, psa_ref <= 1, psa_dif >= 0, psa_dif <= 1,
            S_dir > 0, S_dif > 0)
  if (is.na(S_ref)) {
    S_ref <- a_s * S_dir
  }
  A_dir <- A * psa_dir
  A_dif <- A * psa_dif
  A_ref <- A * psa_ref
  a * A_dir * S_dir + a * A_dif * S_dif + a * A_ref * S_ref
}
for (hour in 1:24) {
  Qabs[hour] <- Qradiation_absorbed(a = a, A = A, psa_dir = psa[hour], psa_dif = 0.5, psa_ref = 0.5 * (1 - psa_g), S_dir = Sdir[hour], S_dif = Sdif[hour], rho = rho)
  # use SOLAR_ecto for NMR prediction, assuming halfway between perpendicular and parallel posture
 Qabs_NMR[hour] <- SOLAR_ecto(QSOLR = metout_sub$SOLR[hour], ATOT = A, FATOSK = 0.5, FATOSB = 0.5, ASIL = (ASILN + ASILP) / 2, ABSAN = a, ABSSB = 1 - rho, ZEN = metout_sub$Z[hour] / 180 * pi, postur = 0, PDIF = p_dif[hour])$QSOLAR
}
```

```{r, echo=FALSE}
#plot
plot(seq(0, 23), Qabs, xlab = 'hour of day', ylim = c(0, 2.5))
points(seq(0, 23), Qabs_NMR, pch = 3, col = "red")

legend("topleft",
       legend = c("TrenchR", "NichemapR"),
       lty = NULL, pch = c(1,3), col = c("black","red"))
```

We then estimate and compare net emitted thermal radiation.
```{r}
epsilon_s <- 0.965 # Surface emmissivity of lizards
Qemit <- rep(NA, 24)
Qemit_NMR <- rep(NA, 24)
for (hour in 1:24) {
  Qemit[hour] <- Qemitted_thermal_radiation(epsilon = epsilon_s, A = A, psa_dir = 0.5, psa_ref = 0.5 * (1 - psa_g), T_b = Ta_liz[hour] + 273.15, T_g = Ts[hour] + 273.15, T_a = Ta[hour] + 273.15, enclosed = TRUE)
  # use RADOUT_ecto for NMR prediction
  Qemit_NMR[hour] <- RADOUT_ecto(TSKIN = Ta_liz[hour], ATOT = A, EMISAN = epsilon_s) - RADIN_ecto(ATOT = A, EMISAN = epsilon_s, EMISSB = 1, EMISSK = 1, TSKY = Ta[hour], TGRD = Ts[hour])
}
```

```{r, echo=FALSE}
#plot
plot(seq(0, 23), Qemit, ylim = c(0, 0.1), xlab = 'hour of day')
points(seq(0, 23), Qemit_NMR, pch = 3, col = 2)

legend("topleft",
       legend = c("TrenchR", "NichemapR"),
       lty = NULL, pch = c(1,3), col = c("black","red"))
```

We then estimate and compare heat transfer coefficients.
```{r}
# Use DRYAIR from NicheMapR to estimate the thermal conductivity of air and kinematic viscosity
ap <- airpressure_from_elev(elev) * 1000 # Barometric pressure (pascal)
DRYAIRout <- DRYAIR(db = Ta, bp = ap, alt = elev)

K  <- mean(DRYAIRout$thcond) # Thermal conductivity (Wm^-2K^-1)
nu <- mean(DRYAIRout$viskin) # Kinematic viscosity (m2 s-1)

# Estimate the characteristic dimension as cube root of volume, assuming density of water as 1000kg/m^3

D <- ((mass / 1000) / 1000) ^ (1 / 3)

# Estimate the heat transfer coefficient using an empirical relationship for lizards
H_L <- heat_transfer_coefficient(u = V_liz, D = D, K = K, nu = nu, taxon = "lizard_surface")
# Estimate the heat transfer coefficient using a spherical approximation
H_L2 <- heat_transfer_coefficient_approximation(u = V_liz, D = D, K = K, nu = nu, taxon = "lizard")
# Estimate the heat transfer coefficient using a simplified version of the approximation
H_L3 <- heat_transfer_coefficient_simple(u = V_liz, D = D, type = "Gates")

# run NMR CONV_ecto to get characteristic dimension
CONV.out <- CONV_ecto(GEOMETRY = 3, AL = GEOM.out$AL)

# compare heat transfer coefficients
H_L_NMR <- CONV.out$HC #NicheMapR
H_L
H_L_NMR
```

We then estimate and compare convection. NichemapR estimates a higher convection coefficient and thus a greater magnitude of convection. 
```{r}
Qconv <- rep(NA, 24)
Qconv_NMR <- Qconv

offset <- 10 # adding an offset to air temperature, otherwise just get zero
for (hour in 1:24) {
  Qconv[hour] <- Qconvection(T_a = Ta_liz[hour] + 273.15 + offset, T_b = Ta_liz[hour] + 273.15, H = H_L, A = A, proportion = 1 - psa_g, ef = 1.3)
  # use CONV_ecto for NMR prediction
  Qconv_NMR[hour] <- CONV_ecto(GEOMETRY = 3, AL = GEOM.out$AL, AV = GEOM.out$AV, ATOT = GEOM.out$AREA, BP = ap, ALT = elev, TA = Ta_liz[hour] + offset, TSKIN = Ta_liz[hour])$QCONV
}
```

```{r, echo=FALSE}
#plot
plot(seq(0, 23), Qconv, ylim = c(-2, 0), xlab = 'hour of day')
points(seq(0, 23), Qconv_NMR, pch = 3, col = 2)

legend("topleft",
       legend = c("TrenchR", "NichemapR"),
       lty = NULL, pch = c(1,3), col = c("black","red"))
```

We then Estimate and compare conduction.
```{r}
Qcond <- rep(NA, 24)
Qcond_NMR <- Qcond

for(hr in 1:24) {
  Qcond[hr] <- Qconduction_animal(T_g = Ts[hr] + 273.15, T_b = Ta_liz[hr] + 273.15, d = 0.025, K = K_sub, A = A, proportion = psa_g)

  # use COND_ecto for NMR prediction
  Qcond_NMR[hr] <- COND_ecto(AV = GEOM.out$AV, TSKIN = Ta_liz[hr], TSUBST = Ts[hr], SUBTK = K_sub)
}
```

```{r, echo=FALSE}
#plot
plot(seq(0, 23), Qcond, xlab = 'hour of day')
points(seq(0, 23), Qcond_NMR, pch = 3, col = 2)

legend("topleft",
       legend = c("TrenchR", "NichemapR"),
       lty = NULL, pch = c(1,3), col = c("black","red"))
```

We estimate operative temperatures as in the manuscript, but add the biophysical model from NichemapR.
```{r}

# General biophysical model from Gates (1980)
Te <- rep(NA, 24)
for (hour in 1:24) {
  Te[hour] <- Tb_Gates(A = A, D = 0.025, psa_dir = psa[hour], psa_ref = 0.5 * (1 - psa_g), psa_air = 1 - psa_g, psa_g = psa_g, T_g = Ts[hour], T_a = Ta_liz[hour], Qabs = Qabs[hour], epsilon = epsilon_s, H_L = H_L, ef = 1.3, K = K_sub) # MRK made D = 2.5 cm and passed K_sub
}

# General biophysical model from Campbell and Norman (2000)
Te2 <- rep(NA, 24)
for (hr in 1:24) {
  # S is solar radiation flux (W m^-2), so we divide by surface area, A
  Te2[hr] <- Tb_CampbellNorman(T_a = Ta_liz[hr], T_g = Ts[hr], S = Qabs[hr] / A, a_l = 0.96, epsilon = epsilon_s, c_p = 29.3, D = svl, u = V_liz)
}

# Lizard specific biophysical model from Buckley (2008)
Te3 <- rep(NA, 24)
for (hour in 1:24) {
  Te3[hour] <- Tb_lizard(T_a = Ta_liz[hour], T_g = Ts[hour], u = V_liz, svl = svl * 1000, m = mass, psi = psi_deg[hour], rho_s = rho, elev = elev, doy = doy, sun = TRUE, surface = TRUE, a_s = a, a_l = 0.965, epsilon_s = epsilon_s, F_d = 0.8, F_r = 0.5, F_a = 0.5, F_g = 0.5)
}

# General biophysical model from NicheMapR, using modular function ectoR_devel()
TC <- unlist(lapply(1:length(Ta_liz), function(x){ectoR_devel(
  Ww_g = mass,
  shape = 3,
  alpha = a,
  M_1 = 0, # turns of metabolism
  pct_wet = 0, # turns of cutaneous evap
  pct_eyes = 0, # turns off ocular evap
  pantmax = 0, # turns of respiration
  alpha_sub = 1 - rho,
  elev = elev,
  pct_cond = psa_g * 100,
  epsilon = epsilon_s,
  epsilon_sub = 1,
  epsilon_sky = 1,
  postur = 0,
  pres = ap,
  K_sub = K,
  fatosk = 0.5,
  fatosb = 0.5,
  TA = Ta_liz[x],
  TGRD = Ts[x],
  TSKY = 1.22 * Ta_liz[x] - 20.4,
  VEL = V_liz,
  RH = 10,
  QSOLR = (Sdir + Sdif)[x],
  PDIF = Sdif[x] / (Sdir + Sdif)[x],
  Z = psi_deg[x]
)$TC})) # run ectotherm_simple across environments
```

We then recreate the plot in the manuscript, adding output from the NicheMapR biophysical model. 
```{r, echo=FALSE}
par(mar = c(5, 4, 4, 8) + 0.1)

plot(1:24, Te, type = "l", xlab = "Hour", ylab = expression("Temperature ("*~degree*C*")"), col = "#5DC863FF", ylim = c(0,70), las = 1) # Gates
points(1:24, Te2, type ="l", col = "#5DC863FF", lty = "dotted") # Campbell and Norman
points(1:24, Te3, type = "l", col = "#5DC863FF", lty = "dashed") # Lizard, Buckley 2008
points(1:24, TC, type = "l", col = "orange", lty = "solid") # NicheMapR

points(1:24, Ta, type = "l", col = "#440154FF")
points(1:24, Ta_liz, type = "l", col = "#3B528BFF")
points(1:24, Ts, type = "l", col = "#21908CFF")

# add additional axis with radiation
par(new = TRUE)
plot(1:24, Sdir, pch = 16, axes = FALSE, xlab = NA, ylab = NA, type = "l", col = "#FDE725FF")
axis(side = 4, las = 1)
mtext(side = 4, line = 3, 'Radiation (W/m^2)')

legend("topleft",
       legend = c("Ta (2m)", "Ta (0.02m)", "Ts", "Te", "Te NichemapR", "Radiation"),
       lty = 1, pch = NA, col = c("#440154FF", "#3B528BFF", "#21908CFF", "#5DC863FF","orange", "#FDE725FF"))
```
**Figure**. Body temperatures (Te) are predicted to drastically exceed air temperature when lizards are exposed to high levels of solar radiation. Air temperatures (Ta, $^\circ$C) at lizard height (0.02 m) are predicted to exceed air temperatures at 2 m and to be similar to surface temperatures (Ts). 
We estimate body temperatures using two general energy budgets [solid: Tb_Gates(); dotted: Tb_CampbellNorman()] and a lizard specific biophysical model [dashed: Tb_lizard()] that differ in how they model heat exchanges. The orange line is NichemapR output.

### 2022-10-28 Manuscript_full.Rmd
update Te and add appendix S1

@@
 ---
-title: 'TrenchR: an R package for transparent microclimate and biophysical ecology'
+title: 'TrenchR: an R package for modular and accessbile microclimate and biophysical ecology'
 author: 
@@
 output:
+  html_document: default
   word_document: default
-  html_document: default
   pdf_document: 
@@
 bibliography: bibliography.bib
-csl: methods-in-ecology-and-evolution.csl
+csl: plos.csl
 editor_options: 
@@
 
--   Within the word count (6000-7000 words for Standard Articles, 3000-4000
-    words for Applications and Practical Tools)
-
--   Continuous line and page numbering throughout
-
 Lauren B. Buckley^1^, Bryan A. Briones Ortiz^1^, Isaac Caruso^1^, Aji John^1^, Ofir Levy^2^, Abigail V. Meyer^1^, Eric A. Riddell^3^, Yutaro Sakairi^1^, and Juniper L. Simonis^4^
@@
 
-Type: applications
+Type: new methods
 
@@
 
-1.
 Much understanding of organismal responses to climate change and variability relies on the assumption that body temperatures are equal to temporally averaged air temperatures high above the ground.
-    However, most organisms live near the ground and acute exposure to solar and thermal radiation and thermal extremes can substantially elevate their body temperatures.
-2.  We introduce the TrenchR package, which aids in Translating Environmental Change into organismal responses. 
+However, most organisms live near the ground and acute exposure to solar and thermal radiation and thermal extremes can substantially elevate or depress their body temperatures.
+We introduce the TrenchR package, which aids in Translating Environmental Change into organismal responses.
 The package includes microclimate models to vertically scale weather station data to organismal heights.
@@
 We offer utility functions to aid in estimating the organismal and environmental parameters needed for biophysical ecology.
-3.  TrenchR focuses on simple and modular functions so users can create transparent and flexible models for biophysical applications.
-4.  The package aims to introduce and enable microclimate and biophysical modeling to improve ecological and evolutionary forecasting. 
+TrenchR focuses on simple and modular functions so users can create transparent and flexible models for biophysical applications.
+The package aims to introduce and enable microclimate and biophysical modeling to improve ecological and evolutionary forecasting.
 We further this aim through a series of educational modules that introduce the field of biophysical ecology.
@@
 Although the Fortran source code was recently released, the complexity of NicheMapR functions can make it difficult to understand and modify the source code.
-The function documentation including variable definitions and units can also be challenging to follow. 
-Many of the approaches are similar between the packages, but TrenchR provides simple functions to aid the understanding and transparency of biophysical approaches. 
+The integrated functions made it difficult to separate and adapt different forms of heat flow, but we note that some modular R functions were made available during the review process for this manuscript that enable comparison with the TrenchR package (Appendix S1).
+Many of the approaches are similar between the packages, but TrenchR provides simple functions to aid the understanding and accessibility of biophysical approaches.
 The TrenchR functions are modular and easily adapt to a variety of organisms and research questions.
@@
 Some energy budget calculation components of biophysical models are also available in the ThermImage package [@tattersall2017]. 
-Microclimate models, with an emphasis on describing spatial variation, are also available in the Microclima package [@maclean2019microclima].
+Microclimate models, with an emphasis on describing spatial variation, are also available in the microclima package [@maclean2019microclima].
 
@@
 
-The TrenchR package (<http://trenchproject.github.io/TrenchR/>) aims to promote transparency and reproducibility. 
+The TrenchR package (<http://trenchproject.github.io/TrenchR/>) aims to promote accessibility and reproducibility. 
 We welcome contributions and corrections from users. 
@@
 
-We adapted functions from biophysical ecology textbooks [@campbell2000introduction; @gates1980biophysical] and research articles. 
+Few of the functions in TrenchR are "state of the art".
+The development of increasingly complex microclimate and biophysical models before open and reproducible science was emphasized or feasible has limited their uptake.
+Yet, such models are crucial to improving understand and prediction of species responses to climate change.
+We contend there is a need for simple and modular functions that can be combined and extended in a null modeling approach until the model adequately describes organismal interactions with their environment while remaining accessible and transparent.
+
+Validation against sensor data including from physical models of organisms is essential to microclimate and biophysical modelling.
+We adapted well established functions from biophysical ecology textbooks [@campbell2000introduction; @gates1980biophysical] and research articles.
 Source references are included in function headers and reference sections.
+Many of the references, particularly those developing taxa-specific biophysical models, contain extensive empirical validation.
+Appendix S2 illustrates validation of biophysical models using a physical model of a grasshopper.
+Appendix S1 compares TrenchR function to NicheMapR functions.
 We have incorporated default parameters (e.g., organism emissivity or heat conduction rates) when general values are available.
 Many functions additionally include comments that describe potential parameterizations.
-The biophysical ecology texts [@campbell2000introduction; @gates1980biophysical] include many tables with parameter values and we provide several data tables in TrenchR.
+The biophysical ecology texts [@campbell2000introduction; @gates1980biophysical] include many tables with parameter values and we provide several data tables describing solar and thermal absorptivity in TrenchR.
 
@@
 We currently provide published biophysical models for lizards, salamanders, butterflies, grasshoppers, limpets, mussels, and snails.
-Most models predict operative environmental temperatures, which are the equilibrium body temperatures of organisms with specified physical properties in a specific microclimate and assume no heat exchange via metabolism or evaporation [@bakken1992measurement].
+Most models predict operative environmental temperatures, which are the steady-state body temperatures of organisms with specified physical properties in a specific microclimate and assume no heat exchange via metabolism or evaporation [@bakken1992measurement].
 However, we also present an analytical function for humid operative temperature that incorporates the effects of evaporative cooling on operative temperatures for wet-skinned ectotherms, such as salamanders.
@@
 We illustrate the use of the TrenchR package by estimating an energy budget for a *Sceloporus* lizard on June 1, 2021 in Santa Fe, New Mexico, USA (35.69$^\circ$N, -105.944$^\circ$W, elevation: 2121 m).
-The example is also incorporated in the *Using energy balances to estimate body temperatures* vignette. 
+The simplified example, which is designed to be self-contained, is also incorporated in the *Using energy balances to estimate body temperatures* vignette. 
 We start by generating environmental inputs (Figure 2).
@@
 Finally, we use an integrated biophysical model to estimate operative environmental temperatures. 
-See Appendix S1 for another example examining a time series of microclimate data and applying and testing the biophysical models.
+See Appendix S2 for a more realistic example examining a time series of microclimate data and applying and testing the biophysical models.
 
-We will use the energy budget to estimate body temperature, $T_b$, which can depart dramatically from the air temperatures due to energy exchange with the environment. 
-Heat is gained from absorbed solar and thermal radiation and from metabolic reactions. 
-Heat is lost through the organism's emission of radiation and the evaporation of water. 
-The organism exchanges heat with the surrounding air or water via convection and with the ground via conduction. 
+We will use the energy budget to estimate body temperature, $T_b$, which can depart dramatically from the air temperatures due to heat exchange with the environment.
+Heat energy is exchanged with the environment by way of solar and thermal radiation, metabolic reactions, and evaporation.
+The organism also exchanges heat with the surrounding air or water via convection and with substrate it is in contact with via conduction.
 The balance of these heat exchanges (omitting metabolism and evaporation, which are often negligible for ectotherms) can be estimated and often referred to as operative environmental temperature, $T_e$ [@bakken1992measurement].
 $T_e$ is an estimate of $T_b$ and the package functions refer to $T_b$ for simplicity. 
-Our exemplar estimation of $T_b$ assumes thermal equilibrium.
+Our exemplar estimation of $T_b$ assumes steady-state thermal conditions.
+Additionally, we assume that the lizard’s body temperature is homogenous, which is generally reasonable for small ectotherms.
+Approaches to account for thermal gradients between the animal’s core and its skin are available elsewhere [@porter2016].
 
@@
 Tmax_s <- 30 # Maximum soil temperature (C)
-V      <- 1  # Wind speed (m/s)
+u      <- 1  # Wind speed (m/s)
 
@@
 Although measured solar radiation is preferable if available, we can estimate hourly solar radiation by discounting incoming solar radiation as it moves through the atmosphere as follows.
-The function partitions radiation into direct, diffuse, and reflected components:
+We use the approach from Cambpell and Norman [-@campbell2000introduction], which uses an empirical relation to partition radiation into direct, diffuse, and reflected components.
+The `partition_solar_radiation()` function includes 8 empirical relationships for and the proportion_diffuse_solar_radiation() includes a more complex numerical approximation  for partitioning radiation components as described in the *Estimating microclimates* vignette.
 
@@
 
-psi_deg <- sapply(1:24, FUN = zenith_angle, doy = doy, lat = lat, lon = lon)
+psi_deg <- sapply(0:23, FUN = zenith_angle, doy = doy, lat = lat, lon = lon)
 
@@
 We then calculate hourly air and soil surface temperatures based on daily minimum and maximum temperatures. 
-We select the sine-exponential model for air temperature and the sine model for surface temperature:
+We select the sine-exponential model for air temperature and the sine model for surface temperature [@wann1985evaluation]:
 
@@
 We assume a surface roughness of $z_0 = 0.2$ m, which corresponds to bare sand and determines the turbulence of airflow. 
-We implement neutral air temperature and wind speed profiles driven by density differences but profiles forced by wind speed are also available.
+We implement free air temperature and wind speed profiles driven by density differences but profiles forced by wind speed are also available.
 
@@
 
-V_liz <- wind_speed_profile_neutral(u_r = V, zr = 2, z0 = 0.2 , z = 0.02) 
+u_liz <- wind_speed_profile_neutral(u_r = u, zr = 2, z0 = 0.2 , z = 0.02) 
 ```
@@
 
-where $Q_{net}$ is the net energy exchange with the environment (W), $Q_{abs}$ is the solar radiation absorbed (W), $Q_{emit}$ is the net thermal radiation emitted (W), $Q_{conv}$ is energy exchange due to convection (W), $Q_{cond}$ is energy exchange due to conduction (W), $Q_{met}$ is the energy emitted due to metabolism (W), and $Q_{evap}$ is the energy emitted due to evaporative water loss (W).
+where $Q_{net}$ is the net energy exchange with the environment (W), $Q_{abs}$ is the solar radiation absorbed (W), $Q_{emit}$ is the net thermal radiation emitted (W), $Q_{conv}$ is energy exchange due to convection (W), $Q_{cond}$ is energy exchange due to conduction (W), $Q_{met}$ is the energy generated by metabolism (W), and $Q_{evap}$ is the energy generated by evaporative water loss (W).
 We will estimate each term on the right side of the equation in turn. 
@@
 mass <- 10  # Mass (g)
+svl  <- 0.006 # Snout vent length (meters)
 a    <- 0.9 # Solar absorptivity (proportion)
+
+#assume 1/3 of surface area is in contact with surface
+psa_g <- 0.33
 
@@
 
-We calculate the hourly solar radiation absorbed (W) as follows:
+We calculate the hourly solar and thermal radiation absorbed (W) as follows:
 
@@
 
-  Qabs[hour] <- Qradiation_absorbed(a = a, A = A, psa_dir = psa[hour], psa_ref = 1 - psa[hour], S_dir = Sdir[hour], S_dif = Sdif[hour], a_s = rho)
+  Qabs[hour] <- Qradiation_absorbed(a = a, A = A, psa_dir = psa[hour], psa_dif = 0.5, psa_ref = 0.5, S_dir = Sdir[hour], S_dif = Sdif[hour], rho = rho)
 
@@
 
-We estimate thermal radiation $Q_{emit}$ (W) for the lizard outdoors, where $psa_{dir}$ and $psa_{ref}$ are the proportions of surface area $A$ ($m^2$) exposed to direct and reflected solar radiation, respectively.
+We estimate thermal radiation $Q_{emit}$ (W) for the lizard outdoors, where $psa_{dir}$ and $psa_{ref}$ are the view factors, also refered to as configuration factors, that indicate the proportions of surface area $A$ ($m^2$) exposed to the sky and ground, respectively.
 We assume the surface emissivity of lizards, $epsilon_s = 0.965$ [@barlett1967].
@@
 
-  Qemit[hour] <- Qemitted_thermal_radiation(epsilon = epsilon_s, A = A, psa_dir = psa[hour], psa_ref = 1-psa[hour], T_b = Ta_liz[hour] + 273.15, T_g = Ts[hour] + 273.15, T_a = Ta_liz[hour] + 273.15, enclosed = FALSE)
+  Qemit[hour] <- Qemitted_thermal_radiation(epsilon = epsilon_s, A = A, psa_dir = 0.5, psa_ref = 0.5, T_b = Ta_liz[hour] + 273.15, T_g = Ts[hour] + 273.15, T_a = Ta_liz[hour] + 273.15, enclosed = FALSE)
 
@@
 We also illustrate a function estimating $H_L$ using a spherical approximation (`heat_transfer_coefficient_approximation()`) and a simplified approximation (`heat_transfer_coefficient_simple()`) for cases when taxon specific relationships for estimating heat transfer coefficients are not available.
+We estimate the characteristic dimension, which determines exposure to convective heat exchange as the the cube root of volume, assuming the animal density approximates that of water [@mitchell1976heat].
+These coefficients assume convection is forced by the wind.
+TrenchR includes approaches for free convection and a function (`free_or_forced_convection()`) that can be used to decide whether free or forced convection is appropriate.
+The function uses dimensionless numbers, which have been developed to describe heat transfer coefficients associated with convection over different geometries and can be estimated using TrenchR (e.g., Grashof, Nusselt, and Reynolds numbers).
 
@@
 
-svl <- 0.006 # Approximate snout vent length (meters) for Sceloporus, used as the characteristic dimension for heat transfer
+# Estimate the characteristic dimension as cube root of volume, assuming density of water as 1000kg/m^3
+
+D <- ((mass / 1000) / 1000) ^ (1 / 3)
 
@@
 
-H_L <- heat_transfer_coefficient(V = V_liz, D = svl, K = K, nu = nu, taxon = "lizard_surface")
+H_L <- heat_transfer_coefficient(u = u_liz, D = D, K = K, nu = nu, taxon = "lizard_surface")
 
@@
 
-H_L2 <- heat_transfer_coefficient_approximation(V = V_liz, D = svl, K = K, nu = nu, taxon = "lizard")
+H_L2 <- heat_transfer_coefficient_approximation(u = u_liz, D = D, K = K, nu = nu, taxon = "lizard")
 
@@
 
-H_L3 <- heat_transfer_coefficient_simple(V = V_liz, D = svl, type = "Gates")
+H_L3 <- heat_transfer_coefficient_simple(u = u_liz, D = svl, type = "Gates")
 ```
@@
 
-We implement the estimate in R assuming that 1/3 of the lizard surface is in contact with the ground and a skin thickness $d = 10^{-6}$:
+We implement the estimate assuming that conductive heat exchange occurs down to a soil depth of 2.5cm.
+We use this value rather than skin thickness, which results in rapid conduction and does not readily reach steady state conditions.
 
@@
 
-  Qcond[hr] <- Qconduction_animal(T_g = Ts[hr] + 273.15, T_b = Ta_liz[hr] + 273.15, d = 10^-6, K = K, A = A, proportion = 0.33)
+  Qcond[hr] <- Qconduction_animal(T_g = Ts[hr] + 273.15, T_b = Ta_liz[hr] + 273.15, d = 0.025, K = K, A = A, proportion = psa_g)
 
@@
 
-We assume, as is generally done for lizards, that heat exchange associated with metabolism and evaporation is negligible:
+We assume, as is generally done for lizards, that heat exchange associated with metabolism and evaporation is negligible.
+however, functions for estimating both forms of heat exchange available in TrenchR.
 
@@
 We constructed the energy balance assuming $T_b = T_a$ for illustration. 
-Thus, the energy balance is dominated by heat gain from contact with the warmer surface. 
+Thus, the energy balance was dominated by heat gain from contact with the warmer surface. 
 We now use a function based on the energy balance above to estimate body temperature given the environmental conditions:
@@
 
-  Te[hour] <- Tb_Gates(A = A, D = svl, psa_dir = psa[hour], psa_ref = 1 - psa[hour], psa_air = 0.67, psa_g = 0.33, T_g = Ts[hour] + 273.15, T_a = Ta_liz[hour] + 273.15, Qabs = Qabs[hour], epsilon = epsilon_s, H_L = H_L, ef = 1.3, K = K)
+  Te[hour] <- Tb_Gates(A = A, D = svl, psa_dir = psa[hour], psa_ref = 1 - psa[hour], psa_air = 0.67, psa_g = 0.25, T_g = Ts[hour], T_a = Ta_liz[hour], Qabs = Qabs[hour], epsilon = epsilon_s, H_L = H_L, ef = 1.3, K = K)
 
@@
 
-  Te2[hr] <- Tb_CampbellNorman(T_a = Ta_liz[hr] + 273.15, T_g = Ts[hr] + 273.15, S = Qabs[hr] / A, alpha_L = 0.96, epsilon = epsilon_s, c_p = 29.3, D = svl, V = V_liz)
+  Te2[hr] <- Tb_CampbellNorman(T_a = Ta_liz[hr], T_g = Ts[hr], S = Qabs[hr] / A, a_l = 0.96, epsilon = epsilon_s, c_p = 29.3, D = D, u = u_liz)
 
@@
 
-  Te3[hour] <- Tb_lizard(T_a = Ta_liz[hour], T_g = Ts[hour], u = V_liz, svl = svl * 1000, m = mass, psi = psi_deg[hour], rho_S = rho, elev = elev, doy = doy, sun = TRUE, surface = TRUE, alpha_S = a, alpha_L = 0.965, epsilon_s = epsilon_s, F_d = 0.8, F_r = 0.5, F_a = 0.5, F_g = 0.5)
+  Te3[hour] <- Tb_lizard(T_a = Ta_liz[hour], T_g = Ts[hour], u = u_liz, svl = svl * 1000, m = mass, psi = psi_deg[hour], rho_s = rho, elev = elev, doy = doy, sun = TRUE, surface = TRUE, a_s = a, a_l = 0.965, epsilon_s = epsilon_s, F_d = 0.8, F_r = 0.5, F_a = 0.5, F_g = 0.5)
 
@@
 The microclimate models indicate that air temperatures at lizard height are similar to surface temperatures (Figure 3). 
-The biophysical models indicate that solar radiation will elevate lizard body temperatures far above air temperatures and that the lizard will face thermal stress if it is unable to seek shade (Figure 3).
-The three biophysical models predict different body temperatures during peak period of solar radiation because they model interactions with the ground differently and users are encouraged to review the details of each biophysical model before selection.
+The biophysical models indicate that solar radiation can elevate lizard body temperatures far above air temperatures and that the lizard will face thermal stress if it is unable to seek shade (Figure 3).
+The three biophysical models predict different body temperatures during peak period of solar radiation because they model interactions with the ground differently and users are encouraged to review the details of each biophysical model and perform empirical validations before selection.
 Differences in estimated body temperatures are accentuated by the high level of solar radiation.
+`Tb_Gates` is a general and comprehensive model that is appropriate for many applications.
+Taxa-specific biophysical models often best account for details of organism environment interactions and have generally been well tested.
 
@@
 
-plot(1:24, Te - 273.15, type = "l", xlab = "Hour", ylab = expression("Temperature ("*~degree*C*")"), col = "#5DC863FF", ylim = c(10,60), las = 1) # Gates
-points(1:24, Te2 - 273.15, type ="l", col = "#5DC863FF", lty = "dotted") # Campbell and Norman
+plot(1:24, Te, type = "l", xlab = "Hour", ylab = expression("Temperature ("*~degree*C*")"), col = "#5DC863FF", ylim = c(10,70), las = 1) # Gates
+points(1:24, Te2, type ="l", col = "#5DC863FF", lty = "dotted") # Campbell and Norman
 points(1:24, Te3, type = "l", col = "#5DC863FF", lty = "dashed") # Lizard, Buckley 2008
@@
 The resultant models are likely to be sufficiently detailed for many applications, but users are referred to NicheMapR for more detailed biophysical models [@kearney2017nichemapr; @kearney2020nichemapr].
-We focus on models that predict equilibrium conditions for simplicity (that is, steady-state conditions). 
+We focus on models that predict steady-state conditions for simplicity (that is, steady-state conditions). 
 Such models generally do not present computational challenges so our models are not optimized for computational efficiency. 
@@
 We thank A. Cannistra, R. Huey, J. Kingsolver, L. Miller, W. Porter, and H. A. Woods for assistance and input and B. Helmuth for providing functions. 
-We particularly thank M. Kearney for extensive input on TrenchR and its alignment with NicheMapR and for contributing two tutorials.
+We particularly thank M. Kearney for extensive input on TrenchR and its alignment with NicheMapR and for contributing two educational modules and code for comparing TrenchR to NicheMapR.
 
@@
 AM, BBO, YS, and LB converted and edited the educational modules.
+Add? Conceptualization, Project Administration, Software, Writing – Original Draft Preparation, Writing – Review & Editing
 

### 2022-10-31 Manuscript_AppS1.Rmd
Ms updates

@@
 psa_g <- 0.33 # fraction of animal surface contacting the ground
-K_sub <- 0.5 # parameter for thermal conductivity of substrate, W/mC
+K_skin <- 0.5 # parameter for thermal conductivity of skin, W/mC
 ```
@@
 
-We then Estimate and compare conduction.
+We then estimate and compare conduction.
 ```{r}
@@
 for(hr in 1:24) {
-  Qcond[hr] <- Qconduction_animal(T_g = Ts[hr] + 273.15, T_b = Ta_liz[hr] + 273.15, d = 0.025, K = K_sub, A = A, proportion = psa_g)
+  Qcond[hr] <- Qconduction_animal(T_g = Ts[hr] + 273.15, T_b = Ta_liz[hr] + 273.15, d = 0.025, K = K_skin, A = A, proportion = psa_g)
 

### 2022-10-31 Manuscript_full.Rmd
Ms updates

@@
 
-Few of the functions in TrenchR are "state of the art".
 The development of increasingly complex microclimate and biophysical models before open and reproducible science was emphasized or feasible has limited their uptake.
@@
   - *conductive heat exchanges* between organisms and solid surfaces (generally the ground) due to physical contact; and 
-  - *evaporative heat exchanges* associated with organisms' evaporative water loss.
+  - *evaporative and metabolic heat exchanges* associated with organisms' evaporative water loss and metabolic heat generation.
 
@@
 We model a 10 gram *Sceloporus* lizard with solar absorptivity $a = 0.9$ [@gates1980biophysical]. 
-We will initially assume $T_b = T_a$ to illustrate the calculations before solving for $T_b$ given the environmental conditions.
+We will initially assume $T_b = T_a + 10$ to illustrate the calculations before solving for $T_b$ given the environmental conditions.
 
@@
 We also illustrate a function estimating $H_L$ using a spherical approximation (`heat_transfer_coefficient_approximation()`) and a simplified approximation (`heat_transfer_coefficient_simple()`) for cases when taxon specific relationships for estimating heat transfer coefficients are not available.
-We estimate the characteristic dimension, which determines exposure to convective heat exchange as the the cube root of volume, assuming the animal density approximates that of water [@mitchell1976heat].
+We estimate the characteristic dimension, which determines exposure to convective heat exchange as the cube root of volume, assuming the animal density approximates that of water [@mitchell1976heat].
+
 These coefficients assume convection is forced by the wind.
@@
 ```{r}
-# Use DRYAIR from NicheMapR to estimate the thermal conductivity of air and kinematic viscosity
+# Use DRYAIR from NicheMapR to estimate the thermal conductivity of air and kinematic viscosity.
 
@@
 We implement the function in R assuming that 2/3 of the lizard's surface area is exchanging heat through convection.
-We currently estimate no convection since we are assuming $T_b = T_a$ for illustration purposes:
 
@@
 
-  Qconv[hour] <- Qconvection(T_a = Ta_liz[hour] + 273.15, T_b = Ta_liz[hour] + 273.15, H = H_L, A = A, proportion = 0.67, ef = 1.3)
+  Qconv[hour] <- Qconvection(T_a = Ta_liz[hour] + 273.15, T_b = Ta_liz[hour] + 10 + 273.15, H = H_L, A = A, proportion = 0.67, ef = 1.3)
 
@@
 
-$$Q_{cond} = A \cdot \mbox{proportion} \cdot K(T_g-T_b)/d. $$
-
+$$Q_{cond} = A \cdot \mbox{proportion} \cdot K_{skin}(T_g-T_b)/d$$ 
+where $K_{skin}$ in the thermal conductivity of lizard skin ($Wm^-2K^-1$).
 We implement the estimate assuming that conductive heat exchange occurs down to a soil depth of 2.5cm.
@@
 
-  Qcond[hr] <- Qconduction_animal(T_g = Ts[hr] + 273.15, T_b = Ta_liz[hr] + 273.15, d = 0.025, K = K, A = A, proportion = psa_g)
+  Qcond[hr] <- Qconduction_animal(T_g = Ts[hr] + 273.15, T_b = Ta_liz[hr] + 10 + 273.15, d = 0.025, K = 0.5, A = A, proportion = psa_g)
 
@@
 ## Authors' Contributions
-
-LB conceived and designed the package and lead all aspect of the packages creation and dissemination.
-All authors wrote and edited functions and associated package resources, contributed critically to manuscript drafts, and gave final approval for publication.
-AJ and JS configured the package.
-AM, BBO, YS, and LB converted and edited the educational modules.
-Add? Conceptualization, Project Administration, Software, Writing – Original Draft Preparation, Writing – Review & Editing
+Lauren Buckley: Conceptualization, Methodology, Project Administration, Writing – Original Draft Preparation. All authors: Software, Writing – Review & Editing.
 

### 2022-11-14 Manuscript_full.Rmd
Ms updates

@@
 ---
-title: 'TrenchR: an R package for modular and accessbile microclimate and biophysical ecology'
+title: 'TrenchR: an R package for modular and accessible microclimate and biophysical ecology'
 author: 
@@
 Much understanding of organismal responses to climate change and variability relies on the assumption that body temperatures are equal to temporally averaged air temperatures high above the ground.
-However, most organisms live near the ground and acute exposure to solar and thermal radiation and thermal extremes can substantially elevate or depress their body temperatures.
+However, most organisms experience microclimates near the ground and acute exposure to solar and thermal radiation and thermal extremes can substantially elevate or depress their body temperatures.
 We introduce the TrenchR package, which aids in Translating Environmental Change into organismal responses.
 The package includes microclimate models to vertically scale weather station data to organismal heights.
-The biophysical modeling tools include both general models for heat flows and specific models to predict body temperatures for a variety of ectothermic taxa.
 Additional functions model and temporally partition air and soil temperatures and solar radiation.
-We offer utility functions to aid in estimating the organismal and environmental parameters needed for biophysical ecology.
+TrenchR biophysical modeling tools include both general models for heat flows and specific models to predict body temperatures for a variety of ectothermic taxa.
+We also offer utility functions to aid in estimating the organismal and environmental parameters needed for biophysical ecology.
 TrenchR focuses on simple and modular functions so users can create transparent and flexible models for biophysical applications.
@@
 However, air temperatures near the ground where most organisms reside are often considerably warmer, and absorption of solar and thermal radiation can raise body temperatures well above air temperatures, resulting in potential discrepancies of tens of degrees Celsius [@sinclair2016; @sunday2014; @kearney2009].
+The rapid drop in performance and onset of thermal stress at high body temperatures can amplify the biological significance of underestimations [@buckley2022asymmetry].
+
 Applications are increasingly demonstrating the importance of accounting for microclimate variation [@bramer2018advances; @butikofer2020problem; @lembrechts2019comparing]. 
@@
 They can also characterize direct and diffuse solar radiation and longwave radiation emanating from the sky, ground, and surrounding objects like vegetation.
-Biophysical models estimate body temperatures using an energy budget to balance heat exchanges between organisms and their environment.
+Microclimate model output is used as input in biophysical models, which estimate body temperatures using an energy budget to balance heat exchanges between organisms and their environment.
 
@@
 The TrenchR package is intended to complement the NicheMapR package, which includes sophisticated microclimate [@kearney2017nichemapr] and biophysical [@kearney2020nichemapr] models.
-Although the Fortran source code was recently released, the complexity of NicheMapR functions can make it difficult to understand and modify the source code.
+Although the Fortran source code was recently released, the complexity of NicheMapR functions can make it difficult to understand and modify the source code and Fortran can be complicated to compile and run.
 The integrated functions made it difficult to separate and adapt different forms of heat flow, but we note that some modular R functions were made available during the review process for this manuscript that enable comparison with the TrenchR package (Appendix S1).
-Many of the approaches are similar between the packages, but TrenchR provides simple functions to aid the understanding and accessibility of biophysical approaches.
+Many of the approaches are similar between the packages, but TrenchR provides simple R functions to aid the understanding and accessibility of biophysical approaches.
 The TrenchR functions are modular and easily adapt to a variety of organisms and research questions.
@@
 Microclimate models, with an emphasis on describing spatial variation, are also available in the microclima package [@maclean2019microclima].
+
+The development of increasingly complex microclimate and biophysical models before open and reproducible science was emphasized or feasible has limited their uptake.
+Yet, such models are crucial to improving understand and prediction of species responses to climate change.
+Simple and modular functions can be combined and extended in a null modeling approach until the model adequately describes organismal interactions with their environment while remaining accessible and transparent.
 
@@
 
-The development of increasingly complex microclimate and biophysical models before open and reproducible science was emphasized or feasible has limited their uptake.
-Yet, such models are crucial to improving understand and prediction of species responses to climate change.
-We contend there is a need for simple and modular functions that can be combined and extended in a null modeling approach until the model adequately describes organismal interactions with their environment while remaining accessible and transparent.
-
 Validation against sensor data including from physical models of organisms is essential to microclimate and biophysical modelling.
-We adapted well established functions from biophysical ecology textbooks [@campbell2000introduction; @gates1980biophysical] and research articles.
-Source references are included in function headers and reference sections.
+We adapted well established and validated functions from biophysical ecology textbooks [@campbell2000introduction; @gates1980biophysical] and research articles.
+Source references are cited in function headers and reference sections.
 Many of the references, particularly those developing taxa-specific biophysical models, contain extensive empirical validation.
@@
 These coefficients assume convection is forced by the wind.
-TrenchR includes approaches for free convection and a function (`free_or_forced_convection()`) that can be used to decide whether free or forced convection is appropriate.
+TrenchR includes approaches for free convection and a function (`free_or_forced_convection()`) that evaluates whether free or forced convection is appropriate.
 The function uses dimensionless numbers, which have been developed to describe heat transfer coefficients associated with convection over different geometries and can be estimated using TrenchR (e.g., Grashof, Nusselt, and Reynolds numbers).
@@
 
-We constructed the energy balance assuming $T_b = T_a$ for illustration. 
-Thus, the energy balance was dominated by heat gain from contact with the warmer surface. 
-We now use a function based on the energy balance above to estimate body temperature given the environmental conditions:
+We now use a function based on the Gates energy balance above to estimate body temperature given the environmental conditions:
 
@@
 
-We thank A. Cannistra, R. Huey, J. Kingsolver, L. Miller, W. Porter, and H. A. Woods for assistance and input and B. Helmuth for providing functions. 
+We thank A. Cannistra, R. Huey, J. Kingsolver, L. Miller, W. Porter, R. Potter, and H. A. Woods for assistance and input and B. Helmuth for providing functions. 
 We particularly thank M. Kearney for extensive input on TrenchR and its alignment with NicheMapR and for contributing two educational modules and code for comparing TrenchR to NicheMapR.

### 2022-11-14 Manuscript_full.Rmd
Minor additional updates

@@
 Appendix S2 illustrates validation of biophysical models using a physical model of a grasshopper.
-Appendix S1 compares TrenchR function to NicheMapR functions.
+Appendix S1 compares TrenchR functions to NicheMapR functions.
 We have incorporated default parameters (e.g., organism emissivity or heat conduction rates) when general values are available.
@@
 We use the approach from Cambpell and Norman [-@campbell2000introduction], which uses an empirical relation to partition radiation into direct, diffuse, and reflected components.
-The `partition_solar_radiation()` function includes 8 empirical relationships for and the proportion_diffuse_solar_radiation() includes a more complex numerical approximation  for partitioning radiation components as described in the *Estimating microclimates* vignette.
+The `partition_solar_radiation()` function includes 8 empirical relationships for, and the proportion_diffuse_solar_radiation() includes a more complex numerical approximation for, partitioning radiation components as described in the *Estimating microclimates* vignette.
 
@@
 We assume, as is generally done for lizards, that heat exchange associated with metabolism and evaporation is negligible.
-however, functions for estimating both forms of heat exchange available in TrenchR.
+However, functions for estimating both forms of heat exchange available in TrenchR.
 
@@
 mtext(side = 4, line = 3, 'Radiation (W/m^2)')
-legend("topleft",
+legend("topleft", bty = "n",
        legend = c("Ta (2m)", "Ta (0.02m)", "Ts", "Te", "Radiation"),

### 2022-11-16 ManuscriptTe.Rmd
minor text updates

@@
 
-  out <- Tb_Gates(A = A, D = 0.0211/3, psa_dir = 0.5, psa_ref = 1-0.5, psa_air = 1, psa_g = 0.0, T_g = dat$SoilTemp[ind], T_a = dat$Tgrass[ind], Qabs = A * 0.66 * dat$Rad[ind], epsilon = 0.95, H_L = dat$H_L[ind], ef = 1.3, K = 0.15)
+  out <- Tb_Gates(A = A, D = 0.0211, psa_dir = 0.5, psa_ref = 0.5, psa_air = 1, psa_g = 0, T_g = dat$SoilTemp[ind], T_a = dat$Tgrass[ind], Qabs = A * 0.66 * dat$Rad[ind], epsilon = 0.95, H_L = dat$H_L[ind], ef = 1.3, K = 0.15)
 
@@
 }
+
+dat$TeGates<- TeGates
 
@@
 
-ggplot(data = dat.pl, aes(x=hour, y = value, color = variable))+ geom_line(aes(lty = variable))+ facet_wrap( ~ elev.lab) + theme_classic()+ theme(legend.position = "bottom") + ylim(15, 65)
+ggplot(data = dat.pl, aes(x=hour, y = value, color = variable))+ geom_line(aes(lty = variable))+ facet_wrap( ~ elev.lab) + theme_classic()+ theme(legend.position = "bottom") + ylim(15, 65)+ylab("temperature (C)")
 

### 2022-11-16 Manuscript_AppS1.Rmd
minor text updates

@@
 #source_url('https://github.com/mrke/NicheMapR/blob/master/R/GEOM_ecto.R')
-#source_url('https://github.com/mrke/NicheMapR/blob/master/R/SOLAR_ecto.R')
 ```
@@
 # Assumptions
-tau <- 0.7 # Atmospheric transmissivity # MRK change from 0.7 to 0.75
+tau <- 0.7 # Atmospheric transmissivity
 rho <- 0.6 # Albedo
@@
 K_skin <- 0.5 # parameter for thermal conductivity of skin, W/mC
+K_sub <- 0.5 # parameter for thermal conductivity of substrate, W/mC
 ```
@@
 for (hour in 1:24) {
-  Te[hour] <- Tb_Gates(A = A, D = 0.025, psa_dir = psa[hour], psa_ref = 0.5 * (1 - psa_g), psa_air = 1 - psa_g, psa_g = psa_g, T_g = Ts[hour], T_a = Ta_liz[hour], Qabs = Qabs[hour], epsilon = epsilon_s, H_L = H_L, ef = 1.3, K = K_sub) # MRK made D = 2.5 cm and passed K_sub
+  Te[hour] <- Tb_Gates(A = A, D = D, psa_dir = psa[hour], psa_ref = 0.5 * (1 - psa_g), psa_air = 1 - psa_g, psa_g = psa_g, T_g = Ts[hour], T_a = Ta_liz[hour], Qabs = Qabs[hour], epsilon = epsilon_s, H_L = H_L, ef = 1.3, K = K_sub) # MRK made D = 2.5 cm and passed K_sub
 }
@@
   # S is solar radiation flux (W m^-2), so we divide by surface area, A
-  Te2[hr] <- Tb_CampbellNorman(T_a = Ta_liz[hr], T_g = Ts[hr], S = Qabs[hr] / A, a_l = 0.96, epsilon = epsilon_s, c_p = 29.3, D = svl, u = V_liz)
+  Te2[hr] <- Tb_CampbellNorman(T_a = Ta_liz[hr], T_g = Ts[hr], S = Qabs[hr] / A, a_l = 0.96, epsilon = epsilon_s, c_p = 29.3, D = D, u = V_liz)
 }
@@
 for (hour in 1:24) {
-  Te3[hour] <- Tb_lizard(T_a = Ta_liz[hour], T_g = Ts[hour], u = V_liz, svl = svl * 1000, m = mass, psi = psi_deg[hour], rho_s = rho, elev = elev, doy = doy, sun = TRUE, surface = TRUE, a_s = a, a_l = 0.965, epsilon_s = epsilon_s, F_d = 0.8, F_r = 0.5, F_a = 0.5, F_g = 0.5)
+  Te3[hour] <- Tb_lizard(T_a = Ta_liz[hour], T_g = Ts[hour], u = V_liz, svl = D * 1000, m = mass, psi = psi_deg[hour], rho_s = rho, elev = elev, doy = doy, sun = TRUE, surface = TRUE, a_s = a, a_l = 0.965, epsilon_s = epsilon_s, F_d = 0.8, F_r = 0.5, F_a = 0.5, F_g = 0.5)
 }
@@
 
-plot(1:24, Te, type = "l", xlab = "Hour", ylab = expression("Temperature ("*~degree*C*")"), col = "#5DC863FF", ylim = c(0,70), las = 1) # Gates
+plot(1:24, Te, type = "l", xlab = "Hour", ylab = expression("Temperature ("*~degree*C*")"), col = "#5DC863FF", ylim = c(0,55), las = 1) # Gates
 points(1:24, Te2, type ="l", col = "#5DC863FF", lty = "dotted") # Campbell and Norman
@@
 
-legend("topleft",
+legend("topleft", bty = "n",
        legend = c("Ta (2m)", "Ta (0.02m)", "Ts", "Te", "Te NichemapR", "Radiation"),

### 2023-06-28 Manuscript_full.Rmd
Update ms

@@
 output:
-  word_document: default
-  html_document: default
   pdf_document: 
@@
       in_header: preamble.tex
+  html_document: default
+  word_document: default
 bibliography: bibliography.bib
@@
 Some of these predictive shortcomings stem from omitting spatial and temporal environmental variation and how it interacts with organismal phenotypes (e.g., size, coloration) [@nadeau2017coarse; @potter2013microclimatic].
-Many analyses assume that organismal body temperatures are equal to air temperatures at weather station height (usually 2 meters). 
+Many analyses assume that organismal body temperatures are equal to shaded air temperatures inside a screen at weather station height (usually 2 meters). 
 However, air temperatures near the ground where most organisms reside are often considerably warmer, and absorption of solar and thermal radiation can raise body temperatures well above air temperatures, resulting in potential discrepancies of tens of degrees Celsius [@sinclair2016; @sunday2014; @kearney2009].
@@
 The TrenchR functions are modular and easily adapt to a variety of organisms and research questions.
-Although many of the TrenchR functions are general, some components of TrenchR are best suited for ectothermic animals.
+
+Although many of the TrenchR functions are general, some components of TrenchR are best suited for small ectothermic animals since body temperatures are generally assumed to be uniform and at steady-state.
 We omit plant specific biophysical models since they are the focus of several R packages [tealeaves: [@muir2019tea]; plantecophys: [@duursma2015plantecophys]].
 Some energy budget calculation components of biophysical models are also available in the ThermImage package [@tattersall2017]. 
-Microclimate models, with an emphasis on describing spatial variation, are also available in the microclima package [@maclean2019microclima].
+Microclimate models, with an emphasis on describing spatial variation, are also available in the microclima package [@maclean2019microclima].The microclimc package enables accounting for how forest canopies alter microclimates [@maclean2021microclimc].
 
@@
 Simple and modular functions can be combined and extended in a null modeling approach until the model adequately describes organismal interactions with their environment while remaining accessible and transparent.
+Many applications will warrant the inclusion of more complex microclimate and biophysical algorithms from R packages such as NicheMapR, microclima, and microclimc.
 
@@
 
-## Example Application
+## Results
 
@@
 
-## Conclusions
+## Discussion
 
@@
 
+## Supporting information captions
+**Appendix S1**.
+We compare TrenchR and NicheMapR implementations.
+We thank Michael Kearney for making modular R functions corresponding to NicheMapR available and providing code comparing TrenchR and NicheMapR functions during the review process.
+Our adaptation of the provided code corresponds to the operative temperature estimation in the manuscript.
+The NicheMapR implementation relies on the micro_global environmental data available through NicheMapR.
+The modular NicheMapR functions are available in version 3.2.1 via GitHub (https://github.com/mrke/NicheMapR/releases).
+
+**Appendix S2**.
+We provide an additional example using microclimate data measured along an elevation gradient in CO, USA to illustrate use of TrenchR.
+We first use a time series of air temperatures and wind speeds collected at multiple heights to examine profiles and surface roughness.
+We then use a time series of air and surface temperatures, wind speeds, and solar radiation collected at a single height to implement a biophysical model for grasshoppers and compare estimates to observations.
+We omit the code to read and process the environmental data here for brevity.
+R markdown files with the full code and the associated environmental data are available at https://github.com/trenchproject/TrenchRmanuscript/.
+
 ## References

### 2023-08-01 TrenchR_TeExample.Rmd
Add simplified Te example

--- 2022-11-14 Manuscript_full.Rmd
@@
 
-```{r setup, include=FALSE}
+```{r setup}
 knitr::opts_chunk$set(echo = TRUE)
@@
 
-Lauren B. Buckley^1^, Bryan A. Briones Ortiz^1^, Isaac Caruso^1^, Aji John^1^, Ofir Levy^2^, Abigail V. Meyer^1^, Eric A. Riddell^3^, Yutaro Sakairi^1^, and Juniper L. Simonis^4^
-
-1.  Department of Biology, University of Washington, Seattle, WA 98195-1800, USA
-2.  School of Zoology, Tel Aviv University, Tel Aviv, Israel
-3.  Department of Ecology, Evolution, and Organismal Biology, Iowa State University, Ames, IA 50011-4009, USA
-4.  DAPPER Stats, 3519 NE 15th Avenue, Suite 467, Portland, Oregon 97212, USA
-
-Corresponding Author: Lauren Buckley, [lbuckley\@uw.edu](mailto:lbuckley@uw.edu){.email}
-
-Running head: TrenchR microclimate and biophysical models
-
-Type: new methods
-
-## Abstract
-
-Much understanding of organismal responses to climate change and variability relies on the assumption that body temperatures are equal to temporally averaged air temperatures high above the ground.
-However, most organisms experience microclimates near the ground and acute exposure to solar and thermal radiation and thermal extremes can substantially elevate or depress their body temperatures.
-We introduce the TrenchR package, which aids in Translating Environmental Change into organismal responses.
-The package includes microclimate models to vertically scale weather station data to organismal heights.
-Additional functions model and temporally partition air and soil temperatures and solar radiation.
-TrenchR biophysical modeling tools include both general models for heat flows and specific models to predict body temperatures for a variety of ectothermic taxa.
-We also offer utility functions to aid in estimating the organismal and environmental parameters needed for biophysical ecology.
-TrenchR focuses on simple and modular functions so users can create transparent and flexible models for biophysical applications.
-The package aims to introduce and enable microclimate and biophysical modeling to improve ecological and evolutionary forecasting.
-We further this aim through a series of educational modules that introduce the field of biophysical ecology.
-
-**Keywords**: biophysical, climate change, ecotherm, energy budget,
-mechanistic models, microclimate, scaling, temperature
-
-## Introduction
-
-Responses of organisms and ecosystems to climate change are heterogeneous and thus inconsistent with current predictive models [@maguire2015modeling]. 
-Some of these predictive shortcomings stem from omitting spatial and temporal environmental variation and how it interacts with organismal phenotypes (e.g., size, coloration) [@nadeau2017coarse; @potter2013microclimatic].
-Many analyses assume that organismal body temperatures are equal to air temperatures at weather station height (usually 2 meters). 
-However, air temperatures near the ground where most organisms reside are often considerably warmer, and absorption of solar and thermal radiation can raise body temperatures well above air temperatures, resulting in potential discrepancies of tens of degrees Celsius [@sinclair2016; @sunday2014; @kearney2009].
-The rapid drop in performance and onset of thermal stress at high body temperatures can amplify the biological significance of underestimations [@buckley2022asymmetry].
-
-Applications are increasingly demonstrating the importance of accounting for microclimate variation [@bramer2018advances; @butikofer2020problem; @lembrechts2019comparing]. 
-Acute thermal stress events are often more relevant to the physiology and energy balance of organisms than daily, monthly, or annually averaged environmental conditions. 
-Yet, daily and seasonal environmental variation is often neglected [@dillon2016introduction; @dillon2016life]. 
-We aim to help remedy these shortcomings by introducing the TrenchR package (<https://github.com/trenchproject/TrenchR>) as a component of the TrEnCh project (<https:www.trenchproject.com>), which builds computational and visualization tools to Translate Environmental Change into organismal responses.
-
-The mathematical tools for leveraging environmental data to predict organismal conditions have long lingered in books [@gates1980biophysical; @campbell2000introduction] and articles [@porter1969thermodynamic; @porter1983biophysical], but adoption of these tools has not kept pace with research on climate change responses.
-In recent decades, adoption of biophysical approaches is expanding with recognition of the importance of spatial and temporal environmental variation to determining climate change responses [@buckley2018leveraging; @kearney2017nichemapr] and the increased dissemination of computational tools.
-
-Estimating how organisms experience their thermal environment generally entails two classes of models [@gates1980biophysical; @campbell2000introduction] (Figure 1).
-Microclimate models allow scaling conditions from sensors to organism height. They can characterize heat and air transport to estimate vertical air and soil temperature and wind profiles. 
-They can also characterize direct and diffuse solar radiation and longwave radiation emanating from the sky, ground, and surrounding objects like vegetation.
-Microclimate model output is used as input in biophysical models, which estimate body temperatures using an energy budget to balance heat exchanges between organisms and their environment.
-
-```{r, echo=FALSE, fig.height=4, fig.align = 'center', out.width = "75%"}
-knitr::include_graphics('./figures/TrenchRBiophys.png')
-```
-
-**Figure 1** A) Microclimate models scale air and soil temperatures and wind speeds from sensor to organismal heights. 
-B) Biophysical models balance heat exchanges between organisms and their environment to estimate body temperatures [@porter1973behavioral]. 
-A thermal image depicts how grasshopper body temperatures vary substantially from vegetation, air, and ground temperatures due to heat exchange.
-
-The TrenchR package is intended to complement the NicheMapR package, which includes sophisticated microclimate [@kearney2017nichemapr] and biophysical [@kearney2020nichemapr] models.
-Although the Fortran source code was recently released, the complexity of NicheMapR functions can make it difficult to understand and modify the source code and Fortran can be complicated to compile and run.
-The integrated functions made it difficult to separate and adapt different forms of heat flow, but we note that some modular R functions were made available during the review process for this manuscript that enable comparison with the TrenchR package (Appendix S1).
-Many of the approaches are similar between the packages, but TrenchR provides simple R functions to aid the understanding and accessibility of biophysical approaches.
-The TrenchR functions are modular and easily adapt to a variety of organisms and research questions.
-Although many of the TrenchR functions are general, some components of TrenchR are best suited for ectothermic animals.
-We omit plant specific biophysical models since they are the focus of several R packages [tealeaves: [@muir2019tea]; plantecophys: [@duursma2015plantecophys]].
-Some energy budget calculation components of biophysical models are also available in the ThermImage package [@tattersall2017]. 
-Microclimate models, with an emphasis on describing spatial variation, are also available in the microclima package [@maclean2019microclima].
-
-The development of increasingly complex microclimate and biophysical models before open and reproducible science was emphasized or feasible has limited their uptake.
-Yet, such models are crucial to improving understand and prediction of species responses to climate change.
-Simple and modular functions can be combined and extended in a null modeling approach until the model adequately describes organismal interactions with their environment while remaining accessible and transparent.
-
-## Methods and Features
-
-The TrenchR package (<http://trenchproject.github.io/TrenchR/>) aims to promote accessibility and reproducibility. 
-We welcome contributions and corrections from users. 
-Our package was built using the devtools methodology (<https://github.com/r-lib/devtools>) with version control managed in Github. 
-Issues and feature requests can be contributed in Github (<https://github.com/trenchproject/TrenchR>). 
-The package is available via CRAN and Github.
-
-Validation against sensor data including from physical models of organisms is essential to microclimate and biophysical modelling.
-We adapted well established and validated functions from biophysical ecology textbooks [@campbell2000introduction; @gates1980biophysical] and research articles.
-Source references are cited in function headers and reference sections.
-Many of the references, particularly those developing taxa-specific biophysical models, contain extensive empirical validation.
-Appendix S2 illustrates validation of biophysical models using a physical model of a grasshopper.
-Appendix S1 compares TrenchR functions to NicheMapR functions.
-We have incorporated default parameters (e.g., organism emissivity or heat conduction rates) when general values are available.
-Many functions additionally include comments that describe potential parameterizations.
-The biophysical ecology texts [@campbell2000introduction; @gates1980biophysical] include many tables with parameter values and we provide several data tables describing solar and thermal absorptivity in TrenchR.
-
-### Components
-
-TrenchR functions are organized into the following categories (Figure 2):
-
-**Utility functions** calculate environmental metrics that form the basis of microclimate and biophysical models (e.g., zenith and declination angles, which describe the angles of incident sunlight), facilitate modeling diurnal temperature variation, and estimate biologically relevant aggregate metrics such as degree days available for organismal development.
-
-**Allometric functions** allow estimating the dimensions of organisms needed for energy balances and other analyses. 
-Available functions can convert between organismal length, mass, surface area, and volume as well as estimate the silhoutte area, which describes the organismal area exposed to solar radiation.
-
-**Microclimate functions** facilitate calculating the environmental conditions experienced by organisms. 
-Temperature and wind profile functions scale environmental conditions from sensors to organismal height. 
-Radiation functions allow estimating incoming solar radiation and partitioning the variation diurnally and across components (i.e., direct, diffuse, reflected).
-
-**General biophysical functions** provide generalized models of heat exchanges between organisms and their environment so that users can build custom biophysical models. 
-The functions allow implementing an energy balance including the following components of heat exchange between organisms and their environment:
-
-  - *radiative heat exchanges* of solar and thermal radiation; 
-  - *convective heat exchanges* between organisms and their surrounding fluid (air or water) driven by fluid flow; 
-  - *conductive heat exchanges* between organisms and solid surfaces (generally the ground) due to physical contact; and 
-  - *evaporative and metabolic heat exchanges* associated with organisms' evaporative water loss and metabolic heat generation.
-
-Additional functions aggregate these forms of heat exchange into energy balances and to use the energy balances to predict body temperatures.
-
-**Specific biophysical functions** provide biophysical models that have been built for particular organisms based on their physical properties, behavior, environment, and life history. 
-We currently provide published biophysical models for lizards, salamanders, butterflies, grasshoppers, limpets, mussels, and snails.
-Most models predict operative environmental temperatures, which are the steady-state body temperatures of organisms with specified physical properties in a specific microclimate and assume no heat exchange via metabolism or evaporation [@bakken1992measurement].
-However, we also present an analytical function for humid operative temperature that incorporates the effects of evaporative cooling on operative temperatures for wet-skinned ectotherms, such as salamanders.
-
-```{r, echo=FALSE, fig.height=4, fig.align = 'center', out.width = "75%"}
-knitr::include_graphics('./figures/TrenchRFlow.png')
-```
-
-**Figure 2** An examplar aim of translating from environmental and phenotypic data to organismal conditions in a given environment (grey boxes) can be achieved in TrenchR via several categories of functions (white boxes).
-We list example functions for each category used in the example application below.
-
-### Vignettes and introductory tutorials
-
-We introduce the functions in several vignettes. 
-A good place to start is the *Allometry and Conversions* vignette, which provides tools for preparing data, such as estimating additional dimensions of organisms from measured dimensions.
-The *Estimating Microclimates* vignette provides resources for estimating the environmental conditions experienced by organisms.
-This includes estimating solar radiation and its components, diurnal variation in temperature and radiation, temperature and wind speed profiles, and soil temperatures and profiles.
-Finally, the core biophysical modeling functions are described in a tutorial on *Using Energy Balances to Estimate Body Temperatures*.
-Components of an energy budget can be estimated using individual functions and then body temperatures can be solved for using either a generic energy balance or taxa specific biophysical models. 
-We additionally offer a *List of Symbols* used in equations.
-
-For additional background on microclimate and biophysical modeling, we have updated a series of tutorials entitled *Physical Processes in Ecosystems*. 
-We intend the tutorials to provide a more contained and accessible introduction to microclimate and biophysical modeling than that included in classic textbooks [@gates1980biophysical; @campbell2000introduction].
-The tutorials provide less detailed coverage of many of the topics contained in the books. 
-The series of 14 tutorials starts with an overview of the calculus and physics principles underlying the modeling. 
-Thermodynamics and energy budget modeling are then introduced with detailed examples. 
-Tutorials address the climate space concept, operative temperatures, biophysical models for leaves and sheep, and heat flow in soils. 
-We have revised the tutorials (which originated from an NSF training grant in 1979 and lacked broad distribution) to include R code and utilization of TrenchR functions. 
-We expanded the original series to include tutorials contributed by M.
-Kearney introducing the Microclim environmental data [@kearney2014microclim] and integrating the Microclim data with biophysical modeling to examine broad scale climatic limits.
-The tutorials are provided in html form and available for download as a pdf at the bookdown server (<https://bookdown.org/huckley/Physical_Processes_In_Ecosystems/>).
-R markdown files for the tutorials are available in Github (<https://github.com/trenchproject/TrenchRmodules>).
-
-## Example Application
-
 We illustrate the use of the TrenchR package by estimating an energy budget for a *Sceloporus* lizard on June 1, 2021 in Santa Fe, New Mexico, USA (35.69$^\circ$N, -105.944$^\circ$W, elevation: 2121 m).
 The simplified example, which is designed to be self-contained, is also incorporated in the *Using energy balances to estimate body temperatures* vignette. 
-We start by generating environmental inputs (Figure 2).
+We start by generating environmental inputs .
 Using these inputs, we estimate the energy budget with component functions. 
 Finally, we use an integrated biophysical model to estimate operative environmental temperatures.
-See Appendix S2 for a more realistic example examining a time series of microclimate data and applying and testing the biophysical models.
-
-We will use the energy budget to estimate body temperature, $T_b$, which can depart dramatically from the air temperatures due to heat exchange with the environment.
-Heat energy is exchanged with the environment by way of solar and thermal radiation, metabolic reactions, and evaporation.
-The organism also exchanges heat with the surrounding air or water via convection and with substrate it is in contact with via conduction.
-The balance of these heat exchanges (omitting metabolism and evaporation, which are often negligible for ectotherms) can be estimated and often referred to as operative environmental temperature, $T_e$ [@bakken1992measurement].
-$T_e$ is an estimate of $T_b$ and the package functions refer to $T_b$ for simplicity. 
-Our exemplar estimation of $T_b$ assumes steady-state thermal conditions.
-Additionally, we assume that the lizard’s body temperature is homogenous, which is generally reasonable for small ectotherms.
-Approaches to account for thermal gradients between the animal’s core and its skin are available elsewhere [@porter2016].
 
@@
 Sref <- Srad[3,] # Reflected solar radiation  (W/m2)
+
+plot(x    = 0:23, 
+     y    = Sdir, 
+     type = "l", 
+     xlab = "hour", 
+     ylab = expression(radiation ~ (W/m^{2})), 
+     ylim = c(0,1200))
+
+points(x    = 0:23, 
+       y    = Sdif, 
+       type = "l", 
+       lty  = "dotted")
+
+points(x    = 0:23, 
+       y    = Sref, 
+       type = "l", 
+       lty  = "dashed")
+
+legend(x      = "topright", 
+       title  = "Solar radiation component", 
+       legend = c("direct", "diffuse", "reflected"), 
+       lty    = c("solid", "dotted", "dashed"))
+
 ```
@@
 u_liz <- wind_speed_profile_neutral(u_r = u, zr = 2, z0 = 0.2 , z = 0.02)
+
+plot(x    = 0:23, 
+     y    = Ta, 
+     type = "l", 
+     xlab = "hour", 
+     ylab = "temperature", 
+     ylim = c(8,32))
+
+points(x    = 0:23, 
+       y    = Ts, 
+       type = "l", 
+       lty  = "dotted")
+
+points(x    = 0:23, 
+       y    = Ta_liz, 
+       type = "l", 
+       lty  = "dashed")
+
+legend(x      = "topright", 
+       title  = "Temperature", 
+       legend = c("Ta", "Ts", "Ta liz"), 
+       lty    = c("solid", "dotted", "dashed"))
 ```
@@
 
-The full heat budget can be calculated as follows [@gates1980biophysical]:
+The full heat budget can be calculated as follows [@gates1980biophysical]. We plot out the energy balance components.
 
@@
 Qnet <- Qnet_Gates(Qabs = Qabs, Qemit = Qemit, Qconv = Qconv, Qcond = Qcond, Qmet = Qmet, Qevap = Qevap)
+
+par(mar = c(5, 5, 3, 5))
+
+plot(x    = 1:24, 
+     y    = Qnet, 
+     type = "l", 
+     xlab = "hour", 
+     ylab = expression("heat flux" ~ (W/m^{2})), 
+     ylim = c(-1,4))
+
+points(x    = 1:24, 
+       y    = Qabs, 
+       type = "l", 
+       lty  = "dotted")
+
+points(x    = 1:24, 
+       y    = Qemit, 
+       type = "l", 
+       lty  = "dashed")
+
+points(x    = 1:24, 
+       y    = Qconv, 
+       type = "l", 
+       lty  = "dotdash")
+
+points(x    = 1:24, 
+       y    = Qcond, 
+       type = "l", 
+       lty  = "twodash")
+
+legend(x      = "topright", 
+       title  = "component", 
+       legend = c("net radition, Qnet", "solar radiation, Qabs", "thermal radiation, Qemit", "convection, Qconv", "conduction, Qcond"), 
+       lty    = c("solid", "dotted", "dashed", "dotdash", "twodash"))
+
 ```
@@
 
-The microclimate models indicate that air temperatures at lizard height are similar to surface temperatures (Figure 3). 
-The biophysical models indicate that solar radiation can elevate lizard body temperatures far above air temperatures and that the lizard will face thermal stress if it is unable to seek shade (Figure 3).
-The three biophysical models predict different body temperatures during peak period of solar radiation because they model interactions with the ground differently and users are encouraged to review the details of each biophysical model and perform empirical validations before selection.
-Differences in estimated body temperatures are accentuated by the high level of solar radiation.
-`Tb_Gates` is a general and comprehensive model that is appropriate for many applications.
-Taxa-specific biophysical models often best account for details of organism environment interactions and have generally been well tested.
+We then plot a comparison of operative temperature estimates.
 
@@
 ```
-
-**Figure 3**. Body temperatures (Te) are predicted to drastically exceed air temperature when lizards are exposed to high levels of solar radiation. 
-Air temperatures (Ta, $^\circ$C) at lizard height (0.02 m) are predicted to exceed air temperatures at 2 m and to be similar to surface temperatures (Ts). 
-We estimate body temperatures using two general energy budgets [solid: Tb_Gates(); dotted: Tb_CampbellNorman()] and a lizard specific biophysical model [dashed: Tb_lizard()] that differ in how they model heat exchanges.
-
-## Conclusions
-
-TrenchR is intended to promote understanding of how organisms interact with their environment and consequences for physiology, energetics, behavior, and demography. 
-Our example implementation highlights the importance of considering organismal body temperatures, rather than air temperatures, when examining thermal stress and other responses to environmental variability and change.
-TrenchR currently focuses on heat balances but may be expanded to include water balances. 
-Simple functions can be combined as needed to produce comprehensive and transparent models for biophysical ecology and evolution. 
-The resultant models are likely to be sufficiently detailed for many applications, but users are referred to NicheMapR for more detailed biophysical models [@kearney2017nichemapr; @kearney2020nichemapr].
-We focus on models that predict steady-state conditions for simplicity (that is, steady-state conditions). 
-Such models generally do not present computational challenges so our models are not optimized for computational efficiency. 
-Making classic biophysical ecology techniques more accessible will allow researchers to take advantage of rapidly accumulating data on environmental conditions and organismal traits to understand and predict ecological and evolutionary responses.
-Considering how organisms experience their environment is central to understanding responses to variable and changing environments [@suggitt2018extinction].
-
-## Acknowledgements
-
-We thank A. Cannistra, R. Huey, J. Kingsolver, L. Miller, W. Porter, R. Potter, and H. A. Woods for assistance and input and B. Helmuth for providing functions. 
-We particularly thank M. Kearney for extensive input on TrenchR and its alignment with NicheMapR and for contributing two educational modules and code for comparing TrenchR to NicheMapR.
-
-## Authors' Contributions
-Lauren Buckley: Conceptualization, Methodology, Project Administration, Writing – Original Draft Preparation. All authors: Software, Writing – Review & Editing.
-
-## Data Availability
-TrenchR is available at CRAN and <https://github.com/trenchproject/TrenchR>. All text, code, and data for this manuscript are available at <https://github.com/trenchproject/TrenchRmanuscript/>.
-
-## References
